### 2023-02-17 oopsla23/paper.tex
init outline

\documentclass[acmsmall,review,anonymous]{acmart}

%\settopmatter{printfolios=true,printccs=false,printacmref=false}

\overfullrule=1mm
\citestyle{acmauthoryear}
%\setcitestyle{round}

\usepackage{alltt}
\usepackage{amssymb}
\usepackage{calc}
\usepackage{listings}
\usepackage{mathpartir}
\usepackage{pifont}
\usepackage{tikz}
\usepackage{wrapfig}
\usepackage{xcolor}
\usetikzlibrary{shapes.geometric}

\begin{document}

\title{Millions of Type Errors}

\author{Ben Greenman}
\orcid{0000-0001-7078-9287}
\affiliation{%
  \institution{Brown University}
  \city{Providence}
  \state{Rhode Island}
  \country{USA}
}
\email{benjaminlgreenman@gmail.com}

\author{Alan Jeffrey}
\orcid{TODO}
\affiliation{%
  \institution{Roblox}
  \city{}
  \state{}
  \country{USA}
}
\email{}

\author{Mitesh Shah}
\orcid{TODO}
\affiliation{%
  \institution{Roblox}
  \city{}
  \state{}
  \country{USA}
}
\email{}

\author{Shriram Krishnamurthi}
\orcid{0000-0001-5184-1975}
\affiliation{%
  \institution{Brown University}
  \city{Providence}
  \state{Rhode Island}
  \country{USA}
}
\email{shriram@brown.edu}

%\renewcommand{\shortauthors}{...}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  TBD
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011039.10011311</concept_id>
<concept_desc>Software and its engineering~Semantics</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011008.10011024.10011032</concept_id>
<concept_desc>Software and its engineering~Constraints</concept_desc>
<concept_significance>100</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
<concept_desc>Software and its engineering~Functional languages</concept_desc>
<concept_significance>100</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Semantics}
\ccsdesc[100]{Software and its engineering~Constraints}
\ccsdesc[100]{Software and its engineering~Functional languages}

\keywords{types, gradual typing, telemetry, user study, large-scale study}

\maketitle

\section{Introduction}
\label{s:introduction}

millions of creators:
 gradual types,
 success types,
 semantic subtyping

millions (?) of telemetry records

takeaways about adoption,
 errors by mode (nocheck, nonstrict, strict),
 patterns of use,

\section{Roblox Contex}

\paragraph{Script, Module}

%% https://create.roblox.com/docs/education/coding-6/intro-to-module-scripts

script = code that runs, can have many scripts in a codebase

module = library, reusable functions, doesn't run directly

scripts are fine for basic games. professional developers are much more likely to use modules.

\paragraph{Data Model}

changes to DM can give huge type error changes

moving (re-parent) a node changes the type of the DM, unsoundness in type system

type checker assumes initial state

\section{Predictions}

nocheck forcestrict keep growing

nonstrict te tend to zero, fixing your program should fix the errors, no false positives

nonstrict fs unclear

\section{Results}
\label{s:data}

\section{Interpretation}

Error 1000 = type mismatch (subtyping etc);
1001 = unknown symbol;
\ldots

% https://github.com/Roblox/luau/blob/master/Analysis/include/Luau/Error.h

\begin{verbatim}
 require(foobar)
  foobar is a module script in the data model
 require(foo.bar.b)
  - could be edit in progress
  - could be renamed module
  - 
\end{verbatim}

\section{Related Work}
\label{s:related}

Mind your language~\cite{mfk-onward-2011}.

\section{Discussion}
\label{s:conclusion}
\label{s:discussion}

\begin{acks}
  TBD

Greenman was supported by
  \grantsponsor{NSF}{NSF}{https://www.nsf.gov} grant
 \href{"https://www.nsf.gov/awardsearch/showAward?AWD_ID=2030859"}{\grantnum{NSF}{CCF 2030859}}
  to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
\end{acks}

\bibliographystyle{ACM-Reference-Format}
\bibliography{bib}

\end{document}
\endinput

### 2023-03-10 oopsla23/paper.tex
roblox context notes
plot telemetry records per hour for the sample dataset

@@
 \usepackage{calc}
+\usepackage{cleveref}
 \usepackage{listings}
@@
 
+\paragraph{Creating Experiences}
+
+more than games
+
+telemetry gets a session id per experience
+
 \paragraph{Script, Module}
@@
 
+\section{Telemetry Design}
+
+No PII whatsoever
+
+Random session selection
+
+Random events during session, except module switch.
+
+No access to other key events: save, exit.
+Don't want access to run, publish.
+
+\url{https://docs.google.com/document/d/1DnKvw8x1jy0EWCbBSM8neKULWz7WAg1OauKy7qqYqkI/edit}
+
 \section{Predictions}
+%%bg: merge with prev section, on telemetry design?
 
@@
 \label{s:data}
+
+\begin{figure}[t]
+  \includegraphics{img/row-distribution.pdf}
+  \Description{TBD: histogram with about 100 records per hour except for a 600-record spike near the end of Jan 12th.}
+  \caption{Telemetry records per hour. Each tick on the $x$-axis marks the start of a new day in California.}
+  \label{f:records-per-hour}
+\end{figure}
+
+\Cref{f:records-per-hour} shows when data arrived across the whole dataset.
 

### 2023-03-17 oopsla23/paper.tex
Fixed TeX compilation errors
Add some affiliation info
Add a comment
Add text to intro

@@
-\documentclass[acmsmall,review,anonymous]{acmart}
+% The paper on using Luau telemetry to measure the effectiveness of type error reporting.
+
+% \documentclass[acmsmall,review,anonymous]{acmart}
+\documentclass[acmsmall,review]{acmart}
 
@@
 \usepackage{alltt}
-\usepackage{amssymb}
+% \usepackage{amssymb} -- already loaded by acmart
 \usepackage{calc}
@@
 \title{Millions of Type Errors}
+\subtitle{Using Telemetry To Measure The User Experience Of Type Error Reporting}
+
+% Alphabetical order for authors?
 
@@
 \author{Alan Jeffrey}
-\orcid{TODO}
+\orcid{0000-0001-6342-0318}
 \affiliation{%
   \institution{Roblox}
-  \city{}
-  \state{}
-  \country{USA}
-}
-\email{}
-
-\author{Mitesh Shah}
-\orcid{TODO}
-\affiliation{%
-  \institution{Roblox}
-  \city{}
-  \state{}
+  \city{San Mateo}
+  \state{California}
   \country{USA}
@@
 
+\author{Mitesh Shah}
+\orcid{TODO}
+\affiliation{%
+  \institution{Roblox}
+  \city{San Mateo}
+  \state{California}
+  \country{USA}
+}
+\email{}
+
 %\renewcommand{\shortauthors}{...}
@@
 \begin{abstract}
-  TBD
+  \anon{Roblox Studio} is a tool which places programming in the hands of
+  millions of creators, ranging from high school students to professional
+  development studios. It has the ability to report telemetry data,
+  which allows large-scale measurement of the creator experience. In
+  this paper, we discuss one use of telemtry, to measure the experience
+  of type error reporting.
 \end{abstract}
@@
 \label{s:introduction}
+
+\anon{Roblox} is a platform for \anon{shared virtual experiences},
+with 56~million Daily Active Users, and 49~billion hours of engagement in
+2022~\anon[(ANONYMIZED CITATION)]{\cite{roblox-quarterly-results}}.
+There are XX~million creators using \anon{Roblox Studio},
+and YY~million creations.
+
+\anon{Roblox experiences} are scripted using the 
+\anon{Luau} programming language~\anon[(ANONYMIZED CITATION)]{\cite{luau-lang.org}},
+an extension of \anon{Lua~5.1~\cite{lua}}.
+The main extension is the addition of a static type system, which uses
+type inference to synthesize types for user code. These types
+are used primarily in type-driven tooling such as autocomplete
+and API documentation~\anon[(ANONYMIZED CITATION)]{\cite{luau-autocomplete}},
+but creators can also opt in to receiving type error reports.
+
+As discussed in~\anon[(ANONYMIZED CITATION)]{\cite{bfj-hatra-2021}},
+the goals of the \anon{Luau} type system are rather different from
+a traditional type system, which focuses on compilation and memory safety.
+\anon{Luau} has a very heterogenous user community, ranging from
+students in code camps to professional development studios. These
+creators have quite different needs, with different emphases on
+enabling rapid creation and ensuring software quality.
+
+In this paper, we investigate methods for measuring the effectiveness
+of the \anon{Luau} type system in development of \anon{Roblox} scripts.
+In comparison to existing studies~\cite{BEN-WE-NEED-CITATIONS}
+which are based on small-scale user studies, we performed
+a large-scale study using \emph{telemetry}.
 

### 2023-03-17 oopsla23/paper.tex
Add more text to intro

@@
 
-millions of creators:
- gradual types,
- success types,
- semantic subtyping
+\anon{Roblox Studio} has a telemetry system, which is used to guage
+the effectiveness of creation features. This system stochastically
+determines which sessions should report telemety, and for those
+sessions, reports telemetry records back with a summary of the
+session. In the case of this study, the telemetry includes data on the
+number of errors at various levels of granularity: in the current edit
+region, in the current file, and in every file which was type
+checked.
 
-millions (?) of telemetry records
+The telemetry data we analyzed does not contain any Personally
+Identifying Information. There is no record of the creator's identity,
+geographic location, IP address, or which creation they worked on.
+Telemetry records are correlated by session, using a pseudonymized
+session identifier.
 
-takeaways about adoption,
- errors by mode (nocheck, nonstrict, strict),
- patterns of use,
+Most users of \anon{Roblox Studio} do not opt in to type error
+reporting, and so they do not see the ``squiggly underlining'' that
+indicates a type error site. Nonetheless, the type inference system
+still runs (since it drives automplete and other type-based tools) and
+so we can record which type errors would have been reported had the
+user enabled type error reporting. As a result, we can investigate
+which type errors are  fixed by users, even if they did not opt in to
+type error reporting.
+
+\textbf{Ben: give an outline of our RQs here.}
+
+This paper is the first to use telemetry as a mechanism for
+large-scale measurement of the effectiveness of type error reporting.
+Our data captures \textbf{XX} sessions, for a total of \textbf{YY}
+hours and \textbf{ZZ} type errors.
+
+Telemetry cannot replace user studies, as there is no way to measure
+creator sentiment, but provide complementary data at scale.
 

### 2023-03-22 oopsla23/paper.tex
Reanonymized
Highlight missing numbers
data: first-glance summary table

@@
 
-% \documentclass[acmsmall,review,anonymous]{acmart}
-\documentclass[acmsmall,review]{acmart}
+\documentclass[
+  acmsmall,
+  review,
+  anonymous,
+]{acmart}
 
@@
 2022~\anon[(ANONYMIZED CITATION)]{\cite{roblox-quarterly-results}}.
-There are XX~million creators using \anon{Roblox Studio},
-and YY~million creations.
+There are \textbf{XX}~million creators using \anon{Roblox Studio},
+and \textbf{YY}~million creations.
 
@@
 
+\begin{table}[t]
+  \caption{Dataset overview}
+  \label{t:dataset-overview}
+\begin{tabular}{rl}
+ 5,108 & total logs \\
+       & 4,574 nocheck + 510 nonstrict + 24 strict \\
+ 182,247 & total forced strict type errors \\
+       & {113,274 in module, 6,425 in edit regions} \\
+  1,521 & total type errors \\
+       & {1,485 in module, 331 in edit regions}
+  \\[2ex]
+
+  1,146 & sessions \\
+  & \begin{tabular}{rl}
+      1,143 & single mode = 1046 nocheck + 94 nonstrict + 3 strict \\
+      7 & multi mode projects \\
+      3 & mode upgrades \\
+      4 & mode downgrades
+    \end{tabular}
+\end{tabular}
+\end{table}
+
 \section{Interpretation}
@@
 
+\paragraph{Research on Errors}
+
 Mind your language~\cite{mfk-onward-2011}.
+
+\paragraph{Telemetry}
+
+Transparent telemetry: explain what it is and contrast our approach.
+What essential, non-transparent things did we collect?
+
+TT =
+no user ID, no machine ID,
+no time-ordered traces,
+public process to decide what to collect
+
+https://research.swtch.com/telemetry-intro
 

### 2023-03-24 oopsla23/paper.tex
Add screenshots

@@
   review,
-  anonymous,
+%  anonymous,
 ]{acmart}
@@
 
-\section{Roblox Contex}
+\section{\anon{Roblox} Context}
+
+\begin{figure}
+  \anon{
+    \includegraphics[width=.45\textwidth]{img/roblox-studio.png}
+    \includegraphics[width=.45\textwidth]{img/roblox-studio-ide.png}
+  }
+  \caption{\anon{Roblox Studio 3D creation} tools (left) and IDE (right)}
+  \label{fig:roblox-studio}
+\end{figure}
+
+Creators of \anon{Roblox experiences} use \anon{Roblox Studio},
+which combines \anon{3D creation} tools as well as an Integrated
+Developer Environment (IDE), as seen in Fig.~\ref{fig:roblox-studio}.
 

### 2023-03-24 oopsla23/paper.tex
Fleshed out Roblox Context

@@
 Developer Environment (IDE), as seen in Fig.~\ref{fig:roblox-studio}.
+The IDE includes an optional ``Script Analysis'' widget, which
+reports syntax errors, type errors, and problems identified by
+lint tools. The script editor also (optionally) highlights
+the location in code where reported errors occur.
 
-\paragraph{Creating Experiences}
+To opt in to type error reporting, creators set a \emph{mode}
+for each script, which is one of:
+\begin{itemize}
+\item \emph{nocheck}: only syntax errors are reported,
+\item \emph{nonstrict}: all syntax errors, and a subset of ``high probability'' type errors, are reported, or
+\item \emph{strict}: all syntax errors and type errors, are reported.
+\end{itemize}
+As an example of nonstrict mode, the following program only reports one error:
+\begin{verbatim}
+--!nonstrict
+local x = { p = 5, q = nil }
+if condition then x.q = 7 end
+local y = x.p + x.q --> no type error
+local z = x.r       --> "Key 'r' not found in table 'x'"
+\end{verbatim}
+but in strict mode it reports two:
+\begin{verbatim}
+--!strict
+local x = { p = 5, q = nil }
+if condition then x.q = 7 end
+local y = x.p + x.q --> "Type 'nil' could not be converted into 'number'"
+local z = x.r       --> "Key 'r' not found in table 'x'"
+\end{verbatim}
+In cases like this, where it is undecidable whether there will be a run-time error,
+strict mode errs on the side of reporting an error, and nonstrict mode errs on
+the side of suppressing the error.
 
-more than games
+Both modes report the \verb|Key 'r' not found in table 'x'| error --
+misspellings of property names are common enough to report in both
+strict and nonstrict mode. See~\anon[ANONYMIZED CITATION]{\cite{bfj-hatra-2021}}
+for a more detailed discussion of the rationale for strict and nonstrict mode.
 
-telemetry gets a session id per experience
+Both modes are opt-in. Creators have the option to make nonstrict mode
+the default rather than nocheck mode.
 
-\paragraph{Script, Module}
+Even in nocheck mode, \anon{Roblox Studio} performs type inference, since
+the results are needed by type-directed tooling such as autocomplete and
+API documentation. This behind-the-scenes typechecking is always performed
+in strict mode, since it is important that the inferred types be as precise
+as possible. The type errors produced by this pass are always discarded,
+so the verbosity of strict mode is not an issue.
 
-%% https://create.roblox.com/docs/education/coding-6/intro-to-module-scripts
+Since this pass is always performed in strict mode, we refer to it as
+\emph{forced strict} mode. Its main use is in autocomplete, so forced
+strict mode is triggered on every keystroke
 
-script = code that runs, can have many scripts in a codebase
+Scripts come in two favors: \emph{module scripts} and \emph{non-module
+scripts}.  Module scripts provide reusable libraries, which may be
+\emph{required} by other scripts. Since module scripts can require
+other module script, modules form a graph (though we consider it to be
+an error in strict mode if the graph is cyclic, and remove edges to
+make it acyclic).
 
-module = library, reusable functions, doesn't run directly
+When typechecking is performed for script analysis, any script that
+has been modified is marked as dirty, then any script that is dirty,
+or which transitively requires a dirty module, is typechecked. More
+commonly, when typechecking is performed for autocomplete, we only
+need to typecheck the current script, since it is the only dirty
+script, and nothing it requires can transitively require it, since we
+have broken cycles.
 
-scripts are fine for basic games. professional developers are much more likely to use modules.
+The state of the world in a \anon{Roblox} experience is captured by
+the \emph{data model}, which is a tree of \emph{instances}, such as
+parts, models, meshes, effects, lighting, audio assets, and physics
+constraints such as forces, springs and joints.
 
-\paragraph{Data Model}
-
-changes to DM can give huge type error changes
-
-moving (re-parent) a node changes the type of the DM, unsoundness in type system
-
-type checker assumes initial state
+While an experience is under development, it is typical for the data
+model to be edited (for example instances to be added, deleted, moved
+or renamed). Since the initial shape of the data model tree is reflected in
+the type system, it is possible for these edits to introduce type errors.
 

### 2023-03-24 oopsla23/paper.tex
Spellcheck

@@
   which allows large-scale measurement of the creator experience. In
-  this paper, we discuss one use of telemtry, to measure the experience
+  this paper, we discuss one use of telemetry, to measure the experience
   of type error reporting.
@@
 a traditional type system, which focuses on compilation and memory safety.
-\anon{Luau} has a very heterogenous user community, ranging from
+\anon{Luau} has a very heterogeneous user community, ranging from
 students in code camps to professional development studios. These
@@
 
-\anon{Roblox Studio} has a telemetry system, which is used to guage
+\anon{Roblox Studio} has a telemetry system, which is used to gauge
 the effectiveness of creation features. This system stochastically
-determines which sessions should report telemety, and for those
+determines which sessions should report telemetry, and for those
 sessions, reports telemetry records back with a summary of the
@@
 indicates a type error site. Nonetheless, the type inference system
-still runs (since it drives automplete and other type-based tools) and
+still runs (since it drives autocomplete and other type-based tools) and
 so we can record which type errors would have been reported had the

### 2023-03-30 oopsla23/paper.tex
draft RQ overview, add table of error labels

@@
 
+\newcommand{\code}[1]{\texttt{#1}}
+\newcommand{\FILL}{\textbf{FILL}}
+
 %%
@@
 of the \anon{Luau} type system in development of \anon{Roblox} scripts.
-In comparison to existing studies~\cite{BEN-WE-NEED-CITATIONS}
-which are based on small-scale user studies, we performed
-a large-scale study using \emph{telemetry}.
+In comparison to prior work~(\cref{s:related}), which is either small in scale
+or depends on personally identifiable information~(PII),
+we performed a large-scale study using pseudonymized \emph{telemetry}.
 
@@
 
-The telemetry data we analyzed does not contain any Personally
-Identifying Information. There is no record of the creator's identity,
-geographic location, IP address, or which creation they worked on.
+The telemetry data we analyzed does not contain any PII:
+no source code;
+no source code locations;
+no error messages (which may contain source code);
+no record of the creator's identity, locale, or IP address;
+and no information about what creation the data came from.
 Telemetry records are correlated by session, using a pseudonymized
@@
 
-\textbf{Ben: give an outline of our RQs here.}
+With this telemetry data, we investigate research questions about
+the adoption and benefits of type analysis.
+First, to learn how many creators use type analysis of any sort
+and to estimate whether they pay attention to the results (\FILL{} really, attention?).
+Once in type-analysis mode, do creators stay, or revert to no-check?
+Second, what errors do creators face and how do they respond.
+Are the error highlights helpful for removing the error?
+Is there any indication that type analysis improves the quality of the development experience?
 
@@
 
+\paragraph{Contributions}
+\begin{itemize}
+  \item
+    Design of a low-overhead, (black-box / PII-safe / impersonal)
+    telemetry method, that other
+    researchers can build on.
+
+  \item
+    Lessons from millions of type errors about
+    the adoption of strict type analysis,
+    the usefulness of type errors,
+    and \FILL{}.
+    These findings are especially important for the
+    gradual typing, success typing, and semantic subtyping communities.
+
+\end{itemize}
+
 \section{\anon{Roblox} Context}
+% https://create.roblox.com/docs/scripting/luau
+
+FILL table = object
 
@@
 the type system, it is possible for these edits to introduce type errors.
+
+\begin{table}
+  \caption{Selected Error Labels}
+  %% from type analysis, but they're not all really type errors
+  %% TODO review selection after analyzing the full data
+  \label{t:type-error-labels}
+
+  %% FILL examples for each error?
+  \begin{tabular}{ll}
+    Label & Interpretation \\\midrule
+    \code{CodeTooComplex} & Type analysis failed, cannot understand the code \\
+    \code{UnificationTooComplex} & Type analysis failed, unification solver hit a limit \\
+    \code{SyntaxError} & Basic parse error, e.g., \code{for if end} \\
+
+    \code{IncorrectGenericParameterCount} & Arity mismatch for a generic type \\
+    \code{CountMismatch} & \FILL{} \\
+    \code{UnknownProperty} & Referenced an invalid field or method  \\
+    \code{OnlyTablesCanHaveMethods} & Tried to attach a method to a non-table \\
+    \code{CannotCallNonFunction} & Called a value that is not a function \\
+    \code{TypesAreUnrelated} & Failed to cast, unify, or check subtyping \\
+    \code{TypeMismatch} & Generic label for other type errors \\
+    \code{GenericError}
+    & Generic label for other errors, e.g., declaring a \\
+    & supertype that is not a class type
+    %% or iterated over a table without specifying how
+
+%    \code{ExtraInformation} & Follow-on error that provides additional information for an error at the same source location. \\
+%    UnknownSymbol &  \\
+%    NotATable &  \\
+%    CannotExtendTable &  \\
+%    DuplicateTypeDefinition &  \\
+%    FunctionDoesNotTakeSelf &  \\
+%    FunctionRequiresSelf &  \\
+%    OccursCheckFailed &  \\
+%    UnknownRequire &  \\
+%    UnknownPropButFoundLikeProp &  \\
+%    InternalError &  \\
+%    DeprecatedApiUsed &  \\
+%    ModuleHasCyclicDependency &  \\
+%    IllegalRequire &  \\
+%    FunctionExitsWithoutReturning &  \\
+%    DuplicateGenericParameter &  \\
+%    CannotInferBinaryOperation &  \\
+%    MissingProperties &  \\
+%    SwappedGenericTypeParameter &  \\
+%    OptionalValueAccess &  \\
+%    MissingUnionProperty &  \\
+%    NormalizationTooComplex &  \\
+%    TypePackMismatch &  \\
+%    DynamicPropertyLookupOnClassesUnsafe &  \\)
+  \end{tabular}
+\end{table}
 

### 2023-03-31 oopsla23/paper.tex
drop in telemetry text from google doc

@@
 
+\subsection{What are we logging (from gdoc)}
+
+\FILL{} change to long-form paragraphs, use more space for examples
+
+\begin{description}
+  \item[D1] metadata: uid, timestamp, session duration
+  \item[D2] mode (strict / nonstrict / nocheck)
+  \item[D3] a 1-bit reason for sending: did the user switch module?
+    \subitem we get logs for two reasons: (1) switch module, (2) randomly with
+    low probability after typechecking
+  \item[D4] size of codebase: num files, num lines
+  \item[D5] size of edit region
+  \item[D6] overall type errors: for old and curr states, have: \# total, \# in module, and \# in region
+  \item[D7] specific errors in the edit region: in the edit region, for old and
+    curr states, how many type errors for each type error code? (There is 1
+    edit region per log.)
+    \subitem Examples, if the old state had one type-A error:
+    \begin{itemize}
+      \item If the latest edits touched it and fixed it, the next log says 1
+        old type-A and 0 new type-A
+      \item touched it, didn't fix => 1 old-A, 1 new-A
+      \item didn't touch (fixed or not) => 0 old-A, 0 new-A
+    \end{itemize}
+  \item[D8] ForcedStrict "what if" errors for nocheck: specific errors for the
+    edit region computed as if the user had strict mode enabled (even though they
+    have no-check enabled)
+    \subitem This info is available because it gives better hints for
+    autocomplete, because it assumes precise types for ALL libraries (rather
+    than Any)
+  \item[D9] TooComplexErrors: project-wide count for this one specific error type
+\end{description}
+
+\subsection{what to study}
+
+RQs + Analysis
+
+\begin{itemize}
+  \item How often does the typechecker give up with CodeTooComplex errors?
+  \subitem [[ For context: the type system is designed to give useful feedback
+    to all sorts of code, whether strict-typed, nonstrict-typed, or nocheck.
+    But it has an escape hatch: CTC. Hope it's rarely needed. ]]
+    \begin{itemize}
+      \item Count the overall CTC errors (in each mode) for a coarse view of
+        how rare they are. [D9 of course, D8 to compare nocheck code with
+        strict-mode code via the same typechecker]
+      \subitem Note: We can't look at the code to learn why the errors
+        happened. Can only look for trends and tell the Roblox team where to
+        look.
+      %% \subitem Count CTC errors within each session to see if these errors are intermittent, or if they stick around.
+[D9 for the count, D8 for nocheck vs strict, D7 because CTC might show up in the edit region]
+    \end{itemize}
+  \item How helpful are highlighted regions?
+    \begin{itemize}
+      \item How often do edits in a highlighted region remove type errors? How
+        often do they add new ones?
+      \subitem [D7 compare specific old vs specific new]
+      \subitem [may want to filter using D2 + D5]
+      \item How often do edits outside the region remove type errors? How often do they add new ones?
+      \subitem [D6 looking for old count vs. new count difference without a matching delta in the D7 counts]
+      \item Which specific errors seem easy / hard to fix via highlights?
+      \subitem [D6 removed by edit => easy; D6 survive edit => hard]
+    \end{itemize}
+  \item In what ways do types improve quality of life?
+  \begin{itemize}
+    \item Does nocheck code have many more latent errors than the others?
+    \subitem [D8 latent errors come from the "what if" analysis]
+  \end{itemize}
+\end{itemize}
+
+% What to study?
+% 
+% Sanity checks
+% How large are edit regions? Plot the distribution.
+% How long between logs? Long times & very short times are problematic.
+% General Roblox trends
+% How many sessions use strict / nonstrict / nocheck?
+% Expect few strict/nonstrict, lots of nocheck
+% How often does mode change within a session?
+% Expect few of these b/c error reports are easy to ignore; non-intrusive
+% If a down-switch happens, expect the current #errors to be high
+% If an up-switch happens, hope to stay up
+% How many CodeTooComplex errors arise, overall?
+% Expect very few. Hopefully zero.
+% How often do module switches happen?
+% Expect not-too-many, with some logs in between each time
+% Compare strict vs. nonstrict vs. nocheck
+% Get median # errors across all snapshots.
+% Expect nocheck median to be higher than the others.
+% Plot #errors vs time, or vs codebase size
+% Expect nocheck errors to keep going up. For other modes, #errors should stay low.
+% General type error trends
+% Plot the distribution of #errors across all edit regions
+% Expect different counts based on error-kind. SEE BELOW
+% Count specific errors that get fixed. For each kind of error, how many times is it old-in-region but not new-in-region? Compare to #old-in-region totals.
+% Expect all errors to have a decent fix rate.
+% If one error has a lower % than others, its error region may be misleading. Or, maybe it's hard to fix
+% Type errors vs. edit region
+% Compare #old vs. #new in region
+% Expect a decrease, #old > #new
+% If there's an increase, look for an explanation. Could be a huge edit region. Could be a 1-line region that we captured 2x in a row. Could be that this user is simply ignoring type errors.
+% Compare [#old-in-region vs. (total #old - total #new)]
+% If #old-in-region = 0, then hope the total # increases
+% Decrease => nonlocal edits fixed an error => type error hilite may be inaccurate
+% If #old-in-region > 0, then hope the total # decreases
+% Increase (in region or otherwise) => made things worse
+
 \section{Predictions}

### 2023-04-01 oopsla23/paper.tex
ideas for 3rd contribution
shorter title, 'non-invasive tel.'

@@
 \title{Millions of Type Errors}
-\subtitle{Using Telemetry To Measure The User Experience Of Type Error Reporting}
+\subtitle{Non-Invasive Telemetry to Measure User Experience}
 
@@
     gradual typing, success typing, and semantic subtyping communities.
+
+  \item
+    \FILL{} any surprises when doing the analysis?
+    Note, no ML for analysis, we do not have a corpus of text.
 
@@
 
-TT =
+https://research.swtch.com/telemetry-intro
+
+TT:
 no user ID, no machine ID,
@@
 
-https://research.swtch.com/telemetry-intro
+kindle track every tap (page turn etc),
+enables whispersync feature,
+drove design of navigation tools,
+opt-out possible~\cite{kindle-telemetry}
+
+vscode: usage data, crash reports, error data (not a crash but unexpected);
+save file, open terminal, copy-paste, autocomplete offered, git queries, machine id,
+session id, timestamp,
+opt-out may be possible, but not for all usage data (per license, sec 2a) and
+every extension can do its own thing~\cite{vscode-telemetry}
+
+.NET SDK and .NET CLI:
+crash reports, command invoked, args, hashed cwd, timings;
+more listed online;
+published in aggregate under CC-BY
+opt-out possible~\cite{dotnet-telemetry}
 

### 2023-04-01 oopsla23/paper.tex
telemetry table

@@
 \usepackage{tikz}
+\usepackage{wasysym}
 \usepackage{wrapfig}
@@
 \title{Millions of Type Errors}
-\subtitle{Non-Invasive Telemetry to Measure User Experience}
+\subtitle{Lite Telemetry to Measure User Experience}
+% lean, lite, modest, non-intrusive, private, ...
 
@@
 \newcommand{\FILL}{\textbf{FILL}}
+\newcommand{\dotscale}[1]{\scalebox{0.72}{#1}}
+\newcommand{\wideas}[2]{\makebox[\widthof{#2}][l]{#1}}
+\newcommand{\twoline}[2]{\parbox[s]{1.4cm}{\flushleft#1\newline#2}}
+\newcommand{\chkYes}{\dotscale{\CIRCLE}}
+\newcommand{\chkMaybe}{\wideas{\dotscale{\Circle}}{\chkYes}}
+\newcommand{\chkNo}{\wideas{}{\chkYes}}
 
@@
 
+\begin{table}[t]
+  \caption{Comparing telemetry systems}
+  \label{t:telemetry-design}
+
+  \begin{tabular}{l@{~}cccccc}
+    &             & PII       & Session ID & Deterministic & Private Data \\\midrule
+    & Transparent & \chkNo    & \chkNo     & \chkYes       & \chkNo      \\
+  * & Lite        & \chkNo    & \chkYes    & \chkNo        & \chkYes     \\
+    & \code{.NET} & \chkMaybe & \chkYes    & \chkYes       & \chkYes     \\
+    & VS Code     & \chkYes   & \chkYes    & \chkYes       & \chkYes     \\
+  \end{tabular}
+\end{table}
+
 Transparent telemetry: explain what it is and contrast our approach.
@@
 
-https://research.swtch.com/telemetry-intro
-
-TT:
+counting only, public decisions about what to count, public data,
+collected weekly for a sample of clients, LastWeek field = prior day when
+system gathered any data;
+eg command invocations, lib stack frames, 
 no user ID, no machine ID,
-no time-ordered traces,
-public process to decide what to collect
+no time-ordered traces~\cite{transparent-telemetry}.
 
@@
 more listed online;
-published in aggregate under CC-BY
+custom builds beware inadvertant disclosure, keep your filepaths clear;
+data published in aggregate under CC-BY;
 opt-out possible~\cite{dotnet-telemetry}

### 2023-04-03 oopsla23/paper.tex
Lite -> Impersonal
redo dataset overview, add brief comments

@@
 \title{Millions of Type Errors}
-\subtitle{Lite Telemetry to Measure User Experience}
+\subtitle{Impersonal Telemetry to Measure User Experience}
 % lean, lite, modest, non-intrusive, private, ...
@@
     & Generic label for other errors, e.g., declaring a \\
-    & supertype that is not a class type
+    & \quad{} supertype that is not a class type
     %% or iterated over a table without specifying how
@@
 
-\begin{table}[t]
-  \caption{Dataset overview}
-  \label{t:dataset-overview}
+\begin{figure}[t]
 \begin{tabular}{rl}
+ 5108 total logs = 4574 nocheck + 510 nonstrict + 24 strict
+ 182247 total forced strict type errors, 83005 in module, 5113 in edit regions
+ 1521 total type errors, 1035 in module, 110 non-stx + 47 syntax in edit regions
+ 1146 sessions
+  1143 single mode = 1046 nocheck + 94 nonstrict + 3 strict
+  2 multi mode projects
+  2 mode upgrades
+  2 mode downgrades
+
  5,108 & total logs \\
@@
  182,247 & total forced strict type errors \\
-       & {113,274 in module, 6,425 in edit regions} \\
+       & {83,005 in module, 5,113 in edit regions} \\
   1,521 & total type errors \\
-       & {1,485 in module, 331 in edit regions}
+       & {1,035 in module, 110 non-syntax + 47 syntax in edit regions}
   \\[2ex]
@@
   1,146 & sessions \\
-  & \begin{tabular}{rl}
-      1,143 & single mode = 1046 nocheck + 94 nonstrict + 3 strict \\
-      7 & multi mode projects \\
-      3 & mode upgrades \\
-      4 & mode downgrades
+  & \begin{tabular}{l}
+      1,143 single mode = 1046 nocheck + 94 nonstrict + 3 strict \\
+      detected 2 multi-mode projects, 2 mode upgrades, 2 mode downgrades
     \end{tabular}
 \end{tabular}
-\end{table}
+  \caption{Dataset overview}
+  \label{f:dataset-overview}
+\end{figure}
+
+\Cref{f:dataset-overview}:
+cannot filter syntax errors from the total and per-module error counts (\cref{s:threats});
+multi-mode project = found a module switch that coincides with a mode switch, which suggests that two modules have different modes;
+upgrade = found a up-mode switch that does not match a module switch (quite possible to miss these);
+downgrade = found a down-mode switch that does not match a module switch
 
@@
 \end{verbatim}
+
+\section{Threats to Validity}
+\label{s:threats}
+
+Error counts per-project and per-module include syntax errors.
+Problematic, because syntax errors are uninteresting (easy to fix),
+but common, which makes it likely that our per-keystroke telemetry will record them.
+Edit region has specific counts.
+Next time, we should do that for the other aggregates.
 

### 2023-04-03 oopsla23/paper.tex
cleanup row dist

@@
 
+\section{Data Cleaning}
+
+Deduplicated records with the same client timestamp,
+kept only the first.
+Timestamps by millisecond.
+
 \section{Results}
@@
 \begin{tabular}{rl}
- 5108 total logs = 4574 nocheck + 510 nonstrict + 24 strict
- 182247 total forced strict type errors, 83005 in module, 5113 in edit regions
- 1521 total type errors, 1035 in module, 110 non-stx + 47 syntax in edit regions
- 1146 sessions
-  1143 single mode = 1046 nocheck + 94 nonstrict + 3 strict
-  2 multi mode projects
-  2 mode upgrades
-  2 mode downgrades
-
- 5,108 & total logs \\
-       & 4,574 nocheck + 510 nonstrict + 24 strict \\
+ 5,001 & total logs \\
+       & 4,471 nocheck + 506 nonstrict + 24 strict \\
  182,247 & total forced strict type errors \\

### 2023-04-03 oopsla23/paper.tex
placeholder for file size

@@
   \includegraphics{img/row-distribution.pdf}
-  \Description{TBD: histogram with about 100 records per hour except for a 600-record spike near the end of Jan 12th.}
+  \Description{FILL Histogram with approximately 100 records per hour, except for a 600-record spike near the end of Jan 12th.}
   \caption{Telemetry records per hour. Each tick on the $x$-axis marks the start of a new day in California.}
   \label{f:records-per-hour}
+\end{figure}
+
+\begin{figure}[t]
+  TODO file size distribution: table or bars %% \includegraphics{img/row-distribution.pdf}
+  \Description{FILL}
+  \caption{Size of analyzed code: number of files, number of lines , and lines in edit region}
+  \label{f:analysis-size}
 \end{figure}
@@
 \section{Interpretation}
+
+\subsection{Code Too Complex}
+
+The (example) code is NEVER too complex.
+
+\subsection{TBD}
 

### 2023-04-05 oopsla23/paper.tex
count mismatch explanation
more overview figures
2 tables
horizontal bars could work too,

@@
 
+    \code{CountMismatch} & Arity mismatch for a function \\
     \code{IncorrectGenericParameterCount} & Arity mismatch for a generic type \\
-    \code{CountMismatch} & \FILL{} \\
     \code{UnknownProperty} & Referenced an invalid field or method  \\
@@
     \code{GenericError}
-    & Generic label for other errors, e.g., declaring a \\
-    & \quad{} supertype that is not a class type
-    %% or iterated over a table without specifying how
+    & Generic label for other non-type errors, such as looping\!\!\! \\
+    & over a table without a clear iteration strategy
+    %% & {}~~declaring a supertype that is not a class type
 
@@
 \end{tabular}
+
+  \begin{minipage}{0.5\textwidth}
+    \begin{tabular}{rr}
+      Session length & \# Sessions \\\midrule
+      1 &  450 \\
+      2 &  199 \\
+      3 &  129 \\
+      4 &  73 \\
+      5 &  60 \\
+      6-10 & 150 \\
+      11-20 & 67 \\
+      21-40 & 11 \\
+      >40 & 7
+      %% max 393
+    \end{tabular}
+  \end{minipage}\begin{minipage}{0.5\textwidth}
+    \begin{tabular}{rr}
+      Time delta & \# Records \\\midrule
+      <=0ms & 0 \\
+      0-5ms & 284 \\
+      5-1000ms & 248 \\
+      1-2s & 153 \\
+      2-5s & 232 \\
+      5-10s & 207 \\
+      10-30s & 524 \\
+      30-60s & 392 \\
+      1-2m & 394 \\
+      2-5m & 559 \\
+      5-10m & 360 \\
+      >10-60m & 502
+    \end{tabular}
+  \end{minipage}
+
   \caption{Dataset overview}

### 2023-04-18 oopsla23/paper.tex
type errors overview, 1st draft

@@
 \end{itemize}
+
+Analysis pipeline to be freely available.
+Data may be available, depends on Roblox.
 
@@
 \begin{figure}[t]
+  %% TODO split counts, total = current + previous
   \begin{tabular}{rl}
@@
 
-  \begin{minipage}{0.5\textwidth}
+  %% TODO these need to be plots, it's too much reading for the quick takeaway
+  %% put exactly this data into bars!
+  \begin{minipage}[t]{0.5\textwidth}
     \begin{tabular}{rr}
@@
     \end{tabular}
-  \end{minipage}\begin{minipage}{0.5\textwidth}
+  \end{minipage}\begin{minipage}[t]{0.5\textwidth}
     \begin{tabular}{rr}
       Time delta & \# Records \\\midrule
-      <=0ms & 0 \\
       0-5ms & 284 \\
@@
   \label{f:dataset-overview}
+\end{figure}
+
+\begin{figure}[t]
+  %% TODO draw bars or curves b/c numbers are too hard to compare horizontally
+
+  TODOD table: mode (fs), project, module, region; cell olde,curr; two rows per actual mode ... or two tables
+
+  Errors in edit region.
+  Each record with an edit region error adds +1 to at least 1 row.
+
+  \begin{tabular}{lrrr}
+    Error Label & \# Added & \# Persist & \# Remove \\\midrule
+    CodeTooComplex
+    & 0 & 0 & 0 \\
+    UnificationTooComplex
+    & 0 & 0 & 0 \\[1ex]
+    SyntaxError
+    & 30 & 0 & 76 \\
+    UnknownSymbol
+    & 12 & 14 & 19 \\
+    UnknownProperty
+    & 3 & 1 & 4 \\
+    UnknownRequire
+    & 1 & 5 & 0 \\
+    GenericError
+    & 1 & 0 & 1 \\
+    CountMismatch
+    &  1 &  0 &  1 \\
+    OptionalValueAccess
+    &  0 &  0 &  1 \\
+  \end{tabular}
+
+  \caption{Errors overview}
+  \label{f:errors-overview}
 \end{figure}

### 2023-04-19 oopsla23/paper.tex
table: rows with error

@@
 \begin{figure}[t]
-  %% TODO split counts, total = current + previous
   \begin{tabular}{rl}
@@
          & 4,471 nocheck + 506 nonstrict + 24 strict \\
+         & 1,628 from module switches \\
    182,247 & total forced strict type errors \\
@@
 \begin{figure}[t]
-  %% TODO draw bars or curves b/c numbers are too hard to compare horizontally
+  %% TODO stacked bars?
+  %% TODO align old;curr
+  \begin{tabular}{lrrrrr}
+    Mode             & \# Rows & E         & Any Error & In Module & In Region \\
+      & & & & (old ; curr) &  (old ; curr) \\\midrule
+    \code{nocheck}   &   4,471 & \code{te} &    5.77\% &  0.20\% ;  2.21\% &  0.16\% ;  0.63\% \\
+                     &         & \code{fs} &   59.18\% & 32.81\% ; 49.16\% &  9.17\% ; 13.13\% \\
+    \code{nonstrict} &     506 & \code{te} &   25.89\% & 12.06\% ; 18.18\% &  6.32\% ;  7.91\% \\
+                     &         & \code{fs} &   62.45\% & 37.94\% ; 53.75\% & 16.60\% ; 20.75\% \\
+    \code{strict}    &      24 & \code{te} &   50.00\% & 29.17\% ; 33.33\% &     0\% ;     0\% \\
+                     &         & \code{fs} &   62.50\% & 50.00\% ; 58.33\% &  4.17\% ;  8.33\% 
+  \end{tabular}
 
-  TODOD table: mode (fs), project, module, region; cell olde,curr; two rows per actual mode ... or two tables
+  \bigskip
 
@@
 
+  %% TODO draw bars or curves b/c numbers are too hard to compare horizontally
   \begin{tabular}{lrrr}
-    Error Label & \# Added & \# Persist & \# Remove \\\midrule
+    Error Label & \# Add & \# Keep & \# Remove \\\midrule
     CodeTooComplex

### 2023-05-03 oopsla23/paper.tex
Update paper.tex
Added email addresses
table, by-mode add/remove error
quick module switch numbers

@@
 \usepackage{pifont}
+\usepackage{siunitx}
 \usepackage{tikz}
@@
 }
-\email{}
+\email{ajeffrey@roblox.com}
 
@@
 }
-\email{}
+\email{mshah@roblox.com}
 
@@
 \newcommand{\chkNo}{\wideas{}{\chkYes}}
+\newcommand{\pct}[1]{\SI{#1}{\percent}}
 
@@
 
-  Errors in edit region.
+  Below, edit region only.
   Each record with an edit region error adds +1 to at least 1 row.
+
+  \smallskip
 
@@
 
+\begin{figure}[t]
+
+  {Few local adds => Few people ignoring script analysis}
+  \smallskip
+
+  \begin{tabular}{lrrrr}
+    & Local Remove & Local Add & Nonlocal Remove & Nonlocal Add \\
+    Mode
+    & \#err (count)
+    & \#err (count)
+    & \#err (count)
+    & \#err (count) \\\midrule
+    \code{nocheck}   & 86 (71) & 0 (0) & 110 (57) & 161 (74) \\
+    \code{nonstrict} & 87 (25) & 9 (3) & 185 (29) & 207 (37) \\
+    \code{strict}    &  1  (1) & 0 (0) &  14  (6) &   6  (3) \\
+  \end{tabular}
+
+  \caption{Per-mode comparison}
+  \label{f:mode-errors}
+\end{figure}
+
 \section{Interpretation}
@@
 \end{verbatim}
+
+\subsection{Module Switches}
+
+FILL need \%s out of all module switches.
+
+How many module switches exit a module with errors?
+nocheck 24, nonstrict 9, strict 3.
+Rare?
+More common for forcedstrict:
+ nocheck 190, nonstrict 16, strict 2.
+
+How many module switches go to a module with errors?
+nocheck 70, nonstrict 9, strict 4
+For FS: nocheck 400, nonstrict 39, strict 6.
 

### 2023-05-03 oopsla23/paper.tex
mode vs. change in error, macros for the modes

@@
 \newcommand{\pct}[1]{\SI{#1}{\percent}}
+\newcommand{\modefont}[1]{\texttt{#1}}
+\newcommand{\mnocheck}{\modefont{nocheck}}
+\newcommand{\mnonstrict}{\modefont{nonstrict}}
+\newcommand{\mstrict}{\modefont{strict}}
 
@@
 \begin{itemize}
-\item \emph{nocheck}: only syntax errors are reported,
-\item \emph{nonstrict}: all syntax errors, and a subset of ``high probability'' type errors, are reported, or
-\item \emph{strict}: all syntax errors and type errors, are reported.
+  \item \mnocheck{}: only syntax errors are reported,
+  \item \mnonstrict{}: all syntax errors, and a subset of ``high probability'' type errors, are reported, or
+  \item \mstrict{}: all syntax errors and type errors, are reported.
 \end{itemize}
@@
 Both modes are opt-in. Creators have the option to make nonstrict mode
-the default rather than nocheck mode.
+the default rather than \mnocheck{} mode.
 
-Even in nocheck mode, \anon{Roblox Studio} performs type inference, since
+Even in \mnocheck{} mode, \anon{Roblox Studio} performs type inference, since
 the results are needed by type-directed tooling such as autocomplete and
@@
   \item[D1] metadata: uid, timestamp, session duration
-  \item[D2] mode (strict / nonstrict / nocheck)
+  \item[D2] mode (\mnocheck{}, \mnonstrict{}, \mstrict{})
   \item[D3] a 1-bit reason for sending: did the user switch module?
@@
     \end{itemize}
-  \item[D8] ForcedStrict "what if" errors for nocheck: specific errors for the
+  \item[D8] ForcedStrict "what if" errors for \mnocheck{}: specific errors for the
     edit region computed as if the user had strict mode enabled (even though they
@@
   \subitem [[ For context: the type system is designed to give useful feedback
-    to all sorts of code, whether strict-typed, nonstrict-typed, or nocheck.
+    to all sorts of code, whether strict-typed, nonstrict-typed, or \mnocheck{}.
     But it has an escape hatch: CTC. Hope it's rarely needed. ]]
@@
       \item Count the overall CTC errors (in each mode) for a coarse view of
-        how rare they are. [D9 of course, D8 to compare nocheck code with
+        how rare they are. [D9 of course, D8 to compare \mnocheck{} code with
         strict-mode code via the same typechecker]
@@
         %% \subitem Count CTC errors within each session to see if these errors are intermittent, or if they stick around.
-[D9 for the count, D8 for nocheck vs strict, D7 because CTC might show up in the edit region]
+[D9 for the count, D8 for \mnocheck{} vs strict, D7 because CTC might show up in the edit region]
     \end{itemize}
@@
   \begin{itemize}
-    \item Does nocheck code have many more latent errors than the others?
+    \item Does \mnocheck{} code have many more latent errors than the others?
     \subitem [D8 latent errors come from the "what if" analysis]
@@
 % General Roblox trends
-% How many sessions use strict / nonstrict / nocheck?
-% Expect few strict/nonstrict, lots of nocheck
+% How many sessions use strict / nonstrict / \mnocheck{}?
+% Expect few strict/nonstrict, lots of \mnocheck{}
 % How often does mode change within a session?
@@
    5,001 & total logs \\
-         & 4,471 nocheck + 506 nonstrict + 24 strict \\
+         & 4,471 \mnocheck{} + 506 \mnonstrict{} + 24 \mstrict{} \\
          & 1,628 from module switches \\
@@
     & \begin{tabular}{l}
-        1,143 single mode = 1046 nocheck + 94 nonstrict + 3 strict \\
+      1,143 single mode = 1046 \mnocheck{} + 94 \mnonstrict{} + 3 \mstrict{} \\
         detected 2 multi-mode projects, 2 mode upgrades, 2 mode downgrades
@@
       & & & & (old ; curr) &  (old ; curr) \\\midrule
-    \code{nocheck}   &   4,471 & \code{te} &    5.77\% &  0.20\% ;  2.21\% &  0.16\% ;  0.63\% \\
+    \mnocheck{}   &   4,471 & \code{te} &    5.77\% &  0.20\% ;  2.21\% &  0.16\% ;  0.63\% \\
                   &         & \code{fs} &   59.18\% & 32.81\% ; 49.16\% &  9.17\% ; 13.13\% \\
-    \code{nonstrict} &     506 & \code{te} &   25.89\% & 12.06\% ; 18.18\% &  6.32\% ;  7.91\% \\
+    \mnonstrict{} &     506 & \code{te} &   25.89\% & 12.06\% ; 18.18\% &  6.32\% ;  7.91\% \\
                   &         & \code{fs} &   62.45\% & 37.94\% ; 53.75\% & 16.60\% ; 20.75\% \\
-    \code{strict}    &      24 & \code{te} &   50.00\% & 29.17\% ; 33.33\% &     0\% ;     0\% \\
+    \mstrict{}    &      24 & \code{te} &   50.00\% & 29.17\% ; 33.33\% &     0\% ;     0\% \\
                   &         & \code{fs} &   62.50\% & 50.00\% ; 58.33\% &  4.17\% ;  8.33\% 
@@
     & \#err (count) \\\midrule
-    \code{nocheck}   & 86 (71) & 0 (0) & 110 (57) & 161 (74) \\
-    \code{nonstrict} & 87 (25) & 9 (3) & 185 (29) & 207 (37) \\
-    \code{strict}    &  1  (1) & 0 (0) &  14  (6) &   6  (3) \\
+    \mnocheck{}   & 86 (71) & 0 (0) & 110 (57) & 161 (74) \\
+    \mnonstrict{} & 87 (25) & 9 (3) & 185 (29) & 207 (37) \\
+    \mstrict{}    &  1  (1) & 0 (0) &  14  (6) &   6  (3) \\
   \end{tabular}
@@
 
+FILL why study, what implications?
+
 FILL need \%s out of all module switches.
@@
 How many module switches exit a module with errors?
-nocheck 24, nonstrict 9, strict 3.
+\mnocheck{} 24, \mnonstrict{} 9, \mstrict{} 3.
 Rare?
 More common for forcedstrict:
- nocheck 190, nonstrict 16, strict 2.
+ \mnocheck{} 190, \mnonstrict{} 16, \mstrict{} 2.
 
 How many module switches go to a module with errors?
-nocheck 70, nonstrict 9, strict 4
-For FS: nocheck 400, nonstrict 39, strict 6.
+\mnocheck{} 70, nonstrict 9, strict 4
+For FS: \mnocheck{} 400, \mnonstrict{} 39, \mstrict{} 6.
+
+\subsection{Types and Quality of Life}
+
+FILL error density: number vs region size
+
+\begin{figure}[h]
+  \begin{tabular}{llrrr}
+    Mode & E & Add & Keep & Remove \\\midrule
+    \mnocheck{}   & \code{te} & - & - & - \\
+                  & \code{fs} & - & - & - \\
+    \mnonstrict{} & \code{te} & 16 & 18 & 21 \\
+                  & \code{fs} & 19 & 18 & 18 \\
+    \mstrict{}    & \code{te} & - & - & - \\
+                  & \code{fs} & - & - & - \\
+  \end{tabular}
+
+  \caption{Does mode influence the change in total error count?}
+  \label{f:edit-deltas}
+\end{figure}
+
+TODO scribble figures ...
 

### 2023-05-10 oopsla23/paper.tex
example timelines

@@
 
+TODO nocheck accumulated errors?
+
+\subsection{Example Sessions}
+
+FILL example sessions.
+
+\Cref{f:ex-session-strict} begins in strict mode
+but jumps to nocheck mode as well.
+The jumps sometimes coincide with module switches (vertical line),
+sometimes not.
+The green line counts type errors.in the codebase.
+
+\begin{figure}[t]
+  \includegraphics[width=0.7\columnwidth]{img/example-session-strict.pdf}
+  \caption{Example strict session}
+  \label{f:ex-session-strict}
+\end{figure}
+
+\Cref{f:ex-session-nonstrict} presents the three sessions from \mnonstrict{} mode
+that have the greatest number of events (not necessarily the longest time spans).
+There are no module or mode switches.
+There are few type errors.
+One high point shows 10 errors, but these errors are gone (probably fixed) at
+the next event.
+
+\begin{figure}[t]
+  \includegraphics[width=0.7\columnwidth]{img/example-session-nonstrict.pdf}
+  \caption{Three most populated nonstrict sessions}
+  \label{f:ex-session-nonstrict}
+\end{figure}
+
+\Cref{f:ex-session-nocheck} shows a roller coaster from \mnocheck{} mode.
+Several hundred ForcedStrict type errors appear and disappear between events.
+There are lots of module switches too \ldots but that makes no difference
+for these project-wide error counts.
+
+\begin{figure}[t]
+  \includegraphics[width=0.7\columnwidth]{img/example-session-nocheck.pdf}
+  \caption{Example nocheck session}
+  \label{f:ex-session-nocheck}
+\end{figure}
+
+\subsection{Aggregate Sessions}
+
 TODO scribble figures ...

### 2023-05-10 oopsla23/paper.tex
anon
aggregate best fits (scribbles)

@@
   review,
-%  anonymous,
+  anonymous,
 ]{acmart}
@@
 
-TODO scribble figures ...
+\begin{figure}[t]
+  \includegraphics[width=0.8\columnwidth]{img/fit-nocheck-Force.pdf}
+  \includegraphics[width=0.8\columnwidth]{img/fit-nonstrict-TypeE.pdf}
+  \includegraphics[width=0.8\columnwidth]{img/fit-strict-TypeE.pdf}
+  \caption{Best-fit curves for errors in each mode (\code{fs} for \mnocheck{}, \code{te} for the others)}
+  \label{f:scribbles}
+\end{figure}
 

### 2023-05-12 oopsla23/paper.tex
note wonky clientTimestamp

@@
 \Cref{f:records-per-hour} shows when data arrived across the whole dataset.
+FILL: currently clientTimestamp, which has caveats.
+Unclear what timezone they are from.
+Some records have a client timestamp much older than the server, up to 1 week behind.
+Some are ahead of the server by a day.
+Why?
+Could be wrong local settings on the computer, or a delayed record due to lost
+internet connection or closed laptop.
 

### 2023-05-15 oopsla23/paper.tex
fig 1: full-size anon version
recolor fit lines, red=fs green=te

@@
 \begin{figure}
-  \anon{
+
+  \makeatletter
+  \if@ACM@anonymous
+    \includegraphics[width=.45\textwidth]{img/anon-studio.png}
+    \includegraphics[width=.45\textwidth]{img/anon-studio-ide.png}
+  \else
     \includegraphics[width=.45\textwidth]{img/roblox-studio.png}
     \includegraphics[width=.45\textwidth]{img/roblox-studio-ide.png}
-  }
+  \fi
+  \makeatother
+
   \caption{\anon{Roblox Studio 3D creation} tools (left) and IDE (right)}
@@
 \begin{figure}[t]
-  \includegraphics[width=0.8\columnwidth]{img/fit-nocheck-Force.pdf}
-  \includegraphics[width=0.8\columnwidth]{img/fit-nonstrict-TypeE.pdf}
-  \includegraphics[width=0.8\columnwidth]{img/fit-strict-TypeE.pdf}
+  \includegraphics[width=0.8\columnwidth]{img/fit-nocheck-Force-3.pdf}
+  \includegraphics[width=0.8\columnwidth]{img/fit-nonstrict-TypeE-3.pdf}
+  \includegraphics[width=0.8\columnwidth]{img/fit-strict-TypeE-3.pdf}
   \caption{Best-fit curves for errors in each mode (\code{fs} for \mnocheck{}, \code{te} for the others)}

### 2023-05-22 oopsla23/paper.tex
relabel tbl col

@@
 
+FILL ignore sessions close to start / end time?
+remove all syntax errors from later stages?
+
 \section{Results}
@@
     Mode
-    & \#err (count)
-    & \#err (count)
-    & \#err (count)
-    & \#err (count) \\\midrule
+    & \#err (\#log)
+    & \#err (\#log)
+    & \#err (\#log)
+    & \#err (\#log) \\\midrule
     \mnocheck{}   & 86 (71) & 0 (0) & 110 (57) & 161 (74) \\

### 2023-05-22 oopsla23/paper.tex
intro, firm up RQs

@@
   of type error reporting.
+  We find FILL.
 \end{abstract}
@@
 a traditional type system, which focuses on compilation and memory safety.
-\anon{Luau} has a very heterogeneous user community, ranging from
+\anon{Luau} has a diverse user community, ranging from
 students in code camps to professional development studios. These
@@
 enabling rapid creation and ensuring software quality.
+\anon{Luau} therefore supports three typing modes---strict types,
+relaxed (nonstrict) types, and no types---and lets users switch
+between modes gradually~\cite{st-sfp-2006,tfffgksst-snapl-2017}, one module at a time.
 
 In this paper, we investigate methods for measuring the effectiveness
-of the \anon{Luau} type system in development of \anon{Roblox} scripts.
+of the \anon{Luau} type system in the development of \anon{Roblox} scripts.
 In comparison to prior work~(\cref{s:related}), which is either small in scale
@@
 indicates a type error site. Nonetheless, the type inference system
-still runs (since it drives autocomplete and other type-based tools) and
-so we can record which type errors would have been reported had the
-user enabled type error reporting. As a result, we can investigate
-which type errors are  fixed by users, even if they did not opt in to
-type error reporting.
+still runs in the background (since it drives autocomplete and other
+type-based tools), letting us see which type errors would have
+been reported and whether users fix these errors over time.
 
 With this telemetry data, we investigate research questions about
-the adoption and benefits of type analysis.
-First, to learn how many creators use type analysis of any sort
-and to estimate whether they pay attention to the results (\FILL{} really, attention?).
-Once in type-analysis mode, do creators stay, or revert to no-check?
-Second, what errors do creators face and how do they respond.
-Are the error highlights helpful for removing the error?
-Is there any indication that type analysis improves the quality of the development experience?
+the adoption and benefits of type analysis:
+\begin{description}
+  \item[RQ1.]
+    How many creators use type analysis,
+    how often does a project use multiple analysis modes,
+    and how often do creators turn the analysis off?
+  \item[RQ2.]
+    For modules that use type analysis: which errors
+    arise, how do users respond, and which errors
+    tend to persist through subsequent edits?
+  \item[RQ3.]
+    What impact, if any, does type analysis have on overall code quality?
+\end{description}
 
-This paper is the first to use telemetry as a mechanism for
+Answers to these questions can inform the design of gradual languages
+in general, and of extensions to \anon{Lua} and languages with semantic
+subtyping in particular.
+
+At a higher level, this paper is a step toward data-driven language
+design. We are the first to use telemetry as a mechanism for
 large-scale measurement of the effectiveness of type error reporting.
@@
 creator sentiment, but provide complementary data at scale.
+FILL can stats answer questions that point-by-point analysis cannot?
 
@@
 
+FILL regrets, lessons learned, advice for next time
+
 \begin{acks}

### 2023-05-24 oopsla23/paper.tex
rq notes, rw notes

@@
   \item[RQ1.]
-    How many creators use type analysis,
-    how often does a project use multiple analysis modes,
-    and how often do creators turn the analysis off?
+    How many creators use type analysis?
+    How often do projects contain modules with different
+    analysis modes?
+    How often do creators turn analysis off?
   \item[RQ2.]
-    For modules that use type analysis: which errors
-    arise, how do users respond, and which errors
-    tend to persist through subsequent edits?
+    For modules that use type analysis:
+    which errors arise,
+    how do creators respond,
+    and which errors tend to persist despite subsequent edits?
   \item[RQ3.]
-    What impact, if any, does type analysis have on overall code quality?
+    What impact does type analysis have on the number of background type
+    errors?
+    For example, do background errors pile up in unanalyzed projects?
 \end{description}
 
-Answers to these questions can inform the design of gradual languages
-in general, and of extensions to \anon{Lua} and languages with semantic
-subtyping in particular.
+Beyond their immediate applications to \anon{Lua},
+answers to these questions have implications for gradual types,
+success types, and semantic subtyping.
 
-At a higher level, this paper is a step toward data-driven language
-design. We are the first to use telemetry as a mechanism for
-large-scale measurement of the effectiveness of type error reporting.
-Our data captures \textbf{XX} sessions, for a total of \textbf{YY}
-hours and \textbf{ZZ} type errors.
+At a higher level, this paper is the first to use telemetry 
+to study a typechecker.
+It thus represents a step toward data-driven language design,
+informed by many users' actual practice.
+Our data captures \textbf{XX} sessions
+for a total of \textbf{YY} hours
+and \textbf{ZZ} type errors.
+By contrast to typical qualitative methods (e.g., surveys and interviews), it
+is not restricted to users' \emph{perceptions} about their work and it is not
+limited to a small number of users.
 
-Telemetry cannot replace user studies, as there is no way to measure
-creator sentiment, but provide complementary data at scale.
 FILL can stats answer questions that point-by-point analysis cannot?
@@
 
+Telemetry cannot replace user studies, as there is no way to measure
+creator sentiment, but provide complementary data at scale.
+
 \section{Related Work}
@@
 
+\citet{lnsmc-usenix-2018} blame-proportional logging,
+scale up ``near'' points of interest,
+could be useful for us in the future.
+\citet{fnm-sigmod-2020} ditto, faster than statistical at finding root-cause,
+deterministic.
+
 \section{Discussion}

### 2023-06-02 oopsla23/paper.tex
CodeTooComplex notes

@@
 
-\subsection{Code Too Complex}
+\subsection{Dead Ends: Too Complex}
 
-The (example) code is NEVER too complex.
+The typechecker has two internal limits regarding the problems it will attempt
+to solving.
+Hitting a limit triggers either a \code{CodeTooComplex} or
+\code{UnificationTooComplex} error.
+A casual user should never see these errors.
+
+Based on the data, these errors are indeed extremely rare.
+Out of 1.5M telemetry records, only 11 report any \code{TooComplex} errors.
+These records came from three sessions (possibly by a single user), and each
+record counted at most 4 \code{CodeTooComplex} errors and zero
+\code{UnificationTooComplex} errors.
+There is no evidence to suggest that the limits are too strict or that the
+typechecker should improve.
 
@@
 Problematic, because syntax errors are uninteresting (easy to fix),
-but common, which makes it likely that our per-keystroke telemetry will record them.
+but common, which makes it likely that our per-keystroke telemetry (1/2k) will record them.
 Edit region has specific counts.
@@
 \label{s:related}
+
+%% zfstt-ieeesensors-2022
 

### 2023-07-29 programming23/paper.tex
reformat for programming

--- 2023-06-02 oopsla23/paper.tex
@@
+% -*- coding: utf-8; -*-
+% vim: set fileencoding=utf-8 :
+
 % The paper on using Luau telemetry to measure the effectiveness of type error reporting.
 
-\documentclass[
-  acmsmall,
-  review,
-  anonymous,
-]{acmart}
-
-%\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[english,submission,cleveref]{programming}
+%% use 'submission' for initial submission, remove it for camera-ready (see 5.1)
 
 \overfullrule=1mm
-\citestyle{acmauthoryear}
-%\setcitestyle{round}
+
+%% BEGIN tobias pape 2021-11-06
+\makeatletter
+\newcommand*\abstractpart[1]{\unskip\par\noindent{\firamedium\color{P@GrayFG}{#1}}\enspace}
+\makeatother
+%% END
 
 \usepackage{alltt}
-% \usepackage{amssymb} -- already loaded by acmart
+\usepackage[backend=biber]{biblatex}
+\usepackage{amssymb}
 \usepackage{calc}
@@
 
+\addbibresource{bib.bib}
+
+%%%%%%%%%%%%%%%%%%
+\paperdetails{
+  %% perspective options are: art, sciencetheoretical, scienceempirical, engineering.
+  %% Choose exactly the one that best describes this work. (see 2.1)
+  perspective=scienceempirical,
+  %% State one or more areas, separated by a comma. (see 2.2)
+  %% Please see list of areas in http://programming-journal.org/cfp/
+  %% The list is open-ended, so use other areas if yours is/are not listed.
+  area={Data mining for programming, Telemetry},
+  %% You may choose the license for your paper (see 3.)
+  %% License options include: cc-by (default), cc-by-nc
+  % license=cc-by,
+}
+%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%
+%% These data are provided by the editors. May be left out on submission.
+%\paperdetails{
+%  submitted=2023-1-31,
+%  published=2016-10-11,
+%  year=2016,
+%  volume=1,
+%  issue=1,
+%  articlenumber=1,
+%}
+%%%%%%%%%%%%%%%%%%
+
 \begin{document}
@@
 
-\author{Ben Greenman}
-\orcid{0000-0001-7078-9287}
-\affiliation{%
-  \institution{Brown University}
-  \city{Providence}
-  \state{Rhode Island}
-  \country{USA}
-}
-\email{benjaminlgreenman@gmail.com}
+\author[a]{Ben Greenman}
+\authorinfo{(\email{benjamin.l.greenman@gmail.com}) is a postdoc at Brown University. He will be joining the University of Utah in Fall~2023.}
+\affiliation[a]{Brown University, Providence, RI, USA}
+% \orcid{0000-0001-7078-9287}
+\affiliation[a]{Brown University, Providence, RI, USA}
 
-\author{Alan Jeffrey}
-\orcid{0000-0001-6342-0318}
-\affiliation{%
-  \institution{Roblox}
-  \city{San Mateo}
-  \state{California}
-  \country{USA}
-}
-\email{ajeffrey@roblox.com}
+\author[b]{Alan Jeffrey}
+\authorinfo{(\email{ajeffrey@roblox.com}) TBD}
+% \orcid{0000-0001-6342-0318}
+\affiliation[b]{Roblox, San Mateo, CA, USA}
 
-\author{Shriram Krishnamurthi}
-\orcid{0000-0001-5184-1975}
-\affiliation{%
-  \institution{Brown University}
-  \city{Providence}
-  \state{Rhode Island}
-  \country{USA}
-}
-\email{shriram@brown.edu}
+\author[a]{Shriram Krishnamurthi}
+% \orcid{0000-0001-5184-1975}
+\authorinfo{(\email{shriram@brown.edu}) is the Vice President of Programming Languages (no, not really) at Brown University.}
 
-\author{Mitesh Shah}
-\orcid{TODO}
-\affiliation{%
-  \institution{Roblox}
-  \city{San Mateo}
-  \state{California}
-  \country{USA}
-}
-\email{mshah@roblox.com}
+\author[b]{Mitesh Shah}
+\authorinfo{(\email{mshah@roblox.com}) TBD.}
 
 %\renewcommand{\shortauthors}{...}
-
-%%
-%% The abstract is a short summary of the work to be presented in the
-%% article.
-\begin{abstract}
-  \anon{Roblox Studio} is a tool which places programming in the hands of
-  millions of creators, ranging from high school students to professional
-  development studios. It has the ability to report telemetry data,
-  which allows large-scale measurement of the creator experience. In
-  this paper, we discuss one use of telemetry, to measure the experience
-  of type error reporting.
-  We find FILL.
-\end{abstract}
 
@@
 
+\begin{abstract}
+  {Roblox Studio} is a tool which places programming in the hands of
+  millions of creators, ranging from high school students to professional
+  development studios. It has the ability to report telemetry data,
+  which allows large-scale measurement of the creator experience. In
+  this paper, we discuss one use of telemetry, to measure the experience
+  of type error reporting.
+  We find FILL.
+
+  %\let\paragraph\abstractpart
+
+  % \paragraph{Context}
+  %% What is the broad context of the work? What is
+  %% the importance of the general research area?
+
+  % \paragraph{Inquiry}
+  %% What problem or question does the paper
+  %% address? How has this problem or question been
+  %% addressed by others (if at all)?
+
+  %\paragraph{Approach}
+  %% What was done that unveiled new knowledge?
+
+  %\paragraph{Knowledge}
+  %% What new facts were uncovered? If the
+  %% research was not results oriented, what new
+  %% capabilities are enabled by the work?
+
+  %\paragraph{Grounding}
+  %% What argument, feasibility proof, artifacts,
+  %% or results and evaluation support this work?
+
+  %\paragraph{Importance}
+  %% Why does this work matter?
+
+\end{abstract}
+
 \section{Introduction}
@@
 
-\anon{Roblox} is a platform for \anon{shared virtual experiences},
+{Roblox} is a platform for {shared virtual experiences},
 with 56~million Daily Active Users, and 49~billion hours of engagement in
-2022~\anon[(ANONYMIZED CITATION)]{\cite{roblox-quarterly-results}}.
-There are \textbf{XX}~million creators using \anon{Roblox Studio},
+2022~\cite{roblox-quarterly-results}.
+There are \textbf{XX}~million creators using {Roblox Studio},
 and \textbf{YY}~million creations.
 
-\anon{Roblox experiences} are scripted using the 
-\anon{Luau} programming language~\anon[(ANONYMIZED CITATION)]{\cite{luau-lang.org}},
-an extension of \anon{Lua~5.1~\cite{lua}}.
+{Roblox experiences} are scripted using the 
+{Luau} programming language~\cite{luau-lang.org},
+an extension of {Lua~5.1~\cite{lua}}.
 The main extension is the addition of a static type system, which uses
@@
 are used primarily in type-driven tooling such as autocomplete
-and API documentation~\anon[(ANONYMIZED CITATION)]{\cite{luau-autocomplete}},
+and API documentation~\cite{luau-autocomplete},
 but creators can also opt in to receiving type error reports.
 
-As discussed in~\anon[(ANONYMIZED CITATION)]{\cite{bfj-hatra-2021}},
-the goals of the \anon{Luau} type system are rather different from
+As discussed in~\cite{bfj-hatra-2021},
+the goals of the {Luau} type system are rather different from
 a traditional type system, which focuses on compilation and memory safety.
-\anon{Luau} has a diverse user community, ranging from
+{Luau} has a diverse user community, ranging from
 students in code camps to professional development studios. These
@@
 enabling rapid creation and ensuring software quality.
-\anon{Luau} therefore supports three typing modes---strict types,
+{Luau} therefore supports three typing modes---strict types,
 relaxed (nonstrict) types, and no types---and lets users switch
@@
 In this paper, we investigate methods for measuring the effectiveness
-of the \anon{Luau} type system in the development of \anon{Roblox} scripts.
+of the {Luau} type system in the development of {Roblox} scripts.
 In comparison to prior work~(\cref{s:related}), which is either small in scale
@@
 
-\anon{Roblox Studio} has a telemetry system, which is used to gauge
+{Roblox Studio} has a telemetry system, which is used to gauge
 the effectiveness of creation features. This system stochastically
@@
 
-Most users of \anon{Roblox Studio} do not opt in to type error
+Most users of {Roblox Studio} do not opt in to type error
 reporting, and so they do not see the ``squiggly underlining'' that
@@
 
-Beyond their immediate applications to \anon{Lua},
+Beyond their immediate applications to {Lua},
 answers to these questions have implications for gradual types,
@@
 
-\section{\anon{Roblox} Context}
+\section{{Roblox} Context}
 % https://create.roblox.com/docs/scripting/luau
@@
 
-  \makeatletter
-  \if@ACM@anonymous
-    \includegraphics[width=.45\textwidth]{img/anon-studio.png}
-    \includegraphics[width=.45\textwidth]{img/anon-studio-ide.png}
-  \else
   \includegraphics[width=.45\textwidth]{img/roblox-studio.png}
   \includegraphics[width=.45\textwidth]{img/roblox-studio-ide.png}
-  \fi
-  \makeatother
 
-  \caption{\anon{Roblox Studio 3D creation} tools (left) and IDE (right)}
+  \caption{{Roblox Studio 3D creation} tools (left) and IDE (right)}
   \label{fig:roblox-studio}
@@
 
-Creators of \anon{Roblox experiences} use \anon{Roblox Studio},
-which combines \anon{3D creation} tools as well as an Integrated
+Creators of {Roblox experiences} use {Roblox Studio},
+which combines {3D creation} tools as well as an Integrated
 Developer Environment (IDE), as seen in Fig.~\ref{fig:roblox-studio}.
@@
 misspellings of property names are common enough to report in both
-strict and nonstrict mode. See~\anon[ANONYMIZED CITATION]{\cite{bfj-hatra-2021}}
+strict and nonstrict mode. See~\cite{bfj-hatra-2021}
 for a more detailed discussion of the rationale for strict and nonstrict mode.
@@
 
-Even in \mnocheck{} mode, \anon{Roblox Studio} performs type inference, since
+Even in \mnocheck{} mode, {Roblox Studio} performs type inference, since
 the results are needed by type-directed tooling such as autocomplete and
@@
 
-The state of the world in a \anon{Roblox} experience is captured by
+The state of the world in a {Roblox} experience is captured by
 the \emph{data model}, which is a tree of \emph{instances}, such as
@@
   \includegraphics{img/row-distribution.pdf}
-  \Description{FILL Histogram with approximately 100 records per hour, except for a 600-record spike near the end of Jan 12th.}
   \caption{Telemetry records per hour. Each tick on the $x$-axis marks the start of a new day in California.}
@@
   TODO file size distribution: table or bars %% \includegraphics{img/row-distribution.pdf}
-  \Description{FILL}
   \caption{Size of analyzed code: number of files, number of lines , and lines in edit region}
@@
 
-\citet{lnsmc-usenix-2018} blame-proportional logging,
+\cite{lnsmc-usenix-2018} blame-proportional logging,
 scale up ``near'' points of interest,
 could be useful for us in the future.
-\citet{fnm-sigmod-2020} ditto, faster than statistical at finding root-cause,
+\cite{fnm-sigmod-2020} ditto, faster than statistical at finding root-cause,
 deterministic.
@@
 
-\begin{acks}
-  TBD
+\acks
 
 Greenman was supported by
-  \grantsponsor{NSF}{NSF}{https://www.nsf.gov} grant
- \href{"https://www.nsf.gov/awardsearch/showAward?AWD_ID=2030859"}{\grantnum{NSF}{CCF 2030859}}
+grant \href{https://nsf.gov/awardsearch/showAward?AWD_ID=2030859&HistoricalAwards=false}{2030859}
 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
-\end{acks}
 
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{bib}
+\newpage
+
+{ \printbibliography }
 
 \end{document}
-\endinput
+
+% Local Variables:
+% TeX-engine: luatex
+% End:

### 2023-07-31 programming23/paper.tex
telemetry design notes

@@
 
-\overfullrule=1mm
+%%\overfullrule=1mm
 
@@
 Beyond their immediate applications to {Lua},
-answers to these questions have implications for gradual types,
-success types, and semantic subtyping.
+answers to these questions have implications for the design
+of gradual types, success types, and semantic subtyping.
+Luau represents the first large-scale combination of these features,
+and lessons from this experience can inform future applications.
 
@@
     Label & Interpretation \\\midrule
-    \code{CodeTooComplex} & Type analysis failed, cannot understand the code \\
-    \code{UnificationTooComplex} & Type analysis failed, unification solver hit a limit \\
+    \code{CodeTooComplex} & Type analysis failed, cannot understand the code\!\!\! \\
+    \code{UnificationTooComplex} & Type analysis failed, unification solver hit a limit\!\!\! \\
     \code{SyntaxError} & Basic parse error, e.g., \code{for if end} \\
@@
     \code{GenericError}
-    & Generic label for other non-type errors, such as looping\!\!\! \\
-    & over a table without a clear iteration strategy
-    %% & {}~~declaring a supertype that is not a class type
+    & Generic label for other non-type errors, such as\\
+    & \hbox{}~~looping over an unordered table
+    %% attempting to extend a type that does not describe a class
+    %% attempting to iterate over a table without a clear order
 
@@
 
-No PII whatsoever
+\paragraph{Goals:}
 
-Random session selection
+\begin{itemize}
+  \item
+    Learn about type system adoption and use
+  \item
+    Discover roadblocks to adoption
+  \item
+    Assess the helpfulness of error highlights
+\end{itemize}
 
-Random events during session, except module switch.
+\paragraph{Limits:}
 
-No access to other key events: save, exit.
-Don't want access to run, publish.
+\begin{itemize}
+\item
+  No personal data (PII)
+  \subitem
+    Fairly obvious.
+    But, we also can't collect directory names, filenames, or code snapshots
+    because these all might have PII throughout.
+\item
+  No data about public events.
+  \subitem
+    Contributions to the Roblox Marketplace, for instance, are a public event.
+    Although it would be interesting to know the number of type errors
+    in published experiences, this data might be used to identify creators
+    who publish often --- or on a distinct schedule.
+\item
+  Low cost
+  \subitem
+    Building, sending, and storing the data must all run quickly.
+    FILL alan: what was our hard cap on Kibana use?
+\end{itemize}
+
+\paragraph{Solutions:}
+
+\begin{itemize}
+  \item
+    Collect metrics instead of source code: lines, modules, num errors
+  \item
+    Track an edit range to compare type errors with edits,
+    use a coarse $[min, max]$ interval for speed
+  \item
+    Count specific errors only for the current edit range
+    (reduces size of logs)
+  \item
+    Log randomly on keystroke instead of per specific event.
+    One exception: log module switches because that helps discover
+    codebases that use several analysis modes.
+    Randomness helps lower the amount of data.
+  \item
+    Observe sessions randomly (again to lower the amount of data).
+\end{itemize}
+
+%% Extra Roblox constraints:
+%% - No access to some events: save, exit
+%% - No idea if analysis widget is open, closed, or obscured
+%% - Don't know which module is the data model
+%%   (no ids for ANY module for that matter)
 

### 2023-08-02 programming23/paper.tex
rm cleveref, its in programming.sty
data review

@@
 \usepackage{calc}
-\usepackage{cleveref}
 \usepackage{listings}
@@
 
+Prior work: transparent telemetry, VS Code telemetry, many others undocumented
+that are probably not cleary documented.
+\Cref{t:telemetry-design} gives an overview.
+
 \paragraph{Goals:}
@@
     (reduces size of logs)
+  \item
+    Within the edit range, collect data before and after the latest edits.
   \item
@@
     & Transparent & \chkNo    & \chkNo     & \chkYes       & \chkNo      \\
-  * & Lite        & \chkNo    & \chkYes    & \chkNo        & \chkYes     \\
+  * & Us          & \chkNo    & \chkYes    & \chkNo        & \chkYes     \\
     & \code{.NET} & \chkMaybe & \chkYes    & \chkYes       & \chkYes     \\

### 2023-08-11 programming23/paper.tex
quick related work notes, lots of unbound cites

@@
 
-%% zfstt-ieeesensors-2022
+%% - zfstt-ieeesensors-2022
+%% - lit review, benefits of static types
+%%   http://danluu.com/empirical-pl/
+%% - thomas kennedy .... generic usage monitoring ascilite 2003
+%% - ahmadzadeh elliman higgins iticse 2005
+%% - buffardi etal 2014 adaptive and social mechanisms
+%% - bddf-icse-2016
+%% -  sh edwards and jason snyder icer 2009 comparing effective and ineffective grading platform
+%% - murphy etal sigcse 2009 Retina: helping students
+%% - ESP workshop on empirical studies of programming
+%%   https://dl.acm.org/doi/proceedings/10.1145/266399
+%%   nothing sounds big ... empirical yes, large no
+%% - Choppella Haynes IU TR 426
+%% - Tip Dinesh tosem 2001
+
+\cite{bgimgm-cse-2016}
+study Java compiler error messages after an intervention, counting num errors,
+counting repeats, counting per user
+
+\cite{anna-russo-kennedy-ms-2006}
+BitFit, web-based java ide log num hint requests num, result of check answer
+requests num compile, num run, no code?! good we have the same constraint
+
+\cite{ab-sigcse-2015}
+frequency, time-to-fix, spread of errors
+ (we can't do time)
+parsing compile error + source code
+ sometimes with custom parsers
+
+\cite{m-masters-2016}
+study errors in blackbox dataset,
+ overwhelmingly syntax
+
+\cite{bkmu-sigcse-2014}
+blackbox, 100k users
+source code, diffs
+error line, column, message;
+keep all comments EXCEPT one above the class declaration
+
+\cite{bask-icer-2018}
+2TB data, Java programs
+18 publications survey, technical challenges in analysis
+ most pubs analyze errors;
+nobody asked:
+ which exns do students encounter (what?! mccall kolling 2014)
+ execution / editing timelines
+(blackbox analysis tutorial in sigcse 2020 on a mini dataset, to help researchers start writing an analysis)
+
+\cite{t-hatra-2021}
+plan for a large user-centered study 
+... not very deep, look at edit sequences, replay for errors
+
+\cite{cdhhjklwya-hatra-2020}
+evaluating the PLIERS design process (Coblenz etal 202X)
+teaching students to use PLIERS; everyone ran a user study!
+6 languages
+ 3 ran usability studies
+recommend talk-alouds
+
+\cite{gstf-hatra-2021}
+liquid types for java
+30 devs user study
+ zoom
+
+\cite{lfgc-pldi-2007}
+typechecker never errors,
+ calls a helper that looks for nearby programs that do typecheck
+tested on
+ 10 students, 2yrs professional experience
+ 5 hw assignments
+ 2122 files collected;
+won \pct{19}, lost \pct{17};
+ill-suited for things like (let x = e1 in ....),
+ change e1,
+ use typechecker feedback to guide edits;
+small data!
+
+\cite{w-popl-1986}
+reporting source, not detection
+tested on 9 type errors
+ all \emph{deletions}, such as forgetting to inject in an enum
+
+\cite{sscwj-oopsla-2017}
+NATE numerical analysis of type errors
+5000 labeled programs
+ errors from class, first fix
+
+\cite{sjw-jfp-2018}
+type error witness generation
+4500 ill-typed student programs
+ \pct{85} of time, synthesis works
+ \pct{70} of time, locate source
+
+\cite{h-dissertiation-2005}
+constraint-based type inference;
+general te criteria:
+- target audience (novice / expert)
+- output format (text / viz)
+- interactivity (yes / no)
+- heuristics (yes / no)
+- primary-or-external (most type errors are easy, some could really use dedicated help)
+
+\cite{hw-scp-2004}
+location of type error as a slice,
+ not a srcloc,
+ not a subtree
 
 \paragraph{Research on Errors}
+
+TODO Liblit dissertiation
 

### 2023-08-11 programming23/paper.tex
quick analysis notes from gdoc

@@
 
+\subsection{Analysis 2}
+
+TODO work this into the above or below.
+
+What can we learn from the telemetry? We have 1.5 million records from Feb 22
+to April 14 (roughly 27K records per day).
+They mention +70 million type errors.
+(Some records point to a large number of errors. FILL median, max. Could be cascading,
+which makes the 70million number uninterpretable.)
+What takeaways are there??!
+
+First,
+users rarely switch type analysis mode. There are 3 modes available and users
+are free to switch between them, or to mix modes in a codebase, but we rarely
+detect these events:
+
+\pct{99} of sessions use a single mode
+
+\pct{0.2} of sessions use different modes in different modules. (More
+precisely, the user switched to a module --- triggering a telemetry record ---
+that used a different mode than the previous one.)
+
+\pct{0.2} of records show a mode switch (equally split between upgrades and
+downgrades, 341 and 320)
+
+Second,
+the stronger analysis modes are roughly 10x less popular than weaker ones
+(across the \pct{99} of sessions that stick with one mode):
+
+\pct{90} of sessions use nocheck mode
+\pct{9.8} use nonstrict mode
+\pct{0.2} use strict mode
+
+Third,
+In the edit range, we know exactly which errors showed up. This leads to a few
+takeaways, such as:
+
+Only 3 sessions (out of +300K) hit the typechecker's internal limits on problem
+size --- CodeTooComplex errors. Good!
+
+About \pct{90} of type errors are probably due to typos (SyntaxError, UnknownSymbol,
+UnknownProperty). These are not propoer type errors.
+
+10 errors never showed up. These include DynamicPropertyLookupOnClassesUnsafe and
+SwappedGenericTypeParameter
+
+Fourth,
+Codebase size and session length tend to follow a power law. 
+One aspect of size for nonstrict sessions: the number of
+lines in the edit range. Over 80 sessions have 20-line ranges, about 40 have a
+30-line range, etc.
+
+TODO must explain edit range carefully.
+
+TODO must show full plot, zoomed plot side by side
+
+Fifth,
+Increases and decreases to the number of type errors are evenly balanced across
+the data.
+In strict mode, \pct{24} of records increase the number of type errors. \pct{26}
+decrease. \pct{50} keep it constant.
+In nonstrict mode, \pct{14} increase errors, \pct{15} decrease, and \pct{71}
+keep it constant
+This plot is for nonstrict. X-axis = time since start of session, cut off at 30
+seconds. Y-axis = change in num errors since last record, cut off at +/- 100.
+
+\begin{figure}[t]
+  \includegraphics[width=0.8\columnwidth]{img/example-error-delta.png}
+  \caption{Errors delta, nonstrict}
+  \label{f:error-delta}
+\end{figure}
+
+Sixth, a mystery. May be a bug in Luau.
+May be DM insensitivity in strict mode
+(if DM has top type that would lead to errors.)
+Context: each of the 3 modes runs 2 kinds of type analysis. One is the regular
+type analysis. The other is Forced Strict (FS), which converts all dependencies
+to strict mode. The reason for FS is that it sometimes gives better
+autocomplete suggestions etc.
+You'd expect that the number of type errors is always <= num FS errors. But we
+found several cases where this is false!
+The plot here "stacks" every strict session into bars. X-axis is a record index
+within a session (bars are higher on the left because we have lots more short
+sessions than long ones). Y-axis counts the number of records where num TE <= num FS
+is true (blue) and false (yellow).
+
+TODO clarify semantic x-axis order
+
+%% FS has: 1 all dependencies in strict, 2 data-model awareness (DM turns into
+%% type graph, ... initial DM, all properties go into a table type; DM -> folder
+%% -> model -> part .. get dot-driven autocomplete) in strict, DM might have type
+%% Any might have type Instance, the latter would mean more errors than FS
+
+\begin{figure}[t]
+  \includegraphics[width=0.8\columnwidth]{img/example-tefs.png}
+  \caption{Are there fewer type errors than forcedstrict errors? Expect true always, but that is not the case.}
+  \label{f:tefs}
+\end{figure}
+
 \section{Interpretation}

### 2023-08-25 programming23/paper.tex
quick note, data model
note, big edit range
note about cutoff, not sure if right idea

@@
 
-FILL ignore sessions close to start / end time?
-remove all syntax errors from later stages?
+Edit range ignore huge numbers (above 4B),
+which are the result of subtraction overflowing.
+%% 4,000,000,000
+
+Sessions, ignore if too close to start or end.
+Deciding if a session is close is tricky because we don't have
+a firm start and end time.
+Um, does this make sense at all?
+Well anyway we took the greater number from mean vs median
+gap (over all points, not just keystroke points)
+and multiplied by Nx to test closeness to a boundary.
+With N=2 we drop 1 nonstrict session, 0 strict (not sure about nocheck).
+With N=4 we drop 3, with N=8 we drop 9.
+
+FILL remove all syntax errors from later stages?
 
@@
 
+Seventh,
+
+\begin{figure}[t]
+  \includegraphics[width=0.8\columnwidth]{img/example-compass.png}
+  \caption{Are there fewer type errors than forcedstrict errors? Expect true always, but that is not the case.}
+  \label{f:tefs}
+\end{figure}
+
 \section{Interpretation}
@@
 \end{figure}
+
+\Cref{f:scribbles} kill this.
+
+\subsection{Asset Creation vs Scripting}
+
+Data model. Point to discussion above.
+
+Big fluctuations in errors.
+FS disagrees with TE.
+Probably data model!
 

### 2023-08-30 programming23/paper.tex
related work, diff private trace analysis
row distro data, quick plot
timeline plot

@@
 \begin{figure}[t]
-  \includegraphics{img/row-distribution.pdf}
-  \caption{Telemetry records per hour. Each tick on the $x$-axis marks the start of a new day in California.}
+  \includegraphics[width=\columnwidth]{img/row-distribution.pdf}
+  \caption{Telemetry records per hour. Each tick on the $x$-axis marks the start of a new day in California. Shaded regions correspond to weekends.}
   \label{f:records-per-hour}
@@
 
+Differential privacy for coverage analysis of software traces,
+adds noise to traces,
+hard because infinitely many traces,
+use CFG abstraction of possible behaviors for callgraph analysis,
+count sketch data structure~\cite{hlzbr-ecoop-2021}.
+
+other Nasko papers:
+https://dblp.org/pid/83/1412.html
+
 %% - zfstt-ieeesensors-2022

### 2023-08-31 programming23/paper.tex
checkpoint, data overview

@@
 \newcommand{\mstrict}{\modefont{strict}}
+\newcommand{\zerowidth}[1]{\makebox[0pt][l]{#1}}
 
@@
 \begin{figure}[t]
+  %% TODO bigger text
+  %% code/row-distribution.rkt
   \includegraphics[width=\columnwidth]{img/row-distribution.pdf}
@@
 
+"89.1417"
+"10.4259"
+"0.4323"
+
 \begin{figure}[t]
-  \begin{tabular}{rl}
-   5,001 & total logs \\
-         & 4,471 \mnocheck{} + 506 \mnonstrict{} + 24 \mstrict{} \\
-         & 1,628 from module switches \\
-   182,247 & total forced strict type errors \\
-         & {83,005 in module, 5,113 in edit regions} \\
-    1,521 & total type errors \\
-         & {1,035 in module, 110 non-syntax + 47 syntax in edit regions}
-    \\[2ex]
-
-    1,146 & sessions \\
-    & \begin{tabular}{l}
-      1,143 single mode = 1046 \mnocheck{} + 94 \mnonstrict{} + 3 \mstrict{} \\
-        detected 2 multi-mode projects, 2 mode upgrades, 2 mode downgrades
+
+  %% TODO subfigure??
+  \begin{tabular}{ll}
+    \begin{tabular}{r@{~~}l@{~~}r}
+       1,341,348 & \mnocheck{}          & [\pct{89.14}] \\
+         156,883 & \mnonstrict{}        & [\pct{10.43}] \\
+           6,505 & \mstrict{}           & [\pct{ 0.43}] \\\midrule
+       1,504,736 & \zerowidth{Total telemetry records}
+    \end{tabular}
+    \begin{tabular}{r@{~~}l@{~~}r}
+         508,572 & due to module switch & [\pct{33.80}] \\
+         996,164 & due to keystrokes    & [\pct{66.20}]
     \end{tabular}
@@
 
-  %% TODO these need to be plots, it's too much reading for the quick takeaway
-  %% put exactly this data into bars!
-  \begin{minipage}[t]{0.5\textwidth}
-    \begin{tabular}{rr}
-      Session length & \# Sessions \\\midrule
-      1 &  450 \\
-      2 &  199 \\
-      3 &  129 \\
-      4 &  73 \\
-      5 &  60 \\
-      6-10 & 150 \\
-      11-20 & 67 \\
-      21-40 & 11 \\
-      >40 & 7
-      %% max 393
-    \end{tabular}
-  \end{minipage}\begin{minipage}[t]{0.5\textwidth}
-    \begin{tabular}{rr}
-      Time delta & \# Records \\\midrule
-      0-5ms & 284 \\
-      5-1000ms & 248 \\
-      1-2s & 153 \\
-      2-5s & 232 \\
-      5-10s & 207 \\
-      10-30s & 524 \\
-      30-60s & 392 \\
-      1-2m & 394 \\
-      2-5m & 559 \\
-      5-10m & 360 \\
-      >10-60m & 502
-    \end{tabular}
-  \end{minipage}
+ 72,235,735 total forced strict type errors, 37,027,281 in module, 1,111,178 in edit regions
+
+ 595,137 total type errors, 289,698 in module, 14,917 non-stx + 16,007 syntax in edit regions
+
+ 347,598 sessions
+  346,956 single mode = 313,509 nocheck + 32,902 nonstrict + 545 strict
+  512 multi mode projects
+  341 mode upgrades
+  320 mode downgrades
+
+%  \begin{tabular}{rl}
+%   182,247 & total forced-strict type errors \\
+%         & {83,005 in module, 5,113 in edit regions} \\
+%    1,521 & total type errors \\
+%         & {1,035 in module, 110 non-syntax + 47 syntax in edit regions}
+%    \\[2ex]
+%
+%    1,146 & sessions \\
+%    & \begin{tabular}{l}
+%      1,143 single mode = 1046 \mnocheck{} + 94 \mnonstrict{} + 3 \mstrict{} \\
+%        detected 2 multi-mode projects, 2 mode upgrades, 2 mode downgrades
+%      \end{tabular}
+%  \end{tabular}
+
+%  %% TODO these need to be plots, it's too much reading for the quick takeaway
+%  %% put exactly this data into bars!
+%  \begin{minipage}[t]{0.5\textwidth}
+%    \begin{tabular}{rr}
+%      Session length & \# Sessions \\\midrule
+%      1 &  450 \\
+%      2 &  199 \\
+%      3 &  129 \\
+%      4 &  73 \\
+%      5 &  60 \\
+%      6-10 & 150 \\
+%      11-20 & 67 \\
+%      21-40 & 11 \\
+%      >40 & 7
+%      %% max 393
+%    \end{tabular}
+%  \end{minipage}\begin{minipage}[t]{0.5\textwidth}
+%    \begin{tabular}{rr}
+%      Time delta & \# Records \\\midrule
+%      0-5ms & 284 \\
+%      5-1000ms & 248 \\
+%      1-2s & 153 \\
+%      2-5s & 232 \\
+%      5-10s & 207 \\
+%      10-30s & 524 \\
+%      30-60s & 392 \\
+%      1-2m & 394 \\
+%      2-5m & 559 \\
+%      5-10m & 360 \\
+%      >10-60m & 502
+%    \end{tabular}
+%  \end{minipage}
 
@@
 \subsection{Dead Ends: Too Complex}
-
-The typechecker has two internal limits regarding the problems it will attempt
-to solving.
+%% data = out/ctc-info.txt
+
+To deal with pathologies such as the worst-case time for ML-style type
+inference~\cite{m-popl-1990,ktu-caap-1990}, the typechecker
+has two internal limits to restrict the problems it will attempt to solve.
 Hitting a limit triggers either a \code{CodeTooComplex} or
@@
 
-Based on the data, these errors are indeed extremely rare.
-Out of 1.5M telemetry records, only 11 report any \code{TooComplex} errors.
-These records came from three sessions (possibly by a single user), and each
-record counted at most 4 \code{CodeTooComplex} errors and zero
-\code{UnificationTooComplex} errors.
-There is no evidence to suggest that the limits are too strict or that the
-typechecker should improve.
+Fortunately, the data rarely contains too-complex errors.
+There are zero \code{UnificationTooComplex} errors and only a
+handful of \code{CodeTooComplex} errors.
+%% 26 total
+It is entirely possible that the \code{CodeTooComplex} errors came from
+a single user, as they appear in eleven records (out of 1.5 million)
+spread across three sessions.
+At any rate, the overall number is low enough to suggest that the internal
+limits are not posing barriers to typechecker adoption.
 

### 2023-09-01 programming23/paper.tex
dataset overview fig

@@
 \usepackage{tikz}
+\usepackage{subcaption}
 \usepackage{wasysym}
-\usepackage{wrapfig}
 \usepackage{xcolor}
@@
 
-"89.1417"
-"10.4259"
-"0.4323"
+FILL te in edit region, 
+ 14,917 non-stx
+ 16,007 syntax in edit regions.
+ Most are typos!
 
 \begin{figure}[t]
-
-  %% TODO subfigure??
-  \begin{tabular}{ll}
-    \begin{tabular}{r@{~~}l@{~~}r}
+  \begin{subfigure}[t]{\columnwidth}
+    \begin{tabular}[t]{ll}
+      \begin{tabular}[t]{r@{~~}l@{~~}r}
          1,341,348 & \mnocheck{}          & [\pct{89.14}] \\
            156,883 & \mnonstrict{}        & [\pct{10.43}] \\
-           6,505 & \mstrict{}           & [\pct{ 0.43}] \\\midrule
-       1,504,736 & \zerowidth{Total telemetry records}
-    \end{tabular}
-    \begin{tabular}{r@{~~}l@{~~}r}
+             6,505 & \mstrict{}           & [\pct{ 0.43}]
+      \end{tabular}
+      \begin{tabular}[t]{r@{~~}l@{~~}r}
            508,572 & due to module switch & [\pct{33.80}] \\
@@
     \end{tabular}
-
- 72,235,735 total forced strict type errors, 37,027,281 in module, 1,111,178 in edit regions
-
- 595,137 total type errors, 289,698 in module, 14,917 non-stx + 16,007 syntax in edit regions
-
- 347,598 sessions
-  346,956 single mode = 313,509 nocheck + 32,902 nonstrict + 545 strict
-  512 multi mode projects
-  341 mode upgrades
-  320 mode downgrades
-
-%  \begin{tabular}{rl}
-%   182,247 & total forced-strict type errors \\
-%         & {83,005 in module, 5,113 in edit regions} \\
-%    1,521 & total type errors \\
-%         & {1,035 in module, 110 non-syntax + 47 syntax in edit regions}
-%    \\[2ex]
-%
-%    1,146 & sessions \\
-%    & \begin{tabular}{l}
-%      1,143 single mode = 1046 \mnocheck{} + 94 \mnonstrict{} + 3 \mstrict{} \\
-%        detected 2 multi-mode projects, 2 mode upgrades, 2 mode downgrades
-%      \end{tabular}
-%  \end{tabular}
+    \caption{Total telemetry records (1,504,736), analysis modes, and reason-for-sending}
+    \label{f:total-records}
+  \end{subfigure}
+
+  \begin{subfigure}[t]{\columnwidth}
+    \begin{tabular}[t]{ll}
+      \begin{tabular}[t]{lrl}
+        595,137 & Type errors \\
+        289,698 & in current module & [\pct{48.68}] \\
+         30,924 & in edit region & [\pct{5.20}]
+      \end{tabular}
+      \begin{tabular}[t]{lrl}
+        72,235,735 & {Forced-strict errors} \\
+        37,027,281 & in current module & [\pct{51.26}] \\
+         1,111,178 & in edit region & [\pct{1.54}]
+      \end{tabular}
+    \end{tabular}
+    \caption{Type errors and forced-strict errors across all telemetry records}
+    \label{f:total-tefs}
+  \end{subfigure}
+
+  \begin{subfigure}[t]{\columnwidth}
+    \begin{tabular}[t]{ll}
+      \begin{tabular}[t]{lrl}
+        313,509 & \mnocheck{}   & [\pct{90.19}] \\
+         32,902 & \mnonstrict{} & [\pct{ 9.47}] \\
+            545 & \mstrict{}    & [\pct{ 0.16}] \\
+            642 & mixed-mode    & [\pct{ 0.18}]
+      \end{tabular}
+      \begin{tabular}[t]{l@{~~}ll}
+        \zerowidth{Among the mixed-mode sessions:} \\
+        & 341 & contain a mode upgrade \\
+        & 320 & contain a mode downgrade \\
+        & 512 & contain modules with different modes
+      \end{tabular}
+    \end{tabular}
+    \caption{Total sessions (347,598), whether/how they switch analysis mode}
+    \label{f:total-sessions}
+  \end{subfigure}
+
+  \caption{Dataset overview}
+  \label{f:dataset-overview}
+\end{figure}
 
@@
 
-  \caption{Dataset overview}
-  \label{f:dataset-overview}
-\end{figure}
-
 \begin{figure}[t]

### 2023-09-01 programming23/paper.tex
region -> range, next step notes

@@
 number of errors at various levels of granularity: in the current edit
-region, in the current file, and in every file which was type
+range, in the current file, and in every file which was type
 checked.
@@
   \item[D4] size of codebase: num files, num lines
-  \item[D5] size of edit region
-  \item[D6] overall type errors: for old and curr states, have: \# total, \# in module, and \# in region
-  \item[D7] specific errors in the edit region: in the edit region, for old and
+  \item[D5] size of edit range
+  \item[D6] overall type errors: for old and curr states, have: \# total, \# in module, and \# in range
+  \item[D7] specific errors in the edit range: in the edit range, for old and
     curr states, how many type errors for each type error code? (There is 1
-    edit region per log.)
+    edit range per log.)
     \subitem Examples, if the old state had one type-A error:
@@
   \item[D8] ForcedStrict "what if" errors for \mnocheck{}: specific errors for the
-    edit region computed as if the user had strict mode enabled (even though they
+    edit range computed as if the user had strict mode enabled (even though they
     have no-check enabled)
@@
         %% \subitem Count CTC errors within each session to see if these errors are intermittent, or if they stick around.
-[D9 for the count, D8 for \mnocheck{} vs strict, D7 because CTC might show up in the edit region]
+[D9 for the count, D8 for \mnocheck{} vs strict, D7 because CTC might show up in the edit range]
     \end{itemize}
-  \item How helpful are highlighted regions?
+  \item How helpful are highlighted ranges?
     \begin{itemize}
-      \item How often do edits in a highlighted region remove type errors? How
+      \item How often do edits in a highlighted range remove type errors? How
         often do they add new ones?
@@
       \subitem [may want to filter using D2 + D5]
-      \item How often do edits outside the region remove type errors? How often do they add new ones?
+      \item How often do edits outside the range remove type errors? How often do they add new ones?
       \subitem [D6 looking for old count vs. new count difference without a matching delta in the D7 counts]
@@
 % Sanity checks
-% How large are edit regions? Plot the distribution.
+% How large are edit ranges? Plot the distribution.
 % How long between logs? Long times & very short times are problematic.
@@
 % General type error trends
-% Plot the distribution of #errors across all edit regions
+% Plot the distribution of #errors across all edit ranges
 % Expect different counts based on error-kind. SEE BELOW
-% Count specific errors that get fixed. For each kind of error, how many times is it old-in-region but not new-in-region? Compare to #old-in-region totals.
+% Count specific errors that get fixed. For each kind of error, how many times is it old-in-range but not new-in-range? Compare to #old-in-range totals.
 % Expect all errors to have a decent fix rate.
-% If one error has a lower % than others, its error region may be misleading. Or, maybe it's hard to fix
-% Type errors vs. edit region
-% Compare #old vs. #new in region
+% If one error has a lower % than others, its error range may be misleading. Or, maybe it's hard to fix
+% Type errors vs. edit range
+% Compare #old vs. #new in range
 % Expect a decrease, #old > #new
-% If there's an increase, look for an explanation. Could be a huge edit region. Could be a 1-line region that we captured 2x in a row. Could be that this user is simply ignoring type errors.
-% Compare [#old-in-region vs. (total #old - total #new)]
-% If #old-in-region = 0, then hope the total # increases
+% If there's an increase, look for an explanation. Could be a huge edit range. Could be a 1-line range that we captured 2x in a row. Could be that this user is simply ignoring type errors.
+% Compare [#old-in-range vs. (total #old - total #new)]
+% If #old-in-range = 0, then hope the total # increases
 % Decrease => nonlocal edits fixed an error => type error hilite may be inaccurate
-% If #old-in-region > 0, then hope the total # decreases
-% Increase (in region or otherwise) => made things worse
+% If #old-in-range > 0, then hope the total # decreases
+% Increase (in range or otherwise) => made things worse
 
@@
   \includegraphics[width=\columnwidth]{img/row-distribution.pdf}
-  \caption{Telemetry records per hour. Each tick on the $x$-axis marks the start of a new day in California. Shaded regions correspond to weekends.}
+  \caption{Telemetry records per hour. Each tick on the $x$-axis marks the start of a new day in California. Shaded ranges correspond to weekends.}
   \label{f:records-per-hour}
@@
   TODO file size distribution: table or bars %% \includegraphics{img/row-distribution.pdf}
-  \caption{Size of analyzed code: number of files, number of lines , and lines in edit region}
+  \caption{Size of analyzed code: number of files, number of lines , and lines in edit range}
   \label{f:analysis-size}
@@
 
-FILL te in edit region, 
+FILL te in edit range, 
  14,917 non-stx
- 16,007 syntax in edit regions.
+ 16,007 syntax in edit ranges.
  Most are typos!
@@
         289,698 & in current module & [\pct{48.68}] \\
-         30,924 & in edit region & [\pct{5.20}]
+         30,924 & in edit range & [\pct{5.20}]
       \end{tabular}
@@
         37,027,281 & in current module & [\pct{51.26}] \\
-         1,111,178 & in edit region & [\pct{1.54}]
+         1,111,178 & in edit range & [\pct{1.54}]
       \end{tabular}
@@
 \end{figure}
+
+%% TODO explain timestamps in prose, fig in appendix?
+
+%% TODO power-law: event count, timespan, files lines editRange /session, 
+%% ... gotta know project size
+
+%% TODO overall stats on type errors
 
@@
   \begin{tabular}{lrrrrr}
-    Mode             & \# Rows & E         & Any Error & In Module & In Region \\
+    Mode             & \# Rows & E         & Any Error & In Module & In range \\
       & & & & (old ; curr) &  (old ; curr) \\\midrule
@@
 
-  Below, edit region only.
-  Each record with an edit region error adds +1 to at least 1 row.
+  Below, edit range only.
+  Each record with an edit range error adds +1 to at least 1 row.
 
@@
 
-FILL error density: number vs region size
+FILL error density: number vs range size
 
@@
 but common, which makes it likely that our per-keystroke telemetry (1/2k) will record them.
-Edit region has specific counts.
+Edit range has specific counts.
 Next time, we should do that for the other aggregates.

### 2023-09-04 programming23/paper.tex
server timestamp less useful than client timestamp

@@
 
-Deduplicated records with the same client timestamp,
+Deduplicated records with the same timestamp,
 kept only the first.
@@
 
-Edit range ignore huge numbers (above 4B),
+Records come with two timestamps, a \emph{client timestamp}
+that the IDE generates before sending the record
+and a \emph{server timestamp} that the telemetry server
+generates when it receives the record.
+We collected timestamps to sort records from earliest to latest.
+The server timestamps, however, are less useful for sorting
+because the server processes groups of records in a batch and
+often assigns equal timestamps to dozens of records.
+Hence, we ignore the server timestamps.
+Throughout the paper a \emph{timestamp} is always a client-side timestamp.
+
+Edit range ignore huge numbers (above 4 billion),
 which are the result of subtraction overflowing.
@@
 
-%% TODO explain timestamps in prose, fig in appendix?
-
 %% TODO power-law: event count, timespan, files lines editRange /session, 
@@
 
-%% TODO overall stats on type errors
+%% results 1 section?!
+
+%% TODO counting stats on type errors
+%% - specific errors
+%% - typos vs rest
+%% - too complex
+%% - mod switch vs other
 

### 2023-09-04 programming23/paper.tex
negative edit range note

@@
 
-Edit range ignore huge numbers (above 4 billion),
-which are the result of subtraction overflowing.
-%% 4,000,000,000
+Edit ranges are negative in a handful of records (1,533).
+%% out/negative-edit-range.txt
+%% actual value is NOT negative, but large positive over 4 billion
+%% example: 4294967293
+This may happen when a user deletes a large block of code.
+Since the issue affects so few records, we omit those ranges
+from the analysis.
 
@@
 
+\appendix
+
+\section{Data Details}
+
+Tips for interpreting the data:
+
+\begin{itemize}
+  \item
+    Edit ranges are non-negative because we
+    implemented the interval arithmetic for them using unsigned integers.
+    To filter out negative edit ranges, we removed all ranges
+    greater than 4 billion.
+
+  \item
+  \item
+  \item
+\end{itemize}
+
+\newpage
+
 { \printbibliography }

### 2023-09-07 programming23/paper.tex
size figs

@@
 
+FILL investigate one tower
+
 \begin{figure}[t]
-  TODO file size distribution: table or bars %% \includegraphics{img/row-distribution.pdf}
+  %% \includegraphics{img/row-distribution.pdf}
+  \begin{tabular}{lrrr}
+           & Files  &     Lines &    Edit Range \\
+    Median &  7,678 &     3,115 &           926 \\
+    Mean   & 11,911 &     6,497 &         3,680 \\
+    Max    & 54,884 & 1,089,963 &     1,156,036 \\
+    & \includegraphics[width=0.2\columnwidth]{img/files-distribution.pdf}
+    & \includegraphics[width=0.2\columnwidth]{img/lines-distribution.pdf}
+    & \includegraphics[width=0.2\columnwidth]{img/editrange-distribution.pdf}
+
+  \end{tabular}
+
   \caption{Size of analyzed code: number of files, number of lines , and lines in edit range}
-  \label{f:analysis-size}
+  \label{f:codebase-size}
 \end{figure}
+
+TODO .. somewhere ... out/error-density-*
+TODO ... somewhere ... te leq fs : use error-density
+
+summary of timespan :: max 1387897376 median 845648 mean 3184317.084881936 stddev 15724745.722577972
+summary of event-count :: max 6079 median 138 mean 286.5851851851852 stddev 582.5752201972966
 
@@
 
+  \begin{subfigure}[t]{\columnwidth}
+    \caption{TODO session size}
+    \label{f:sessions-size}
+  \end{subfigure}
+
+  %% TODO out/size-dist*rktd
+
   \caption{Dataset overview}
@@
 
-%  %% TODO these need to be plots, it's too much reading for the quick takeaway
-%  %% put exactly this data into bars!
-%  \begin{minipage}[t]{0.5\textwidth}
-%    \begin{tabular}{rr}
-%      Session length & \# Sessions \\\midrule
-%      1 &  450 \\
-%      2 &  199 \\
-%      3 &  129 \\
-%      4 &  73 \\
-%      5 &  60 \\
-%      6-10 & 150 \\
-%      11-20 & 67 \\
-%      21-40 & 11 \\
-%      >40 & 7
-%      %% max 393
-%    \end{tabular}
-%  \end{minipage}\begin{minipage}[t]{0.5\textwidth}
-%    \begin{tabular}{rr}
-%      Time delta & \# Records \\\midrule
-%      0-5ms & 284 \\
-%      5-1000ms & 248 \\
-%      1-2s & 153 \\
-%      2-5s & 232 \\
-%      5-10s & 207 \\
-%      10-30s & 524 \\
-%      30-60s & 392 \\
-%      1-2m & 394 \\
-%      2-5m & 559 \\
-%      5-10m & 360 \\
-%      >10-60m & 502
-%    \end{tabular}
-%  \end{minipage}
-
 \begin{figure}[t]
@@
   %% TODO align old;curr
+  %% TODO sep23 remove this??!
   \begin{tabular}{lrrrrr}
@@
 
+  %% TODO import code from tmp to nsa
   {Few local adds => Few people ignoring script analysis}

### 2023-09-07 programming23/paper.tex
session size

@@
 \newcommand{\zerowidth}[1]{\makebox[0pt][l]{#1}}
+\newcommand{\stddev}[1]{[#1\textsc{k}]}
 
@@
   %% \includegraphics{img/row-distribution.pdf}
-  \begin{tabular}{lrrr}
-           & Files  &     Lines &    Edit Range \\
-    Median &  7,678 &     3,115 &           926 \\
-    Mean   & 11,911 &     6,497 &         3,680 \\
-    Max    & 54,884 & 1,089,963 &     1,156,036 \\
+  \begin{tabular}{l@{}r@{~}l@{}r@{~}l@{}r@{~}l}
+    & Files  & &     Lines & &    Edit Range & \\
+    Mean [std]   & 11,911 & \stddev{31}  &     6,497 & \stddev{22} &         3,680 & \stddev{12} \\
+    Median       &  7,678 &              &     3,115 &             &          926  & \\
+    Max          & 54,884 &              & 1,089,963 &             &    1,156,036  & \\
     & \includegraphics[width=0.2\columnwidth]{img/files-distribution.pdf}
-    & \includegraphics[width=0.2\columnwidth]{img/lines-distribution.pdf}
-    & \includegraphics[width=0.2\columnwidth]{img/editrange-distribution.pdf}
+    & & \includegraphics[width=0.2\columnwidth]{img/lines-distribution.pdf}
+    & & \includegraphics[width=0.2\columnwidth]{img/editrange-distribution.pdf}
 
@@
 \end{figure}
-
-TODO .. somewhere ... out/error-density-*
-TODO ... somewhere ... te leq fs : use error-density
-
-summary of timespan :: max 1387897376 median 845648 mean 3184317.084881936 stddev 15724745.722577972
-summary of event-count :: max 6079 median 138 mean 286.5851851851852 stddev 582.5752201972966
 
@@
 
-FILL te in edit range, 
- 14,917 non-stx
- 16,007 syntax in edit ranges.
- Most are typos!
+FILL power-law files lines editrange.
+Incredible max numbers.
 
@@
     \begin{tabular}[t]{ll}
-      \begin{tabular}[t]{r@{~~}l@{~~}r}
+      \begin{tabular}[t]{r@{~~}l@{~}r}
          1,341,348 & \mnocheck{}          & [\pct{89.14}] \\
@@
   \begin{subfigure}[t]{\columnwidth}
-    \begin{tabular}[t]{ll}
-      \begin{tabular}[t]{lrl}
+    \begin{tabular}[t]{ll} \\
+      \begin{tabular}[t]{l@{~~}r@{~}l}
         595,137 & Type errors \\
@@
       \end{tabular}
-      \begin{tabular}[t]{lrl}
+      \begin{tabular}[t]{l@{~~}r@{~}l}
         72,235,735 & {Forced-strict errors} \\
@@
   \begin{subfigure}[t]{\columnwidth}
-    \begin{tabular}[t]{ll}
-      \begin{tabular}[t]{lrl}
+    \begin{tabular}[t]{ll} \\
+      \begin{tabular}[t]{l@{~~}r@{~}l}
         313,509 & \mnocheck{}   & [\pct{90.19}] \\
@@
   \begin{subfigure}[t]{\columnwidth}
-    \caption{TODO session size}
+    \begin{tabular}{l@{}r@{~}l@{}r@{~}l} \\
+                   & Time Span (sec)  &       & Record Count  \\
+      Mean [std]   &     3,184 & \stddev{16}  &     286 & [583] \\
+      Median       &       845 &              &     138        \\
+      Max          & 1,387,897 &              &   6,079        \\
+      & \includegraphics[width=0.2\columnwidth]{img/timespan-distribution.pdf}
+      & & \includegraphics[width=0.2\columnwidth]{img/event-count-distribution.pdf}
+
+    \end{tabular}
+
+    \caption{Session length in seconds and in number of records}
     \label{f:sessions-size}
   \end{subfigure}
-
-  %% TODO out/size-dist*rktd
 
@@
 
-%% TODO power-law: event count, timespan, files lines editRange /session, 
-%% ... gotta know project size
+%% TODO power-law: event count, timespan
+
+TODO .. somewhere ... out/error-density-*
+TODO ... somewhere ... te leq fs : use error-density
+
+FILL te in edit range, 
+ 14,917 non-stx
+ 16,007 syntax in edit ranges.
+ Most are typos!
 

### 2023-09-07 programming23/paper.tex
counting specific type errors

@@
 \usepackage{calc}
+\usepackage{colortbl}
 \usepackage{listings}
@@
 \newcommand{\stddev}[1]{[#1\textsc{k}]}
+\newcommand{\ycell}[1]{\cellcolor{yellow!18}#1}
+\newcommand{\ocell}[1]{\cellcolor{orange!29}#1}
+\newcommand{\rcell}[1]{\cellcolor{red!30}$\!\!$#1$\!\!$}
 
@@
     Building, sending, and storing the data must all run quickly.
-    FILL alan: what was our hard cap on Kibana use?
+    FILL what was our hard cap on Kibana use?
 \end{itemize}
@@
     \end{tabular}
-    \caption{Total sessions (347,598), whether/how they switch analysis mode}
+    \caption{Total sessions (347,598), whether and how they switch analysis mode}
     \label{f:total-sessions}
@@
 
-%% TODO power-law: event count, timespan
-
-TODO .. somewhere ... out/error-density-*
-TODO ... somewhere ... te leq fs : use error-density
-
-FILL te in edit range, 
- 14,917 non-stx
- 16,007 syntax in edit ranges.
- Most are typos!
-
-%% results 1 section?!
-
-%% TODO counting stats on type errors
-%% - specific errors
-%% - typos vs rest
-%% - too complex
-%% - mod switch vs other
+\Cref{f:dataset-overview}
+
+FILL nocheck nonstrict strict, order-of-magnitude
+
+FILL type error counts
+Mostly syntax errors though!
+Estimate \pct{40} are substantial type errors rather than syntax
+errors based on the edit range type errors~(\cref{s:type-error-count})
+but unfortunately we cannot get an exact number for the overall
+or per-module type error counts.
+
+FILL single-mode sessions, mostly
+
+FILL details on upgrades, downgrades, then ignore these for rest of paper.
+multi-mode project = found a module switch that coincides with a mode switch, which suggests that two modules have different modes;
+upgrade = found a up-mode switch that does not match a module switch (quite possible to miss these);
+downgrade = found a down-mode switch that does not match a module switch
+
+FILL most sessions fairly long in time (order of minutes),
+short in event count (dozens).
+
+\section{Type Errors in Edit Range}
+\label{s:type-error-count}
+
+\begin{table}[t]
+  %% Looks the same when restricted to modules
+  \begin{tabular}{l@{~~}l}
+    \begin{tabular}[t]{lr}
+      \code{SyntaxError} & 60.79\% \\
+      \code{UnknownSymbol} & 27.34\% \\
+      \code{UnknownProperty} & 4.08\% \\
+      \code{TypeMismatch} & 1.92\% \\
+      \code{UnknownRequire} & 1.40\% \\
+      \code{CountMismatch} & 1.08\% \\
+      \code{OptionalValueAccess} & 0.99\% \\
+      \code{GenericError} & 0.93\% \\
+      \code{CannotExtendTable} & 0.21\% \\
+      \code{UnknownPropButFoundLikeProp} & 0.20\% \\
+      \code{FunctionExitsWithoutReturning} & 0.16\% \\
+      \code{ExtraInformation} & 0.14\% \\
+      \code{CannotInferBinaryOperation} & 0.14\% \\
+      \code{MissingProperties} & 0.13\% \\
+      \code{ModuleHasCyclicDependency} & 0.10\% \\
+      \code{IllegalRequire} & 0.09\% \\
+      \code{NotATable} & 0.09\% \\
+      \code{MissingUnionProperty} & 0.08\%
+    \end{tabular} &
+    \begin{tabular}[t]{lr}
+      \code{CannotCallNonFunction} & 0.07\% \\
+      \code{FunctionDoesNotTakeSelf} & 0.03\% \\
+      \code{IncorrectGenericParameterCount} & 0.01\% \\
+      \code{OnlyTablesCanHaveMethods} & 0.00\% \\
+      \code{TypesAreUnrelated} & 0.00\% \\
+      \code{DuplicateTypeDefinition} & 0.00\% \\
+      \code{OccursCheckFailed} & 0.00\% \\
+      \code{TypePackMismatch} & 0.00\% \\
+      \code{FunctionRequiresSelf} & 0.00\% \\
+      \code{DuplicateGenericParameter} & 0.00\% \\
+      \code{InternalError} & 0.00\% \\
+      \code{CodeTooComplex} & 0.00\% \\
+      \code{DeprecatedApiUsed} & 0.00\% \\
+      \code{NormalizationTooComplex} & 0.00\% \\
+      \code{SwappedGenericTypeParameter} & 0.00\% \\
+      \code{DynamicPropertyLookupOnClassesUnsafe} & 0.00\% \\
+      \code{UnificationTooComplex} & 0.00\% \\
+    \end{tabular}
+  \end{tabular}
+  \caption{Type error popularity}
+  \label{t:type-error-count}
+\end{table}
+
+FILL popularity over all old and new in edit range.
+Syntax error by far the most common.
+UnknownSymbol is likely a syntax error --- so \pct{90} syntax error!
+
+Fourteen errors never showed up.
+Nothing for \code{CodeTooComplex} or \code{UnificationTooComplex} is great news.
+But what about the others? FILL
+
+=== BEN START HERE
+
+TODO get counts for each mode. Are there fewer stx errors in strict mode??!
+
+TODO which type errors at end of session?
 
 \begin{figure}[t]
-  %% TODO stacked bars?
-  %% TODO align old;curr
-  %% TODO sep23 remove this??!
-  \begin{tabular}{lrrrrr}
-    Mode             & \# Rows & E         & Any Error & In Module & In range \\
-      & & & & (old ; curr) &  (old ; curr) \\\midrule
-    \mnocheck{}   &   4,471 & \code{te} &    5.77\% &  0.20\% ;  2.21\% &  0.16\% ;  0.63\% \\
-                  &         & \code{fs} &   59.18\% & 32.81\% ; 49.16\% &  9.17\% ; 13.13\% \\
-    \mnonstrict{} &     506 & \code{te} &   25.89\% & 12.06\% ; 18.18\% &  6.32\% ;  7.91\% \\
-                  &         & \code{fs} &   62.45\% & 37.94\% ; 53.75\% & 16.60\% ; 20.75\% \\
-    \mstrict{}    &      24 & \code{te} &   50.00\% & 29.17\% ; 33.33\% &     0\% ;     0\% \\
-                  &         & \code{fs} &   62.50\% & 50.00\% ; 58.33\% &  4.17\% ;  8.33\% 
-  \end{tabular}
-
-  \bigskip
-
   Below, edit range only.
@@
 
-\Cref{f:dataset-overview}:
-cannot filter syntax errors from the total and per-module error counts (\cref{s:threats});
-multi-mode project = found a module switch that coincides with a mode switch, which suggests that two modules have different modes;
-upgrade = found a up-mode switch that does not match a module switch (quite possible to miss these);
-downgrade = found a down-mode switch that does not match a module switch
+TODO .. somewhere ... out/error-density-*
+TODO ... somewhere ... te leq fs : use error-density
 
@@
   \label{f:mode-errors}
+\end{figure}
+
+\section{Analysis}
+
+\begin{figure}[thb]
+  %% TODO stacked bars?
+  %% TODO align old;curr
+  %% TODO sep'23 remove this??!
+  \begin{tabular}{lrrrrr}
+    Mode             & \# Rows & E         & Any Error & In Module & In range \\
+      & & & & (old ; curr) &  (old ; curr) \\\midrule
+    \mnocheck{}   &   4,471 & \code{te} &    5.77\% &  0.20\% ;  2.21\% &  0.16\% ;  0.63\% \\
+                  &         & \code{fs} &   59.18\% & 32.81\% ; 49.16\% &  9.17\% ; 13.13\% \\
+    \mnonstrict{} &     506 & \code{te} &   25.89\% & 12.06\% ; 18.18\% &  6.32\% ;  7.91\% \\
+                  &         & \code{fs} &   62.45\% & 37.94\% ; 53.75\% & 16.60\% ; 20.75\% \\
+    \mstrict{}    &      24 & \code{te} &   50.00\% & 29.17\% ; 33.33\% &     0\% ;     0\% \\
+                  &         & \code{fs} &   62.50\% & 50.00\% ; 58.33\% &  4.17\% ;  8.33\% 
+  \end{tabular}
+
+  \caption{TODO}
+
 \end{figure}

### 2023-09-08 programming23/paper.tex
type error survival table: add keep drop

@@
 \newcommand{\stddev}[1]{[#1\textsc{k}]}
+\newcommand{\gcell}[1]{\cellcolor{green!20}#1}
 \newcommand{\ycell}[1]{\cellcolor{yellow!18}#1}
@@
 
-=== BEN START HERE
-
 TODO get counts for each mode. Are there fewer stx errors in strict mode??!
 
-TODO which type errors at end of session?
-
-\begin{figure}[t]
-  Below, edit range only.
-  Each record with an edit range error adds +1 to at least 1 row.
-
-  \smallskip
-
-  %% TODO draw bars or curves b/c numbers are too hard to compare horizontally
-  \begin{tabular}{lrrr}
-    Error Label & \# Add & \# Keep & \# Remove \\\midrule
-    CodeTooComplex
-    & 0 & 0 & 0 \\
-    UnificationTooComplex
-    & 0 & 0 & 0 \\[1ex]
-    SyntaxError
-    & 30 & 0 & 76 \\
-    UnknownSymbol
-    & 12 & 14 & 19 \\
-    UnknownProperty
-    & 3 & 1 & 4 \\
-    UnknownRequire
-    & 1 & 5 & 0 \\
-    GenericError
-    & 1 & 0 & 1 \\
-    CountMismatch
-    &  1 &  0 &  1 \\
-    OptionalValueAccess
-    &  0 &  0 &  1 \\
-  \end{tabular}
-
-  \caption{Errors overview}
-  \label{f:errors-overview}
-\end{figure}
+NOTE can't ask about type errors at end because we don't know
+the end.
+The final record can be sent minutes before the end of session.
+
+\begin{table}[t]
+
+    \begin{tabular}{lr@{~}r@{~}rr@{~}r@{~}rr@{~}r@{~}r}
+      & \zerowidth{\mnocheck{}} & & & \zerowidth{\mnonstrict{}} & & & \zerowidth{\mstrict{}} & & \\
+      & \rcell{Add} & \ycell{Keep} & \gcell{Drop} & \rcell{Add} & \ycell{Keep} & \gcell{Drop} & \rcell{Add} & \ycell{Keep} & \gcell{Drop} \\\midrule
+      TypeMismatch & {0} & {0} & {0} & {159} & {62} & {129} & {21} & {10} & {28} \\
+      UnknownSymbol & {0} & {0} & {0} & {2348} & {496} & {2973} & {49} & {19} & {55} \\
+      UnknownProperty & {0} & {0} & {0} & {347} & {180} & {385} & {19} & {19} & {35} \\
+      NotATable & {0} & {0} & {0} & {7} & {2} & {8} & {2} & {0} & {1} \\
+      CannotExtendTable & {0} & {0} & {0} & {7} & {9} & {10} & {0} & {0} & {1} \\
+      OnlyTablesCanHaveMethods & {0} & {0} & {0} & {1} & {0} & {2} & {0} & {0} & {0} \\
+      DuplicateTypeDefinition & {0} & {0} & {0} & {1} & {0} & {0} & {0} & {0} & {0} \\
+      CountMismatch & {0} & {0} & {0} & {223} & {48} & {269} & {12} & {2} & {17} \\
+      FunctionDoesNotTakeSelf & {0} & {0} & {0} & {4} & {6} & {5} & {0} & {0} & {0} \\
+      OccursCheckFailed & {0} & {0} & {0} & {0} & {0} & {0} & {0} & {0} & {1} \\
+      UnknownRequire & {0} & {0} & {0} & {91} & {39} & {85} & {12} & {3} & {4} \\
+      IncorrectGenericParameterCount & {0} & {0} & {0} & {0} & {0} & {0} & {1} & {0} & {1} \\
+      SyntaxError & {8349} & {314} & {19220} & {983} & {46} & {2658} & {16} & {0} & {71} \\
+      UnknownPropButFoundLikeProp & {0} & {0} & {0} & {26} & {18} & {21} & {1} & {0} & {1} \\
+      GenericError & {0} & {0} & {0} & {206} & {52} & {203} & {8} & {1} & {9} \\
+      CannotCallNonFunction & {0} & {0} & {0} & {17} & {4} & {20} & {1} & {0} & {1} \\
+      ExtraInformation & {0} & {0} & {0} & {48} & {7} & {39} & {2} & {0} & {1} \\
+      ModuleHasCyclicDependency & {0} & {0} & {0} & {20} & {9} & {15} & {0} & {0} & {1} \\
+      IllegalRequire & {0} & {0} & {0} & {16} & {3} & {33} & {0} & {0} & {2} \\
+      FunctionExitsWithoutReturning & {0} & {0} & {0} & {9} & {3} & {6} & {11} & {5} & {8} \\
+      CannotInferBinaryOperation & {0} & {0} & {0} & {1} & {2} & {1} & {7} & {6} & {9} \\
+      MissingProperties & {0} & {0} & {0} & {11} & {3} & {11} & {6} & {4} & {6} \\
+      OptionalValueAccess & {0} & {0} & {0} & {40} & {52} & {27} & {6} & {2} & {7} \\
+      MissingUnionProperty & {0} & {0} & {0} & {1} & {0} & {2} & {0} & {0} & {0} \\
+      TypesAreUnrelated & {0} & {0} & {0} & {1} & {0} & {0} & {0} & {0} & {1} \\
+    \end{tabular}
+
+  \caption{Type error survival}
+  \label{t:type-error-survival}
+\end{table}
 
@@
 TODO ... somewhere ... te leq fs : use error-density
-
-\begin{figure}[t]
-
-  %% TODO import code from tmp to nsa
-  {Few local adds => Few people ignoring script analysis}
-  \smallskip
-
-  \begin{tabular}{lrrrr}
-    & Local Remove & Local Add & Nonlocal Remove & Nonlocal Add \\
-    Mode
-    & \#err (\#log)
-    & \#err (\#log)
-    & \#err (\#log)
-    & \#err (\#log) \\\midrule
-    \mnocheck{}   & 86 (71) & 0 (0) & 110 (57) & 161 (74) \\
-    \mnonstrict{} & 87 (25) & 9 (3) & 185 (29) & 207 (37) \\
-    \mstrict{}    &  1  (1) & 0 (0) &  14  (6) &   6  (3) \\
-  \end{tabular}
-
-  \caption{Per-mode comparison}
-  \label{f:mode-errors}
-\end{figure}
 

### 2023-09-09 programming23/paper.tex
error density plots

@@
 
-%% Extra Roblox constraints:
-%% - No access to some events: save, exit
-%% - No idea if analysis widget is open, closed, or obscured
-%% - Don't know which module is the data model
-%%   (no ids for ANY module for that matter)
+TODO must explain edit range carefully.
+
+Extra Roblox constraints:
+- No access to some events: save, exit
+- No idea if analysis widget is open, closed, or obscured
+- Don't know which module is the data model
+  (no ids for ANY module for that matter)
 
@@
     \end{tabular}
+
+    15 days
 
@@
 
-TODO .. somewhere ... out/error-density-*
-TODO ... somewhere ... te leq fs : use error-density
-
 \section{Analysis}
-
-\begin{figure}[thb]
-  %% TODO stacked bars?
-  %% TODO align old;curr
-  %% TODO sep'23 remove this??!
-  \begin{tabular}{lrrrrr}
-    Mode             & \# Rows & E         & Any Error & In Module & In range \\
-      & & & & (old ; curr) &  (old ; curr) \\\midrule
-    \mnocheck{}   &   4,471 & \code{te} &    5.77\% &  0.20\% ;  2.21\% &  0.16\% ;  0.63\% \\
-                  &         & \code{fs} &   59.18\% & 32.81\% ; 49.16\% &  9.17\% ; 13.13\% \\
-    \mnonstrict{} &     506 & \code{te} &   25.89\% & 12.06\% ; 18.18\% &  6.32\% ;  7.91\% \\
-                  &         & \code{fs} &   62.45\% & 37.94\% ; 53.75\% & 16.60\% ; 20.75\% \\
-    \mstrict{}    &      24 & \code{te} &   50.00\% & 29.17\% ; 33.33\% &     0\% ;     0\% \\
-                  &         & \code{fs} &   62.50\% & 50.00\% ; 58.33\% &  4.17\% ;  8.33\% 
-  \end{tabular}
-
-  \caption{TODO}
-
-\end{figure}
 
@@
 30-line range, etc.
-
-TODO must explain edit range carefully.
 
@@
 
+% TODO .. somewhere ... out/error-density-*
+%   row->te nocheck :: 48378 ++ , 9440 ==, 45479 --
+%   row->te nonstrict :: 19491 ++ , 9567 ==, 20121 --
+%   row->te strict :: 733 ++ , 368 ==, 770 --
+% 
+% TODO ... somewhere ... te leq fs : use error-density
+
 \begin{figure}[t]
-  \includegraphics[width=0.8\columnwidth]{img/example-error-delta.png}
-  \caption{Errors delta, nonstrict}
+  \includegraphics[width=\columnwidth]{img/error-count-nocheck-row->te-density-diff.pdf}
+  \includegraphics[width=\columnwidth]{img/error-count-nonstrict-row->te-density-diff.pdf}
+  \includegraphics[width=\columnwidth]{img/error-count-strict-row->te-density-diff.pdf}
+
+  \begin{tabular}{lr@{}rr@{}rr@{}r}
+    & \zerowidth{Add} & & \zerowidth{Keep} & & \zerowidth{Drop} \\\midrule
+    \mnocheck{} & 48378 & [\pct{46.83}] & 9440 & [\pct{9.14}] & 45479 & [\pct{44.03}] \\
+    \mnonstrict{} & 19491 & [\pct{39.63}] & 9567 & [\pct{19.45}] & 20121 & [\pct{40.91}] \\
+    \mstrict{} & 733 & [\pct{39.18}] & 368 & [\pct{19.67}] & 770 & [\pct{41.15}] \\
+  \end{tabular}
+  \caption{FILL error density}
   \label{f:error-delta}

### 2023-09-09 programming23/paper.tex
cleanup old figs

@@
 \begin{table}[t]
-  %% Looks the same when restricted to modules
+  \caption{Type error survival}
+  \label{t:type-error-survival}
+
+    \begin{tabular}{lr@{~}r@{~}rr@{~}r@{~}rr@{~}r@{~}r}
+      & \zerowidth{\mnocheck{}} & & & \zerowidth{\mnonstrict{}} & & & \zerowidth{\mstrict{}} & & \\
+      & \rcell{Add} & \ycell{Keep} & \gcell{Drop} & \rcell{Add} & \ycell{Keep} & \gcell{Drop} & \rcell{Add} & \ycell{Keep} & \gcell{Drop} \\\midrule
+      TypeMismatch & {0} & {0} & {0} & {159} & {62} & {129} & {21} & {10} & {28} \\
+      UnknownSymbol & {0} & {0} & {0} & {2348} & {496} & {2973} & {49} & {19} & {55} \\
+      UnknownProperty & {0} & {0} & {0} & {347} & {180} & {385} & {19} & {19} & {35} \\
+      NotATable & {0} & {0} & {0} & {7} & {2} & {8} & {2} & {0} & {1} \\
+      CannotExtendTable & {0} & {0} & {0} & {7} & {9} & {10} & {0} & {0} & {1} \\
+      OnlyTablesCanHaveMethods & {0} & {0} & {0} & {1} & {0} & {2} & {0} & {0} & {0} \\
+      DuplicateTypeDefinition & {0} & {0} & {0} & {1} & {0} & {0} & {0} & {0} & {0} \\
+      CountMismatch & {0} & {0} & {0} & {223} & {48} & {269} & {12} & {2} & {17} \\
+      FunctionDoesNotTakeSelf & {0} & {0} & {0} & {4} & {6} & {5} & {0} & {0} & {0} \\
+      OccursCheckFailed & {0} & {0} & {0} & {0} & {0} & {0} & {0} & {0} & {1} \\
+      UnknownRequire & {0} & {0} & {0} & {91} & {39} & {85} & {12} & {3} & {4} \\
+      IncorrectGenericParameterCount & {0} & {0} & {0} & {0} & {0} & {0} & {1} & {0} & {1} \\
+      SyntaxError & {8349} & {314} & {19220} & {983} & {46} & {2658} & {16} & {0} & {71} \\
+      UnknownPropButFoundLikeProp & {0} & {0} & {0} & {26} & {18} & {21} & {1} & {0} & {1} \\
+      GenericError & {0} & {0} & {0} & {206} & {52} & {203} & {8} & {1} & {9} \\
+      CannotCallNonFunction & {0} & {0} & {0} & {17} & {4} & {20} & {1} & {0} & {1} \\
+      ExtraInformation & {0} & {0} & {0} & {48} & {7} & {39} & {2} & {0} & {1} \\
+      ModuleHasCyclicDependency & {0} & {0} & {0} & {20} & {9} & {15} & {0} & {0} & {1} \\
+      IllegalRequire & {0} & {0} & {0} & {16} & {3} & {33} & {0} & {0} & {2} \\
+      FunctionExitsWithoutReturning & {0} & {0} & {0} & {9} & {3} & {6} & {11} & {5} & {8} \\
+      CannotInferBinaryOperation & {0} & {0} & {0} & {1} & {2} & {1} & {7} & {6} & {9} \\
+      MissingProperties & {0} & {0} & {0} & {11} & {3} & {11} & {6} & {4} & {6} \\
+      OptionalValueAccess & {0} & {0} & {0} & {40} & {52} & {27} & {6} & {2} & {7} \\
+      MissingUnionProperty & {0} & {0} & {0} & {1} & {0} & {2} & {0} & {0} & {0} \\
+      TypesAreUnrelated & {0} & {0} & {0} & {1} & {0} & {0} & {0} & {0} & {1} \\
+    \end{tabular}
+
+\end{table}
+
+FILL popularity over all old and new in edit range.
+Syntax error by far the most common.
+UnknownSymbol is likely a syntax error --- so \pct{90} syntax error!
+
+Fourteen errors never showed up.
+Nothing for \code{CodeTooComplex} or \code{UnificationTooComplex} is great news.
+But what about the others? FILL
+
+TODO get counts for each mode. Are there fewer stx errors in strict mode??!
+
+NOTE can't ask about type errors at end because we don't know
+the end.
+The final record can be sent minutes before the end of session.
+
+\begin{table}[t]
+  %% NOTE: looks the same when restricted to modules
+  TODO recompute split by mode : te-editrange-ss-nocheck.rktd etc
+
   \begin{tabular}{l@{~~}l}
@@
 
-FILL popularity over all old and new in edit range.
-Syntax error by far the most common.
-UnknownSymbol is likely a syntax error --- so \pct{90} syntax error!
-
-Fourteen errors never showed up.
-Nothing for \code{CodeTooComplex} or \code{UnificationTooComplex} is great news.
-But what about the others? FILL
-
-TODO get counts for each mode. Are there fewer stx errors in strict mode??!
-
-NOTE can't ask about type errors at end because we don't know
-the end.
-The final record can be sent minutes before the end of session.
-
-\begin{table}[t]
-
-    \begin{tabular}{lr@{~}r@{~}rr@{~}r@{~}rr@{~}r@{~}r}
-      & \zerowidth{\mnocheck{}} & & & \zerowidth{\mnonstrict{}} & & & \zerowidth{\mstrict{}} & & \\
-      & \rcell{Add} & \ycell{Keep} & \gcell{Drop} & \rcell{Add} & \ycell{Keep} & \gcell{Drop} & \rcell{Add} & \ycell{Keep} & \gcell{Drop} \\\midrule
-      TypeMismatch & {0} & {0} & {0} & {159} & {62} & {129} & {21} & {10} & {28} \\
-      UnknownSymbol & {0} & {0} & {0} & {2348} & {496} & {2973} & {49} & {19} & {55} \\
-      UnknownProperty & {0} & {0} & {0} & {347} & {180} & {385} & {19} & {19} & {35} \\
-      NotATable & {0} & {0} & {0} & {7} & {2} & {8} & {2} & {0} & {1} \\
-      CannotExtendTable & {0} & {0} & {0} & {7} & {9} & {10} & {0} & {0} & {1} \\
-      OnlyTablesCanHaveMethods & {0} & {0} & {0} & {1} & {0} & {2} & {0} & {0} & {0} \\
-      DuplicateTypeDefinition & {0} & {0} & {0} & {1} & {0} & {0} & {0} & {0} & {0} \\
-      CountMismatch & {0} & {0} & {0} & {223} & {48} & {269} & {12} & {2} & {17} \\
-      FunctionDoesNotTakeSelf & {0} & {0} & {0} & {4} & {6} & {5} & {0} & {0} & {0} \\
-      OccursCheckFailed & {0} & {0} & {0} & {0} & {0} & {0} & {0} & {0} & {1} \\
-      UnknownRequire & {0} & {0} & {0} & {91} & {39} & {85} & {12} & {3} & {4} \\
-      IncorrectGenericParameterCount & {0} & {0} & {0} & {0} & {0} & {0} & {1} & {0} & {1} \\
-      SyntaxError & {8349} & {314} & {19220} & {983} & {46} & {2658} & {16} & {0} & {71} \\
-      UnknownPropButFoundLikeProp & {0} & {0} & {0} & {26} & {18} & {21} & {1} & {0} & {1} \\
-      GenericError & {0} & {0} & {0} & {206} & {52} & {203} & {8} & {1} & {9} \\
-      CannotCallNonFunction & {0} & {0} & {0} & {17} & {4} & {20} & {1} & {0} & {1} \\
-      ExtraInformation & {0} & {0} & {0} & {48} & {7} & {39} & {2} & {0} & {1} \\
-      ModuleHasCyclicDependency & {0} & {0} & {0} & {20} & {9} & {15} & {0} & {0} & {1} \\
-      IllegalRequire & {0} & {0} & {0} & {16} & {3} & {33} & {0} & {0} & {2} \\
-      FunctionExitsWithoutReturning & {0} & {0} & {0} & {9} & {3} & {6} & {11} & {5} & {8} \\
-      CannotInferBinaryOperation & {0} & {0} & {0} & {1} & {2} & {1} & {7} & {6} & {9} \\
-      MissingProperties & {0} & {0} & {0} & {11} & {3} & {11} & {6} & {4} & {6} \\
-      OptionalValueAccess & {0} & {0} & {0} & {40} & {52} & {27} & {6} & {2} & {7} \\
-      MissingUnionProperty & {0} & {0} & {0} & {1} & {0} & {2} & {0} & {0} & {0} \\
-      TypesAreUnrelated & {0} & {0} & {0} & {1} & {0} & {0} & {0} & {0} & {1} \\
-    \end{tabular}
-
-  \caption{Type error survival}
-  \label{t:type-error-survival}
-\end{table}
-
 \section{Analysis}
@@
 
-% TODO .. somewhere ... out/error-density-*
-%   row->te nocheck :: 48378 ++ , 9440 ==, 45479 --
-%   row->te nonstrict :: 19491 ++ , 9567 ==, 20121 --
-%   row->te strict :: 733 ++ , 368 ==, 770 --
-% 
-% TODO ... somewhere ... te leq fs : use error-density
-
-\begin{figure}[t]
+\begin{figure}[t]\centering
+TODO diff vs 1st, not previous
+
   \includegraphics[width=\columnwidth]{img/error-count-nocheck-row->te-density-diff.pdf}
@@
   \includegraphics[width=\columnwidth]{img/error-count-strict-row->te-density-diff.pdf}
+
+  TODO bars instead, percent only, don't care about counts
 
@@
 
-Sixth, a mystery. May be a bug in Luau.
-May be DM insensitivity in strict mode
+Sixth, DM story.
+FS has: 1 all dependencies in strict, 2 data-model awareness (DM turns into
+type graph, ... initial DM, all properties go into a table type; DM -> folder
+-> model -> part .. get dot-driven autocomplete) in strict, DM might have type
+Any might have type Instance, the latter would mean more errors than FS
+
+DM insensitivity in strict mode
 (if DM has top type that would lead to errors.)
@@
 
-TODO clarify semantic x-axis order
-
-%% FS has: 1 all dependencies in strict, 2 data-model awareness (DM turns into
-%% type graph, ... initial DM, all properties go into a table type; DM -> folder
-%% -> model -> part .. get dot-driven autocomplete) in strict, DM might have type
-%% Any might have type Instance, the latter would mean more errors than FS
+% TODO ... somewhere ... te leq fs : use error-density
+%%
 
 \begin{figure}[t]
+  TODO use 2 bars? ask a few similar questions?
   \includegraphics[width=0.8\columnwidth]{img/example-tefs.png}
@@
 \begin{figure}[t]
-  \includegraphics[width=0.8\columnwidth]{img/example-compass.png}
+TODO smaller
+
+  \includegraphics[width=0.5\columnwidth]{img/example-compass.png}
   \caption{Are there fewer type errors than forcedstrict errors? Expect true always, but that is not the case.}
@@
 
-\begin{figure}[h]
-  \begin{tabular}{llrrr}
-    Mode & E & Add & Keep & Remove \\\midrule
-    \mnocheck{}   & \code{te} & - & - & - \\
-                  & \code{fs} & - & - & - \\
-    \mnonstrict{} & \code{te} & 16 & 18 & 21 \\
-                  & \code{fs} & 19 & 18 & 18 \\
-    \mstrict{}    & \code{te} & - & - & - \\
-                  & \code{fs} & - & - & - \\
-  \end{tabular}
-
-  \caption{Does mode influence the change in total error count?}
-  \label{f:edit-deltas}
-\end{figure}
-
 TODO nocheck accumulated errors?
@@
 FILL example sessions.
-
-\Cref{f:ex-session-strict} begins in strict mode
-but jumps to nocheck mode as well.
-The jumps sometimes coincide with module switches (vertical line),
-sometimes not.
-The green line counts type errors.in the codebase.
-
-\begin{figure}[t]
-  \includegraphics[width=0.7\columnwidth]{img/example-session-strict.pdf}
-  \caption{Example strict session}
-  \label{f:ex-session-strict}
-\end{figure}
-
-\Cref{f:ex-session-nonstrict} presents the three sessions from \mnonstrict{} mode
-that have the greatest number of events (not necessarily the longest time spans).
-There are no module or mode switches.
-There are few type errors.
-One high point shows 10 errors, but these errors are gone (probably fixed) at
-the next event.
-
-\begin{figure}[t]
-  \includegraphics[width=0.7\columnwidth]{img/example-session-nonstrict.pdf}
-  \caption{Three most populated nonstrict sessions}
-  \label{f:ex-session-nonstrict}
-\end{figure}
-
-\Cref{f:ex-session-nocheck} shows a roller coaster from \mnocheck{} mode.
-Several hundred ForcedStrict type errors appear and disappear between events.
-There are lots of module switches too \ldots but that makes no difference
-for these project-wide error counts.
-
-\begin{figure}[t]
-  \includegraphics[width=0.7\columnwidth]{img/example-session-nocheck.pdf}
-  \caption{Example nocheck session}
-  \label{f:ex-session-nocheck}
-\end{figure}
-
-\subsection{Aggregate Sessions}
-
-\begin{figure}[t]
-  \includegraphics[width=0.8\columnwidth]{img/fit-nocheck-Force-3.pdf}
-  \includegraphics[width=0.8\columnwidth]{img/fit-nonstrict-TypeE-3.pdf}
-  \includegraphics[width=0.8\columnwidth]{img/fit-strict-TypeE-3.pdf}
-  \caption{Best-fit curves for errors in each mode (\code{fs} for \mnocheck{}, \code{te} for the others)}
-  \label{f:scribbles}
-\end{figure}
-
-\Cref{f:scribbles} kill this.
 
@@
 
+Thanks to Ben(jamin?) Chung for feedback on plots and data analysis advice.
 Greenman was supported by

### 2023-09-11 programming23/paper.tex
te fs agree/disagree
type error popularity by mode

@@
 
-\begin{table}[t]
-  %% NOTE: looks the same when restricted to modules
-  TODO recompute split by mode : te-editrange-ss-nocheck.rktd etc
-
-  \begin{tabular}{l@{~~}l}
-    \begin{tabular}[t]{lr}
-      \code{SyntaxError} & 60.79\% \\
-      \code{UnknownSymbol} & 27.34\% \\
-      \code{UnknownProperty} & 4.08\% \\
-      \code{TypeMismatch} & 1.92\% \\
-      \code{UnknownRequire} & 1.40\% \\
-      \code{CountMismatch} & 1.08\% \\
-      \code{OptionalValueAccess} & 0.99\% \\
-      \code{GenericError} & 0.93\% \\
-      \code{CannotExtendTable} & 0.21\% \\
-      \code{UnknownPropButFoundLikeProp} & 0.20\% \\
-      \code{FunctionExitsWithoutReturning} & 0.16\% \\
-      \code{ExtraInformation} & 0.14\% \\
-      \code{CannotInferBinaryOperation} & 0.14\% \\
-      \code{MissingProperties} & 0.13\% \\
-      \code{ModuleHasCyclicDependency} & 0.10\% \\
-      \code{IllegalRequire} & 0.09\% \\
-      \code{NotATable} & 0.09\% \\
-      \code{MissingUnionProperty} & 0.08\%
-    \end{tabular} &
-    \begin{tabular}[t]{lr}
-      \code{CannotCallNonFunction} & 0.07\% \\
-      \code{FunctionDoesNotTakeSelf} & 0.03\% \\
-      \code{IncorrectGenericParameterCount} & 0.01\% \\
-      \code{OnlyTablesCanHaveMethods} & 0.00\% \\
-      \code{TypesAreUnrelated} & 0.00\% \\
-      \code{DuplicateTypeDefinition} & 0.00\% \\
-      \code{OccursCheckFailed} & 0.00\% \\
-      \code{TypePackMismatch} & 0.00\% \\
-      \code{FunctionRequiresSelf} & 0.00\% \\
-      \code{DuplicateGenericParameter} & 0.00\% \\
-      \code{InternalError} & 0.00\% \\
-      \code{CodeTooComplex} & 0.00\% \\
-      \code{DeprecatedApiUsed} & 0.00\% \\
-      \code{NormalizationTooComplex} & 0.00\% \\
-      \code{SwappedGenericTypeParameter} & 0.00\% \\
-      \code{DynamicPropertyLookupOnClassesUnsafe} & 0.00\% \\
-      \code{UnificationTooComplex} & 0.00\% \\
-    \end{tabular}
-  \end{tabular}
-  \caption{Type error popularity}
+\begin{table}[t]\centering
+  %% NOTE: popularity looks similar when [over all records] vs. [restricted to module switch records]
+
+  \begin{tabular}{ll}
+    \mnonstrict{} & \mstrict{} \\
+    \begin{tabular}[t]{l@{~}r}
+      \code{UnknownSymbol} & \pct{62.13} \\
+      \code{SyntaxError} & \pct{15.42} \\
+      \code{UnknownProperty} & \pct{8.28} \\
+      \code{UnknownRequire} & \pct{3.13} \\
+      \code{TypeMismatch} & \pct{2.44} \\
+      \code{CountMismatch} & \pct{2.26} \\
+      \code{OptionalValueAccess} & \pct{2.07} \\
+      \code{GenericError} & \pct{2.06} \\
+      \code{UnknownPropButFoundLikeProp} & \pct{0.44} \\
+      \code{CannotExtendTable} & \pct{0.42} \\
+      \code{ExtraInformation} & \pct{0.32} \\
+      \code{ModuleHasCyclicDependency} & \pct{0.23} \\
+      \code{IllegalRequire} & \pct{0.20} \\
+      \code{NotATable} & \pct{0.15} \\
+      \code{CannotCallNonFunction} & \pct{0.15} \\
+      \code{MissingProperties} & \pct{0.09} \\
+      \code{FunctionExitsWithoutReturning} & \pct{0.07} \\
+      \code{FunctionDoesNotTakeSelf} & \pct{0.07} \\
+      \code{MissingUnionProperty} & \pct{0.02} \\
+      \code{CannotInferBinaryOperation} & \pct{0.02} \\
+      \code{OnlyTablesCanHaveMethods} & \pct{0.01} \\
+      \code{DuplicateTypeDefinition} & {<\pct{0.01}} \\
+      \code{TypesAreUnrelated} & {<\pct{0.01}} \\
+    \end{tabular}
+        &
+    \begin{tabular}[t]{l@{~}r}
+      \code{UnknownSymbol} & \pct{23.97} \\
+      \code{TypeMismatch} & \pct{20.46} \\
+      \code{UnknownProperty} & \pct{18.88} \\
+      \code{SyntaxError} & \pct{9.31} \\
+      \code{CannotInferBinaryOperation} & \pct{6.94} \\
+      \code{MissingProperties} & \pct{4.04} \\
+      \code{CountMismatch} & \pct{2.99} \\
+      \code{OptionalValueAccess} & \pct{2.99} \\
+      \code{FunctionExitsWithoutReturning} & \pct{2.99} \\
+      \code{UnknownRequire} & \pct{2.37} \\
+      \code{GenericError} & \pct{2.28} \\
+      \code{NotATable} & \pct{1.32} \\
+      \code{ExtraInformation} & \pct{0.35} \\
+      \code{UnknownPropButFoundLikeProp} & \pct{0.26} \\
+      \code{IncorrectGenericParameterCount} & \pct{0.18} \\
+      \code{CannotCallNonFunction} & \pct{0.18} \\
+      \code{IllegalRequire} & \pct{0.18} \\
+      \code{ModuleHasCyclicDependency} & \pct{0.09} \\
+      \code{CannotExtendTable} & \pct{0.09} \\
+      \code{OccursCheckFailed} & \pct{0.09} \\
+      \code{TypesAreUnrelated} & \pct{0.09} \\
+    \end{tabular}
+  \end{tabular}
+
+  \caption{Type error popularity for \mnonstrict{} and \mstrict{} modes.
+  In \mnocheck{}, \code{SyntaxError} accounts for \pct{100} of the error.}
   \label{t:type-error-count}
@@
 
-% TODO ... somewhere ... te leq fs : use error-density
-%%
-
-\begin{figure}[t]
-  TODO use 2 bars? ask a few similar questions?
-  \includegraphics[width=0.8\columnwidth]{img/example-tefs.png}
-  \caption{Are there fewer type errors than forcedstrict errors? Expect true always, but that is not the case.}
+\begin{figure}[t]\centering
+  %% TODO
+  %% - thin bars
+  %% - colors for good , bad (disagree) , neutral
+  \begin{tabular}{lll}
+    \mnocheck{} & \mnonstrict{} & \mstrict{} \\
+    \includegraphics[width=0.3\columnwidth]{img/compass-nocheck.pdf}
+    &
+    \includegraphics[width=0.3\columnwidth]{img/compass-nonstrict.pdf}
+    &
+    \includegraphics[width=0.3\columnwidth]{img/compass-strict.pdf}
+  \end{tabular}
+  \caption{FILL Are there fewer type errors than forcedstrict errors? Expect true always, but that is not the case.}
   \label{f:tefs}
@@
 
-Seventh,
-
-\begin{figure}[t]
-TODO smaller
-
-  \includegraphics[width=0.5\columnwidth]{img/example-compass.png}
-  \caption{Are there fewer type errors than forcedstrict errors? Expect true always, but that is not the case.}
-  \label{f:tefs}
-\end{figure}
+TODO individual sessions
 

### 2023-09-11 programming23/paper.tex
Getting LaTeX to run without error
Can't use greater-than as a filename char

@@
 \usepackage{pifont}
-\usepackage{siunitx}
+% Errors with
+% ! Undefined control sequence.
+% <argument> \l__siunitx_number_implicit_plus_bool 
+% l.61 \begin{document}
+% using Mac TeXlive 20230313_2
+% \usepackage{siunitx}
 \usepackage{tikz}
@@
 
-  \includegraphics[width=\columnwidth]{img/error-count-nocheck-row->te-density-diff.pdf}
-  \includegraphics[width=\columnwidth]{img/error-count-nonstrict-row->te-density-diff.pdf}
-  \includegraphics[width=\columnwidth]{img/error-count-strict-row->te-density-diff.pdf}
+  % > is not a valid filename char on some OSs!
+  \includegraphics[width=\columnwidth]{img/error-count-nocheck-row--te-density-diff.pdf}
+  \includegraphics[width=\columnwidth]{img/error-count-nonstrict-row--te-density-diff.pdf}
+  \includegraphics[width=\columnwidth]{img/error-count-strict-row--te-density-diff.pdf}
 
@@
     \mnocheck{} & \mnonstrict{} & \mstrict{} \\
-    \includegraphics[width=0.3\columnwidth]{img/compass-nocheck.pdf}
+    % Missing file
+    % \includegraphics[width=0.3\columnwidth]{img/compass-nocheck.pdf}
     &
-    \includegraphics[width=0.3\columnwidth]{img/compass-nonstrict.pdf}
+    % Missing file
+    % \includegraphics[width=0.3\columnwidth]{img/compass-nonstrict.pdf}
     &
-    \includegraphics[width=0.3\columnwidth]{img/compass-strict.pdf}
+    % Missing file
+    % \includegraphics[width=0.3\columnwidth]{img/compass-strict.pdf}
   \end{tabular}

### 2023-09-11 programming23/paper.tex
Use BibTeX rather than biber
General tidying up
CCS classification

@@
 %% BEGIN tobias pape 2021-11-06
-\makeatletter
-\newcommand*\abstractpart[1]{\unskip\par\noindent{\firamedium\color{P@GrayFG}{#1}}\enspace}
-\makeatother
+% Why do we need this?
+%\makeatletter
+%\newcommand*\abstractpart[1]{\unskip\par\noindent{\firamedium\color{P@GrayFG}{#1}}\enspace}
+%\makeatother
 %% END
@@
 \usepackage{alltt}
-\usepackage[backend=biber]{biblatex}
+% Not sure why <Programming> is recommending non-standard tools
+% \usepackage[backend=biber]{biblatex}
 \usepackage{amssymb}
@@
 
-\addbibresource{bib.bib}
-
 %%%%%%%%%%%%%%%%%%
@@
 % \orcid{0000-0001-7078-9287}
-\affiliation[a]{Brown University, Providence, RI, USA}
 
@@
 <concept>
-<concept_id>10011007.10011006.10011039.10011311</concept_id>
-<concept_desc>Software and its engineering~Semantics</concept_desc>
+<concept_id>10002944.10011123.10010916</concept_id>
+<concept_desc>General and reference~Measurement</concept_desc>
 <concept_significance>500</concept_significance>
-</concept>
-<concept>
-<concept_id>10011007.10011006.10011008.10011024.10011032</concept_id>
-<concept_desc>Software and its engineering~Constraints</concept_desc>
-<concept_significance>100</concept_significance>
-</concept>
-<concept>
-<concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
-<concept_desc>Software and its engineering~Functional languages</concept_desc>
-<concept_significance>100</concept_significance>
 </concept>
@@
 
-\ccsdesc[500]{Software and its engineering~Semantics}
-\ccsdesc[100]{Software and its engineering~Constraints}
-\ccsdesc[100]{Software and its engineering~Functional languages}
+\ccsdesc[500]{General and reference~Measurement}
 
@@
 
-{ \printbibliography }
+\bibliography{bib}
 

### 2023-09-11 programming23/paper.tex
Updated Roblox stats

@@
 
-{Roblox} is a platform for {shared virtual experiences},
-with 56~million Daily Active Users, and 49~billion hours of engagement in
-2022~\cite{roblox-quarterly-results}.
-There are \textbf{XX}~million creators using {Roblox Studio},
-and \textbf{YY}~million creations.
+{Roblox} is a platform for {shared virtual experiences}, with
+65~million Daily Active Users, and 14~billion hours of engagement in
+April--June 2023; there are 3~million creators using {Roblox Studio},
+creating 5~million experiences~\cite{corp.roblox.com}.
 

### 2023-09-12 programming23/paper.tex
More citations

@@
 
-Beyond their immediate applications to {Lua},
+Beyond their immediate applications to {Luau},
 answers to these questions have implications for the design
-of gradual types, success types, and semantic subtyping.
+of gradual types~\cite{st-sfp-2006}, success types~\cite{lindahl2006practical},
+and semantic subtyping~\cite{CF05:GentleIntroduction,Jef22:SemanticSubtyping}.
 Luau represents the first large-scale combination of these features,

### 2023-09-12 programming23/paper.tex
Tidying up

@@
 the location in code where reported errors occur.
-
 To opt in to type error reporting, creators set a \emph{mode}
@@
   \item \mnocheck{}: only syntax errors are reported,
-  \item \mnonstrict{}: all syntax errors, and a subset of ``high probability'' type errors, are reported, or
+  \item \mnonstrict{}: all syntax errors, and a subset of ``high confidence'' type errors, are reported, or
   \item \mstrict{}: all syntax errors and type errors, are reported.
@@
 
-Both modes are opt-in. Creators have the option to make nonstrict mode
+Both modes are opt-in. Creators have the option to make \mnonstrict{} mode
 the default rather than \mnocheck{} mode.
@@
 \emph{required} by other scripts. Since module scripts can require
-other module script, modules form a graph (though we consider it to be
-an error in strict mode if the graph is cyclic, and remove edges to
-make it acyclic).
+other module script, modules form a graph (though it is
+an error in strict mode if the graph is cyclic, and edges are removed to
+make the graph acyclic).
 
@@
 or which transitively requires a dirty module, is typechecked. More
-commonly, when typechecking is performed for autocomplete, we only
-need to typecheck the current script, since it is the only dirty
-script, and nothing it requires can transitively require it, since we
-have broken cycles.
+commonly, when typechecking is performed for autocomplete, only
+the current script is typechecked, since it is the only dirty
+script, and nothing it requires can transitively require it, since
+the module graph is acyclic.
 

### 2023-09-12 programming23/paper.tex
Tidying up
Tidying up telemetry intro

@@
 constraints such as forces, springs and joints.
-
+The data model can be seen in the Explorer view of Fig.~\ref{fig:roblox-studio}.
 While an experience is under development, it is typical for the data
@@
 
-Prior work: transparent telemetry, VS Code telemetry, many others undocumented
-that are probably not cleary documented.
-\Cref{t:telemetry-design} gives an overview.
-
-\paragraph{Goals:}
-
+\subsection{Limits of telemetry}
+
+Telemetry allows an application to ``phone home'' with data
+summarizing usage patterns, such performance data, crash reporting,
+and feature uptake. A typical usage is in deciding whether an API can
+be deprecated: without telemetry it may be difficult to know how
+popular a feature is with users, but with telemetry it is
+straightforward.
+
+IDEs use telemetry in a similar fashion to most user-facing
+applications, for example VSCode~\cite{vsc-telemetry} and
+IntelliJ~\cite{intellij-telemetry} report telemetry.
+
+Telemetry for programming languages can be controversial, for example
+the lively discussion around telemetry in the Go
+toolchain~\cite{golang-telemetry}. 
+The Luau open source toolchain does \emph{not} report telemetry,
+as it is designed to be used in build environments such as Continuous Integration
+servers, where hermetic deterministic builds are expected.
+
+Much of the concern about telemetry is because of its potential impact on
+privacy, by revealing Personally Identifiable Information (PII).
+For this reason, it is important that any experiments using telemetry
+not reveal PII:
+no source code;
+no source code locations;
+no error messages (which may contain source code);
+no record of the creator's identity, locale, or IP address;
+and no information about what creation the data came from.
+A good introduction to the privacy implications and tradeoffs
+involved with telemetry is~\cite{transparent-telemetry}.
+
+For performance reasons, there are limits to telemetry data:
+telemetry records have size limits, and the performance impact of
+recording telemetry should be minimal.
+
+In summary, the limitations of telemetry are:
 \begin{itemize}
-  \item
-    Learn about type system adoption and use
-  \item
-    Discover roadblocks to adoption
-  \item
-    Assess the helpfulness of error highlights
+  \item Data which reveals no PII.
+  \item Fixed-size telemetry records.
+  \item Minimal overhead to record telemtry.
 \end{itemize}
-
-\paragraph{Limits:}
-
+In addition, due to the architecture of Roblox Studio
+there is some information that is not available to the
+typechecker:
 \begin{itemize}
-\item
-  No personal data (PII)
-  \subitem
-    Fairly obvious.
-    But, we also can't collect directory names, filenames, or code snapshots
-    because these all might have PII throughout.
-\item
-  No data about public events.
-  \subitem
-    Contributions to the Roblox Marketplace, for instance, are a public event.
-    Although it would be interesting to know the number of type errors
-    in published experiences, this data might be used to identify creators
-    who publish often --- or on a distinct schedule.
-\item
-  Low cost
-  \subitem
-    Building, sending, and storing the data must all run quickly.
-    FILL what was our hard cap on Kibana use?
+  \item The typechecker does not have access to some lifecycle events,
+    such as save, quit and publish.
+  \item The typechecker does not have access to GUI state, for example
+    whether the Script Analysis widget is visible.
 \end{itemize}
 
-\paragraph{Solutions:}
+\subsection{When are we logging?}
 
@@
 
-Extra Roblox constraints:
-- No access to some events: save, exit
-- No idea if analysis widget is open, closed, or obscured
-- Don't know which module is the data model
-  (no ids for ANY module for that matter)
-
-\url{https://docs.google.com/document/d/1DnKvw8x1jy0EWCbBSM8neKULWz7WAg1OauKy7qqYqkI/edit}
-
-\subsection{What are we logging (from gdoc)}
+\subsection{What are we logging?}
 

### 2023-09-13 programming23/paper.tex
related: placeholder cite for liblit
add missing files, compass pdfs
checkpoint: start to programming-ize the abstract

@@
 \begin{abstract}
-  {Roblox Studio} is a tool which places programming in the hands of
-  millions of creators, ranging from high school students to professional
-  development studios. It has the ability to report telemetry data,
+  \let\paragraph\abstractpart
+
+  \paragraph{Context}
+  %% What is the broad context of the work? What is
+  %% the importance of the general research area?
+  {Roblox Studio} is a tool that lets millions of creators
+  build interactive experiences by programming in a variant
+  of Lua called Luau.
+  The creators form a broad group, ranging from novices writing
+  their first script to professional Luau developers, thus Luau
+  must support a wide audience.
+  As part of its efforts to support all kinds programmers, Luau includes a
+  gradual type system that goes to great lengths to minimize false positive
+  errors.
+
+  \paragraph{Inquiry}
+  %% What problem or question does the paper
+  %% address? How has this problem or question been
+  %% addressed by others (if at all)?
+  Since Luau is in the hands of so many creators,
+  how do they use it?
+  Of central importance, however, is protecting the privacy
+  of Luau creators.
+  FILL
+
+%  TYPE ERRORS
+%  investigate
+
+  It has the ability to report telemetry data,
   which allows large-scale measurement of the creator experience. In
@@
 
-  %\let\paragraph\abstractpart
-
-  % \paragraph{Context}
-  %% What is the broad context of the work? What is
-  %% the importance of the general research area?
-
-  % \paragraph{Inquiry}
-  %% What problem or question does the paper
-  %% address? How has this problem or question been
-  %% addressed by others (if at all)?
-
-  %\paragraph{Approach}
+  \paragraph{Approach}
   %% What was done that unveiled new knowledge?
-
-  %\paragraph{Knowledge}
+  FILL impersonal telemetry, N months.
+
+  \paragraph{Knowledge}
   %% What new facts were uncovered? If the
@@
   %% capabilities are enabled by the work?
-
-  %\paragraph{Grounding}
+  FILL
+
+  \paragraph{Grounding}
   %% What argument, feasibility proof, artifacts,
   %% or results and evaluation support this work?
-
-  %\paragraph{Importance}
+  Our findings are supported by a publicly-available dataset
+  of over 1.5 million telemetry records.
+
+  \paragraph{Importance}
   %% Why does this work matter?
+  FILL improve Luau, offer lessons to similar developments (epic?, elixir?).
 
@@
 
-TODO must show full plot, zoomed plot side by side
-
 Fifth,
@@
 \begin{figure}[t]\centering
-TODO diff vs 1st, not previous
-
-  % > is not a valid filename char on some OSs!
+% TODO diff vs 1st, not previous
+
+  \mnocheck{}
   \includegraphics[width=\columnwidth]{img/error-count-nocheck-row--te-density-diff.pdf}
+  \medskip
+  \mstrict{}
   \includegraphics[width=\columnwidth]{img/error-count-nonstrict-row--te-density-diff.pdf}
+  \medskip
+  \mstrict{}
   \includegraphics[width=\columnwidth]{img/error-count-strict-row--te-density-diff.pdf}
@@
     \mnocheck{} & \mnonstrict{} & \mstrict{} \\
-    % Missing file
-    % \includegraphics[width=0.3\columnwidth]{img/compass-nocheck.pdf}
+    \includegraphics[width=0.3\columnwidth]{img/compass-nocheck.pdf}
     &
-    % Missing file
-    % \includegraphics[width=0.3\columnwidth]{img/compass-nonstrict.pdf}
+    \includegraphics[width=0.3\columnwidth]{img/compass-nonstrict.pdf}
     &
-    % Missing file
-    % \includegraphics[width=0.3\columnwidth]{img/compass-strict.pdf}
+    \includegraphics[width=0.3\columnwidth]{img/compass-strict.pdf}
   \end{tabular}
@@
 \label{s:related}
+
+Liblit dissertation~\cite{liblit-thesis}
+
+RAMSS workshop series.
 

### 2023-09-13 programming23/paper.tex
max -> 99 percentile, fix abstract

@@
 %% BEGIN tobias pape 2021-11-06
-% Why do we need this?
-%\makeatletter
-%\newcommand*\abstractpart[1]{\unskip\par\noindent{\firamedium\color{P@GrayFG}{#1}}\enspace}
-%\makeatother
+% Why do we need this? To have headings in the abstract.
+\makeatletter
+\newcommand*\abstractpart[1]{\unskip\par\noindent{\firamedium\color{P@GrayFG}{#1}}\enspace}
+\makeatother
 %% END
@@
 \begin{figure}[t]
-  %% \includegraphics{img/row-distribution.pdf}
+  % ;; (list* (max* vv) (median < vv) mm (stddev/mean mm vv) (percentile* vv)))
+  % ;; percentile* = 0.95 -- 0.99
+  %#hash((editrange . (1156036 926 3007043/817 30975.40553032358 (0.95 8220) (0.96 9858) (0.97 13656) (0.98 18956) (0.99 34725)))
+  %      (event-count . (6079 138 38689/135 582.5752201972966 (0.95 960) (0.96 1174) (0.97 1304) (0.98 1836) (0.99 3302)))
+  %      (files . (54884 7678 55257996/4639 11853.650098459446 (0.95 40029) (0.96 42771) (0.97 45417) (0.98 48688) (0.99 51761)))
+  %      (lines . (1089963 3115 38928947/5992 22364.75846303099 (0.95 18561) (0.96 25892) (0.97 27014) (0.98 29725) (0.99 50547)))
+  %      (timespan . (1387897376 845648 591343604248/185705 15724745.722577972 (0.95 10696064) (0.96 12715468) (0.97 15596940) (0.98 21153924) (0.99 35450460))))
+
   \begin{tabular}{l@{}r@{~}l@{}r@{~}l@{}r@{~}l}
@@
     Median       &  7,678 &              &     3,115 &             &           926 & \\
-    Max          & 54,884 &              & 1,089,963 &             &    1,156,036  & \\
+    \pct{99}     & 51,761 &              &    50,547 &             &        34,725 & \\
     & \includegraphics[width=0.2\columnwidth]{img/files-distribution.pdf}
@@
       Median       &       845 &              &     138        \\
-      Max          & 1,387,897 &              &   6,079        \\
+      \pct{99}     &    35,450 &              &   3,302        \\
       & \includegraphics[width=0.2\columnwidth]{img/timespan-distribution.pdf}

### 2023-09-13 programming23/paper.tex
note some outliner telemetry records

@@
   %      (timespan . (1387897376 845648 591343604248/185705 15724745.722577972 (0.95 10696064) (0.96 12715468) (0.97 15596940) (0.98 21153924) (0.99 35450460))))
+  %% NOTE: 3 records have +1M lines, all from one nocheck session, all with 3 files
+  %% - session "30,762,216,447,324"
+  %% NOTE: 133 records have +52K files, all from one nocheck session
+  %% - session "330,172,870,224,081" 
+  %% (2nd biggest has 41K files "31,650,473,913,529")
 
@@
   \begin{subfigure}[t]{\columnwidth}
+    %% NOTE max time = 1 million seconds = 15 days
+
     \begin{tabular}{l@{}r@{~}l@{}r@{~}l} \\
@@
     \end{tabular}
-
-    15 days
 
@@
 
-Fourteen errors never showed up.
+FILL N errors never showed up.
 Nothing for \code{CodeTooComplex} or \code{UnificationTooComplex} is great news.
@@
 
+\subsection{Internal Error}
+%% out/te-editrange-*
+
+FILL hope to never see \code{InternalError}.
+Have zero, yay!
+
 \subsection{Dead Ends: Too Complex}
@@
 inference~\cite{m-popl-1990,ktu-caap-1990}, the typechecker
-has two internal limits to restrict the problems it will attempt to solve.
-Hitting a limit triggers either a \code{CodeTooComplex} or
-\code{UnificationTooComplex} error.
+has internal limits that restrict the problems it will attempt to solve.
+Hitting a limit triggers one of the following errors:
+\code{CodeTooComplex},
+\code{NormalizationTooComplex}, or
+\code{UnificationTooComplex}.
 A casual user should never see these errors.
@@
 
-Thanks to Ben(jamin?) Chung for feedback on plots and data analysis advice.
+Thanks to Benjamin Chung for feedback on plots and data analysis advice.
 Greenman was supported by

### 2023-09-13 programming23/paper.tex
Tidying up telemetry design

@@
 
-\subsection{When are we logging?}
-
+\subsection{Telemetry records}
+
+The telemetry for a Roblox Studio session is gathered as a series of
+telemtry records, all with the same psedonynimized session
+identifier. This allows us to correlate telemetry across a single
+session, but not between sessions.
+
+In order to avoid swamping our telemtry servers, we randomly throttle
+the generation of telemetry records:
 \begin{itemize}
   \item
-    Collect metrics instead of source code: lines, modules, num errors
+    10\% of Roblox Studio sessions were enrolled in generating telemetry, and
   \item
-    Track an edit range to compare type errors with edits,
-    use a coarse $[min, max]$ interval for speed
+    5\% of runs of the typechecker in an enrolled session generated a telemetry record.
+\end{itemize}
+In order to find how file-switching correlated with type errors, we additionally have:
+\begin{itemize}
   \item
-    Count specific errors only for the current edit range
-    (reduces size of logs)
-  \item
-    Within the edit range, collect data before and after the latest edits.
-  \item
-    Log randomly on keystroke instead of per specific event.
-    One exception: log module switches because that helps discover
-    codebases that use several analysis modes.
-    Randomness helps lower the amount of data.
-  \item
-    Observe sessions randomly (again to lower the amount of data).
+    every run of the typechecker in an enrolled session which switched file generated a telemetry record.
 \end{itemize}
-
-TODO must explain edit range carefully.
-
-\subsection{What are we logging?}
-
-\FILL{} change to long-form paragraphs, use more space for examples
-
+Each telemetry record contains:
 \begin{description}
-  \item[D1] metadata: uid, timestamp, session duration
-  \item[D2] mode (\mnocheck{}, \mnonstrict{}, \mstrict{})
-  \item[D3] a 1-bit reason for sending: did the user switch module?
-    \subitem we get logs for two reasons: (1) switch module, (2) randomly with
-    low probability after typechecking
-  \item[D4] size of codebase: num files, num lines
+  \item[D1] metadata: a psedonymized session identifier, timestamp and session duration,
+  \item[D2] mode of the current file: (\mnocheck{}, \mnonstrict{}, or \mstrict{}),
+  \item[D3] reason for sending: (random selection, or file switch),
+  \item[D4] size of codebase: number of files, and number of lines of code type checked,
   \item[D5] size of edit range
-  \item[D6] overall type errors: for old and curr states, have: \# total, \# in module, and \# in range
-  \item[D7] specific errors in the edit range: in the edit range, for old and
-    curr states, how many type errors for each type error code? (There is 1
-    edit range per log.)
-    \subitem Examples, if the old state had one type-A error:
-    \begin{itemize}
-      \item If the latest edits touched it and fixed it, the next log says 1
-        old type-A and 0 new type-A
-      \item touched it, didn't fix => 1 old-A, 1 new-A
-      \item didn't touch (fixed or not) => 0 old-A, 0 new-A
-    \end{itemize}
-  \item[D8] ForcedStrict "what if" errors for \mnocheck{}: specific errors for the
-    edit range computed as if the user had strict mode enabled (even though they
-    have no-check enabled)
-    \subitem This info is available because it gives better hints for
-    autocomplete, because it assumes precise types for ALL libraries (rather
-    than Any)
+  \item[D6] overall type errors: for previous and current states, we have the total number, the nunber in the current file, and number in the edit range,
+  \item[D7] specific errors in the edit range: for the previous and current states, for each type error code, the number type errors in the edit range,
+  \item[D8] forced strict errors: specific errors for the
+    edit range computed as if the user had strict mode enabled.
   \item[D9] TooComplexErrors: project-wide count for this one specific error type
 \end{description}
+To track the edit range, we record a start and end position, which we
+update appropriately on every edit. This can result in very large edit
+ranges (for example if the user edits an the beginning and end of the
+file) but fits in a fixed-size telemetry record.
 

### 2023-09-13 programming23/paper.tex
Add short bios
Not luatex
huge ctime comment (15 days), fix overlapping color boxes

@@
 \author[b]{Alan Jeffrey}
-\authorinfo{(\email{ajeffrey@roblox.com}) TBD}
+\authorinfo{(\email{ajeffrey@roblox.com}) is a Principal Software Engineer at Roblox.}
 % \orcid{0000-0001-6342-0318}
@@
 \author[b]{Mitesh Shah}
-\authorinfo{(\email{mshah@roblox.com}) TBD.}
+\authorinfo{(\email{mshah@roblox.com}) is Senior Engineering Director, Programmability, at Roblox.}
 
@@
 \newcommand{\rcell}[1]{\cellcolor{red!30}$\!\!$#1$\!\!$}
+\newcommand{\gbox}[1]{\colorbox{green!20}{#1}}
+\newcommand{\ybox}[1]{\colorbox{yellow!18}{#1\vphantom{Ap}}}
+\newcommand{\obox}[1]{\colorbox{orange!29}{#1\vphantom{Ap}}}
+\newcommand{\rbox}[1]{\colorbox{red!30}{#1\vphantom{Ap}}}
 
@@
   \begin{subfigure}[t]{\columnwidth}
-    %% NOTE max time = 1 million seconds = 15 days
+    %% NOTE max time = 1 million seconds = 15 days ...  +100 are over 2 days
 
@@
 
-    \begin{tabular}{lr@{~}r@{~}rr@{~}r@{~}rr@{~}r@{~}r}
+    \begin{tabular}{lr@{}r@{}rr@{}r@{}rr@{}r@{}r}
       & \zerowidth{\mnocheck{}} & & & \zerowidth{\mnonstrict{}} & & & \zerowidth{\mstrict{}} & & \\
-      & \rcell{Add} & \ycell{Keep} & \gcell{Drop} & \rcell{Add} & \ycell{Keep} & \gcell{Drop} & \rcell{Add} & \ycell{Keep} & \gcell{Drop} \\\midrule
+      & \rbox{Add} & \ybox{Keep} & \gbox{Drop} & \rbox{Add} & \ybox{Keep} & \gbox{Drop} & \rbox{Add} & \ybox{Keep} & \gbox{Drop} \\\midrule
       TypeMismatch & {0} & {0} & {0} & {159} & {62} & {129} & {21} & {10} & {28} \\
@@
 
+\Cref{f:tefs}~FILL story is edits to DM vs edits to code.
+Data model awareness, call for future work.
+
+\begin{figure}[t]
+  \mnocheck{}
+
+  \mnonstrict{}
+
+  \mstrict{}
+
+  \caption{Example sessions}
+  \label{f:indy-session}
+\end{figure}
+
 TODO individual sessions
+
+TODO correlations, compare modes!!!
 
@@
 \end{document}
-
-% Local Variables:
-% TeX-engine: luatex
-% End:

### 2023-09-13 programming23/paper.tex
more abstract

@@
   %% addressed by others (if at all)?
-  Since Luau is in the hands of so many creators,
-  how do they use it?
-  Of central importance, however, is protecting the privacy
-  of Luau creators.
-  FILL
-
-%  TYPE ERRORS
-%  investigate
-
-  It has the ability to report telemetry data,
-  which allows large-scale measurement of the creator experience. In
-  this paper, we discuss one use of telemetry, to measure the experience
-  of type error reporting.
-  We find FILL.
+  Our main goal is to collect feedback from creators to improve the design of
+  Luau.
+  We want to do so at a large scale, as creators work on their actual projects,
+  which suggests we deploy client-side telemetry to report on notable events.
+  However, we must respect creators privacy.
+  Finding ways to harness data from telemetery without exposing any
+  personally-identifiable information (such as source code) is the
+  central research challenge.
 
@@
   %% What was done that unveiled new knowledge?
-  FILL impersonal telemetry, N months.
+  We designed and implemented a privacy-preserving telemetry system for Luau.
+  We deployed it for approximately four months in Spring 2023, during which
+  time we collected over one million telemetry records.
+  The telemetry focuses on the type errors that creators encountered.
 
@@
   of over 1.5 million telemetry records.
+  We also release the code used to analyze the data.
 

### 2023-09-14 programming23/paper.tex
abstract draft

--- 2023-09-13 programming23/paper.tex
@@
   %% the importance of the general research area?
-  {Roblox Studio} is a tool that lets millions of creators
+  {Roblox Studio} lets millions of creators
   build interactive experiences by programming in a variant
@@
   The creators form a broad group, ranging from novices writing
-  their first script to professional Luau developers, thus Luau
+  their first script to professional developers, thus Luau
   must support a wide audience.
   As part of its efforts to support all kinds programmers, Luau includes a
-  gradual type system that goes to great lengths to minimize false positive
+  gradual type system and goes to great lengths to minimize false positive
   errors.
@@
   %% addressed by others (if at all)?
-  Since Luau is in the hands of so many creators,
-  how do they use it?
-  Of central importance, however, is protecting the privacy
-  of Luau creators.
-  FILL
-
-%  TYPE ERRORS
-%  investigate
-
-  It has the ability to report telemetry data,
-  which allows large-scale measurement of the creator experience. In
-  this paper, we discuss one use of telemetry, to measure the experience
-  of type error reporting.
-  We find FILL.
+  Since Luau is in the hands of many creators, we want to collect their feedback
+  and ultimately improve the language design (in particular, the type system).
+  We want feedback at a large scale, as creators work on their actual projects,
+  which suggests the deployment of client-side telemetry.
+  %% ^ awkward
+  However, we must respect creators privacy.
+  Whatever telemetry we use to study the Luau language cannot include
+  personally-identifiable information.
+  For example, we cannot monitor source code or error messages as they
+  may contain personal data.
+  Finding ways to harness telemetry in a private way is the central research challenge.
 
@@
   %% What was done that unveiled new knowledge?
-  FILL impersonal telemetry, N months.
+  We designed and implemented a privacy-preserving telemetry system for Luau.
+  Telemetry records include a timestamp, a (randomized) session id, a reason
+  for sending, and a sequence of counter values summarizing the results of type
+  analysis.
+  Over four months in Spring 2023, we deployed the telemetry system and collected
+  1.5 million telemetry records.
+  We analyzed the data to improve Luau.
 
@@
   %% capabilities are enabled by the work?
-  FILL
+  We present several findings about Luau:
+  some confirm hypotheses by the Luau team, others are rather surprising, but
+  all findings support our claim that privacy-preserving telemetry is a useful
+  way to collect feedback.
+  One expected finding is that types are unpopular; there is an order-of-maginitude
+  difference in the number of sessions that use different type analysis modes.
+  One surprise is that the strict type analysis mode evidently makes it difficult
+  to connect a script with data assets.
+  Lastly, one reassuring finding is that type analysis rarely hits its internal
+  limits on problem size.
 
@@
   Our findings are supported by a publicly-available dataset
-  of over 1.5 million telemetry records.
+  of over 1.5 million telemetry records
+  and by free software for analyzing the data.
 
@@
   %% Why does this work matter?
-  FILL improve Luau, offer lessons to similar developments (epic?, elixir?).
+  Beyond the immediate benefits to Luau,
+  our findings about types and type errors have implications
+  for adoption and ergonomics in other gradually typed languages.
+  %% mention Elixir, Verse?
+  FILL specifics.
+  Our telemetry design is of broad interest, as it gathers useful
+  data without exposing personal information.
 
@@
   \item
-    10\% of Roblox Studio sessions were enrolled in generating telemetry, and
+    1\% of Roblox Studio sessions were enrolled in generating telemetry, and
   \item
-    5\% of runs of the typechecker in an enrolled session generated a telemetry record.
+    0.5\% of runs of the typechecker (approximately 1 in 200 keystrokes)
+      in an enrolled session generated a telemetry record.
 \end{itemize}
@@
   \medskip
-  \mstrict{}
+  \mnonstrict{}
   \includegraphics[width=\columnwidth]{img/error-count-nonstrict-row--te-density-diff.pdf}
@@
 Problematic, because syntax errors are uninteresting (easy to fix),
-but common, which makes it likely that our per-keystroke telemetry (1/2k) will record them.
+but common, which makes it likely that our per-keystroke telemetry (1/200) will record them.
 Edit range has specific counts.
@@
 
+\paragraph{Do we need timestamps?}
+
+Our telemetry records include timestamps.
+While timestamps are not personal data per se, they are in the ballpark.
+Telemetry would be even more private without them.
+Do we need them?
+
+Timestamps are not the focus of our analysis, but they do play a critical role.
+Consider~\cref{f:error-delta}.
+With timestamps, it is clear that error density stays close to zero.
+Without timestamps, we would be left with the counts at the bottom
+of~\cref{f:error-delta}.
+Many distributions could fit those counts, such as a ``mountain'' session that first
+grows and grows the number of type errors then shrinks and shrinks down to zero.
+
 \acks

### 2023-09-14 programming23/paper.tex
intro pass

--- 2023-09-14 programming23/paper.tex
@@
 
-\title{Millions of Type Errors}
-\subtitle{Impersonal Telemetry to Measure User Experience}
+\title{Type Error Telemetry at Scale: Pseudonymized and Stochiastic}
 % lean, lite, modest, non-intrusive, private, ...
+% {Millions of Type Errors} = 72M forced strict, 1/2M type error is disappointing!
+%% \subtitle{Impersonal Telemetry to Measure User Experience}
 
@@
   %% Why does this work matter?
+  Luau is the first large-scale combination of gradual types and semantic
+  subtyping.
   Beyond the immediate benefits to Luau,
   our findings about types and type errors have implications
-  for adoption and ergonomics in other gradually typed languages.
+  for adoption and ergonomics in other gradual languages.
   %% mention Elixir, Verse?
@@
 The main extension is the addition of a static type system, which uses
-type inference to synthesize types for user code. These types
-are used primarily in type-driven tooling such as autocomplete
-and API documentation~\cite{luau-autocomplete},
-but creators can also opt in to receiving type error reports.
+type inference to synthesize types for user code on the fly.
+These types are used primarily in type-driven tooling such as autocomplete and
+API documentation~\cite{luau-autocomplete}, but creators can also opt in to
+receiving type error reports.
 
@@
 enabling rapid creation and ensuring software quality.
-{Luau} therefore supports three typing modes---strict types,
-relaxed (nonstrict) types, and no types---and lets users switch
+{Luau} therefore supports three typing modes---\mstrict{} types,
+relaxed (\mnonstrict{}) types, and no types (\mnocheck{})---and lets users switch
 between modes gradually~\cite{st-sfp-2006,tfffgksst-snapl-2017}, one module at a time.
@@
 of the {Luau} type system in the development of {Roblox} scripts.
-In comparison to prior work~(\cref{s:related}), which is either small in scale
-or depends on personally identifiable information~(PII),
+We want to collect measurements at a large scale, with thousands
+of participants, and we must do so in a way that protects creators' privacy.
+In comparison to prior work~(\cref{s:related}), which
+with few exceptions~\cite{zhlbr-cc-2020,zhlbr-oopsla-2020,hlzbr-ecoop-2021} is either small in scale
+or collects personally identifiable information~(PII),
 we performed a large-scale study using pseudonymized \emph{telemetry}.
@@
 
-The telemetry data we analyzed does not contain any PII:
-no source code;
-no source code locations;
+The telemetry data we analyzed does not contain any PII.
+It has no source code;
+ no source locations;
 no error messages (which may contain source code);
+no error messages;
 no record of the creator's identity, locale, or IP address;
@@
 informed by many users' actual practice.
-Our data captures \textbf{XX} sessions
-for a total of \textbf{YY} hours
-and \textbf{ZZ} type errors.
+Our data captures over 340,000 sessions
+that occured between February and April 2023
+and includes 72 million type analysis errors.
 By contrast to typical qualitative methods (e.g., surveys and interviews), it
@@
 
-other Nasko papers:
-https://dblp.org/pid/83/1412.html
+Nasko~\cite{zhlbr-cc-2020,zhlbr-oopsla-2020,hlzbr-ecoop-2021}
 

### 2023-09-17 programming23/paper.tex
example timelines
compress up/down sessions into counts
talk about timestamps, cleaning

@@
 IntelliJ~\cite{intellij-telemetry} report telemetry.
+See~\cref{s:related} for futher comparisons among telemetry designs.
 
@@
 \subsection{Telemetry records}
+\label{s:telemetry-records}
 
@@
 
-In order to avoid swamping our telemtry servers, we randomly throttle
+In order to avoid swamping our telemetry servers, we randomly throttle
 the generation of telemetry records:
@@
 update appropriately on every edit. This can result in very large edit
-ranges (for example if the user edits an the beginning and end of the
-file) but fits in a fixed-size telemetry record.
-
-\subsection{what to study}
-
-RQs + Analysis
-
-\begin{itemize}
-  \item How often does the typechecker give up with CodeTooComplex errors?
-  \subitem [[ For context: the type system is designed to give useful feedback
-    to all sorts of code, whether strict-typed, nonstrict-typed, or \mnocheck{}.
-    But it has an escape hatch: CTC. Hope it's rarely needed. ]]
-    \begin{itemize}
-      \item Count the overall CTC errors (in each mode) for a coarse view of
-        how rare they are. [D9 of course, D8 to compare \mnocheck{} code with
-        strict-mode code via the same typechecker]
-      \subitem Note: We can't look at the code to learn why the errors
-        happened. Can only look for trends and tell the Roblox team where to
-        look.
-        %% \subitem Count CTC errors within each session to see if these errors are intermittent, or if they stick around.
-[D9 for the count, D8 for \mnocheck{} vs strict, D7 because CTC might show up in the edit range]
-    \end{itemize}
-  \item How helpful are highlighted ranges?
-    \begin{itemize}
-      \item How often do edits in a highlighted range remove type errors? How
-        often do they add new ones?
-      \subitem [D7 compare specific old vs specific new]
-      \subitem [may want to filter using D2 + D5]
-      \item How often do edits outside the range remove type errors? How often do they add new ones?
-      \subitem [D6 looking for old count vs. new count difference without a matching delta in the D7 counts]
-      \item Which specific errors seem easy / hard to fix via highlights?
-      \subitem [D6 removed by edit => easy; D6 survive edit => hard]
-    \end{itemize}
-  \item In what ways do types improve quality of life?
-  \begin{itemize}
-    \item Does \mnocheck{} code have many more latent errors than the others?
-    \subitem [D8 latent errors come from the "what if" analysis]
-  \end{itemize}
-\end{itemize}
-
-% What to study?
-% 
-% Sanity checks
-% How large are edit ranges? Plot the distribution.
-% How long between logs? Long times & very short times are problematic.
-% General Roblox trends
-% How many sessions use strict / nonstrict / \mnocheck{}?
-% Expect few strict/nonstrict, lots of \mnocheck{}
-% How often does mode change within a session?
-% Expect few of these b/c error reports are easy to ignore; non-intrusive
-% If a down-switch happens, expect the current #errors to be high
-% If an up-switch happens, hope to stay up
-% How many CodeTooComplex errors arise, overall?
-% Expect very few. Hopefully zero.
-% How often do module switches happen?
-% Expect not-too-many, with some logs in between each time
-% Compare strict vs. nonstrict vs. nocheck
-% Get median # errors across all snapshots.
-% Expect nocheck median to be higher than the others.
-% Plot #errors vs time, or vs codebase size
-% Expect nocheck errors to keep going up. For other modes, #errors should stay low.
-% General type error trends
-% Plot the distribution of #errors across all edit ranges
-% Expect different counts based on error-kind. SEE BELOW
-% Count specific errors that get fixed. For each kind of error, how many times is it old-in-range but not new-in-range? Compare to #old-in-range totals.
-% Expect all errors to have a decent fix rate.
-% If one error has a lower % than others, its error range may be misleading. Or, maybe it's hard to fix
-% Type errors vs. edit range
-% Compare #old vs. #new in range
-% Expect a decrease, #old > #new
-% If there's an increase, look for an explanation. Could be a huge edit range. Could be a 1-line range that we captured 2x in a row. Could be that this user is simply ignoring type errors.
-% Compare [#old-in-range vs. (total #old - total #new)]
-% If #old-in-range = 0, then hope the total # increases
-% Decrease => nonlocal edits fixed an error => type error hilite may be inaccurate
-% If #old-in-range > 0, then hope the total # decreases
-% Increase (in range or otherwise) => made things worse
-
-\section{Predictions}
-%%bg: merge with prev section, on telemetry design?
-
-nocheck forcestrict keep growing
-
-nonstrict te tend to zero, fixing your program should fix the errors, no false positives
-
-nonstrict fs unclear
-
-\section{Data Cleaning}
-
-Deduplicated records with the same timestamp,
-kept only the first.
-Timestamps by millisecond.
-
-Records come with two timestamps, a \emph{client timestamp}
-that the IDE generates before sending the record
-and a \emph{server timestamp} that the telemetry server
-generates when it receives the record.
-We collected timestamps to sort records from earliest to latest.
-The server timestamps, however, are less useful for sorting
-because the server processes groups of records in a batch and
-often assigns equal timestamps to dozens of records.
-Hence, we ignore the server timestamps.
-Throughout the paper a \emph{timestamp} is always a client-side timestamp.
-
-Edit ranges are negative in a handful of records (1,533).
-%% out/negative-edit-range.txt
-%% actual value is NOT negative, but large positive over 4 billion
-%% example: 4294967293
-This may happen when a user deletes a large block of code.
-Since the issue affects so few records, we omit those ranges
-from the analysis.
-
-Sessions, ignore if too close to start or end.
-Deciding if a session is close is tricky because we don't have
-a firm start and end time.
-Um, does this make sense at all?
-Well anyway we took the greater number from mean vs median
-gap (over all points, not just keystroke points)
-and multiplied by Nx to test closeness to a boundary.
-With N=2 we drop 1 nonstrict session, 0 strict (not sure about nocheck).
-With N=4 we drop 3, with N=8 we drop 9.
-
-FILL remove all syntax errors from later stages?
-
-\section{Results}
+ranges, for example, if the user edits at the beginning and end of the
+file.
+The main benefit of this strategy is that it ensures constant-size telemetry records.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%  STOP!  ROUGH TEXT BELOW     %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{The Data}
 \label{s:data}
+
+Roblox Studio collected type error telemetry in Spring 2023,
+between February and April.
+Every Studio session had a small random chance of generating
+telemetry~(\cref{s:telemetry-records}).
+The chosen sessions generated a record whenever the user switched modules and
+randomly on each keystroke.
+
+\Cref{f:records-per-hour} provides a time-ordered distribution of the data.
+Each vertical bar counts the number of records generated per hour according to
+a client-side timestamp.
+The bars are labeled with a California time zone because that is the locale of
+the Roblox servers, which parsed the timestamps.
+Most buckets contain roughly one thousand records, but some have very few
+and a handful of others exceed three thousand.
+%% \FILL{} investigate one tower, what's happening?? all one session?
+There is a sharp drop in early April because we disabled telemetry at that point
+but some clients continued to generate data
+(\FILL{} Q. Alan, is that an ok sentence? I'm not sure why there's a long tail,
+guess it could be people not updating studio or people with strange time settings
+on their machine.)
+We do not know which timezone the records originated in, but since the counts
+tend to peak midday California time it seems that the majority
+of Roblox developers are following a Western Hemisphere schedule.
+%% \FILL{} probably North America, but do we really want to guess?
+The counts for weekends (shaded regions) are often the tallest; these days presumably
+have a larger number of sessions for telemetry to observe.
 
@@
   %% TODO bigger text
+  %% TODO weird artifact (|) in x-min label
   %% code/row-distribution.rkt
@@
 
-FILL investigate one tower
+\subsection{Data Cleaning}
+\label{s:data-cleaning}
+
+A close inspection of the data revealed a few anomalies.
+For the most part, we omit the strange data from futher analysis.
+
+First, some records have odd timestamps according to the server's records.
+Every telemetry record comes with two timestamps: a client-side timestamp
+from when Roblox Studio created it and a server-side timestamp from when
+the server posted it to its database.
+The server timestamps are little use for our analysis because they bunch
+together; it is common for several records to have identical server
+timestamps.
+However, server timestamps are useful for auditing because they should
+be relatively close in value to the client timestamps.
+For some records (<100? \FILL{}), though, the client is off by a large
+margin: up to one week behind and up to one day ahead of the server.
+We attribute the large delays to issues when sending telemetry (lost internet
+connection, etc.) and smaller offsets to incorrect time settings on the client
+machine.
+In any event, we do not filter these records from further analysis.
+
+Second, some records have identical (client) timestamps.
+This can happen legitimately if type analysis runs several times
+in the same millisecond, but it is unlikely that a human would
+enter keystrokes or switch modules quickly enough.
+We keep the first record with a distinct timestamp and discard the others.
+%% TODO wait, better deduplicate by session and not globally!!!
+
+%% out/negative-edit-range.txt
+%% actual value is NOT negative, but large positive over 4 billion
+%% example: 4294967293
+Third, the edit ranges are negative in a few hundred records~(1,533).
+This is likely due to the user deleting a large block of code.
+Since the issue affects so few records, we omit the negative ranges
+from analysis (but, we do use other parts of the records; for example
+the timestamps influence~\cref{f:records-per-hour}).
+(\FILL{} Alan, double-check reasoning)
+
+\subsection{Codebase Size}
+\label{s:codebase-size}
 
@@
 
-\Cref{f:records-per-hour} shows when data arrived across the whole dataset.
-FILL: currently clientTimestamp, which has caveats.
-Unclear what timezone they are from.
-Some records have a client timestamp much older than the server, up to 1 week behind.
-Some are ahead of the server by a day.
-Why?
-Could be wrong local settings on the computer, or a delayed record due to lost
-internet connection or closed laptop.
+TODO findings lessons etc.
 
@@
 Incredible max numbers.
+
+\subsection{??? records}
 
@@
 downgrade = found a down-mode switch that does not match a module switch
+
+wow 619 up-pairs, 583 down-pairs
+=== UP
+min -3 max 57 median 0 mean 2.62 stddev 6.8390518952707335
+=== DOWN
+min -48 max 30 median 0 mean -0.31 stddev 4.498748355271854
 
@@
 
-\begin{figure}[t]
-  \mnocheck{}
-
-  \mnonstrict{}
-
-  \mstrict{}
+\begin{figure}[t]\centering
+  %% TODO can we 
+  \mnonstrict{}\\
+  \includegraphics[width=\columnwidth]{img/timeline-nonstrict.pdf}
+
+  \medskip{}
+  \mstrict{}\\
+  \includegraphics[width=\columnwidth]{img/timeline-strict.pdf}
 
@@
 \end{figure}
-
-TODO individual sessions
 

### 2023-09-17 programming23/paper.tex
codebase size

@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%  STOP!  ROUGH TEXT BELOW     %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \section{The Data}
@@
 randomly on each keystroke.
+In total we collected 1.5 million telemetry records.
 
@@
 tend to peak midday California time it seems that the majority
-of Roblox developers are following a Western Hemisphere schedule.
-%% \FILL{} probably North America, but do we really want to guess?
+of Roblox developers are following a USA schedule (or rather, a Western
+Hemisphere schedule).
 The counts for weekends (shaded regions) are often the tallest; these days presumably
@@
   % ;; percentile* = 0.95 -- 0.99
-  %#hash((editrange . (1156036 926 3007043/817 30975.40553032358 (0.95 8220) (0.96 9858) (0.97 13656) (0.98 18956) (0.99 34725)))
+  % #hash((editrange . (1156036 926 3007043/817 30975.40553032358 (0.95 8220) (0.96 9858) (0.97 13656) (0.98 18956) (0.99 34725)))
   %       (event-count . (6079 138 38689/135 582.5752201972966 (0.95 960) (0.96 1174) (0.97 1304) (0.98 1836) (0.99 3302)))
@@
                  & Files  &              &     Lines &             &    Edit Range & \\
-    Mean [std]   & 11,911 & \stddev{31}  &     6,497 & \stddev{22} &         3,680 & \stddev{12} \\
+    Mean [std]   & 11,911 & \stddev{12}  &     6,497 & \stddev{22} &         3,680 & \stddev{31} \\
     Median       &  7,678 &              &     3,115 &             &           926 & \\
@@
 
-TODO findings lessons etc.
-
-FILL power-law files lines editrange.
-Incredible max numbers.
-
-\subsection{??? records}
+\Cref{f:codebase-size} summarizes the size of projects in the dataset.
+The top half of the figure reports the mean, standard deviation, median,
+and {99}th percentile for the number of files, lines, and lines in edit range.
+The bottom half shows zoomed-in distributions of the file, line, and edit range
+numbers.
+For example, the x-axis of the first plot ranges from 0 to 1,000 files and
+the y-axis counts up to 1,000 telemetry records.
+
+There is a huge amount of variation across projects.
+The largest ones have over 50k files (!) and 50k lines of code, and
+there are many small ones with 1 file and/or 1 line of code.
+Unsurprisingly, the numbers come with large standard deviations.
+The median values are more reasonable, with roughly 3,000 lines of code
+and 1,000 lines in edit ranges, but still have a large number of files: over 7,000.
+The high number of files may be because creators define mavy components to
+build a Roblox experience and put these components in separate files.
+(\FILL{} still that's a lot of components ... any better guesses+)
+
+\subsection{Type Analysis Modes}
 
@@
 
-\Cref{f:dataset-overview}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%  STOP!  ROUGH TEXT BELOW     %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\Cref{f:dataset-overview} breaks down the dataset ...
 

### 2023-09-17 programming23/paper.tex
session size

@@
 \newcommand{\pct}[1]{\SI{#1}{\percent}}
+\newcommand{\percentile}[1]{{#1}th~percentile}
 \newcommand{\modefont}[1]{\texttt{#1}}
@@
     \pct{99}     & 51,761 &              &    50,547 &             &        34,725 & \\
+    %% TODO x-axis starts at 1, not zero ... there are no zeros right??!
     & \includegraphics[width=0.2\columnwidth]{img/files-distribution.pdf}
@@
 The top half of the figure reports the mean, standard deviation, median,
-and {99}th percentile for the number of files, lines, and lines in edit range.
+and \percentile{99} for the number of files, lines, and lines in edit range.
 The bottom half shows zoomed-in distributions of the file, line, and edit range
@@
 
+\subsection{Session Size}
+
+\begin{figure}[t]
+  %% NOTE max time = 1 million seconds = 15 days ...  +100 are over 2 days
+
+  \begin{tabular}{l@{}r@{~}l@{}r@{~}l} \\
+                 & Time Span (sec)  &       & Record Count  \\
+    Mean [std]   &     3,184 & \stddev{16}  &     286 & [583] \\
+    Median       &       845 &              &     138        \\
+    \pct{99}     &    35,450 &              &   3,302        \\
+    %% TODO x-axis starts at 1, not zero ... there are no zeros right??!
+    & \includegraphics[width=0.2\columnwidth]{img/timespan-distribution.pdf}
+    & & \includegraphics[width=0.2\columnwidth]{img/event-count-distribution.pdf}
+
+  \end{tabular}
+
+  \caption{Session size in seconds and in number of records}
+  \label{f:session-size}
+\end{figure}
+
+FILL say more about why this data is interesting. What's the significance?
+
+\Cref{f:session-size} outlines the size of the Roblox Studio sessions in the
+data.
+There are two useful notions of size: real time in seconds (between the first
+and last record) and the total number of records.
+The top of~\cref{f:session-size} reports the mean, standard deviation,
+median, and \percentile{99}; the bottom shows size distributions, zoomed in
+to focus on at most 1,000 units (x-axis) and 1,000 sessions (y-axis).
+
+Similar to the data on project size, there is huge variation in session
+length.
+The longest sessions last several days and/or contain thousands of records,
+while the shortest last a few milliseconds or include just one record.
+Likewise, both plots in the figure follow a power-law distribution
+(especially for record count).
+The median session time is approximately 14 minutes.
+The median number of
+records is 138.
+These numbers are reasonable.
+
 \subsection{Type Analysis Modes}
@@
 
-  \begin{subfigure}[t]{\columnwidth}
-    %% NOTE max time = 1 million seconds = 15 days ...  +100 are over 2 days
-
-    \begin{tabular}{l@{}r@{~}l@{}r@{~}l} \\
-                   & Time Span (sec)  &       & Record Count  \\
-      Mean [std]   &     3,184 & \stddev{16}  &     286 & [583] \\
-      Median       &       845 &              &     138        \\
-      \pct{99}     &    35,450 &              &   3,302        \\
-      & \includegraphics[width=0.2\columnwidth]{img/timespan-distribution.pdf}
-      & & \includegraphics[width=0.2\columnwidth]{img/event-count-distribution.pdf}
-
-    \end{tabular}
-
-    \caption{Session length in seconds and in number of records}
-    \label{f:sessions-size}
-  \end{subfigure}
-
-  \caption{Dataset overview}
+  \caption{FILL Type analysis overview}
   \label{f:dataset-overview}

### 2023-09-17 programming23/paper.tex
overview, size text

@@
 
-Both modes are opt-in. Creators have the option to make \mnonstrict{} mode
-the default rather than \mnocheck{} mode.
+Both modes are opt-in. The default for a new script is \mnocheck{} by
+default, but creators can change this default to \mnonstrict{}.
 
@@
 randomly on each keystroke.
-In total we collected 1.5 million telemetry records.
+In total, we collected 1.5 million telemetry records.
+Roughly two-thirds of records are due to keystrokes.
 
@@
 
-\begin{figure}[t]
+\begin{figure}[t]\centering
   %% TODO bigger text
@@
   \includegraphics[width=\columnwidth]{img/row-distribution.pdf}
+  \begin{tabular}{ll}
+    Reason for sending:
+    &
+    \begin{tabular}[t]{r@{~~}l@{~~}r}
+      996,164 & due to keystrokes  & [\pct{66.20}] \\
+      508,572 & due to module switch & [\pct{33.80}]
+    \end{tabular}
+  \end{tabular}
   \caption{Telemetry records per hour. Each tick on the $x$-axis marks the start of a new day in California. Shaded ranges correspond to weekends.}
@@
 
-\subsection{Codebase Size}
-\label{s:codebase-size}
+\subsection{Overall Size and Shape}
+
+The first step toward understanding the data is to have a sense of what
+is in it.
+Three important characteristics are the size of the codebases in the data,
+the length of the sessions we captured, and the number of type analysis
+errors.
+We discuss these in turn.
+Codebase size and session length follow power-law distributions, with
+many small items and a few very large items.
+The number of type analysis errors is somewhat low~(explained in the next
+section), but sufficient to proceed with futher analysis.
+
+\paragraph{Codebase Size}
+%\label{s:codebase-size}
 
@@
 build a Roblox experience and put these components in separate files.
-(\FILL{} still that's a lot of components ... any better guesses+)
-
-\subsection{Session Size}
+(\FILL{} still that's a lot of components ... any better guesses?)
+
+\paragraph{Session Size}
+%\label{s:session-size}
 
@@
 
-FILL say more about why this data is interesting. What's the significance?
-
 \Cref{f:session-size} outlines the size of the Roblox Studio sessions in the
@@
 (especially for record count).
+A typical session runs on the order of minutes and consists of a few dozen events.
 The median session time is approximately 14 minutes.
@@
 
+\paragraph{Number of Type Analysis Errors}
+%\label{s:count-analysis-errors}
+
+\begin{figure}[t]
+  \begin{tabular}[t]{ll} \\
+    \begin{tabular}[t]{l@{~~}r@{~}l}
+      595,137 & type errors \\
+      289,698 & in current module & [\pct{48.68}] \\
+       30,924 & in edit range & [\pct{5.20}]
+    \end{tabular}
+    \begin{tabular}[t]{l@{~~}r@{~}l}
+      72,235,735 & {forced strict errors} \\
+      37,027,281 & in current module & [\pct{51.26}] \\
+       1,111,178 & in edit range & [\pct{1.54}]
+    \end{tabular}
+  \end{tabular}
+  \caption{Type errors and forced strict errors across all telemetry records}
+  \label{f:count-analysis-errors}
+\end{figure}
+
+\FILL{} need old and current in edit range
+
+\Cref{f:count-analysis-errors} lists the number of type analysis errors
+and breaks them down by location in the codebase.
+In total, there are nearly 600K type errors.
+These appeared in modules that opted in to typechecking with either the
+\mnonstrict{} or \mstrict{} mode.
+The number of forced strict errors is much larger, 72M, because this analysis
+ran on every module using strict type.
+(Furthermore, forced strict runs silently, so creators have no incentive to
+fix these errors.)
+Approximately half the analysis errors occurred in the current module,
+and a small fraction overlapped with the current edit range~(\pct{5} for
+type errors, \pct{1} for forced strict).
+
+The large fraction of errors in the current module is encouraging.
+It suggests that errors appear locally, as the result of edits to nearby
+code (as opposed to edits in one module raising errors in another), and
+that creators fix the errors before switching focus to another
+module.
+The low fraction of errors in the edit range suggests that errors rarely
+persist through nearby edits.
+
 \subsection{Type Analysis Modes}
-
-\begin{figure}[t]
+\label{s:type-analysis-modes}
+
+\begin{figure}[t]\centering
   \begin{subfigure}[t]{\columnwidth}
     \begin{tabular}[t]{ll}
+      %% FILL bar chart? rectangle?
       \begin{tabular}[t]{r@{~~}l@{~}r}
@@
       \begin{tabular}[t]{r@{~~}l@{~~}r}
-           508,572 & due to module switch & [\pct{33.80}] \\
-           996,164 & due to keystrokes    & [\pct{66.20}]
-      \end{tabular}
-    \end{tabular}
-    \caption{Total telemetry records (1,504,736), analysis modes, and reason-for-sending}
+      \end{tabular}
+    \end{tabular}
+    \caption{Analysis modes across the 1,504,736 records}
     \label{f:total-records}
-  \end{subfigure}
-
-  \begin{subfigure}[t]{\columnwidth}
-    \begin{tabular}[t]{ll} \\
-      \begin{tabular}[t]{l@{~~}r@{~}l}
-        595,137 & Type errors \\
-        289,698 & in current module & [\pct{48.68}] \\
-         30,924 & in edit range & [\pct{5.20}]
-      \end{tabular}
-      \begin{tabular}[t]{l@{~~}r@{~}l}
-        72,235,735 & {Forced-strict errors} \\
-        37,027,281 & in current module & [\pct{51.26}] \\
-         1,111,178 & in edit range & [\pct{1.54}]
-      \end{tabular}
-    \end{tabular}
-    \caption{Type errors and forced-strict errors across all telemetry records}
-    \label{f:total-tefs}
   \end{subfigure}
@@
     \end{tabular}
-    \caption{Total sessions (347,598), whether and how they switch analysis mode}
+    \caption{Analysis mode(s) for the 347,598 sessions}
     \label{f:total-sessions}
@@
 
-  \caption{FILL Type analysis overview}
+  \caption{Overview of type analysis modes}
   \label{f:dataset-overview}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%  STOP!  ROUGH TEXT BELOW     %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\Cref{f:dataset-overview} breaks down the dataset ...
-
-FILL nocheck nonstrict strict, order-of-magnitude
-
-FILL type error counts
-Mostly syntax errors though!
-Estimate \pct{40} are substantial type errors rather than syntax
-errors based on the edit range type errors~(\cref{s:type-error-count})
-but unfortunately we cannot get an exact number for the overall
-or per-module type error counts.
-
-FILL single-mode sessions, mostly
-
-FILL details on upgrades, downgrades, then ignore these for rest of paper.
-multi-mode project = found a module switch that coincides with a mode switch, which suggests that two modules have different modes;
-upgrade = found a up-mode switch that does not match a module switch (quite possible to miss these);
-downgrade = found a down-mode switch that does not match a module switch
-
-wow 619 up-pairs, 583 down-pairs
-=== UP
-min -3 max 57 median 0 mean 2.62 stddev 6.8390518952707335
-=== DOWN
-min -48 max 30 median 0 mean -0.31 stddev 4.498748355271854
-
-FILL most sessions fairly long in time (order of minutes),
-short in event count (dozens).
-
-\section{Type Errors in Edit Range}
+Creators have three analysis modes to choose from.
+Furthermore, they can switch between modes at any time.
+According to~\cref{s:type-analysis-modes}, however, usage is
+extremely skewed toward the type-free \mnocheck{} mode.
+Nearly \pct{90} of all records use the \mnocheck{} mode
+and \pct{90} of all sessions use \mnocheck{} exclusively.
+Among the rest, roughly \pct{10} of all records use \mnonstrict{} and only a
+tiny fraction (\pct{0.4}) use \mstrict{}.
+The per-session data is similar: \pct{9} use \mnonstrict{} exclusively
+and \pct{0.2} use \mstrict{}.
+
+These adoption numbers are concerning.
+Creators either do not know about type analysis, or they choose not
+to use it in the code they are currently maintaining.
+Uncovering the reason for low adaption is an important topic for future work on
+Luau that has implications for the gradual typing community~(\cref{s:discussion}).
+
+%% mostly pure, rare mixes
+Most sessions (\pct{99.82}) stick to a single analysis mode.
+These sessions never change the mode in the current module and never switch
+focus to a module with a different analysis mode.
+Among the other sessions, most (\pct{80}) switch to modules with
+a different mode, about half contain at least one edit that upgrades the
+current mode, and about half contain a downgrade.
+There are 619 upgrades in total and 583 downgrades.
+\FILL{} Ben check how often the downgrades follow the upgrades.
+
+\paragraph{Are Upgrades Discouraging?}
+
+%% 619 up-pairs, 583 down-pairs
+%% === UP
+%% min -3 max 57 median 0 mean 2.62 stddev 6.8390518952707335
+%% === DOWN
+%% min -48 max 30 median 0 mean -0.31 stddev 4.498748355271854
+
+A possible explanation for the low adoption of \mnonstrict{} and
+\mstrict{} mode is that upgrading to these modes leads to a large
+number of type errors.
+Creators might get discouraged by a high error count and revert to
+\mnocheck{} mode.
+
+The data does not support this explanation.
+On average, the mode-upgrades in the data resulted in 3 additional
+type errors (stddev 7, median 0).
+The worst-case increase was 57 type errors, which is very high
+but exceptional.
+In another exception, upgrading modes removed three type errors
+(presumably due to the increased precision of \mstrict{} mode
+\FILL{} Alan, sounds plausible?).
+
+Similarly, mode downgrades have only a small negative effect on the
+number of errors (mean -0.3, stddev 4, median 0, max -48).
+%% curious outlier: downgrade added 30 errors!
+Creators apparently tend switch modes only when the code is already
+well typed, or close to it.
+
+\subsection{Type Errors in Edit Range}
 \label{s:type-error-count}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%  STOP!  ROUGH TEXT BELOW     %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+FILL specific type errors, 1st add/drop, 2nd popularity
+
+FILL type error counts
+Mostly syntax errors though!
+Estimate \pct{40} are substantial type errors rather than syntax
+errors based on the edit range type errors~(\cref{s:type-error-count})
+but unfortunately we cannot get an exact number for the overall
+or per-module type error counts.
+
 FILL popularity over all old and new in edit range.
@@
 
+Findings:
+
+\begin{itemize}
+  \item
+    \pct{50} of errors in current module, so probably errors appear locally (not in a distant module);
+    \pct{5} of (current? old??) type errors overlap with edit range, so they don't stick around.
+  \item
+    \pct{90} \mnocheck{},  \pct{10} \mnonstrict{},  \pct{0.1} \mstrict{}, wow.
+  \item
+  \item
+  \item
+    DM awareness is an important and surprising gap between strict
+    and forced strict, undoubtedly a usability barrier for \mstrict{}.
+  \item
+\end{itemize}
+
 FILL regrets, lessons learned, advice for next time
 
+\paragraph{Future work: Low adaption}
+
+why so few using nonstrict and strict?
+needs advertising?
+needs usability?
+
 \paragraph{Do we need timestamps?}
+
+FILL these were helpful for the overview
 

### 2023-09-18 programming23/paper.tex
checkpoint: te edit range

@@
 and includes 72 million type analysis errors.
-By contrast to typical qualitative methods (e.g., surveys and interviews), it
+By contrast to typical qualitative methods such as surveys and interviews, it
 is not restricted to users' \emph{perceptions} about their work and it is not
@@
   \item[D7] specific errors in the edit range: for the previous and current states, for each type error code, the number type errors in the edit range,
+    \FILL{} explain previous/current in detail, explain that this data takes up most of the telemetry record
   \item[D8] forced strict errors: specific errors for the
@@
 %% \FILL{} investigate one tower, what's happening?? all one session?
-There is a sharp drop in early April because we disabled telemetry at that point
-but some clients continued to generate data
-(\FILL{} Q. Alan, is that an ok sentence? I'm not sure why there's a long tail,
-guess it could be people not updating studio or people with strange time settings
-on their machine.)
+There is a sharp drop in early April when the experiment was finished
+and the telemetry probe was removed (though there is a long tail as
+developers uopdated Roblox Studio).
 We do not know which timezone the records originated in, but since the counts
 tend to peak midday California time it seems that the majority
-of Roblox developers are following a USA schedule (or rather, a Western
-Hemisphere schedule).
-The counts for weekends (shaded regions) are often the tallest; these days presumably
-have a larger number of sessions for telemetry to observe.
+of Roblox developers are following a Western Hemisphere schedule.
+The counts for weekends (shaded regions) are often the tallest,
+as Roblox has a significant school-aged creator community.
 
@@
 be relatively close in value to the client timestamps.
-For some records (<100? \FILL{}), though, the client is off by a large
+For a small number of records (<100? \FILL{}), though, the client is off by a large
 margin: up to one week behind and up to one day ahead of the server.
@@
 the timestamps influence~\cref{f:records-per-hour}).
-(\FILL{} Alan, double-check reasoning)
+%% (\FILL{} Alan, double-check reasoning)
+%% Seems fine by me
 
@@
 There is a huge amount of variation across projects.
-The largest ones have over 50k files (!) and 50k lines of code, and
+The largest ones have over 50k files (!) and 50k lines of code, and %50kLoC and 50k files????
 there are many small ones with 1 file and/or 1 line of code.
@@
 and 1,000 lines in edit ranges, but still have a large number of files: over 7,000.
-The high number of files may be because creators define mavy components to
+The high number of files may be because creators define many components to
 build a Roblox experience and put these components in separate files.
 (\FILL{} still that's a lot of components ... any better guesses?)
+
+% That seems way off, not sure we're seeing so many files :/
 
@@
 
-These adoption numbers are concerning.
-Creators either do not know about type analysis, or they choose not
-to use it in the code they are currently maintaining.
-Uncovering the reason for low adaption is an important topic for future work on
-Luau that has implications for the gradual typing community~(\cref{s:discussion}).
+% I think this is too negative!
+%% These adoption numbers are concerning.
+%% Creators either do not know about type analysis, or they choose not
+%% to use it in the code they are currently maintaining.
+%% Uncovering the reason for low adaption is an important topic for future work on
+%% Luau that has implications for the gradual typing community~(\cref{s:discussion}).
+
+These adoption numbers indicate that there is a communication gap;
+as long as type checking is opt-in, most developers stay opted out.
+Once Luau type checking is opt-out rather than opt-in, it will be interesting
+to revisit these numbers.
 
@@
 focus to a module with a different analysis mode.
-Among the other sessions, most (\pct{80}) switch to modules with
+Among the mixed-mode sessions, most (\pct{80}) switch to modules with
 a different mode, about half contain at least one edit that upgrades the
@@
 In another exception, upgrading modes removed three type errors
-(presumably due to the increased precision of \mstrict{} mode
-\FILL{} Alan, sounds plausible?).
+% (presumably due to the increased precision of \mstrict{} mode
+% \FILL{} Alan, sounds plausible?).
+% My guess as to what;s going on
+(possibly due to ther edits being rolled in with the
+mode change).
 
@@
 
-\subsection{Type Errors in Edit Range}
-\label{s:type-error-count}
+\subsection{Type Errors vs. Program Edits}
+\label{s:type-error-survival}
+
+Equipped with an overall picture of the dataset and the skewed usage
+of \mnocheck{}, \mnonstrict{}, and \mstrict{} modes, we are ready to explore
+the type error data.
+Our main focus is on the intersection between type errors
+and edits to the programs.
+When a type error highlights part of the code and the next edit range includes
+that code, we want to know whether the error survives the edit.
+Similarly, we want to know when an edit introduces a type error.
+
+\Cref{t:type-error-survival} counts the number of times that types
+errors overlap with the current edit range.
+There are three sets of columns, corresponding to the \mnocheck{},
+\mnonstrict{}, and \mstrict{} type analysis modes.
+Within each set, the Add columns counts the number of telemetry records
+that introduce a certain type error.
+For example, seventeen of the \mnonstrict{} records add
+at least one \code{CannotCallNonFunction} error to the current edit range.
+The Keep columns counts records that have the same, non-zero number of type errors
+before and after.
+Lastly, the Remove columns count records that decrease the number of type errors.
+
+All three counts are imprecise because our telemetry does not track the exact
+identity of a type error.
+Suppose a creator removes one \code{CannotCallNonFunction} error but then
+immediately introduces another on a different line.
+\Cref{t:type-error-survival} reports this pair of edits as a Keep rather than
+an Add and a Remove because we cannot distinguish two errors of the same kind.
+
+\paragraph{Observations}
+
+%% \paragraph{Most Type Errors are Typos}
+One striking aspect of~\cref{t:type-error-survival} is the large number
+of \code{SyntaxError}s and the related errors \code{UnknownSymbol} and
+\code{UnknownProperty}.
+Most type errors in the data are likely due to typos.
+\Cref{s:type-error-count} discusses this point in detail.
+For now, we remark that \code{SyntaxError} is the only possible error
+in \mnocheck{} mode (by design) and observe that these errors are
+usually in the Drop column and seldom in the Keep column.
+
+A second observation is that there are relatively few errors that overlap
+with the edit range; the numbers are low in general.
+The highest value in the \mstrict{} columns is 71, meaning 71 records
+drop a syntax error.
+This number is very small relative to the total number of \mstrict{} records
+from~\cref{f:dataset-overview} (6,505), and similarly for
+the \mnonstrict{} and \mnocheck{} modes.
+%% FILL show percentages for all three modes?
+Edits seldom overlap with type errors.
+\FILL{} what does this mean? Not sure if bad, or inconclusive.
+
+When edits do overlap, however, ... good?
+
+\paragraph{Do Error Highlights Help?}
+
+\paragraph{Do Types Improve Quality of Life?}
+
+%% use the ... percent of all records??
+%% total 1.3M nocheck, 156K nonstrict, 6K strict
+%% ... uhhh wow so few errors!
 
@@
       & \rbox{Add} & \ybox{Keep} & \gbox{Drop} & \rbox{Add} & \ybox{Keep} & \gbox{Drop} & \rbox{Add} & \ybox{Keep} & \gbox{Drop} \\\midrule
-      TypeMismatch & {0} & {0} & {0} & {159} & {62} & {129} & {21} & {10} & {28} \\
-      UnknownSymbol & {0} & {0} & {0} & {2348} & {496} & {2973} & {49} & {19} & {55} \\
-      UnknownProperty & {0} & {0} & {0} & {347} & {180} & {385} & {19} & {19} & {35} \\
-      NotATable & {0} & {0} & {0} & {7} & {2} & {8} & {2} & {0} & {1} \\
-      CannotExtendTable & {0} & {0} & {0} & {7} & {9} & {10} & {0} & {0} & {1} \\
-      OnlyTablesCanHaveMethods & {0} & {0} & {0} & {1} & {0} & {2} & {0} & {0} & {0} \\
-      DuplicateTypeDefinition & {0} & {0} & {0} & {1} & {0} & {0} & {0} & {0} & {0} \\
-      CountMismatch & {0} & {0} & {0} & {223} & {48} & {269} & {12} & {2} & {17} \\
-      FunctionDoesNotTakeSelf & {0} & {0} & {0} & {4} & {6} & {5} & {0} & {0} & {0} \\
-      OccursCheckFailed & {0} & {0} & {0} & {0} & {0} & {0} & {0} & {0} & {1} \\
-      UnknownRequire & {0} & {0} & {0} & {91} & {39} & {85} & {12} & {3} & {4} \\
-      IncorrectGenericParameterCount & {0} & {0} & {0} & {0} & {0} & {0} & {1} & {0} & {1} \\
-      SyntaxError & {8349} & {314} & {19220} & {983} & {46} & {2658} & {16} & {0} & {71} \\
-      UnknownPropButFoundLikeProp & {0} & {0} & {0} & {26} & {18} & {21} & {1} & {0} & {1} \\
-      GenericError & {0} & {0} & {0} & {206} & {52} & {203} & {8} & {1} & {9} \\
-      CannotCallNonFunction & {0} & {0} & {0} & {17} & {4} & {20} & {1} & {0} & {1} \\
-      ExtraInformation & {0} & {0} & {0} & {48} & {7} & {39} & {2} & {0} & {1} \\
-      ModuleHasCyclicDependency & {0} & {0} & {0} & {20} & {9} & {15} & {0} & {0} & {1} \\
-      IllegalRequire & {0} & {0} & {0} & {16} & {3} & {33} & {0} & {0} & {2} \\
-      FunctionExitsWithoutReturning & {0} & {0} & {0} & {9} & {3} & {6} & {11} & {5} & {8} \\
-      CannotInferBinaryOperation & {0} & {0} & {0} & {1} & {2} & {1} & {7} & {6} & {9} \\
-      MissingProperties & {0} & {0} & {0} & {11} & {3} & {11} & {6} & {4} & {6} \\
-      OptionalValueAccess & {0} & {0} & {0} & {40} & {52} & {27} & {6} & {2} & {7} \\
-      MissingUnionProperty & {0} & {0} & {0} & {1} & {0} & {2} & {0} & {0} & {0} \\
-      TypesAreUnrelated & {0} & {0} & {0} & {1} & {0} & {0} & {0} & {0} & {1} \\
+      \code{CannotCallNonFunction} & {0} & {0} & {0} & {17} & {4} & {20} & {1} & {0} & {1} \\
+      \code{CannotExtendTable} & {0} & {0} & {0} & {7} & {9} & {10} & {0} & {0} & {1} \\
+      \code{CannotInferBinaryOperation} & {0} & {0} & {0} & {1} & {2} & {1} & {7} & {6} & {9} \\
+      \code{CountMismatch} & {0} & {0} & {0} & {223} & {48} & {269} & {12} & {2} & {17} \\
+      \code{DuplicateTypeDefinition} & {0} & {0} & {0} & {1} & {0} & {0} & {0} & {0} & {0} \\
+      \code{ExtraInformation} & {0} & {0} & {0} & {48} & {7} & {39} & {2} & {0} & {1} \\
+      \code{FunctionDoesNotTakeSelf} & {0} & {0} & {0} & {4} & {6} & {5} & {0} & {0} & {0} \\
+      \code{FunctionExitsWithoutReturning} & {0} & {0} & {0} & {9} & {3} & {6} & {11} & {5} & {8} \\
+      \code{GenericError} & {0} & {0} & {0} & {206} & {52} & {203} & {8} & {1} & {9} \\
+      \code{IllegalRequire} & {0} & {0} & {0} & {16} & {3} & {33} & {0} & {0} & {2} \\
+      \code{IncorrectGenericParameterCount} & {0} & {0} & {0} & {0} & {0} & {0} & {1} & {0} & {1} \\
+      \code{MissingProperties} & {0} & {0} & {0} & {11} & {3} & {11} & {6} & {4} & {6} \\
+      \code{MissingUnionProperty} & {0} & {0} & {0} & {1} & {0} & {2} & {0} & {0} & {0} \\
+      \code{ModuleHasCyclicDependency} & {0} & {0} & {0} & {20} & {9} & {15} & {0} & {0} & {1} \\
+      \code{NotATable} & {0} & {0} & {0} & {7} & {2} & {8} & {2} & {0} & {1} \\
+      \code{OccursCheckFailed} & {0} & {0} & {0} & {0} & {0} & {0} & {0} & {0} & {1} \\
+      \code{OnlyTablesCanHaveMethods} & {0} & {0} & {0} & {1} & {0} & {2} & {0} & {0} & {0} \\
+      \code{OptionalValueAccess} & {0} & {0} & {0} & {40} & {52} & {27} & {6} & {2} & {7} \\
+      \code{SyntaxError} & {8349} & {314} & {19220} & {983} & {46} & {2658} & {16} & {0} & {71} \\
+      \code{TypeMismatch} & {0} & {0} & {0} & {159} & {62} & {129} & {21} & {10} & {28} \\
+      \code{TypesAreUnrelated} & {0} & {0} & {0} & {1} & {0} & {0} & {0} & {0} & {1} \\
+      \code{UnknownPropButFoundLikeProp} & {0} & {0} & {0} & {26} & {18} & {21} & {1} & {0} & {1} \\
+      \code{UnknownProperty} & {0} & {0} & {0} & {347} & {180} & {385} & {19} & {19} & {35} \\
+      \code{UnknownRequire} & {0} & {0} & {0} & {91} & {39} & {85} & {12} & {3} & {4} \\
+      \code{UnknownSymbol} & {0} & {0} & {0} & {2348} & {496} & {2973} & {49} & {19} & {55}
     \end{tabular}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%  STOP!  ROUGH TEXT BELOW     %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-FILL specific type errors, 1st add/drop, 2nd popularity
-
-FILL type error counts
+\subsection{Type Error Popularity}
+\label{s:type-error-count}
+
+FILL type error popularity
 Mostly syntax errors though!
@@
 
-FILL popularity over all old and new in edit range.
-Syntax error by far the most common.
-UnknownSymbol is likely a syntax error --- so \pct{90} syntax error!
-
 FILL N errors never showed up.
@@
 
-TODO get counts for each mode. Are there fewer stx errors in strict mode??!
-
-NOTE can't ask about type errors at end because we don't know
-the end.
-The final record can be sent minutes before the end of session.
-
 \begin{table}[t]\centering
   %% NOTE: popularity looks similar when [over all records] vs. [restricted to module switch records]
-
+  \caption{Type error popularity for \mnonstrict{} and \mstrict{} modes.
+    All of the type errors in \mnocheck{} mode are \code{SyntaxError}s.}
+  \label{t:type-error-count}
   \begin{tabular}{ll}
@@
   \end{tabular}
-
-  \caption{Type error popularity for \mnonstrict{} and \mstrict{} modes.
-  In \mnocheck{}, \code{SyntaxError} accounts for \pct{100} of the error.}
-  \label{t:type-error-count}
 \end{table}
 
-\section{Analysis}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%  STOP!  ROUGH TEXT BELOW     %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2023-09-18 programming23/paper.tex
Minor edits

--- 2023-09-17 programming23/paper.tex
@@
 (\FILL{} still that's a lot of components ... any better guesses?)
+
+% That seems way off, not sure we're seeing so many files :/
 

### 2023-09-19 programming23/paper.tex
quick half-draft report on specific type errors

--- 2023-09-18 programming23/paper.tex
@@
 
-Equipped with an overall picture of the dataset and the skewed usage
-of \mnocheck{}, \mnonstrict{}, and \mstrict{} modes, we are ready to explore
-the type error data.
-Our main focus is on the intersection between type errors
-and edits to the programs.
-When a type error highlights part of the code and the next edit range includes
-that code, we want to know whether the error survives the edit.
-Similarly, we want to know when an edit introduces a type error.
-
-\Cref{t:type-error-survival} counts the number of times that types
-errors overlap with the current edit range.
-There are three sets of columns, corresponding to the \mnocheck{},
-\mnonstrict{}, and \mstrict{} type analysis modes.
-Within each set, the Add columns counts the number of telemetry records
-that introduce a certain type error.
-For example, seventeen of the \mnonstrict{} records add
-at least one \code{CannotCallNonFunction} error to the current edit range.
-The Keep columns counts records that have the same, non-zero number of type errors
-before and after.
-Lastly, the Remove columns count records that decrease the number of type errors.
-
-All three counts are imprecise because our telemetry does not track the exact
-identity of a type error.
-Suppose a creator removes one \code{CannotCallNonFunction} error but then
-immediately introduces another on a different line.
-\Cref{t:type-error-survival} reports this pair of edits as a Keep rather than
-an Add and a Remove because we cannot distinguish two errors of the same kind.
-
-\paragraph{Observations}
-
-%% \paragraph{Most Type Errors are Typos}
-One striking aspect of~\cref{t:type-error-survival} is the large number
-of \code{SyntaxError}s and the related errors \code{UnknownSymbol} and
-\code{UnknownProperty}.
-Most type errors in the data are likely due to typos.
-\Cref{s:type-error-count} discusses this point in detail.
-For now, we remark that \code{SyntaxError} is the only possible error
-in \mnocheck{} mode (by design) and observe that these errors are
-usually in the Drop column and seldom in the Keep column.
-
-A second observation is that there are relatively few errors that overlap
-with the edit range; the numbers are low in general.
-The highest value in the \mstrict{} columns is 71, meaning 71 records
-drop a syntax error.
-This number is very small relative to the total number of \mstrict{} records
-from~\cref{f:dataset-overview} (6,505), and similarly for
-the \mnonstrict{} and \mnocheck{} modes.
-%% FILL show percentages for all three modes?
-Edits seldom overlap with type errors.
-\FILL{} what does this mean? Not sure if bad, or inconclusive.
-
-When edits do overlap, however, ... good?
-
-\paragraph{Do Error Highlights Help?}
-
-\paragraph{Do Types Improve Quality of Life?}
-
-%% use the ... percent of all records??
-%% total 1.3M nocheck, 156K nonstrict, 6K strict
-%% ... uhhh wow so few errors!
-
 \begin{table}[t]
-  \caption{Type error survival}
+  \caption{Number of records that add, keep, or drop (remove) specific type errors from their edit range.}
   \label{t:type-error-survival}
-
   \begin{tabular}{lr@{}r@{}rr@{}r@{}rr@{}r@{}r}
@@
 
+Equipped with an overall picture of the dataset and the skewed usage
+of \mnocheck{}, \mnonstrict{}, and \mstrict{} modes, we are ready to explore
+the type error data.
+Our main focus is on the intersection between type errors
+and edits to the programs.
+When a type error highlights part of the code and the next edit range includes
+that code, we want to know whether the error survives the edit.
+Similarly, we want to know when an edit introduces a type error.
+
+\Cref{t:type-error-survival} counts the number of times that types
+errors overlap with the current edit range.
+There are three sets of columns, corresponding to the \mnocheck{},
+\mnonstrict{}, and \mstrict{} type analysis modes.
+Within each set, the Add columns counts the number of telemetry records
+that introduce a certain type error.
+For example, seventeen of the \mnonstrict{} records add
+at least one \code{CannotCallNonFunction} error to the current edit range.
+The Keep columns counts records that have the same, non-zero number of type errors
+before and after.
+Lastly, the Remove columns count records that decrease the number of type errors.
+
+All three counts are imprecise because our telemetry does not track the exact
+identity of a type error.
+Suppose a creator removes one \code{CannotCallNonFunction} error but then
+immediately introduces another on a different line.
+\Cref{t:type-error-survival} reports this pair of edits as a Keep rather than
+an Add and a Remove because we cannot distinguish two errors of the same kind.
+
+\paragraph{Observations}
+
+%% \paragraph{Most Type Errors are Typos}
+\begin{itemize}
+  \item
+    One striking aspect of~\cref{t:type-error-survival} is the large number
+    of \code{SyntaxError}s and the related errors \code{UnknownSymbol} and
+    \code{UnknownProperty}.
+    Most type errors in the data are likely due to typos.
+    \Cref{s:type-error-count} discusses this point in detail.
+    For now, we remark that \code{SyntaxError} is the only possible error
+    in \mnocheck{} mode (by design) and observe that these errors are
+    usually in the Drop column and seldom in the Keep column.
+
+  \item
+    The numbers are relatively low in general, which means that
+    few errors in the dataset overlap with the edit range
+    The highest value in the \mstrict{} columns is 71, meaning 71 records
+    out of 6,505 total~(\cref{f:dataset-overview} drop a syntax error.
+    The \mnonstrict{} and \mnocheck{} columns are similar.
+    %% FILL show percentages for all three modes?
+    \FILL{} what does this mean? Not sure if bad, or inconclusive.
+
+  \item
+    %% Drop >= Add >= Rem ~= 27 triples
+    %% other nonzero      ~= 13 triples
+    When type errors overlap with the edit range,
+    we would hope the number of Adds and Drops is roughly
+    equal and the number of Keeps is low.
+    This is indeed the case.
+    In most Add/Keep/Drop triples, the Drop column has the highest
+    value and the Keep column has the lowest value.
+    The optimistic conclusion is that creators are fixing the code
+    highlighted by errors.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%  STOP!  ROUGH TEXT BELOW     %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+  \item
+    After the syntax errors, \code{CountMismatch},
+    \code{TypeMismatch}, and \code{GenericError} are the most common
+    \mnonstrict{} and \mstrict{} errors.
+    \FILL{} so what.
+
+  \item
+    \FILL{} which errors break the Drop > Add > Keep hope?
+
+  \item
+    (NOTE may want to save relative popularity for next section.)
+    Some errors occur more often in \mnonstrict{} than \mstrict{}:
+    \code{CannotCallNonFunction},
+    \code{CannotExtendTable},
+    \code{ExtraInformation},
+    \code{FunctionDoesNotTakeSelf},
+    \code{IllegalRequire},
+    \code{ModuleHasCyclicDependency},
+    \code{OptionalValueAccess},
+    \code{UnknownPropButFoundLikeProp}.
+    \FILL{} is this surprising, or expected from the bigger number of
+    \mnonstrict{} records?
+
+  \item
+    A few errors are more common in \mstrict{} than \mnonstrict{}:
+    \code{CannotInferBinaryOperation},
+    \code{FunctionExitsWithoutReturning},
+    \code{IncorrectGenericParameterCount}.
+    \FILL{} so what.
+
+\end{itemize}
+
+\paragraph{Do Error Highlights Help?}
+
+\FILL{} data seems to say yes.
+
+\paragraph{Do Types Improve Quality of Life?}
+
+\FILL{} compare \mnonstrict{} and \mstrict{}.
+Can't compare to \mnocheck{} with this table, though, so maybe it make sense to
+put this question later.
+
 \subsection{Type Error Popularity}
 \label{s:type-error-count}
-
-FILL type error popularity
-Mostly syntax errors though!
-Estimate \pct{40} are substantial type errors rather than syntax
-errors based on the edit range type errors~(\cref{s:type-error-count})
-but unfortunately we cannot get an exact number for the overall
-or per-module type error counts.
-
-FILL N errors never showed up.
-Nothing for \code{CodeTooComplex} or \code{UnificationTooComplex} is great news.
-But what about the others? FILL
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%  STOP!  ROUGH TEXT BELOW     %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+On a coarser level, \cref{t:type-error-count} reports the percent of
+type errors that overlap with the edit range.
+For each error type that ever overlaps, the table
+shows what percentage it takes relative to all overlaps.
+
+The table does not report \mnocheck{} because all its type errors are
+by definition \code{SyntaxError}s.
+
+\paragraph{Observations}
+
+\begin{itemize}
+  \item
+    The most popular errors are syntax errors.
+    This is undoubtedly due to the stochiastic nature
+    of our telemetry, which has no way to avoid
+    reporting edits that happen in the middle of things.
+    For example, a creator who writes a method call
+    \code{bucket.countFish()} letter-by-letter will
+    get type analysis to find errors like \code{UnknownSymbol(buck)}
+    and \code{UnknownProperty(bucket.c)}.
+    These are transient issues, but telemetry may catch them.
+
+  \item
+    \FILL{} relative mismatches between NS and S?
+
+  \item
+    \FILL{} which errors that NEVER showed up?
+
+FILL N errors never showed up.
+Nothing for \code{CodeTooComplex} or \code{UnificationTooComplex} is great news.
+But what about the others? FILL
+
+\end{itemize}
+
+\paragraph{Internal Limits, Code Too Complex}
+%% data = out/ctc-info.txt
+
+To deal with pathologies such as the worst-case time for ML-style type
+inference~\cite{m-popl-1990,ktu-caap-1990}, the typechecker
+has internal limits that restrict the problems it will attempt to solve.
+Hitting a limit triggers one of the following errors:
+\code{CodeTooComplex},
+\code{NormalizationTooComplex}, or
+\code{UnificationTooComplex}.
+A casual user should never see these errors.
+
+Fortunately, the data rarely contains too-complex errors.
+There are zero \code{UnificationTooComplex} errors and only a
+handful of \code{CodeTooComplex} errors.
+%% 26 total
+It is entirely possible that the \code{CodeTooComplex} errors came from
+a single user, as they appear in eleven records (out of 1.5 million)
+spread across three sessions.
+At any rate, the overall number is low enough to suggest that the internal
+limits are not posing barriers to typechecker adoption.
 
@@
 Have zero, yay!
-
-\subsection{Dead Ends: Too Complex}
-%% data = out/ctc-info.txt
-
-To deal with pathologies such as the worst-case time for ML-style type
-inference~\cite{m-popl-1990,ktu-caap-1990}, the typechecker
-has internal limits that restrict the problems it will attempt to solve.
-Hitting a limit triggers one of the following errors:
-\code{CodeTooComplex},
-\code{NormalizationTooComplex}, or
-\code{UnificationTooComplex}.
-A casual user should never see these errors.
-
-Fortunately, the data rarely contains too-complex errors.
-There are zero \code{UnificationTooComplex} errors and only a
-handful of \code{CodeTooComplex} errors.
-%% 26 total
-It is entirely possible that the \code{CodeTooComplex} errors came from
-a single user, as they appear in eleven records (out of 1.5 million)
-spread across three sessions.
-At any rate, the overall number is low enough to suggest that the internal
-limits are not posing barriers to typechecker adoption.
 

### 2023-09-20 programming23/paper.tex
resolve a few comments
type error popularity notes

@@
 the timestamps influence~\cref{f:records-per-hour}).
-%% (\FILL{} Alan, double-check reasoning)
-%% Seems fine by me
 
@@
 
+(\FILL{} doublecheck file sizes)
 % That seems way off, not sure we're seeing so many files :/
@@
 
-% I think this is too negative!
-%% These adoption numbers are concerning.
-%% Creators either do not know about type analysis, or they choose not
-%% to use it in the code they are currently maintaining.
-%% Uncovering the reason for low adaption is an important topic for future work on
-%% Luau that has implications for the gradual typing community~(\cref{s:discussion}).
-
 These adoption numbers indicate that there is a communication gap;
@@
 In another exception, upgrading modes removed three type errors
-% (presumably due to the increased precision of \mstrict{} mode
-% \FILL{} Alan, sounds plausible?).
-% My guess as to what;s going on
-(possibly due to ther edits being rolled in with the
+(possibly due to other edits being rolled in with the
 mode change).
@@
 
-On a coarser level, \cref{t:type-error-count} reports the percent of
+\Cref{t:type-error-count} reports the percent of
 type errors that overlap with the edit range.
-For each error type that ever overlaps, the table
-shows what percentage it takes relative to all overlaps.
-
+This table is based on the same data as the previous table~(\cref{t:type-error-survival}),
+but focuses on the overall popularity of type errors
+rather than their relation to edits.
+The left column in each row names one kind of type error
+that overlaps with an edit region.
+The right column shows what percentage of all overlapping
+errors the current error took up.
+
+There are two subtables, for \mnonstrict{} and \mstrict{} mode.
 The table does not report \mnocheck{} because all its type errors are
@@
   \item
-    The most popular errors are syntax errors.
+    As we noted above, the most popular errors are syntax errors
+    (\pct{FILL} \mnonstrict{}, \pct{FILL} \mstrict{})..
     This is undoubtedly due to the stochiastic nature
@@
   \item
-    \FILL{} relative mismatches between NS and S?
+    \code{TypeMismatch} is far more common in \mstrict{} (\pct{20.46})
+    than \mnonstrict{} (\pct{2.44}) because \mstrict{} mode
+    has additional type constraints.
+    Several other errors have higher \mstrict{} percentages for the
+    same reason, for example, \code{CannotInferBinaryOperation}.
+
+  \item
+    Two errors that appear in the \mstrict{} records
+    never appear in \mnonstrict{}: \code{OccursCheckFailed} and \code{IncorrectGenericParameterCount}.
+    We attribute these to more advanced uses of types in \mstrict{} mode.
+
+  \item
+    Four errors from the \mnonstrict{} records never appeared in \mstrict{} mode:
+    \code{DuplicateTypeDefinition}, \code{OnlyTablesCanHaveMethods},
+    \code{MissingUnionProperty}, and \code{FunctionDoesNotTakeSelf}.
+    (\FILL{} Alan is this interesting, or just a result of having less strict data?)
 
@@
     \FILL{} which errors that NEVER showed up?
-
-FILL N errors never showed up.
     Nothing for \code{CodeTooComplex} or \code{UnificationTooComplex} is great news.
@@
 Fortunately, the data rarely contains too-complex errors.
-There are zero \code{UnificationTooComplex} errors and only a
-handful of \code{CodeTooComplex} errors.
-%% 26 total
-It is entirely possible that the \code{CodeTooComplex} errors came from
-a single user, as they appear in eleven records (out of 1.5 million)
-spread across three sessions.
-At any rate, the overall number is low enough to suggest that the internal
-limits are not posing barriers to typechecker adoption.
-
-\subsection{Analysis 2}
-
-TODO work this into the above or below.
-
-What can we learn from the telemetry? We have 1.5 million records from Feb 22
-to April 14 (roughly 27K records per day).
-They mention +70 million type errors.
-(Some records point to a large number of errors. FILL median, max. Could be cascading,
-which makes the 70million number uninterpretable.)
-What takeaways are there??!
-
-First,
-users rarely switch type analysis mode. There are 3 modes available and users
-are free to switch between them, or to mix modes in a codebase, but we rarely
-detect these events:
-
-\pct{99} of sessions use a single mode
-
-\pct{0.2} of sessions use different modes in different modules. (More
-precisely, the user switched to a module --- triggering a telemetry record ---
-that used a different mode than the previous one.)
-
-\pct{0.2} of records show a mode switch (equally split between upgrades and
-downgrades, 341 and 320)
+None of these errors appear in edit regions~(\cref{t:type-error-count}).
+For the specific case of \code{CodeTooComplex}, our telemetry tracked project-wide
+counts and found only 26 of these errors, which were spread across eleven
+records in three sessions.
+It is entirely possible that all these errors came from one creator who
+deliberately pushed the limits of type analysis.
+Whatever the reason, the overall number is low enough to suggest that the
+internal limits are not posing barriers to typechecker adoption.
+
+\subsection{NEXT}
 

### 2023-09-20 programming23/paper.tex
outline rest of Data section

@@
   \caption{Selected Error Labels}
+  % https://github.com/Roblox/luau/blob/master/Analysis/include/Luau/Error.h
   %% from type analysis, but they're not all really type errors
@@
 well typed, or close to it.
+
+%% \paragraph{Module Switches vs Errors}
+%% FILL need \%s out of all module switches.
+%% How many module switches exit a module with errors?
+%% > OLD DATA
+%% > \mnocheck{} 24, \mnonstrict{} 9, \mstrict{} 3.
+%% > Rare?
+%% > More common for forcedstrict:
+%% >  \mnocheck{} 190, \mnonstrict{} 16, \mstrict{} 2.
+%% >
+%% > How many module switches go to a module with errors?
+%% > \mnocheck{} 70, nonstrict 9, strict 4
+%% > For FS: \mnocheck{} 400, \mnonstrict{} 39, \mstrict{} 6.
 
@@
 
-\paragraph{Do Types Improve Quality of Life?}
-
-\FILL{} compare \mnonstrict{} and \mstrict{}.
-Can't compare to \mnocheck{} with this table, though, so maybe it make sense to
-put this question later.
-
 \subsection{Type Error Popularity}
@@
     As we noted above, the most popular errors are syntax errors
-    (\pct{FILL} \mnonstrict{}, \pct{FILL} \mstrict{})..
+    (\FILL{} \pct{0} \mnonstrict{}, \pct{0} \mstrict{})..
     This is undoubtedly due to the stochiastic nature
@@
     Nothing for \code{CodeTooComplex} or \code{UnificationTooComplex} is great news.
-    But what about the others? FILL
+    But what about the others?
+    \code{DynamicPropertyLookupOnClassesUnsafe}?
+    \code{SwappedGenericTypeParameter}?
 
@@
 
-\subsection{NEXT}
-
-Second,
-the stronger analysis modes are roughly 10x less popular than weaker ones
-(across the \pct{99} of sessions that stick with one mode):
-
-\pct{90} of sessions use nocheck mode
-\pct{9.8} use nonstrict mode
-\pct{0.2} use strict mode
-
-Third,
-In the edit range, we know exactly which errors showed up. This leads to a few
-takeaways, such as:
-
-Only 3 sessions (out of +300K) hit the typechecker's internal limits on problem
-size --- CodeTooComplex errors. Good!
-
-About \pct{90} of type errors are probably due to typos (SyntaxError, UnknownSymbol,
-UnknownProperty). These are not propoer type errors.
-
-10 errors never showed up. These include DynamicPropertyLookupOnClassesUnsafe and
-SwappedGenericTypeParameter
-
-Fourth,
-Codebase size and session length tend to follow a power law. 
-One aspect of size for nonstrict sessions: the number of
-lines in the edit range. Over 80 sessions have 20-line ranges, about 40 have a
-30-line range, etc.
+\subsection{Forced Strict Errors}
+
+\FILL{} I (ben) think we need to talk more about forced strict errors here, or soon.
+Otherwise people will forget they exist in the data.
+(But if they're not useful to us, remove them!)
+
+\subsection{Type Error Density}
+
+\begin{figure}[t]\centering
+  % TODO try diff vs first record instead of previous?
+
+  \mnocheck{}
+  \includegraphics[width=\columnwidth]{img/error-count-nocheck-row--te-density-diff.pdf}
+  \medskip
+  \mnonstrict{}
+  \includegraphics[width=\columnwidth]{img/error-count-nonstrict-row--te-density-diff.pdf}
+  \medskip
+  \mstrict{}
+  \includegraphics[width=\columnwidth]{img/error-count-strict-row--te-density-diff.pdf}
+
+  % TODO use bars instead, percent only, don't care about counts
+  \begin{tabular}{lr@{}rr@{}rr@{}r}
+    & \zerowidth{Add} & & \zerowidth{Keep} & & \zerowidth{Drop} \\\midrule
+    \mnocheck{} & 48378 & [\pct{46.83}] & 9440 & [\pct{9.14}] & 45479 & [\pct{44.03}] \\
+    \mnonstrict{} & 19491 & [\pct{39.63}] & 9567 & [\pct{19.45}] & 20121 & [\pct{40.91}] \\
+    \mstrict{} & 733 & [\pct{39.18}] & 368 & [\pct{19.67}] & 770 & [\pct{41.15}] \\
+  \end{tabular}
+  \caption{FILL error density}
+  \label{f:error-delta}
+\end{figure}
 
@@
 
+\subsection{Data Model}
+
 \begin{figure}[t]\centering
-% TODO diff vs 1st, not previous
-
-  \mnocheck{}
-  \includegraphics[width=\columnwidth]{img/error-count-nocheck-row--te-density-diff.pdf}
-  \medskip
-  \mnonstrict{}
-  \includegraphics[width=\columnwidth]{img/error-count-nonstrict-row--te-density-diff.pdf}
-  \medskip
-  \mstrict{}
-  \includegraphics[width=\columnwidth]{img/error-count-strict-row--te-density-diff.pdf}
-
-  TODO bars instead, percent only, don't care about counts
-
-  \begin{tabular}{lr@{}rr@{}rr@{}r}
-    & \zerowidth{Add} & & \zerowidth{Keep} & & \zerowidth{Drop} \\\midrule
-    \mnocheck{} & 48378 & [\pct{46.83}] & 9440 & [\pct{9.14}] & 45479 & [\pct{44.03}] \\
-    \mnonstrict{} & 19491 & [\pct{39.63}] & 9567 & [\pct{19.45}] & 20121 & [\pct{40.91}] \\
-    \mstrict{} & 733 & [\pct{39.18}] & 368 & [\pct{19.67}] & 770 & [\pct{41.15}] \\
-  \end{tabular}
-  \caption{FILL error density}
-  \label{f:error-delta}
+  %% TODO
+  %% - thin bars
+  %% - colors for good , bad (disagree) , neutral
+  \begin{tabular}{lll}
+    \mnocheck{} & \mnonstrict{} & \mstrict{} \\
+    \includegraphics[width=0.3\columnwidth]{img/compass-nocheck.pdf}
+    &
+    \includegraphics[width=0.3\columnwidth]{img/compass-nonstrict.pdf}
+    &
+    \includegraphics[width=0.3\columnwidth]{img/compass-strict.pdf}
+  \end{tabular}
+  \caption{FILL Are there fewer type errors than forcedstrict errors? Expect true always, but that is not the case.}
+  \label{f:tefs}
 \end{figure}
+
+% -\subsection{Asset Creation vs Scripting}
+% -
+% -Data model. Point to discussion above.
+% -
+% -Big fluctuations in errors.
+% -FS disagrees with TE.
+% -Probably data model!
 
@@
 
-\begin{figure}[t]\centering
-  %% TODO
-  %% - thin bars
-  %% - colors for good , bad (disagree) , neutral
-  \begin{tabular}{lll}
-    \mnocheck{} & \mnonstrict{} & \mstrict{} \\
-    \includegraphics[width=0.3\columnwidth]{img/compass-nocheck.pdf}
-    &
-    \includegraphics[width=0.3\columnwidth]{img/compass-nonstrict.pdf}
-    &
-    \includegraphics[width=0.3\columnwidth]{img/compass-strict.pdf}
-  \end{tabular}
-  \caption{FILL Are there fewer type errors than forcedstrict errors? Expect true always, but that is not the case.}
-  \label{f:tefs}
-\end{figure}
-
 \Cref{f:tefs}~FILL story is edits to DM vs edits to code.
 Data model awareness, call for future work.
+
+\subsection{Timeline}
 
@@
 
-\section{Interpretation}
-
-\subsection{Internal Error}
-%% out/te-editrange-*
+\section{Discussion}
+\label{s:discussion}
+
+\FILL{} good things first,
+then regrets about the noisy stochiastic telemetry,
+then lessons?
+then threats.
+
+\paragraph{Good: No Internal Error}
 
@@
 
-\subsection{TBD}
-
-Error 1000 = type mismatch (subtyping etc);
-1001 = unknown symbol;
-\ldots
-
-% https://github.com/Roblox/luau/blob/master/Analysis/include/Luau/Error.h
+\paragraph{Do Types Improve Quality of Life?}
+
+\FILL{} compare \mnonstrict{} and \mstrict{}.
+Use edit range~(\cref{t:type-error-survival} for NS vs S)
+and error density~(\cref{f:error-delta})
+
+TODO nocheck accumulated fs errors?
+
+\paragraph{Bad: Noisy Data}
 
@@
 
-\subsection{Module Switches}
-
-FILL why study, what implications?
-
-FILL need \%s out of all module switches.
-
-How many module switches exit a module with errors?
-\mnocheck{} 24, \mnonstrict{} 9, \mstrict{} 3.
-Rare?
-More common for forcedstrict:
- \mnocheck{} 190, \mnonstrict{} 16, \mstrict{} 2.
-
-How many module switches go to a module with errors?
-\mnocheck{} 70, nonstrict 9, strict 4
-For FS: \mnocheck{} 400, \mnonstrict{} 39, \mstrict{} 6.
-
-\subsection{Types and Quality of Life}
-
-FILL error density: number vs range size
-
-TODO nocheck accumulated errors?
-
-\subsection{Example Sessions}
-
-FILL example sessions.
-
-\subsection{Asset Creation vs Scripting}
-
-Data model. Point to discussion above.
-
-Big fluctuations in errors.
-FS disagrees with TE.
-Probably data model!
-
-\section{Threats to Validity}
+\subsection{Threats to Validity}
 \label{s:threats}
@@
   \begin{tabular}{l@{~}cccccc}
-    &             & PII       & Session ID & Deterministic & Private Data \\\midrule
+    &             & PII       & Session ID & Deterministic & Private Data (???) \\\midrule
     & Transparent & \chkNo    & \chkNo     & \chkYes       & \chkNo      \\
   * & Us          & \chkNo    & \chkYes    & \chkNo        & \chkYes     \\
+    & Diff.Priv   & $\epsilon$ & \chkYes   & \chkYes       & \chkYes     \\
+    & Liblit      & ?         & ?          & ?             & ?           \\
     & \code{.NET} & \chkMaybe & \chkYes    & \chkYes       & \chkYes     \\
@@
 
-\section{Discussion}
+\section{Conclusion}
 \label{s:conclusion}
-\label{s:discussion}
-
-Findings:
+
+Quickly summarize findings:
 
@@
 
-FILL regrets, lessons learned, advice for next time
+Quick regrets, lessons learned, advice for next time
 
@@
 
-Thanks to Benjamin Chung for feedback on plots and data analysis advice.
+Thanks to Benjamin Chung for several helpful discussions about data analysis
+and effective plotting.
 Greenman was supported by

### 2023-09-20 programming23/paper.tex
more text for more figures, 2 left for a first cut

@@
   \includegraphics[width=\columnwidth]{img/error-count-strict-row--te-density-diff.pdf}
-
+  \caption{Changes over time (seconds) to type error density (total errors / lines in project)}
+  \label{f:error-density}
+\end{figure}
+
+\FILL{} what does the FS plot look like? is that a better nocheck-vs-rest comparison??
+
+Moving on to type errors across the projects, and not just in the edit range,
+a first interesting question is to study how the overall number of errors
+changes over time.
+But, the sheer number of errors can be skewed by codebase size.
+To give a better basis for project-to-project comparisons, we focus
+on the density of errors rather than the count, where \emph{density} is
+the total number of type errors divided by the lines of code in the project.
+
+\Cref{f:error-density} presents a zoomed-in view of type error density as it
+changes over time in every session of the dataset.
+The x-axis counts time in seconds from the start of the session.
+The y-axis shows the change in density from the previous telemetry record
+in the session to the next.
+Changes can be positive or negative depending on whether the edits added
+or removed errors.
+The plots cut off at \pm{}50 density and 150 seconds.to focus on the
+bulk of the data.
+%% Sadly even with the zoom, it's too hard to follow one session ... the plot only gives a big-picture impression.
+
+\paragraph{Observations}
+
+Due to the relative adoption of type analysis, there is much more
+data for the weak \mnocheck{} mode than for \mnonstrict{},
+and in turn for \mnonstrict{} relative to \mstrict{}.
+Still, the plots reveal interesting points:
+
+\begin{itemize}
+  \item
+    All three plots are roughly balanced,
+    with equal ``mass'' above and below the x-axis.
+    The next section explores this point in detail.
+
+  \item
+    After the 40-second mark, changes to type error density become much
+    smaller.
+    Why?
+    There are also fewer sessions (\mstrict{} makes that very clear).
+
+  \item
+    \mnonstrict{} has the biggest number of up/down deltas in the 10+ range.
+\end{itemize}
+
+\subsection{Changes to Errors by Mode}
+
+\begin{figure}[t]\centering
   % TODO use bars instead, percent only, don't care about counts
@@
   \end{tabular}
-  \caption{FILL error density}
-  \label{f:error-delta}
+  \caption{How often do edits increase, maintain, or decrease the total number of type errors?}
+  \label{f:error-changes}
 \end{figure}
+
+\FILL{}
 
@@
 \end{figure}
+
+\FILL{}
 
@@
 
-\subsection{Timeline}
+\subsection{Session Timelines}
 
 \begin{figure}[t]\centering
-  %% TODO can we 
-  \mnonstrict{}\\
-  \includegraphics[width=\columnwidth]{img/timeline-nonstrict.pdf}
+  %% TODO any way to cluster timelines to avoid the "creepiness" of spying on one user?
+%  \mnonstrict{}\\
+%  \includegraphics[width=\columnwidth]{img/timeline-nonstrict.pdf}
 
@@
 
-TODO correlations, compare modes!!!
+Another insightful visualization that the telemetry supports is plots of individual sessions
+over time.
+\Cref{f:indy-session} shows the type errors, forced strict errors, and module switches
+in two \mstrict{} mode sessions.
+The x-axis shows time in seconds from the start of the session.
+The y-axis counts errors.
+Type errors appear as orange plus marks.
+Forced strict errors appear as green x marks.
+The vertical lines show module switches.
+
+In the first plot, the error counts are low until about the 3,000s mark,
+at which point an edit (not a module switch) leads to 15 errors.
+These 15 errors disappear two records later.
+Then, after two module switches and some edits, the number of errors spikes again.
+In the second plot, the number of errors spikes in the beginning due to edits,
+then remains low for the next few thousand seconds.
+
+A general observation is that the number of type errors is always less than the number of forced strict
+errors, though the numbers are close.
+This should happen in \mstrict{} mode.
 

### 2023-09-20 programming23/paper.tex
math mode
draft rest of figure prose

@@
 \newcommand{\rbox}[1]{\colorbox{red!30}{#1\vphantom{Ap}}}
+\newcommand{\tefsfont}[1]{\textsc{#1}}
+\newcommand{\tekey}{\tefsfont{te}}
+\newcommand{\fskey}{\tefsfont{fs}}
 
@@
 constraints such as forces, springs and joints.
-The data model can be seen in the Explorer view of Fig.~\ref{fig:roblox-studio}.
+The data model can be seen in the Explorer view of~\Cref{fig:roblox-studio}.
 While an experience is under development, it is typical for the data
@@
 or removed errors.
-The plots cut off at \pm{}50 density and 150 seconds.to focus on the
+The plots cut off at$\pm{}50$ density and 150 seconds.to focus on the
 bulk of the data.
@@
 
-\subsection{Changes to Errors by Mode}
+\subsection{Do Edits Tend to Add Type Errors?}
 
@@
   % TODO use bars instead, percent only, don't care about counts
+  %% TODO old version had +-20% add/drop and a bigger % of keeps. What changed?
   \begin{tabular}{lr@{}rr@{}rr@{}r}
@@
   \end{tabular}
-  \caption{How often do edits increase, maintain, or decrease the total number of type errors?}
+  \caption{How often do edits increase, maintain, or decrease the number of type errors?}
   \label{f:error-changes}
@@
 
-\FILL{}
-
-Fifth,
-Increases and decreases to the number of type errors are evenly balanced across
-the data.
-In strict mode, \pct{24} of records increase the number of type errors. \pct{26}
-decrease. \pct{50} keep it constant.
-In nonstrict mode, \pct{14} increase errors, \pct{15} decrease, and \pct{71}
-keep it constant
-This plot is for nonstrict. X-axis = time since start of session, cut off at 30
-seconds. Y-axis = change in num errors since last record, cut off at +/- 100.
-
-\subsection{Data Model}
+The apparent visual symmetry in the density plots~(\cref{f:error-density})
+suggests that edits add and remove errors with rougly equal frequency,
+regardless of mode.
+No matter whether a creator is using \mnocheck{}, \mnonstrict{},
+or \mstrict{} mode, edits seem to balance increases and decreases in the number
+of type errors.
+\Cref{f:error-changes} explores this question of balance in detail.
+For each of the three modes, it reports the percent of edits that increase (Add),
+maintain at a nonzero level (Keep), or decrease (Drop) the number of type
+errors in the project.
+
+\paragraph{Observations}
+
+\begin{itemize}
+  \item
+    The Adds and Drops are remarkably balanced: across all three modes, they are
+    at most \pct{2} apart from one another.
+    This confirms that changes to the number of errors are balanced.
+    Creators tend to fix errors highlighted by type analysis
+    even in the stricter analysis modes.
+
+  \item
+    The percentage of Keeps is somewhat high: nearly \pct{10} in \mnocheck{}
+    and \pct{20} in \mnonstrict{} and \mstrict{}.
+    Evidently, the mode plays a role here.
+    Outside of \mnocheck{}, type errors are more likely to persist through edits.
+
+\end{itemize}
+
+\subsection{Two Activities: Asset Creation and Scripting}
 
@@
   \caption{FILL Are there fewer type errors than forcedstrict errors? Expect true always, but that is not the case.}
-  \label{f:tefs}
+  \label{f:tefs-compass}
 \end{figure}
 
-\FILL{}
-
-% -\subsection{Asset Creation vs Scripting}
-% -
-% -Data model. Point to discussion above.
-% -
-% -Big fluctuations in errors.
-% -FS disagrees with TE.
-% -Probably data model!
-
-Sixth, DM story.
-FS has: 1 all dependencies in strict, 2 data-model awareness (DM turns into
-type graph, ... initial DM, all properties go into a table type; DM -> folder
--> model -> part .. get dot-driven autocomplete) in strict, DM might have type
-Any might have type Instance, the latter would mean more errors than FS
-
-DM insensitivity in strict mode
-(if DM has top type that would lead to errors.)
-Context: each of the 3 modes runs 2 kinds of type analysis. One is the regular
-type analysis. The other is Forced Strict (FS), which converts all dependencies
-to strict mode. The reason for FS is that it sometimes gives better
-autocomplete suggestions etc.
-You'd expect that the number of type errors is always <= num FS errors. But we
-found several cases where this is false!
-The plot here "stacks" every strict session into bars. X-axis is a record index
-within a session (bars are higher on the left because we have lots more short
-sessions than long ones). Y-axis counts the number of records where num TE <= num FS
-is true (blue) and false (yellow).
-
-\Cref{f:tefs}~FILL story is edits to DM vs edits to code.
-Data model awareness, call for future work.
+Since Roblox Studio runs each codebase through two passes of type analysis,
+it is interesting to compare their results.
+Of special interest is the \mstrict{} mode case, because the number of
+type errors and forced strict errors should be nearly the same.
+But, there are two differences between \mstrict{} type checking
+and forced strict checking:
+\begin{itemize}
+  \item
+    Forced strict checks all dependencies in \mstrict{} mode, no matter
+    what mode they declare.
+    So, the number of forced strict errors may be higher.
+  \item
+    Forced strict has looser rules for the \emph{data model};
+    that is, for the static assets that a program manipulates.
+    So, the number of forced strict errors may be lower.
+    %% initial DM, all properties go into a table type; DM -> folder -> model -> part
+    %% get dot-driven autocomplete
+\end{itemize}
+Our data contains a surprising number of records where forced strict
+reports fewer errors.
+Data model edits are apparently a regular occurrence in Luau programs.
+
+\Cref{f:tefs-compass} compares changes in type errors (\tekey{}) to changes in forced
+strict errors (\fskey{}) for the three analysis modes.
+To a first approximation, the changes either agree with one another
+or disagree.
+For example, if type errors increase but forced strict errors decrease,
+we have a disagreement.
+The plots futher categorize agreements and disagreements into
+the following five cases:
+\begin{enumerate}
+  \item
+    \emph{zero}: neither \tekey{} nor \fskey{} changed
+  \item
+    \emph{te only}: \tekey{} changed but \fskey{} remained the same
+  \item 
+    \emph{fs only}: \tekey{} remained the same but \fskey{} changed
+  \item
+    \emph{dis}: \tekey{} and \fskey{} changed in opposite directions
+  \item 
+    \emph{agree}: \tekey{} and \fskey{} changed in the same direction
+\end{enumerate}
+Both \emph{zero} and \emph{agree} are agreements.
+The others are disagreements.
+The plots show what percent of all telemetry records fell into these five categories.
+
+\paragraph{Observations}
+
+\begin{itemize}
+  \item
+    All three plots have a nonzero percentage of \emph{te only} items.
+    In \mstrict{} mode, the percentage jumps up to nearly \pct{25}.
+    Since \fskey{} is more lenient only for data model changes, these must
+    be edits that affect the model in a way that raises a type error.
+    Naturally, \mstrict{} mode has the largest number of such edits because
+    its type checker is the most careful.
+
+  \item
+    In the first plot, for \mnocheck{},
+    half the records are \emph{zero\/}s, nearly half are \emph{fs only}, and only a few
+    land in the other categories.
+    While the number of zeros is a bit high,
+    the \emph{fs. only} spike makes sense because type checking is supposed to complain
+    sparingly in \mnocheck{} mode.
+
+  \item
+    In the \mnonstrict{} plot, the number of agreements jumps up.
+    This is reassuring.
+    \FILL{}.
+
+\end{itemize}
+
+Users of \mstrict{} mode may find it frustrating that the type checker complains
+about data model edits.
+Though we cannot see the exact reason for the complaints, the checker errs on
+the conservative side.
+Thus we expect the number of useful reports is low.
+This may be a barrier to adaption.
 

### 2023-09-21 programming23/paper.tex
stochastic
sk edits: less marketing, clarify Luau vs types vs nocheck distinction

@@
 
-\title{Type Error Telemetry at Scale: Pseudonymized and Stochiastic}
+\title{Type Error Telemetry at Scale: Pseudonymized and Stochastic}
 % lean, lite, modest, non-intrusive, private, ...
@@
   must support a wide audience.
-  As part of its efforts to support all kinds programmers, Luau includes a
-  gradual type system and goes to great lengths to minimize false positive
-  errors.
+  As part of its efforts to support all kinds programmers, Luau includes an
+  optional, gradual type system and goes to great lengths to minimize false
+  positive errors.
 
@@
   %% addressed by others (if at all)?
-  Since Luau is in the hands of many creators, we want to collect their feedback
-  and ultimately improve the language design (in particular, the type system).
-  We want feedback at a large scale, as creators work on their actual projects,
-  which suggests the deployment of client-side telemetry.
-  %% ^ awkward
-  However, we must respect creators privacy.
+  Since Luau is currently being used by many creators, we want to collect their feedback
+  to improve the language and, in particular, the type system.
+  The standard way to collect feedback at a large scale from creators working on
+  realistic projects is to deploy client-side telemetry; however, we must do
+  so in a way that respects creators' privacy.
   Whatever telemetry we use to study the Luau language cannot include
@@
   Finding ways to harness telemetry in a private way is the central research challenge.
+  %% TODO awkward
 
@@
   %% What was done that unveiled new knowledge?
-  We designed and implemented a privacy-preserving telemetry system for Luau.
-  Telemetry records include a timestamp, a (randomized) session id, a reason
-  for sending, and a sequence of counter values summarizing the results of type
-  analysis.
+  We designed and implemented a pseudonymized telemetry system for Luau.
+  Telemetry records include a timestamp, a session id, a reason for sending,
+  and a numeric summary of the most recent type analysis.
+  This information lets us follow type analysis over time within a codebase
+  without revealing private information.
   Over four months in Spring 2023, we deployed the telemetry system and collected
   1.5 million telemetry records.
-  We analyzed the data to improve Luau.
 
@@
   %% capabilities are enabled by the work?
-  We present several findings about Luau:
-  some confirm hypotheses by the Luau team, others are rather surprising, but
-  all findings support our claim that privacy-preserving telemetry is a useful
-  way to collect feedback.
-  One expected finding is that types are unpopular; there is an order-of-maginitude
-  difference in the number of sessions that use different type analysis modes.
-  One surprise is that the strict type analysis mode evidently makes it difficult
-  to connect a script with data assets.
-  Lastly, one reassuring finding is that type analysis rarely hits its internal
+  We present several findings about Luau.
+  Some confirm hypotheses by the Luau team while others are rather surprising,
+  but all findings support our claim that pseudonymized telemetry is a
+  useful way to collect feedback.
+  One of the less-surprising findings is that types are unpopular:
+  there is an 100x gap between the number of untyped Luau sessions and the number
+  of typed ones.
+  One surprise is that the strict mode for type analysis is very conservative
+  about interactions with data assets.
+  Instead of using the flexible dynamic type, it uses a top type and requires
+  creators to down-cast assets.
+  Lastly, a reassuring finding is that type analysis rarely hits its internal
   limits on problem size.
@@
 65~million Daily Active Users, and 14~billion hours of engagement in
-April--June 2023; there are 3~million creators using {Roblox Studio},
-creating 5~million experiences~\cite{corp.roblox.com}.
-
-{Roblox experiences} are scripted using the 
+April--June 2023.
+There are over 5~million experiences available on the platform thanks to a
+worldwide community of 3~million creators~\cite{corp.roblox.com}.
+
+Creators build experiences using the
 {Luau} programming language~\cite{luau-lang.org},
 an extension of {Lua~5.1~\cite{lua}}.
-The main extension is the addition of a static type system, which uses
-type inference to synthesize types for user code on the fly.
+The main addition in Luau is a static type system, which infers
+types for all Luau programs on the fly, as creators modify the code.
 These types are used primarily in type-driven tooling such as autocomplete and
-API documentation~\cite{luau-autocomplete}, but creators can also opt in to
-receiving type error reports.
+API documentation~\cite{luau-autocomplete}; the typical creator is unaware
+that a type system powers these tools.
+However, creators may write their own types, and they can opt in
+to receiving ype error reports.
 
@@
 enabling rapid creation and ensuring software quality.
-{Luau} therefore supports three typing modes---\mstrict{} types,
-relaxed (\mnonstrict{}) types, and no types (\mnocheck{})---and lets users switch
-between modes gradually~\cite{st-sfp-2006,tfffgksst-snapl-2017}, one module at a time.
+{Luau} therefore supports three typing modes for user code:
+the hands-off default \mnocheck{},
+a relaxed \mnonstrict{} mode,
+and a rigorous \mstrict{} mode.
+Creators can switch between modes
+gradually~\cite{st-sfp-2006,tfffgksst-snapl-2017}, one module at a time.
 
@@
 Most users of {Roblox Studio} do not opt in to type error
-reporting, and so they do not see the ``squiggly underlining'' that
+reporting (\pct{90},~\cref{s:data}), and so they do not see the ``squiggly underlining'' that
 indicates a type error site. Nonetheless, the type inference system
@@
 
-Analysis pipeline to be freely available.
-Data may be available, depends on Roblox.
+The data and the analysis pipeline are freely available.
+\FILL{} link.
 
@@
 
-FILL table = object
-
-\begin{figure}
-
+\begin{figure}[t]\centering
   \includegraphics[width=.45\textwidth]{img/roblox-studio.png}
@@
 
-Creators of {Roblox experiences} use {Roblox Studio},
-which combines {3D creation} tools as well as an Integrated
-Developer Environment (IDE), as seen in Fig.~\ref{fig:roblox-studio}.
+%% \FILL{} explain the name "creators"?
+
+{Roblox Studio} is a workbench that
+combines {3D creation} tools and an Integrated
+Developer Environment (IDE), as seen in \Cref{fig:roblox-studio}.
 The IDE includes an optional ``Script Analysis'' widget, which
@@
 constraints such as forces, springs and joints.
-The data model can be seen in the Explorer view of~\Cref{fig:roblox-studio}.
+%% The data model can be seen in the Explorer view of~\Cref{fig:roblox-studio}.
+%% ben: where is it? very hard to see
 While an experience is under development, it is typical for the data
@@
 the type system, it is possible for these edits to introduce type errors.
+
+%% FILL table = object
 
@@
     \code{CodeTooComplex} & Type analysis failed, cannot understand the code\!\!\! \\
-    \code{UnificationTooComplex} & Type analysis failed, unification solver hit a limit\!\!\! \\
+    %% \code{UnificationTooComplex} & Type analysis failed, unification solver hit a limit\!\!\! \\
     \code{SyntaxError} & Basic parse error, e.g., \code{for if end} \\
@@
 \end{itemize}
-Each telemetry record contains:
-\begin{description}
-  \item[D1] metadata: a psedonymized session identifier, timestamp and session duration,
-  \item[D2] mode of the current file: (\mnocheck{}, \mnonstrict{}, or \mstrict{}),
-  \item[D3] reason for sending: (random selection, or file switch),
-  \item[D4] size of codebase: number of files, and number of lines of code type checked,
-  \item[D5] size of edit range
-  \item[D6] overall type errors: for previous and current states, we have the total number, the nunber in the current file, and number in the edit range,
-  \item[D7] specific errors in the edit range: for the previous and current states, for each type error code, the number type errors in the edit range,
+Each telemetry record contains nine kinds of data:
+record 
+\begin{enumerate}
+  \item metadata: a psedonymized session identifier, timestamp and session duration,
+  \item mode of the current file: (\mnocheck{}, \mnonstrict{}, or \mstrict{}),
+  \item reason for sending: (random selection, or file switch),
+  \item size of codebase: number of files, and number of lines of code type checked,
+  \item size of edit range
+  \item overall type errors: for previous and current states, we have the total number, the nunber in the current file, and number in the edit range,
+  \item specific errors in the edit range: for the previous and current states, for each type error code, the number type errors in the edit range,
     \FILL{} explain previous/current in detail, explain that this data takes up most of the telemetry record
-  \item[D8] forced strict errors: specific errors for the
+  \item forced strict errors: specific errors for the
     edit range computed as if the user had strict mode enabled.
-  \item[D9] TooComplexErrors: project-wide count for this one specific error type
-\end{description}
+  \item TooComplexErrors: project-wide count for this one specific error type
+\end{enumerate}
 To track the edit range, we record a start and end position, which we
@@
 
-\begin{figure}[t]
+\begin{figure}[t]\centering
   %% NOTE max time = 1 million seconds = 15 days ...  +100 are over 2 days

### 2023-09-22 programming23/paper.tex
Minor edits

@@
 Each telemetry record contains nine kinds of data:
-record 
 \begin{enumerate}
@@
 There is a huge amount of variation across projects.
-The largest ones have over 50k files (!) and 50k lines of code, and %50kLoC and 50k files????
+The largest ones have over 50,000 files, and
 there are many small ones with 1 file and/or 1 line of code.
@@
 and 1,000 lines in edit ranges, but still have a large number of files: over 7,000.
-The high number of files may be because creators define many components to
-build a Roblox experience and put these components in separate files.
-(\FILL{} still that's a lot of components ... any better guesses?)
-
-(\FILL{} doublecheck file sizes)
+One source of variation is that some developers use Luau scripts as a way
+to store data, writing a module that exports one huge array, typically
+defined on a single line. Such data storage scripts provide a challenge for
+script tooling.
+
+%% The high number of files may be because creators define many components to
+%% build a Roblox experience and put these components in separate files.
+%% (\FILL{} still that's a lot of components ... any better guesses?)
+
+%% (\FILL{} doublecheck file sizes)
 % That seems way off, not sure we're seeing so many files :/
@@
 records is 138.
-These numbers are reasonable.
+% These numbers are reasonable.
 

### 2023-09-22 programming23/paper.tex
delete stochastic, use 'QALAN' instead of 'FILL Alan'

@@
 
-\title{Type Error Telemetry at Scale: Pseudonymized and Stochastic}
+\title{Type Error Telemetry at Scale}
 % lean, lite, modest, non-intrusive, private, ...
@@
 \newcommand{\fskey}{\tefsfont{fs}}
+\newcommand{\QALAN}{\FILL{} Alan:}
 
@@
 {Roblox Studio} has a telemetry system, which is used to gauge
-the effectiveness of creation features. This system stochastically
+the effectiveness of creation features. This system randomly
 determines which sessions should report telemetry, and for those
@@
   \item
-    \FILL{} which errors break the Drop > Add > Keep hope?
+    \FILL{} which errors break the Drop > Add > Keep ``preferred order''?
 
@@
     \code{UnknownPropButFoundLikeProp}.
-    \FILL{} is this surprising, or expected from the bigger number of
+    \QALAN{} are any of these surprising, or expected from the bigger number of
     \mnonstrict{} records?
@@
     \code{MissingUnionProperty}, and \code{FunctionDoesNotTakeSelf}.
-    (\FILL{} Alan is this interesting, or just a result of having less strict data?)
+    (\QALAN{} is this interesting, or just a result of having less strict data?)
 
@@
 Otherwise people will forget they exist in the data.
-(But if they're not useful to us, remove them!)
 

### 2023-09-22 programming23/paper.tex
comments and a bit of fixing

@@
 %% use 'submission' for initial submission, remove it for camera-ready (see 5.1)
+
+\usepackage{todonotes}
 
@@
 Each vertical bar counts the number of records generated per hour according to
-a client-side timestamp.
+a client-side timestamp.\todo{Do we need all this detail?}
 The bars are labeled with a California time zone because that is the locale of
@@
 A close inspection of the data revealed a few anomalies.
-For the most part, we omit the strange data from futher analysis.
+For the most part, we omit the strange data from futher
+analysis.\todo{I'm inclined to rewrite this to be much more concise.}
 
@@
 machine.
-In any event, we do not filter these records from further analysis.
+In any event, we do not filter these records from further
+analysis.\todo{To be clear, you do NOT filter? Because the section is
+  about what you DO filter.}
 
@@
 
-The first step toward understanding the data is to have a sense of what
-is in it.
-Three important characteristics are the size of the codebases in the data,
-the length of the sessions we captured, and the number of type analysis
+Three important characteristics of the recorded data are the size of the codebases,
+the length of the sessions, and the number of type
 errors.
 We discuss these in turn.
-Codebase size and session length follow power-law distributions, with
+Codebase size and session length follow power-law
+distributions,\todo{Please confirm w/ a statistical test. See Clauset
+  et al.: arXiv:0706.1062.} with
 many small items and a few very large items.
@@
 
-  \caption{Size of analyzed code: number of files, number of lines , and lines in edit range}
+  \caption{Size of analyzed code: number of files, number of lines, and lines in edit range}
   \label{f:codebase-size}
@@
 The top half of the figure reports the mean, standard deviation, median,
-and \percentile{99} for the number of files, lines, and lines in edit range.
+and \percentile{99} for the number of files,\todo{There are
+  1000s of files?!? More files than lines? Makes no sense.} lines, and lines in edit range.
 The bottom half shows zoomed-in distributions of the file, line, and edit range
@@
 there are many small ones with 1 file and/or 1 line of code.
-Unsurprisingly, the numbers come with large standard deviations.
+Unsurprisingly, the numbers come with large standard
+deviations.\todo{Which is not shown!}
 The median values are more reasonable, with roughly 3,000 lines of code
@@
 defined on a single line. Such data storage scripts provide a challenge for
-script tooling.
+script tooling.\todo{No idea what's said here.}
 
@@
 median, and \percentile{99}; the bottom shows size distributions, zoomed in
-to focus on at most 1,000 units (x-axis) and 1,000 sessions (y-axis).
+to focus on at most 1,000 units (x-axis) and 1,000 sessions
+(y-axis).\todo{Again, confused.}
 
@@
   \begin{tabular}[t]{ll} \\
-    \begin{tabular}[t]{l@{~~}r@{~}l}
+    \begin{tabular}[t]{r@{~~}r@{~}l}
       595,137 & type errors \\
@@
     \end{tabular}
-    \begin{tabular}[t]{l@{~~}r@{~}l}
+    \begin{tabular}[t]{r@{~~}r@{~}l}
       72,235,735 & {forced strict errors} \\
@@
 ran on every module using strict type.
-(Furthermore, forced strict runs silently, so creators have no incentive to
+(Furthermore, forced strict runs silently, so creators have no
+awareness of or incentive to
 fix these errors.)
@@
     \begin{tabular}[t]{ll} \\
-      \begin{tabular}[t]{l@{~~}r@{~}l}
+      \begin{tabular}[t]{r@{~~}r@{~}r}
         313,509 & \mnocheck{}   & [\pct{90.19}] \\
@@
 Among the mixed-mode sessions, most (\pct{80}) switch to modules with
-a different mode, about half contain at least one edit that upgrades the
+a different mode, about half contain at least one edit that upgrades
+the typing strictness in the
 current mode, and about half contain a downgrade.
@@
 number of type errors.
-Creators might get discouraged by a high error count and revert to
+Creators might get discouraged or overwhelmed by a high error count and revert to
 \mnocheck{} mode.
@@
 Creators apparently tend switch modes only when the code is already
-well typed, or close to it.
+well typed, or close to it.\todo{What incentive is there to do so?}
 
@@
 Our main focus is on the intersection between type errors
-and edits to the programs.
+and edits to the programs.\todo{The next two sentences are in some
+  sense our main RQs, right? I would call them out more explicitly and
+  w/ more justification.}
 When a type error highlights part of the code and the next edit range includes
@@
     in \mnocheck{} mode (by design) and observe that these errors are
-    usually in the Drop column and seldom in the Keep column.
+    usually in the Drop column and seldom in the Keep column.\todo{So?}
 
@@
 \code{UnificationTooComplex}.
-A casual user should never see these errors.
+A casual user should never see these errors.\todo{IDEALLY never see?
+  They CAN see them, right?}
 

### 2023-09-22 programming23/paper.tex
related work: telemetry, privacy ... more coming

@@
 
-Liblit dissertation~\cite{liblit-thesis}
+\begin{table}[t]\centering
+  \caption{Comparing telemetry systems}
+  \label{t:telemetry-design}
+
+  \FILL{}
+
+  \begin{tabular}{l@{~}cccccc}
+    &               & Code      & Session ID & Timestamp     & Deterministic (???) & Private Data (???) \\\midrule
+    & Transparent   & \chkNo    & \chkNo     & \chkNo        & \chkYes       & \chkNo      \\
+  * & Roblox Studio & \chkNo    & \chkYes    & \chkNo        & \chkNo        & \chkYes     \\
+%% orthogonal %%    & Diff.Priv     & $\epsilon$ & \chkYes   & \chkYes       & \chkYes       & \chkYes     \\
+%% orthogonal %%    & CBI           & \chkMaybe & \chkMaybe  & \chkNo        & \chkNo        & ?           \\
+    & Classic       & \chkYes   & \chkYes    & \chkYes       & \chkYes       & \chkYes     \\
+  \end{tabular}
+\end{table}
+
+% kindle track every tap (page turn etc),
+% enables whispersync feature,
+% drove design of navigation tools,
+% opt-out possible~\cite{kindle-telemetry}
+
+Type error telemetry in Roblox Studio has two distinguishing characteristics:
+it includes no personal information (PII), and it sends data on
+randomly-selected runs of the type checker rather than specific events of interest.
+To the best of our knowledge, this design occupies a unique position relative to prior
+work~(\cref{t:telemetry-design}).
+
+In other professional-grade IDEs, such as VSCode~\cite{vscode-telemetry}
+and \code{.NET}~\cite{dotnet-telemetry}, telemetry may include all sorts of
+data: error reports, source code fragments, timestamps, and filepaths.
+Users may be able to opt out of some telemetry, but the details depend on the
+license agreement.
+Furthermore, at least in VSCode, IDE extensions can report their own telemetry.
+While this data can be invaluable for discovering bugs in production, it
+must be handled with extreme care.
+
+The BlackBox project similarly collects telemetry in broad strokes.
+Every Java program submitted for compilation goes into the BlackBox database (minus
+any comments above the main class) along with the compiler's response~\cite{bkmu-sigcse-2014}.
+Though somewhat intrusive, the millions of programs in the BlackBox dataset
+have enabled dozens of contributions to the area of CS
+education~\cite{bask-icer-2018}, ranging from categorization of novice
+errors~\cite{mk-fie-2014,m-masters-2016} to analyses of error frequency and
+time-to-fix~\cite{ab-sigcse-2015}.
+
+At the opposite end of the privacy spectrum, Go's design for \emph{transparent
+telemetry} reports only counter values~\cite{transparent-telemetry}.
+Unlike Roblox Studio, transparent telemetry includes no timestamps and no 
+session ids.
+The data is useful for learning about the frequency of specific events and the
+distribution of events across clients (e.g., how many \code{CodeTooComplex}
+errors arise? what is the average number of type errors?).
+However, the lack of timestamps and ids means that it cannot track edits
+over time (a feature that is quite useful to us, for example, in~\cref{f:error-density}).
+% In~\cref{f:error-density}, for example, we see that increases to the number
+% of type errors are quickly followed by decreases.
+% Without timestamps, we would be unable to rule out the possibility that users
+% build up a mountain of type errors and then slowly reduce them to zero.
+
+One way to futher strengthen the privacy of any telemetry design, including
+transparent telemetry, is to fuzz the data using techniques from
+differential privacy~\cite{zhlbr-cc-2020,epk-ccs-2014,wblj-usenix-2017}.
+Recent work on fuzzing traces~\cite{zhlbr-cc-2020},
+accounting for known relationships between events~\cite{zhlbr-oopsla-2020},
+and performing the fuzzing on the client side~\cite{hlzbr-ecoop-2021} is
+especially relevant to the session-based data that Roblox Studio collects.
+Applying differential privacy is subtle, however.
+While we can likely fuzz the total number of type errors, if we were to
+fuzz the number of \code{CodeTooComplex} errors carelessly, it might drastically
+affect our conclusions.
+
+Lastly, cooperative bug isolation is a related method for designing
+telemetry systems~\cite{liblit-thesis}.
+%% \FILL{} adaptive bug isolation~\cite{nl-icse-2010}, search on the control dependence graph.
+%% 1\% overhead vs 87\% for sampling and 300\% for binary instrumentation
+The focus is not on privacy, but rather on how to collect a small amount of
+data from each user and perform statistically-sound analyses.
+Each feature of interest \emph{within} a user's codebase has a uniform-random
+probability of contributing telemetry.
+The system-builder can analyze this telemetry with predicates to identify
+notable events and logistic regression to narrow down nondeterministic bugs.
+There is a high burden on designers to decide what to capture,
+but a careful design can minimize the data from each individual user.
+%% more in diss: how to deal with native compilers, threads, dynamic loading
+%% ... privacy is future work, can maybe avoid private data with a type system
+
+\paragraph{Telemetry Rates}
+
+We chose randomly, uniform distribution.
+
+\cite{lnsmc-usenix-2018} blame-proportional logging,
+scale up ``near'' points of interest,
+could be useful for us in the future.
+\cite{fnm-sigmod-2020} ditto, faster than statistical at finding root-cause,
+deterministic.
+
+\paragraph{TBD}
 
 RAMSS workshop series.
-
-Differential privacy for coverage analysis of software traces,
-adds noise to traces,
-hard because infinitely many traces,
-use CFG abstraction of possible behaviors for callgraph analysis,
-count sketch data structure~\cite{hlzbr-ecoop-2021}.
-
-Nasko~\cite{zhlbr-cc-2020,zhlbr-oopsla-2020,hlzbr-ecoop-2021}
 
@@
 
-\cite{bgimgm-cse-2016}
-study Java compiler error messages after an intervention, counting num errors,
-counting repeats, counting per user
-
-\cite{anna-russo-kennedy-ms-2006}
-BitFit, web-based java ide log num hint requests num, result of check answer
-requests num compile, num run, no code?! good we have the same constraint
-
-\cite{ab-sigcse-2015}
-frequency, time-to-fix, spread of errors
- (we can't do time)
-parsing compile error + source code
- sometimes with custom parsers
-
-\cite{m-masters-2016}
-study errors in blackbox dataset,
- overwhelmingly syntax
-
-\cite{bkmu-sigcse-2014}
-blackbox, 100k users
-source code, diffs
-error line, column, message;
-keep all comments EXCEPT one above the class declaration
-
-\cite{bask-icer-2018}
-2TB data, Java programs
-18 publications survey, technical challenges in analysis
- most pubs analyze errors;
-nobody asked:
- which exns do students encounter (what?! mccall kolling 2014)
- execution / editing timelines
-(blackbox analysis tutorial in sigcse 2020 on a mini dataset, to help researchers start writing an analysis)
-
-\cite{t-hatra-2021}
-plan for a large user-centered study 
-... not very deep, look at edit sequences, replay for errors
+\paragraph{Contrast: User Studies and Types and/or Proofs}
+
+\cite{h-dissertiation-2005}
+heeren
+constraint-based type inference;
+general te criteria:
+- target audience (novice / expert)
+- output format (text / viz)
+- interactivity (yes / no)
+- heuristics (yes / no)
+- primary-or-external (most type errors are easy, some could really use dedicated help)
+
+\cite{sscwj-oopsla-2017}
+learning to blame localizing type errors
+NATE numerical analysis of type errors
+5000 labeled programs
+ errors from class, first fix
+
+\cite{sjw-jfp-2018}
+type error witness generation
+4500 ill-typed student programs
+ \pct{85} of time, synthesis works
+ \pct{70} of time, locate source
+
+\cite{rsgl-cpp-2020}
+user study coq proof assistant
+N devs, XXX
 
@@
 
-\cite{sscwj-oopsla-2017}
-NATE numerical analysis of type errors
-5000 labeled programs
- errors from class, first fix
-
-\cite{sjw-jfp-2018}
-type error witness generation
-4500 ill-typed student programs
- \pct{85} of time, synthesis works
- \pct{70} of time, locate source
-
-\cite{h-dissertiation-2005}
-constraint-based type inference;
-general te criteria:
-- target audience (novice / expert)
-- output format (text / viz)
-- interactivity (yes / no)
-- heuristics (yes / no)
-- primary-or-external (most type errors are easy, some could really use dedicated help)
-
 \cite{hw-scp-2004}
+haack wells
 location of type error as a slice,
@@
 
-\paragraph{Research on Errors}
-
-TODO Liblit dissertiation
+\paragraph{Errors}
 
 Mind your language~\cite{mfk-onward-2011}.
-
-\paragraph{Telemetry}
-
-\begin{table}[t]
-  \caption{Comparing telemetry systems}
-  \label{t:telemetry-design}
-
-  \begin{tabular}{l@{~}cccccc}
-    &             & PII       & Session ID & Deterministic & Private Data (???) \\\midrule
-    & Transparent & \chkNo    & \chkNo     & \chkYes       & \chkNo      \\
-  * & Us          & \chkNo    & \chkYes    & \chkNo        & \chkYes     \\
-    & Diff.Priv   & $\epsilon$ & \chkYes   & \chkYes       & \chkYes     \\
-    & Liblit      & ?         & ?          & ?             & ?           \\
-    & \code{.NET} & \chkMaybe & \chkYes    & \chkYes       & \chkYes     \\
-    & VS Code     & \chkYes   & \chkYes    & \chkYes       & \chkYes     \\
-  \end{tabular}
-\end{table}
-
-Transparent telemetry: explain what it is and contrast our approach.
-What essential, non-transparent things did we collect?
-
-counting only, public decisions about what to count, public data,
-collected weekly for a sample of clients, LastWeek field = prior day when
-system gathered any data;
-eg command invocations, lib stack frames, 
-no user ID, no machine ID,
-no time-ordered traces~\cite{transparent-telemetry}.
-
-kindle track every tap (page turn etc),
-enables whispersync feature,
-drove design of navigation tools,
-opt-out possible~\cite{kindle-telemetry}
-
-vscode: usage data, crash reports, error data (not a crash but unexpected);
-save file, open terminal, copy-paste, autocomplete offered, git queries, machine id,
-session id, timestamp,
-opt-out may be possible, but not for all usage data (per license, sec 2a) and
-every extension can do its own thing~\cite{vscode-telemetry}
-
-.NET SDK and .NET CLI:
-crash reports, command invoked, args, hashed cwd, timings;
-more listed online;
-custom builds beware inadvertant disclosure, keep your filepaths clear;
-data published in aggregate under CC-BY;
-opt-out possible~\cite{dotnet-telemetry}
-
-\cite{lnsmc-usenix-2018} blame-proportional logging,
-scale up ``near'' points of interest,
-could be useful for us in the future.
-\cite{fnm-sigmod-2020} ditto, faster than statistical at finding root-cause,
-deterministic.
+60 students (from 120, self-selected),
+50 min per week for 6 weeks,
+edit sequences,
+followup interviews,
+huge problem: vocabulary (function body, arity),
+problem: semantics of highlight,
+
+\cite{anna-russo-kennedy-ms-2006}
+BitFit, web-based java ide log num hint requests num, result of check answer
+requests num compile, num run, no code?! good we have the same constraint
+115 bitfit users, 728 sessions
+
+\cite{bgimgm-cse-2016}
+user study,
+study Java compiler error messages after an intervention, counting num errors,
+counting repeats, counting per user
+changing errors helps
+
+\cite{dlc-citicse-2014}
+enhanced stx error seems ineffectual
+
+- - -
+
+\cite{t-hatra-2021}
+plan for a large user-centered study of proof assistants,
+events: up, down, to cursor, interrupt, reset.
+... not very deep, look at edit sequences, replay for errors
+
+\cite{wk-koli-2020,lgk-pj-2022} 
+even more edit sequences
+b2t2, examplar, ... must be others outside brown!!
 

### 2023-09-23 programming23/paper.tex
more related work

@@
 \begin{table}[t]\centering
-  \caption{Comparing telemetry systems}
+  \caption{Comparison of telemetry designs}
   \label{t:telemetry-design}
@@
 
-The BlackBox project similarly collects telemetry in broad strokes.
-Every Java program submitted for compilation goes into the BlackBox database (minus
+The Blackbox project similarly collects telemetry in broad strokes.
+Every Java program submitted for compilation goes into the Blackbox database (minus
 any comments above the main class) along with the compiler's response~\cite{bkmu-sigcse-2014}.
-Though somewhat intrusive, the millions of programs in the BlackBox dataset
+Though somewhat intrusive, the millions of programs in the Blackbox dataset
 have enabled dozens of contributions to the area of CS
@@
 
+\paragraph{Telemetry Enhancements: What and How to Sample}
+
 One way to futher strengthen the privacy of any telemetry design, including
-transparent telemetry, is to fuzz the data using techniques from
-differential privacy~\cite{zhlbr-cc-2020,epk-ccs-2014,wblj-usenix-2017}.
-Recent work on fuzzing traces~\cite{zhlbr-cc-2020},
-accounting for known relationships between events~\cite{zhlbr-oopsla-2020},
-and performing the fuzzing on the client side~\cite{hlzbr-ecoop-2021} is
-especially relevant to the session-based data that Roblox Studio collects.
+transparent telemetry, is to systematically add noise to the data using
+techniques from differential
+privacy~\cite{zhlbr-cc-2020,epk-ccs-2014,wblj-usenix-2017}.
+Recent work that is especially relevant to the session-based data collected
+by Roblox Studio shows how to add noise to traces~\cite{zhlbr-cc-2020},
+how to account for known relationships between events~\cite{zhlbr-oopsla-2020},
+and how to choose the noise amount on the client side~\cite{hlzbr-ecoop-2021}.
 Applying differential privacy is subtle, however.
 While we can likely fuzz the total number of type errors, if we were to
-fuzz the number of \code{CodeTooComplex} errors carelessly, it might drastically
+fuzz the number of \code{CodeTooComplex} errors, it might drastically
 affect our conclusions.
 
-Lastly, cooperative bug isolation is a related method for designing
+Cooperative bug isolation is a method for designing
 telemetry systems~\cite{liblit-thesis}.
-%% \FILL{} adaptive bug isolation~\cite{nl-icse-2010}, search on the control dependence graph.
-%% 1\% overhead vs 87\% for sampling and 300\% for binary instrumentation
 The focus is not on privacy, but rather on how to collect a small amount of
@@
 
-\paragraph{Telemetry Rates}
-
-We chose randomly, uniform distribution.
-
-\cite{lnsmc-usenix-2018} blame-proportional logging,
-scale up ``near'' points of interest,
-could be useful for us in the future.
-\cite{fnm-sigmod-2020} ditto, faster than statistical at finding root-cause,
-deterministic.
-
-\paragraph{TBD}
-
-RAMSS workshop series.
-
-%% - zfstt-ieeesensors-2022
-%% - lit review, benefits of static types
-%%   http://danluu.com/empirical-pl/
-%% - thomas kennedy .... generic usage monitoring ascilite 2003
-%% - ahmadzadeh elliman higgins iticse 2005
-%% - buffardi etal 2014 adaptive and social mechanisms
-%% - bddf-icse-2016
-%% -  sh edwards and jason snyder icer 2009 comparing effective and ineffective grading platform
-%% - murphy etal sigcse 2009 Retina: helping students
-%% - ESP workshop on empirical studies of programming
-%%   https://dl.acm.org/doi/proceedings/10.1145/266399
-%%   nothing sounds big ... empirical yes, large no
-%% - Choppella Haynes IU TR 426
-%% - Tip Dinesh tosem 2001
-
-\paragraph{Contrast: User Studies and Types and/or Proofs}
+Two further refinements, which reduce the burden on experts to
+select points of interest, are adaptive bug isolation~\cite{nl-icse-2010}
+and blame-proportional logging~\cite{lnsmc-usenix-2018}.
+Adaptive bug isolation starts with a set of predicates, studies which are
+most correlated with failures, and experiments with adding telemetry to nearby
+predicates; this technique can reduce the overhead of telemetry by two orders
+of magnitude relative to naive binary instrumentation.
+Blame-proportional logging starts with lightweight telemetry to recognize
+defects, assigns ranks to methods estimating their likelihood as root causes,
+and uses the ranks and future observations to narrow down the cause.
+Deep transfer learning is another promising way to automatically hone in on events of
+interest~\cite{zfstt-ieeesensors-2022}.
+
+%% \cite{fnm-sigmod-2020}
+%% adaptive interventional debugging
+%% failure -> correlated predicates -> temporal properties to overapprox relationships -> fault injection to discover true relationships
+
+%% RAMSS workshop (remote measurement), sounds related but i haven't found proceedings
+%% \cite{op-ramss-2003,op-ramss-2004}
+
+\paragraph{User Studies: Errors and Type Errors}
+
+Foundational:
+
+Mind your language~\cite{mfk-onward-2011}.
+60 students (from 120, self-selected),
+50 min per week for 6 weeks,
+edit sequences,
+followup interviews,
+huge problem: vocabulary (function body, arity),
+problem: semantics of highlight,
+
+\cite{wk-koli-2020,lgk-pj-2022,mcpcspss-abz-2020}
+more edit sequences
+
+Ballpark:
+
+\cite{t-hatra-2021}
+plan for a large user-centered study of proof assistants,
+events: up, down, to cursor, interrupt, reset.
+... not very deep, look at edit sequences, replay for errors
+
+\cite{anna-russo-kennedy-ms-2006}
+BitFit, web-based java ide log num hint requests num, result of check answer
+requests num compile, num run, no code?! good we have the same constraint
+115 bitfit users, 728 sessions
+
+\cite{bgimgm-cse-2016}
+user study,
+study Java compiler error messages after an intervention, counting num errors,
+counting repeats, counting per user
+changing errors helps
+
+\cite{dlc-citicse-2014}
+enhanced stx error seems ineffectual
 
@@
 
-\paragraph{Errors}
-
-Mind your language~\cite{mfk-onward-2011}.
-60 students (from 120, self-selected),
-50 min per week for 6 weeks,
-edit sequences,
-followup interviews,
-huge problem: vocabulary (function body, arity),
-problem: semantics of highlight,
-
-\cite{anna-russo-kennedy-ms-2006}
-BitFit, web-based java ide log num hint requests num, result of check answer
-requests num compile, num run, no code?! good we have the same constraint
-115 bitfit users, 728 sessions
-
-\cite{bgimgm-cse-2016}
-user study,
-study Java compiler error messages after an intervention, counting num errors,
-counting repeats, counting per user
-changing errors helps
-
-\cite{dlc-citicse-2014}
-enhanced stx error seems ineffectual
-
-- - -
-
-\cite{t-hatra-2021}
-plan for a large user-centered study of proof assistants,
-events: up, down, to cursor, interrupt, reset.
-... not very deep, look at edit sequences, replay for errors
-
-\cite{wk-koli-2020,lgk-pj-2022} 
-even more edit sequences
-b2t2, examplar, ... must be others outside brown!!
+bibliography of static type studies:
+http://danluu.com/empirical-pl/
+
+\paragraph{TBD}
+
+thomas kennedy .... generic usage monitoring ascilite 2003
+
+ahmadzadeh elliman higgins iticse 2005
+
+buffardi etal 2014 adaptive and social mechanisms
+
+bddf-icse-2016
+ sh edwards and jason snyder icer 2009 comparing effective and ineffective grading platform
+
+murphy etal sigcse 2009 Retina: helping students
+
+Choppella Haynes IU TR 426
+
+Tip Dinesh tosem 2001
+
+% ESP workshop on empirical studies of programming
+% https://dl.acm.org/doi/proceedings/10.1145/266399
+% nothing sounds big ... empirical yes, large no
+% https://link.springer.com/content/pdf/10.1023/a:1008040416962.pdf
 
@@
 
-\newpage
-
-\appendix
-
-\section{Data Details}
-
-Tips for interpreting the data:
-
-\begin{itemize}
-  \item
-    Edit ranges are non-negative because we
-    implemented the interval arithmetic for them using unsigned integers.
-    To filter out negative edit ranges, we removed all ranges
-    greater than 4 billion.
-
-  \item
-  \item
-  \item
-\end{itemize}
-
-\newpage
+%\newpage
+%
+%\appendix
+%
+%\section{Data Details}
+%
+%Tips for interpreting the data:
+%
+%\begin{itemize}
+%  \item
+%    Edit ranges are non-negative because we
+%    implemented the interval arithmetic for them using unsigned integers.
+%    To filter out negative edit ranges, we removed all ranges
+%    greater than 4 billion.
+%
+%  \item
+%  \item
+%  \item
+%\end{itemize}
+%
+%\newpage
 

### 2023-09-23 programming23/paper.tex
related work table, again

@@
 
-  \FILL{}
-
-  \begin{tabular}{l@{~}cccccc}
-    &               & Code      & Session ID & Timestamp     & Deterministic (???) & Private Data (???) \\\midrule
-    & Transparent   & \chkNo    & \chkNo     & \chkNo        & \chkYes       & \chkNo      \\
-  * & Roblox Studio & \chkNo    & \chkYes    & \chkNo        & \chkNo        & \chkYes     \\
-%% orthogonal %%    & Diff.Priv     & $\epsilon$ & \chkYes   & \chkYes       & \chkYes       & \chkYes     \\
-%% orthogonal %%    & CBI           & \chkMaybe & \chkMaybe  & \chkNo        & \chkNo        & ?           \\
-    & Classic       & \chkYes   & \chkYes    & \chkYes       & \chkYes       & \chkYes     \\
+  \begin{tabular}{l@{~~}c@{~~}c@{~~}c@{~~}c@{~~}c}
+    ~             & Event Counts & Random  & Timestamps & Error Msgs. & Source Code \\\midrule
+    Roblox Studio & \chkYes      & \chkYes & \chkYes    & \chkNo        & \chkNo    \\
+    Transparent   & \chkYes      & \chkNo  & \chkNo     & \chkNo        & \chkNo    \\
+    Classic       & \chkYes      & \chkNo  & \chkYes    & \chkYes       & \chkYes   \\
+%% orthogonal %%    & Diff.Priv
+%% orthogonal %%    & CBI
   \end{tabular}

### 2023-09-23 programming23/paper.tex
blackbox cites

@@
 any comments above the main class) along with the compiler's response~\cite{bkmu-sigcse-2014}.
-Though somewhat intrusive, the millions of programs in the Blackbox dataset
+%% Though somewhat intrusive,
+The millions of programs in the Blackbox dataset
 have enabled dozens of contributions to the area of CS

### 2023-09-23 programming23/paper.tex
related work, drafted

@@
 Use edit range~(\cref{t:type-error-survival} for NS vs S)
-and error density~(\cref{f:error-delta})
+and error density~(\cref{f:error-density})
 
@@
 and uses the ranks and future observations to narrow down the cause.
-Deep transfer learning is another promising way to automatically hone in on events of
+Deep transfer learning is another promising way to hone in on events of
 interest~\cite{zfstt-ieeesensors-2022}.
@@
 
-\paragraph{User Studies: Errors and Type Errors}
-
-Foundational:
-
-Mind your language~\cite{mfk-onward-2011}.
-60 students (from 120, self-selected),
-50 min per week for 6 weeks,
-edit sequences,
-followup interviews,
-huge problem: vocabulary (function body, arity),
-problem: semantics of highlight,
-
-\cite{wk-koli-2020,lgk-pj-2022,mcpcspss-abz-2020}
-more edit sequences
-
-Ballpark:
-
-\cite{t-hatra-2021}
-plan for a large user-centered study of proof assistants,
-events: up, down, to cursor, interrupt, reset.
-... not very deep, look at edit sequences, replay for errors
-
-\cite{anna-russo-kennedy-ms-2006}
-BitFit, web-based java ide log num hint requests num, result of check answer
-requests num compile, num run, no code?! good we have the same constraint
-115 bitfit users, 728 sessions
-
-\cite{bgimgm-cse-2016}
-user study,
-study Java compiler error messages after an intervention, counting num errors,
-counting repeats, counting per user
-changing errors helps
-
-\cite{dlc-citicse-2014}
-enhanced stx error seems ineffectual
-
-\cite{h-dissertiation-2005}
-heeren
-constraint-based type inference;
-general te criteria:
-- target audience (novice / expert)
-- output format (text / viz)
-- interactivity (yes / no)
-- heuristics (yes / no)
-- primary-or-external (most type errors are easy, some could really use dedicated help)
-
-\cite{sscwj-oopsla-2017}
-learning to blame localizing type errors
-NATE numerical analysis of type errors
-5000 labeled programs
- errors from class, first fix
-
-\cite{sjw-jfp-2018}
-type error witness generation
-4500 ill-typed student programs
- \pct{85} of time, synthesis works
- \pct{70} of time, locate source
-
-\cite{rsgl-cpp-2020}
-user study coq proof assistant
-N devs, XXX
-
-\cite{cdhhjklwya-hatra-2020}
-evaluating the PLIERS design process (Coblenz etal 202X)
-teaching students to use PLIERS; everyone ran a user study!
-6 languages
- 3 ran usability studies
-recommend talk-alouds
-
-\cite{gstf-hatra-2021}
-liquid types for java
-30 devs user study
- zoom
-
-\cite{lfgc-pldi-2007}
-typechecker never errors,
- calls a helper that looks for nearby programs that do typecheck
-tested on
- 10 students, 2yrs professional experience
- 5 hw assignments
- 2122 files collected;
-won \pct{19}, lost \pct{17};
-ill-suited for things like (let x = e1 in ....),
- change e1,
- use typechecker feedback to guide edits;
-small data!
-
-\cite{w-popl-1986}
-reporting source, not detection
-tested on 9 type errors
- all \emph{deletions}, such as forgetting to inject in an enum
-
-\cite{hw-scp-2004}
-haack wells
-location of type error as a slice,
- not a srcloc,
- not a subtree
-
-bibliography of static type studies:
-http://danluu.com/empirical-pl/
-
-\paragraph{TBD}
-
-thomas kennedy .... generic usage monitoring ascilite 2003
-
-ahmadzadeh elliman higgins iticse 2005
-
-buffardi etal 2014 adaptive and social mechanisms
-
-bddf-icse-2016
- sh edwards and jason snyder icer 2009 comparing effective and ineffective grading platform
-
-murphy etal sigcse 2009 Retina: helping students
-
-Choppella Haynes IU TR 426
-
-Tip Dinesh tosem 2001
+\paragraph{User Studies, Errors, and Type Errors}
+
+Our approach to data analysis is inspired by Marceau
+et~al.~\cite{mfk-onward-2011,mfk-sigcse-2011}, who used keystroke-level data
+collected over 6 weeks to analyze students' responses to runtime errors
+in DrRacket~\cite{fcffksf-jfp-2002}.
+Similarly, Macedo et~al.~\cite{mcpcsprs-abz-2020,mcpcsprs-scp-2021} capture
+edit sessions from the online Alloy4Fun editor and present several automated
+analyses.
+A major challenge in our setting, however, is that edit sequences are
+incomplete---there may be minutes-long gaps between two telemetry records.
+Several more-recent works have studied edit sequences
+to draw conclusions about errors and intent, e.g.,~\cite{wk-koli-2020,lgk-pj-2022,rsgl-cpp-2020}.
+
+The BitFit project collects data about beginning students interactions with a
+Java IDE~\cite{ekc-wccce-2016,anna-russo-kennedy-ms-2006}; however, unlike Blackbox, it does
+not collect source code.
+BitFit data consists of a timestamp, a session id, and counter values showing
+the number of attempts at a question, the number of hints requested, the number
+of compiles and runs, and the number of submission attempts.
+On one hand, this data is more informative than Roblox Studio's because it
+targets specific events rather than random keystrokes.
+On the other hand, there are no details about errors to show what students
+struggled with.
+Other logging-based studies of Java programs use the full error message to
+draw conclusions~\cite{bgimgm-cse-2016,dlc-iticse-2014}.
+Roblox Studio cannot follow this approach because strings in errors may leak
+personal information.
+
+Much work on the benefits of static types focuses on a small number
+of case studies~\cite{w-popl-1986,hw-scp-2004,td-tosem-2001},
+inverviews~\cite{cdhhjklwya-hatra-2020,gstf-hatra-2021,cams-oopsla-2020},
+or static corpuses~\cite{rppd-fse-2014,bhmvv-toplas-2019,bmvv-arxiv-2019}
+(for literature reviews, see~\cite{empirical-types,heeren-thesis}).
+Controlled experiments, for example comparing statically-typed Java and dynamically-typed
+Groovy~\cite{khrts-icpc-2012}, are a notable exception.
+Lerner et~al.~\cite{lfgc-pldi-2007} apply type error repair tool on over two thousand
+of programs written during five homework exercises.
+Seidel et~al.~\cite{sscwj-oopsla-2017,sjw-jfp-2018} use five thousand programs
+to train data-driven methods for localizing errors.
+These large studies with several thousands of programs are
+nevertheless on a much smaller scale than Roblox Studio.
+
+%\paragraph{TBD}
+%
+%thomas kennedy .... generic usage monitoring ascilite 2003
+%
+%ahmadzadeh elliman higgins iticse 2005
+%
+%buffardi etal 2014 adaptive and social mechanisms
+%
+%bddf-icse-2016
+% sh edwards and jason snyder icer 2009 comparing effective and ineffective grading platform
+%
+%murphy etal sigcse 2009 Retina: helping students
+%
+%Choppella Haynes IU TR 426
 
@@
 
+% \cite{}
+% heeren
+% constraint-based type inference;
+% general te criteria:
+% - target audience (novice / expert)
+% - output format (text / viz)
+% - interactivity (yes / no)
+% - heuristics (yes / no)
+% - primary-or-external (most type errors are easy, some could really use dedicated help)
+
+% \cite{t-hatra-2021}
+% plan for a large user-centered study of proof assistants,
+% events: up, down, to cursor, interrupt, reset.
+% ... not very deep, look at edit sequences, replay for errors
+
 \section{Conclusion}
@@
 Timestamps are not the focus of our analysis, but they do play a critical role.
-Consider~\cref{f:error-delta}.
+Consider~\cref{f:error-density}.
 With timestamps, it is clear that error density stays close to zero.
 Without timestamps, we would be left with the counts at the bottom
-of~\cref{f:error-delta}.
+of~\cref{f:error-density}.
 Many distributions could fit those counts, such as a ``mountain'' session that first

### 2023-09-24 programming23/paper.tex
move file count to footnote

@@
 
-\subsection{Telemetry records}
+\subsection{Telemetry Records}
 \label{s:telemetry-records}
+
+\begin{figure}
+  \begin{tabular}{|l|l|l|l|l|l|l|l|l|}\midrule
+    meta & size & te count & fs count & te 0 & te 1 & te 2 \ldots
+    \\\midrule
+  \end{tabular}
+  \caption{Struture of a telemetry record}
+  \label{f:tdata}
+\end{figure}
 
@@
 
-\begin{figure}[t]
+\begin{figure}[t]\centering
   % ;; (list* (max* vv) (median < vv) mm (stddev/mean mm vv) (percentile* vv)))
@@
 
-  \begin{tabular}{l@{}r@{~}l@{}r@{~}l@{}r@{~}l}
-                 & Files  &              &     Lines &             &    Edit Range & \\
-    Mean [std]   & 11,911 & \stddev{12}  &     6,497 & \stddev{22} &         3,680 & \stddev{31} \\
-    Median       &  7,678 &              &     3,115 &             &           926 & \\
-    \pct{99}     & 51,761 &              &    50,547 &             &        34,725 & \\
+%% Files has HUGE numbers thanks to data assets. Forget it.
+% Files  &              &
+% 11,911 & \stddev{12}  &
+%  7,678 &              &
+% 51,761 &              &
+%   & \includegraphics[width=0.2\columnwidth]{img/files-distribution.pdf}
+
+  \begin{tabular}{l@{}r@{~}l@{}r@{~}l}
+                 &  Lines &             &    Edit Range & \\
+    Mean [std]   &  6,497 & \stddev{22} &         3,680 & \stddev{31} \\
+    Median       &  3,115 &             &           926 & \\
+    \pct{99}     & 50,547 &             &        34,725 & \\
     %% TODO x-axis starts at 1, not zero ... there are no zeros right??!
-    & \includegraphics[width=0.2\columnwidth]{img/files-distribution.pdf}
-    & & \includegraphics[width=0.2\columnwidth]{img/lines-distribution.pdf}
+      & \includegraphics[width=0.2\columnwidth]{img/lines-distribution.pdf}
     & & \includegraphics[width=0.2\columnwidth]{img/editrange-distribution.pdf}
@@
 The top half of the figure reports the mean, standard deviation, median,
-and \percentile{99} for the number of files,\todo{There are
-  1000s of files?!? More files than lines? Makes no sense.} lines, and lines in edit range.
-The bottom half shows zoomed-in distributions of the file, line, and edit range
-numbers.
-For example, the x-axis of the first plot ranges from 0 to 1,000 files and
+and \percentile{99} for the number of lines
+and lines in edit range.\footnote{The data also contains the number of files in
+each project, but these numbers are difficult to relate to typechecking because
+projects may contain hundreds of files that define data assets.  For the
+record, the median file count is 7,678 and the 99th percentile is 51,761 files.
+}
+The bottom half shows zoomed-in distributions of the line and edit range
+sizes.
+For example, the x-axis of the first plot ranges from 0 to 1,000 lines and
 the y-axis counts up to 1,000 telemetry records.
@@
 There is a huge amount of variation across projects.
-The largest ones have over 50,000 files, and
-there are many small ones with 1 file and/or 1 line of code.
+The largest ones have over 50,000 lines of code
+while the smallest have merely 1 line of code.
 Unsurprisingly, the numbers come with large standard
-deviations.\todo{Which is not shown!}
+deviations~([std] in~\cref{f:codebase-size}).
 The median values are more reasonable, with roughly 3,000 lines of code
-and 1,000 lines in edit ranges, but still have a large number of files: over 7,000.
-One source of variation is that some developers use Luau scripts as a way
-to store data, writing a module that exports one huge array, typically
-defined on a single line. Such data storage scripts provide a challenge for
-script tooling.\todo{No idea what's said here.}
-
-%% The high number of files may be because creators define many components to
-%% build a Roblox experience and put these components in separate files.
-%% (\FILL{} still that's a lot of components ... any better guesses?)
-
-%% (\FILL{} doublecheck file sizes)
-% That seems way off, not sure we're seeing so many files :/
+and 1,000 lines in edit ranges.
 
@@
 %% > For FS: \mnocheck{} 400, \mnonstrict{} 39, \mstrict{} 6.
+
+\subsection{Forced Strict Errors by Mode}
+
+FILL
+
+How do the 72 million forced strict divide across modes?
+
+What about type errors, same distribution?!
 
@@
 
-\subsection{Forced Strict Errors}
-
-\FILL{} I (ben) think we need to talk more about forced strict errors here, or soon.
-Otherwise people will forget they exist in the data.
-
 \subsection{Type Error Density}

### 2023-09-24 programming23/paper.tex
ben affiliation

@@
 
-\author[a]{Ben Greenman}
-\authorinfo{(\email{benjamin.l.greenman@gmail.com}) is a postdoc at Brown University. He will be joining the University of Utah in Fall~2023.}
+\author[a,b]{Ben Greenman}
+\authorinfo{(\email{benjamin.l.greenman@gmail.com}) was a CIFellows 2020
+postdoc at Brown Univesity and is now an assistant professor
+at the recently-named Kahlert School of Computing at the University of Utah.}
 \affiliation[a]{Brown University, Providence, RI, USA}
+\affiliation[b]{University of Utah, Salt Lake City, UT, USA}
 % \orcid{0000-0001-7078-9287}
 
-\author[b]{Alan Jeffrey}
+\author[c]{Alan Jeffrey}
 \authorinfo{(\email{ajeffrey@roblox.com}) is a Principal Software Engineer at Roblox.}
 % \orcid{0000-0001-6342-0318}
-\affiliation[b]{Roblox, San Mateo, CA, USA}
+\affiliation[c]{Roblox, San Mateo, CA, USA}
 
@@
 
-\author[b]{Mitesh Shah}
+\author[c]{Mitesh Shah}
 \authorinfo{(\email{mshah@roblox.com}) is Senior Engineering Director, Programmability, at Roblox.}

### 2023-09-24 programming23/paper.tex
rename error labels: shorten, put ExtraInfo next to GenError where it belongs

@@
 \newcommand{\zerowidth}[1]{\makebox[0pt][l]{#1}}
-\newcommand{\stddev}[1]{[#1\textsc{k}]}
+\newcommand{\nK}[1]{#1\textsc{k}}
+\newcommand{\stddev}[1]{[\nK{#1}]}
 \newcommand{\gcell}[1]{\cellcolor{green!20}#1}
@@
     \code{CountMismatch} & Arity mismatch for a function \\
-    \code{IncorrectGenericParameterCount} & Arity mismatch for a generic type \\
+    \code{IncorrectGenericParamCount} & Arity mismatch for a generic type \\
     \code{UnknownProperty} & Referenced an invalid field or method  \\
@@
 %    UnknownRequire &  \\
-%    UnknownPropButFoundLikeProp &  \\
+%    UnknownPropButGotLikeProp &  \\
 %    InternalError &  \\
@@
 %    IllegalRequire &  \\
-%    FunctionExitsWithoutReturning &  \\
+%    FunctionExitsWithoutReturn &  \\
 %    DuplicateGenericParameter &  \\
@@
 \begin{figure}
+  \fbox{
+  }
   \begin{tabular}{|l|l|l|l|l|l|l|l|l|}\midrule
@@
 \begin{table}[t]
-  \caption{Number of records that add, keep, or drop (remove) specific type errors from their edit range.}
+  \caption{Number of telemetry records that add, keep, or drop (remove) specific type errors from their edit range.}
   \label{t:type-error-survival}
-  \begin{tabular}{lr@{}r@{}rr@{}r@{}rr@{}r@{}r}
+  \begin{tabular}{l@{}r@{}r@{}rr@{}r@{}rr@{}r@{}r}
     & \zerowidth{\mnocheck{}} & & & \zerowidth{\mnonstrict{}} & & & \zerowidth{\mstrict{}} & & \\
@@
     \code{DuplicateTypeDefinition} & {0} & {0} & {0} & {1} & {0} & {0} & {0} & {0} & {0} \\
-    \code{ExtraInformation} & {0} & {0} & {0} & {48} & {7} & {39} & {2} & {0} & {1} \\
     \code{FunctionDoesNotTakeSelf} & {0} & {0} & {0} & {4} & {6} & {5} & {0} & {0} & {0} \\
-    \code{FunctionExitsWithoutReturning} & {0} & {0} & {0} & {9} & {3} & {6} & {11} & {5} & {8} \\
+    \code{FunctionExitsWithoutReturn} & {0} & {0} & {0} & {9} & {3} & {6} & {11} & {5} & {8} \\
     \code{GenericError} & {0} & {0} & {0} & {206} & {52} & {203} & {8} & {1} & {9} \\
+    \code{GenericExtraInformation} & {0} & {0} & {0} & {48} & {7} & {39} & {2} & {0} & {1} \\
     \code{IllegalRequire} & {0} & {0} & {0} & {16} & {3} & {33} & {0} & {0} & {2} \\
-    \code{IncorrectGenericParameterCount} & {0} & {0} & {0} & {0} & {0} & {0} & {1} & {0} & {1} \\
+    \code{IncorrectGenericParamCount} & {0} & {0} & {0} & {0} & {0} & {0} & {1} & {0} & {1} \\
     \code{MissingProperties} & {0} & {0} & {0} & {11} & {3} & {11} & {6} & {4} & {6} \\
@@
     \code{TypesAreUnrelated} & {0} & {0} & {0} & {1} & {0} & {0} & {0} & {0} & {1} \\
-    \code{UnknownPropButFoundLikeProp} & {0} & {0} & {0} & {26} & {18} & {21} & {1} & {0} & {1} \\
+    \code{UnknownPropButGotLikeProp} & {0} & {0} & {0} & {26} & {18} & {21} & {1} & {0} & {1} \\
     \code{UnknownProperty} & {0} & {0} & {0} & {347} & {180} & {385} & {19} & {19} & {35} \\
@@
     \code{CannotExtendTable},
-    \code{ExtraInformation},
+    \code{GenericExtraInformation},
     \code{FunctionDoesNotTakeSelf},
@@
     \code{OptionalValueAccess},
-    \code{UnknownPropButFoundLikeProp}.
+    \code{UnknownPropButGotLikeProp}.
     \QALAN{} are any of these surprising, or expected from the bigger number of
@@
     \code{CannotInferBinaryOperation},
-    \code{FunctionExitsWithoutReturning},
-    \code{IncorrectGenericParameterCount}.
+    \code{FunctionExitsWithoutReturn},
+    \code{IncorrectGenericParamCount}.
     \FILL{} so what.
@@
       \code{GenericError} & \pct{2.06} \\
-      \code{UnknownPropButFoundLikeProp} & \pct{0.44} \\
+      \code{UnknownPropButGotLikeProp} & \pct{0.44} \\
       \code{CannotExtendTable} & \pct{0.42} \\
-      \code{ExtraInformation} & \pct{0.32} \\
+      \code{GenericExtraInformation} & \pct{0.32} \\
       \code{ModuleHasCyclicDependency} & \pct{0.23} \\
@@
       \code{MissingProperties} & \pct{0.09} \\
-      \code{FunctionExitsWithoutReturning} & \pct{0.07} \\
+      \code{FunctionExitsWithoutReturn} & \pct{0.07} \\
       \code{FunctionDoesNotTakeSelf} & \pct{0.07} \\
@@
       \code{OptionalValueAccess} & \pct{2.99} \\
-      \code{FunctionExitsWithoutReturning} & \pct{2.99} \\
+      \code{FunctionExitsWithoutReturn} & \pct{2.99} \\
       \code{UnknownRequire} & \pct{2.37} \\
@@
       \code{NotATable} & \pct{1.32} \\
-      \code{ExtraInformation} & \pct{0.35} \\
-      \code{UnknownPropButFoundLikeProp} & \pct{0.26} \\
-      \code{IncorrectGenericParameterCount} & \pct{0.18} \\
+      \code{GenericExtraInformation} & \pct{0.35} \\
+      \code{UnknownPropButGotLikeProp} & \pct{0.26} \\
+      \code{IncorrectGenericParamCount} & \pct{0.18} \\
       \code{CannotCallNonFunction} & \pct{0.18} \\
@@
     Two errors that appear in the \mstrict{} records
-    never appear in \mnonstrict{}: \code{OccursCheckFailed} and \code{IncorrectGenericParameterCount}.
+    never appear in \mnonstrict{}: \code{OccursCheckFailed} and \code{IncorrectGenericParamCount}.
     We attribute these to more advanced uses of types in \mstrict{} mode.

### 2023-09-24 programming23/paper.tex
checkpoint: diagram telemetry record, rewrite sec 3

@@
 
-\begin{figure}
-  \fbox{
-  }
-  \begin{tabular}{|l|l|l|l|l|l|l|l|l|}\midrule
-    meta & size & te count & fs count & te 0 & te 1 & te 2 \ldots
-    \\\midrule
-  \end{tabular}
-  \caption{Struture of a telemetry record}
+\begin{figure}[t]\centering
+  \begin{tabular}{|l|l|l|l|l|l|l|}\hline
+    ID & Time & Mode & Reason & Size Info & Overall Counts & Edit Range Counts
+    \\\toprule
+  \end{tabular}
+  \caption{Structure of a telemetry record. There are 13 overall counts and up to 70 edit range counts in each record.}
   \label{f:tdata}
@@
 \end{itemize}
-Each telemetry record contains nine kinds of data:
+Each telemetry record contains seven kinds of data.
+Refer to~\cref{f:tdata} for an illustration:
 \begin{enumerate}
-  \item metadata: a psedonymized session identifier, timestamp and session duration,
-  \item mode of the current file: (\mnocheck{}, \mnonstrict{}, or \mstrict{}),
+  \item
+    ID: a psedonynimized identifier for the current session
+  \item
+    Time: one timestamp from the client and one from the server.
+    %% timestamp explanation \cref{s:data-cleaning}
+  \item
+    Mode: type analysis mode of the current file, which is either \mnocheck{},
+    \mnonstrict{}, or \mstrict{}.
+%% Reason & Size Info & Overall Counts & Edit Range Counts
+
   \item reason for sending: (random selection, or file switch),
@@
   \item TooComplexErrors: project-wide count for this one specific error type
+     \cref{s:code-too-complex}
 \end{enumerate}
@@
 \paragraph{Internal Limits, Code Too Complex}
+\label{s:code-too-complex}
 %% data = out/ctc-info.txt

### 2023-09-24 programming23/paper.tex
sec 3 relist telemetry fields, misc other fixes (captions)

@@
 \newcommand{\QALAN}{\FILL{} Alan:}
+\newcommand{\panon}{pseudonymized}
 
@@
 or collects personally identifiable information~(PII),
-we performed a large-scale study using pseudonymized \emph{telemetry}.
+we performed a large-scale study using \panon{} \emph{telemetry}.
 
@@
 and no information about what creation the data came from.
-Telemetry records are correlated by session, using a pseudonymized
+Telemetry records are correlated by session, using a \panon{}
 session identifier.
@@
 \section{{Roblox} Context}
+\label{s:context}
 % https://create.roblox.com/docs/scripting/luau
@@
 
-  \caption{{Roblox Studio 3D creation} tools (left) and IDE (right)}
+  \caption{{Roblox Studio 3D creation} tools (left) and IDE (right).}
   \label{fig:roblox-studio}
@@
 Developer Environment (IDE), as seen in \Cref{fig:roblox-studio}.
-The IDE includes an optional ``Script Analysis'' widget, which
+The IDE includes an optional \emph{Script Analysis} widget that
 reports syntax errors, type errors, and problems identified by
@@
 the location in code where reported errors occur.
-To opt in to type error reporting, creators set a \emph{mode}
-for each script, which is one of:
+Creators have three \emph{analysis modes} (modes) to choose from
+for every script they write:
 \begin{itemize}
-  \item \mnocheck{}: only syntax errors are reported,
-  \item \mnonstrict{}: all syntax errors, and a subset of ``high confidence'' type errors, are reported, or
-  \item \mstrict{}: all syntax errors and type errors, are reported.
+  \item \mnocheck{}: report only syntax errors,
+  \item \mnonstrict{}: report syntax errors and a subset of ``high confidence'' type errors, or
+  \item \mstrict{}: report syntax errors and all type errors.
 \end{itemize}
+Each run of type analysis can report several errors.
+There is no guarantee that creators read every error.
+In fact, creators who close the Script Analysis widget
+will see highlights in their code but no further details.
+
 As an example of nonstrict mode, the following program only reports one error:
@@
 \begin{table}
-  \caption{Selected Error Labels}
+  \caption{Sample of type errors.}
   % https://github.com/Roblox/luau/blob/master/Analysis/include/Luau/Error.h
   %% from type analysis, but they're not all really type errors
+  \label{t:type-error-labels}
+
   %% TODO review selection after analyzing the full data
-  \label{t:type-error-labels}
 
@@
 The telemetry for a Roblox Studio session is gathered as a series of
-telemtry records, all with the same psedonynimized session
+telemtry records, all with the same \panon{} session
 identifier. This allows us to correlate telemetry across a single
@@
 \end{itemize}
-Each telemetry record contains seven kinds of data.
-Refer to~\cref{f:tdata} for an illustration:
-\begin{enumerate}
-  \item
-    ID: a psedonynimized identifier for the current session
-  \item
-    Time: one timestamp from the client and one from the server.
-    %% timestamp explanation \cref{s:data-cleaning}
-  \item
-    Mode: type analysis mode of the current file, which is either \mnocheck{},
-    \mnonstrict{}, or \mstrict{}.
-%% Reason & Size Info & Overall Counts & Edit Range Counts
-
-  \item reason for sending: (random selection, or file switch),
-  \item size of codebase: number of files, and number of lines of code type checked,
-  \item size of edit range
-  \item overall type errors: for previous and current states, we have the total number, the nunber in the current file, and number in the edit range,
-  \item specific errors in the edit range: for the previous and current states, for each type error code, the number type errors in the edit range,
-    \FILL{} explain previous/current in detail, explain that this data takes up most of the telemetry record
-  \item forced strict errors: specific errors for the
-    edit range computed as if the user had strict mode enabled.
-  \item TooComplexErrors: project-wide count for this one specific error type
-     \cref{s:code-too-complex}
-\end{enumerate}
+
 To track the edit range, we record a start and end position, which we
@@
 The main benefit of this strategy is that it ensures constant-size telemetry records.
+
+Each telemetry record contains seven kinds of data.
+Refer to~\cref{f:tdata} for an illustration:
+\begin{enumerate}
+  \item
+    ID: a \panon{} identifier for the current session
+  \item
+    Time: one timestamp from the client and one from the server.
+    %% timestamp explanation \cref{s:data-cleaning}
+  \item
+    Mode: type analysis mode of the current file, which is either \mnocheck{},
+    \mnonstrict{}, or \mstrict{}.
+  \item
+    Reason: a flag that explains whether this record was sent due to a
+    randomly-selected keystroke or a module switch.
+  \item
+    Size Info: number of lines in the codebase and number of lines in the edit range.
+  \item
+    Overall Counts: summarizes type errors and forced strict errors during the
+    last two invocations of type analysis, and additionally reports the number
+    of times type analysis hit an internal limit on problem
+    size~(see~\cref{s:code-too-complex} for details).
+    For each invocation and each kind of
+    analysis error, there are three summary counts:
+    %% grand total = 12
+    \begin{itemize}
+      \item total number of errors,
+      \item errors in the current module, and
+      \item errors in the current edit range.
+    \end{itemize}
+  \item
+    Edit Range Counts: these list specific errors that arose in the last two
+    type analyses and overlap with the current edit range.
+    Since there are 35 possible errors~(\cref{s:context}), there can be up to 70 counts
+    in a telemetry record.
+    Exactly how to interpret this data depends on which analysis it came from:
+    \begin{itemize}
+      \item
+        Errors in the latest type analysis are currently visible to the
+        creator.
+        These may have been introduced by the changes in the edit range.
+      \item
+        Errors from the previous type analysis were visible before the creator
+        made the latest edits.
+        These may have motivated some of the changes in the edit range.
+    \end{itemize}
+\end{enumerate}
+
+\paragraph{Reflections}
+This kind of telemetry data is coarse-grained.
+For instance, it does not let us distinguish between edits that ignore a type
+error from edits that remove one error while introducing another.
+We acknowledge this and other threats in~\cref{s:threats}.
+The main advantages of the telemetry design are the small size of each record
+and complete the lack of PII.
+Furthermore, despite the limitations, Studio's telemetry supports a variety of
+interesting conclusions that we showcase in the next section.
 
@@
 
-  \caption{Session size in seconds and in number of records}
+  \caption{Session size in seconds and in number of records.}
   \label{f:session-size}
@@
   \end{tabular}
-  \caption{Type errors and forced strict errors across all telemetry records}
+  \caption{Type errors and forced strict errors across all telemetry records.}
   \label{f:count-analysis-errors}
@@
     \end{tabular}
-    \caption{Analysis modes across the 1,504,736 records}
+    \caption{Analysis modes across the 1,504,736 records.}
     \label{f:total-records}
@@
     \end{tabular}
-    \caption{Analysis mode(s) for the 347,598 sessions}
+    \caption{Analysis mode(s) for the 347,598 sessions.}
     \label{f:total-sessions}
@@
 
-  \caption{Overview of type analysis modes}
+  \caption{Overview of type analysis modes.}
   \label{f:dataset-overview}
@@
   %% NOTE: popularity looks similar when [over all records] vs. [restricted to module switch records]
-  \caption{Type error popularity for \mnonstrict{} and \mstrict{} modes.
-    All of the type errors in \mnocheck{} mode are \code{SyntaxError}s.}
+  \caption{Type error popularity for \mnonstrict{} and \mstrict{} modes.  All of the type errors in \mnocheck{} mode are \code{SyntaxError}s.}
   \label{t:type-error-count}
@@
 The left column in each row names one kind of type error
-that overlaps with an edit region.
+that overlaps with an edit range.
 The right column shows what percentage of all overlapping
@@
 Fortunately, the data rarely contains too-complex errors.
-None of these errors appear in edit regions~(\cref{t:type-error-count}).
+None of these errors appear in edit ranges~(\cref{t:type-error-count}).
 For the specific case of \code{CodeTooComplex}, our telemetry tracked project-wide
@@
   \includegraphics[width=\columnwidth]{img/error-count-strict-row--te-density-diff.pdf}
-  \caption{Changes over time (seconds) to type error density (total errors / lines in project)}
+  \caption{Changes over time (seconds) to type error density (total errors / lines in project).}
   \label{f:error-density}
@@
   \end{tabular}
-  \caption{FILL Are there fewer type errors than forcedstrict errors? Expect true always, but that is not the case.}
+  \caption{Comparing the change in type errors to the change in forced strict errors.}
   \label{f:tefs-compass}
@@
 
-  \caption{Example sessions}
+  \caption{Two example sessions.}
   \label{f:indy-session}
@@
 
+We are sampling sessions. Could miss interesting ones.
+
+Sampling by keystroke means different things for different creators.
+Depends how they type.
+Scripted interactions that generate a large number of keystrokes per second
+can swamp our data and are hard to distinguish from genuine interactions.
+Copy paste might just look like a bot.
+\QALAN{} I think you said a paste counts as 1 keystroke; is that right?
+
+We have no knowledge whatsoever about interesting events like save run publish.
+Publish events are private.
+
+Changes to errors in the edit range are hard to follow.
+Could add one error but remove another of the same kind,
+then we can't distinguish from ignoring the error.
+
+Edit ranges are coarse, merely the top and bottom edit.
+No idea what happened in between.
+Long time between keystroke events makes this even worse.
+
 Telemetry cannot replace user studies, as there is no way to measure
 creator sentiment, but provide complementary data at scale.
+No idea about intent is a big problem, we don't know which error the creator wanted to fix.
+
+Type analysis reports many errors at once.
+No way to know what the creator is looking at.
+Not even sure they have the analysis widget open.
+
+Internal validity?
+
+External validity?
+
+Construct validity?
+
+Conclusion validity
+
+Ecological validity is high.
 
@@
 \begin{table}[t]\centering
-  \caption{Comparison of telemetry designs}
+  \caption{Comparison of telemetry designs. Two orthogonal concerns are when to collect data and whether to add noise to it.}
   \label{t:telemetry-design}
@@
   \begin{tabular}{l@{~~}c@{~~}c@{~~}c@{~~}c@{~~}c}
-    ~             & Event Counts & Random  & Timestamps & Error Msgs. & Source Code \\\midrule
-    Roblox Studio & \chkYes      & \chkYes & \chkYes    & \chkNo        & \chkNo    \\
-    Transparent   & \chkYes      & \chkNo  & \chkNo     & \chkNo        & \chkNo    \\
-    Classic       & \chkYes      & \chkNo  & \chkYes    & \chkYes       & \chkYes   \\
-%% orthogonal %%    & Diff.Priv
-%% orthogonal %%    & CBI
+    ~             & Event Counts & Timestamps & Error Msgs. & Source Code \\\midrule
+    Roblox Studio & \chkYes      & \chkYes    & \chkNo        & \chkNo    \\
+    Transparent   & \chkYes      & \chkNo     & \chkNo        & \chkNo    \\
+    Classic       & \chkYes      & \chkYes    & \chkYes       & \chkYes   \\
   \end{tabular}

### 2023-09-24 programming23/paper.tex
split type err / forced strict by mode

@@
 
-\subsection{Forced Strict Errors by Mode}
-
-FILL
-
-How do the 72 million forced strict divide across modes?
-
-What about type errors, same distribution?!
+\subsection{Errors by Mode}
+
+\begin{figure}[t]
+  %% strict
+  %%  te total 22771
+  %%  mean 6.9 std 26.24592591869705 med 1 max 385
+  %% 
+  %%  fs total 114259
+  %%  mean 34.9 std 155.41955823164017 med 1 max 4789
+  %% 
+  %% nonstrict
+  %%  te total 362964
+  %%  mean 2.3 std 66.44456626891912 med 0 max 13511
+  %% 
+  %%  fs total 7546175
+  %%  mean 48.6 std 270.60578740369266 med 3 max 16712
+
+  \begin{tabular}{l}
+    600K type errors         = \pct{35} \mnocheck{}, \pct{61} \mnonstrict{}, \pct{4} \mstrict{}
+    \\
+    72M forced scrict errors = \pct{89} \mnocheck{}, \pct{10} \mnonstrict{}, \pct{1} \mstrict{}
+    %% really 0.16 strict, very low!!!
+  \end{tabular}
+  \caption{}
+  \label{f:error-by-mode}
+\end{figure}
+
+\Cref{f:error-by-mode} shows how the type errors and forced strict errors
+in the data are split across type analysis modes.
+\FILL{} add averages.
+
+Most type errors come from \mnonstrict{} mode, despite it having an order
+of magnitude fewer users that \mnocheck{}.
+This is unsurprising because \mnocheck{} reports only syntax errors,
+whereas \mnonstrict{} reports several others kinds~(detailed in the next section).
+Finally, \mstrict{} mode accounts for a small fraction of type errors, despite
+being more picky, because so few creators use it.
+
+Forced strict analysis applies the same typechecker to every module, regardless
+of mode, in the hope of enabling IDE tools.
+Consequently, the breakdown of forced strict errors matches the overall usage
+data~(\cref{f:dataset-overview}): almost \pct{90} for \mnocheck{},
+\pct{10} for \mnonstrict{}, and \pct{1} for \mstrict{}.
+Type analysis finds a large number of errors in \mnocheck{} code;
+unfortunately, we have no insight into whether those errors are true
+positives or whether the analysis led to useful IDE suggestions.
 
@@
 \code{UnificationTooComplex}.
-A casual user should never see these errors.\todo{IDEALLY never see?
-  They CAN see them, right?}
+These errors can appear in Script Analysis just like any other
+error, but one design goal of Luau is that code rarely hits the limits.
+A typical creator should never see these errors.
 

### 2023-09-24 programming23/paper.tex
fix a mode-upgrades todo

@@
 
-Similarly, mode downgrades have only a small negative effect on the
+In the other direction, mode downgrades have only a small negative effect on the
 number of errors (mean -0.3, stddev 4, median 0, max -48).
 %% curious outlier: downgrade added 30 errors!
-Creators apparently tend switch modes only when the code is already
-well typed, or close to it.\todo{What incentive is there to do so?}
+We would see a much larger effect here if creators used downgrades
+as a quick way to silence the typechecker.
+But, creators evidently switch modes only when the code is already
+in well-typed shape.
 
@@
 
-\section{Conclusion}
+\section{Lessons Learned}
 \label{s:conclusion}

### 2023-09-24 programming23/paper.tex
quick sec 5 notes

@@
 For the most part, we omit the strange data from futher
-analysis.\todo{I'm inclined to rewrite this to be much more concise.}
+analysis.\todo{I'm inclined to rewrite this to be much more concise.}\todo{agreed}
 
@@
   %%  mean 48.6 std 270.60578740369266 med 3 max 16712
+  %%
+  %% nocheck
+  %%  te total 176309
+  %%  mean 0.13 std 1.9040337635815197 med 0 max 289
+  %% 
+  %%  fs total 63784284
+  %%  mean 47.7 std 885.8225471699477 med 1 max 251131
 
@@
 positives or whether the analysis led to useful IDE suggestions.
+
+\FILL{} average number of FS errors is
+worst in \mnonstrict{}, 49, vs 48 nocheck and 7 strict.
+median is 3 NS, 1 nocheck, 1 strict.
+We might optimistically think that nocheck accumulates
+lots of errors, but no that does not seem to be the case,
+NS and nocheck both apparently fine with latent FS errors.
 
@@
 immediately introduces another on a different line.
+\FILL{} beware, don't say this same thing Nx throughout the paper.
 \Cref{t:type-error-survival} reports this pair of edits as a Keep rather than
@@
 
+... very concerning if this is the top takeaway
+
+\paragraph{Good: No Apparent Turmoil With Upgrades / Downgrades}
+
+But, mode switches are very rare!
+Really too soon to draw any conclusion, this is only a lack of bad.
+
 \paragraph{Do Types Improve Quality of Life?}
@@
 TODO nocheck accumulated fs errors?
+Need this (density fs) to say more about whether errors get fixed,
+same goes for fig 10.
+
+\paragraph{... Something About Add/Drop in Edit Range}
+
+\paragraph{Bad: Strict too Strict}
+
+Comparing NS vs S, the syntax error \% is about the same (15 vs 9)
+but the type mismatch \% is far off (2 vs 20).
+S is probably noisy.
+
+Data model is probably part of this.
 
@@
 \end{verbatim}
+
+\paragraph{Big Bad: Low Adoption}
+
+poor advertising,
+unclear value proposition,
+nocheck code is apparently doing fine despite lots of forced strict errors,
+... can we say this outright?!
 

### 2023-09-25 programming23/paper.tex
conclusion

@@
 \newcommand{\mstrict}{\modefont{strict}}
+\newcommand{\FS}{background}
 \newcommand{\zerowidth}[1]{\makebox[0pt][l]{#1}}
@@
 
+%% TODO migratory
+
 In this paper, we investigate methods for measuring the effectiveness
@@
 or collects personally identifiable information~(PII),
+%% TODO careful, proprietary matters too!!!
 we performed a large-scale study using \panon{} \emph{telemetry}.
@@
 % https://create.roblox.com/docs/scripting/luau
+
+% TODO
+% \begin{table}
+%   \caption{Modes and analyses}
+%   \label{t:mode-analyze}
+% 
+%   Mode: nocheck ...
+%   Analysis: nocheck ... forced strict
+% 
+% \end{table}
 
@@
 
-\section{Lessons Learned}
+\section{Reflections and Lessons Learned}
 \label{s:conclusion}
 
-Quickly summarize findings:
-
+This paper presents the first large-scale analysis of developers' interactions with
+an industrial strength typechecker.
+With +1.5 million telemetry records from over three hundred thousand sessions
+that took place during a three-month timespan, the data has plenty to say about
+type errors and their overlap with program edits.
+At the same time, the data is carefully restricted to avoid leaking personal
+and proprietary information: we know the number and kind of type errors
+that occurred, but nothing about the line(s) of code that the typechecker
+complained about.
+
+On the object level, the analysis offers several lessons for the Luau
+developers and other maintainers of gradually-typed languages:
 \begin{itemize}
   \item
-    \pct{50} of errors in current module, so probably errors appear locally (not in a distant module);
-    \pct{5} of (current? old??) type errors overlap with edit range, so they don't stick around.
-  \item
-    \pct{90} \mnocheck{},  \pct{10} \mnonstrict{},  \pct{0.1} \mstrict{}, wow.
-  \item
-  \item
-  \item
-    DM awareness is an important and surprising gap between strict
-    and forced strict, undoubtedly a usability barrier for \mstrict{}.
-  \item
+    Nearly all creators (\pct{90}) use the default ``no check''
+    typechecking mode~(\cref{f:dataset-overview}).
+    Since the next typing mode (\mnonstrict{}) requires no annotations and is
+    very conservative about the errors it reports, we conjecture that
+    developers would have no objections to using it and are simply unaware that
+    this mode exists.
+    \QALAN{} nonstrict requires no annotations, right?
+%  \item
+%    \pct{50} of all errors are in current module, so probably errors appear locally (not in a distant module);
+%    \pct{5} of (current? old??) type errors overlap with edit range, so they don't stick around.
+  \item
+    Creators tend to fix them immediately rather than dealing with them in a batch
+    or ignoring the errors altogether~(\cref{f:error-density,f:error-changes}).
+    % Changes to the number of errors balance one another~(\cref{f:error-density});
+    % edits are equally likely to add or remove errors~(\cref{f:error-changes}).
+  \item
+    All Luau scripts interact with data assets, but no typechecking mode
+    has a precise view of the data.
+    Improving precision is an important direction for future work.
+    Until then, the \mstrict{} mode clearly needs to relax its data-asset
+    analysis because it raises even more errors than \FS{} checking~(\cref{f:tefs-compass}).
 \end{itemize}
 
-Quick regrets, lessons learned, advice for next time
-
-\paragraph{Future work: Low adaption}
-
-why so few using nonstrict and strict?
-needs advertising?
-needs usability?
-
-\paragraph{Do we need timestamps?}
-
-FILL these were helpful for the overview
-
-Our telemetry records include timestamps.
-While timestamps are not personal data per se, they are in the ballpark.
-Telemetry would be even more private without them.
-Do we need them?
-
-Timestamps are not the focus of our analysis, but they do play a critical role.
-Consider~\cref{f:error-density}.
-With timestamps, it is clear that error density stays close to zero.
-Without timestamps, we would be left with the counts at the bottom
-of~\cref{f:error-density}.
-Many distributions could fit those counts, such as a ``mountain'' session that first
-grows and grows the number of type errors then shrinks and shrinks down to zero.
-
+On the meta level, the main takeaway is that the lightweight, counter-based telemetry
+in Roblox Studio supports a variety of useful analyses:
+\begin{itemize}
+  \item
+    Counting specific type errors helps to identify trends and to confirm that
+    undesirable events rarely happen (e.g.,\code{UnificationTooComplex}).
+  \item
+    Testing the relationship between two counts can yield insights;
+    in~\cref{f:tefs-compass}, the gap between \mstrict{} vs \FS{} errors
+    revealed an issue with \mstrict{} checking.
+  \item
+    Reporting the overlap between type errors and the current (approximate) edit
+    range lets us assess the outcome of edits without revealing any source code.
+    It was especially useful to know the overlap for old and for current type errors.
+  \item
+    Timestamps give useful, low-level insights about the frequency of keystrokes,
+    the length of sessions, and actions over time~(\cref{f:indy-session}).
+    The event ordering that timestamps provide was invaluable; e.g., for learning
+    that error deltas bounce up and down~(\cref{f:error-density}).
+\end{itemize}
+
+One aspect of the telemetry that we would change for next time is to provide
+more metadata.
+Luau recently adopted semantic subtyping to reduce false-positive type
+errors~\cite{luau-subt}, but there is no way to tell if this
+internal extension is enabled without an extra metadata field.
+Two other aspects are worth rethinking.
+First, it would be useful to have fine-grained counts for \FS{} analysis
+errors and to know the specific reason behind errors such as \code{TypeMismatch}.
+But, this additional data could double the size of telemetry records.
+Second, tracking edit ranges made the telemetry system much more difficult to
+build and maintain.
+It would be much easier to track old and current errors at the module level,
+but it remains to be seen how coarse of an approximation that would be.
+
+In future work, it would be interesting to build statistical models of
+the ``average'' programmer using each typing mode and compare their error
+rates.
+It would be possible to start modeling with our \mnocheck{} and \mnonstrict{}
+data, but we have too few \mstrict{} records at the moment due to our uniform
+sampling rate.
+More data is needed  for a robust model of \mstrict{} programmers.
+
+%% PS downloading from Kibana was a HUGE pain and source of errors. Awful!
+
+%% -----------------------------------------------------------------------------
 \acks

### 2023-09-26 programming23/paper.tex
try FS error density, uhoh its an even split 1/3 each

@@
   FILL specifics.
+  %% TODO remove semantic subtyping
   Our telemetry design is of broad interest, as it gathers useful
@@
 
-%% TODO migratory
+%% TODO Luau is migratory, clearer about this
 
@@
 
-% TODO
+% TODO quick reference table
 % \begin{table}
@@
 \begin{figure}[t]\centering
-  % TODO try diff vs first record instead of previous?
+  %% FS plots look similar
+  %% - much higher densities, same basic story
+  %% - nocheck indeed the highest > NS > S
 
@@
   \includegraphics[width=\columnwidth]{img/error-count-strict-row--te-density-diff.pdf}
-  \caption{Changes over time (seconds) to type error density (total errors / lines in project).}
+  \caption{Changes in type error density (errors in module / lines in module) over time (seconds).}
   \label{f:error-density}
 \end{figure}
-
-\FILL{} what does the FS plot look like? is that a better nocheck-vs-rest comparison??
 
@@
   % TODO use bars instead, percent only, don't care about counts
-  %% TODO old version had +-20% add/drop and a bigger % of keeps. What changed?
+
+  % TE: nocheck & 48378 & [\pct{46.83}] & 9440 & [\pct{9.14}] & 45479 & [\pct{44.03}] \\
+  % TE: nonstrict & 19491 & [\pct{39.63}] & 9567 & [\pct{19.45}] & 20121 & [\pct{40.91}] \\
+  % TE: strict & 733 & [\pct{39.18}] & 368 & [\pct{19.67}] & 770 & [\pct{41.15}] \\
+  %
+  % TE mod: nocheck & 20085 & [\pct{49.83}] & 906 & [\pct{2.25}] & 19317 & [\pct{47.92}] \\
+  % TE mod: nonstrict & 13030 & [\pct{40.22}] & 5513 & [\pct{17.02}] & 13852 & [\pct{42.76}] \\
+  % TE mod: strict & 419 & [\pct{39.31}] & 230 & [\pct{21.58}] & 417 & [\pct{39.12}] \\
+  %
+  % FS: nocheck & 270763 & [\pct{36.99}] & 238696 & [\pct{32.61}] & 222588 & [\pct{30.41}] \\
+  % FS: nonstrict & 35330 & [\pct{37.28}] & 29483 & [\pct{31.11}] & 29955 & [\pct{31.61}] \\
+  % FS: strict & 574 & [\pct{33.71}] & 561 & [\pct{32.94}] & 568 & [\pct{33.35}] \\
+  %
+  % FS mod: nocheck & 259865 & [\pct{39.09}] & 175643 & [\pct{26.42}] & 229271 & [\pct{34.49}] \\
+  % FS mod: nonstrict & 33988 & [\pct{39.74}] & 20738 & [\pct{24.25}] & 30798 & [\pct{36.01}] \\
+  % FS mod: strict & 488 & [\pct{36.53}] & 344 & [\pct{25.75}] & 504 & [\pct{37.72}] \\
   \begin{tabular}{lr@{}rr@{}rr@{}r}
     & \zerowidth{Add} & & \zerowidth{Keep} & & \zerowidth{Drop} \\\midrule
-    \mnocheck{} & 48378 & [\pct{46.83}] & 9440 & [\pct{9.14}] & 45479 & [\pct{44.03}] \\
-    \mnonstrict{} & 19491 & [\pct{39.63}] & 9567 & [\pct{19.45}] & 20121 & [\pct{40.91}] \\
-    \mstrict{} & 733 & [\pct{39.18}] & 368 & [\pct{19.67}] & 770 & [\pct{41.15}] \\
-  \end{tabular}
-  \caption{How often do edits increase, maintain, or decrease the number of type errors?}
+    \mnocheck{}   & 20085 & [\pct{49.83}] & 906 & [\pct{2.25}] & 19317 & [\pct{47.92}] \\
+    \mnonstrict{} & 13030 & [\pct{40.22}] & 5513 & [\pct{17.02}] & 13852 & [\pct{42.76}] \\
+    \mstrict{}    & 419 & [\pct{39.31}] & 230 & [\pct{21.58}] & 417 & [\pct{39.12}] \\
+  \end{tabular}
+  \caption{How often do edits increase, maintain, or decrease the number of errors?}
   \label{f:error-changes}
@@
 
-\section{Reflections and Lessons Learned}
+\section{Lessons Learned and Reflections}
 \label{s:conclusion}
@@
 rates.
-It would be possible to start modeling with our \mnocheck{} and \mnonstrict{}
-data, but we have too few \mstrict{} records at the moment due to our uniform
+While one could start modeling with the \mnocheck{} and \mnonstrict{}
+data, but there are too few \mstrict{} records at the moment due to our uniform
 sampling rate.
-More data is needed  for a robust model of \mstrict{} programmers.
-
-%% PS downloading from Kibana was a HUGE pain and source of errors. Awful!
+More data is needed to experiment with models of \mstrict{}-mode programmers.
+
+%% PS
+%% - downloading from Kibana was a HUGE pain and source of errors. Awful!
+%% - worth finding ways to reduce / ignore syntax errors?
 

### 2023-09-26 programming23/paper.tex
checkpoint: sec 5 next

@@
 
-\paragraph{Reflections}
+\paragraph{NOTES} %% FILL
 This kind of telemetry data is coarse-grained.
@@
   % FS mod: strict & 488 & [\pct{36.53}] & 344 & [\pct{25.75}] & 504 & [\pct{37.72}] \\
-  \begin{tabular}{lr@{}rr@{}rr@{}r}
-    & \zerowidth{Add} & & \zerowidth{Keep} & & \zerowidth{Drop} \\\midrule
-    \mnocheck{}   & 20085 & [\pct{49.83}] & 906 & [\pct{2.25}] & 19317 & [\pct{47.92}] \\
-    \mnonstrict{} & 13030 & [\pct{40.22}] & 5513 & [\pct{17.02}] & 13852 & [\pct{42.76}] \\
-    \mstrict{}    & 419 & [\pct{39.31}] & 230 & [\pct{21.58}] & 417 & [\pct{39.12}] \\
+  \begin{tabular}{l@{~}lr@{}rr@{}rr@{}r}
+    & & \zerowidth{Add} & & \zerowidth{Keep} & & \zerowidth{Drop} \\\midrule
+    type err. & \mnocheck{}   & 20085 & [\pct{49.83}] & 906 & [\pct{2.25}] & 19317 & [\pct{47.92}] \\
+              & \mnonstrict{} & 13030 & [\pct{40.22}] & 5513 & [\pct{17.02}] & 13852 & [\pct{42.76}] \\
+              & \mstrict{}    & 419 & [\pct{39.31}] & 230 & [\pct{21.58}] & 417 & [\pct{39.12}] \\
+    \\
+    \FS       & \mnocheck{}   & 259865 & [\pct{39.09}] & 175643 & [\pct{26.42}] & 229271 & [\pct{34.49}] \\
+              & \mnonstrict{} & 33988 & [\pct{39.74}] & 20738 & [\pct{24.25}] & 30798 & [\pct{36.01}] \\
+              & \mstrict{}    & 488 & [\pct{36.53}] & 344 & [\pct{25.75}] & 504 & [\pct{37.72}] \\
   \end{tabular}
@@
 
-\FILL{} good things first,
-then regrets about the noisy stochiastic telemetry,
-then lessons?
-then threats.
+To sum up, the data supports the following conclusions:
+
+\subsection{The Good News}
 
@@
 
-... very concerning if this is the top takeaway
+um ... very concerning if this is the top takeaway
 
@@
 
-TODO nocheck accumulated fs errors?
-Need this (density fs) to say more about whether errors get fixed,
-same goes for fig 10.
+Uh oh, no accumulated FS errors in nocheck.
+Edits vs FS counts are the same in all three modes.
 
@@
 
+\FILL{}
+
+\subsection{The Bad News}
+
 \paragraph{Bad: Strict too Strict}
+
+\FILL{}
 
@@
 Data model is probably part of this.
-
-\paragraph{Bad: Noisy Data}
-
-\begin{verbatim}
- require(foobar)
-  foobar is a module script in the data model
- require(foo.bar.b)
-  - could be edit in progress
-  - could be renamed module
-  - 
-\end{verbatim}
 
@@
 Copy paste might just look like a bot.
-\QALAN{} I think you said a paste counts as 1 keystroke; is that right?
+%% \QALAN{} I think you said a paste counts as 1 keystroke; is that right?
 
@@
     this mode exists.
-    \QALAN{} nonstrict requires no annotations, right?
 %  \item
@@
 Luau recently adopted semantic subtyping to reduce false-positive type
-errors~\cite{luau-subt}, but there is no way to tell if this
-internal extension is enabled without an extra metadata field.
-Two other aspects are worth rethinking.
+errors~\cite{luau-subt}, but the current telemetry has no way to tell if this
+internal extension is enabled or not.
+Three other aspects are worth rethinking.
 First, it would be useful to have fine-grained counts for \FS{} analysis
@@
 But, this additional data could double the size of telemetry records.
-Second, tracking edit ranges made the telemetry system much more difficult to
-build and maintain.
-It would be much easier to track old and current errors at the module level,
-but it remains to be seen how coarse of an approximation that would be.
+Second, although tracking edit ranges was useful, it made the telemetry system
+much more difficult to build and maintain.
+Tracking old and current errors at the module level would be more maintainable,
+but there is a risk that it is too coarse.
+Third, the massive number of syntax errors begs the question of how to reduce
+them.
+Shifting focus from arbitrary keystrokes to selected ones, such as closing
+parentheses or whitespaces, might increase the likelihood of well-formed
+code without losing the ``middle of things'' nature of the data.
+Another option is to ignore records that have a syntax error in the edit range.
 
@@
 More data is needed to experiment with models of \mstrict{}-mode programmers.
+
+% \paragraph{Bad: Noisy Data}
+% 
+% \begin{verbatim}
+%  require(foobar)
+%   foobar is a module script in the data model
+%  require(foo.bar.b)
+%   - could be edit in progress
+%   - could be renamed module
+%   - 
+% \end{verbatim}
 

### 2023-09-27 programming23/paper.tex
related work ++
sec5 threats

@@
 However, creators may write their own types, and they can opt in
-to receiving ype error reports.
+to receiving type error reports.
 
@@
   \item[RQ3.]
-    What impact does type analysis have on the number of background type
+    What impact does type analysis have on the number of \FS{}
     errors?
-    For example, do background errors pile up in unanalyzed projects?
+    For example, do \FS{} errors pile up in unanalyzed projects?
 \end{description}
@@
 limited to a small number of users.
-
-FILL can stats answer questions that point-by-point analysis cannot?
 
@@
   \end{tabular}
-  \caption{}
+  \caption{FILL}
   \label{f:error-by-mode}
@@
   \item
+    \FILL{} which survive the longest, most keeps???
+
+  \item
     One striking aspect of~\cref{t:type-error-survival} is the large number
@@
   %% NOTE: popularity looks similar when [over all records] vs. [restricted to module switch records]
-  \caption{Type error popularity for \mnonstrict{} and \mstrict{} modes.  All of the type errors in \mnocheck{} mode are \code{SyntaxError}s.}
+  \caption{Type error popularity for \mnonstrict{} and \mstrict{} modes (all of the type errors in \mnocheck{} mode are \code{SyntaxError}s).}
   \label{t:type-error-count}
@@
 
+%% UNUSED CODES
+%% \mnonstrict{}
+%%   CodeTooComplex
+%%   DeprecatedApiUsed
+%%   DuplicateGenericParameter
+%%   DynamicPropertyLookupOnClassesUnsafe
+%%   FunctionRequiresSelf
+%%   IncorrectGenericParameterCount
+%%   InternalError
+%%   NormalizationTooComplex
+%%   OccursCheckFailed
+%%   SwappedGenericTypeParameter
+%%   TypePackMismatch
+%%   UnificationTooComplex
+%% 
+%% \mstrict{}
+%%  CodeTooComplex
+%%  DeprecatedApiUsed
+%%  DuplicateGenericParameter
+%%  DuplicateTypeDefinition
+%%  DynamicPropertyLookupOnClassesUnsafe
+%%  FunctionDoesNotTakeSelf
+%%  FunctionRequiresSelf
+%%  InternalError
+%%  MissingUnionProperty
+%%  NormalizationTooComplex
+%%  OnlyTablesCanHaveMethods
+%%  SwappedGenericTypeParameter
+%%  TypePackMismatch
+%%  UnificationTooComplex
+
+%% both NS S
+%%  CodeTooComplex
+%%  DeprecatedApiUsed
+%%  DuplicateGenericParameter
+%%  DynamicPropertyLookupOnClassesUnsafe
+%%  FunctionRequiresSelf
+%%  InternalError
+%%  NormalizationTooComplex
+%%  OnlyTablesCanHaveMethods
+%%  SwappedGenericTypeParameter
+%%  TypePackMismatch
+%%  UnificationTooComplex
+
 \Cref{t:type-error-count} reports the percent of
@@
   \item
+    \FILL{} huge syntax error gap between S nonstrict.
+
+  \item
     \code{TypeMismatch} is far more common in \mstrict{} (\pct{20.46})
@@
   \includegraphics[width=\columnwidth]{img/error-count-nocheck-row--te-density-diff.pdf}
-  \medskip
+  \smallskip
+
   \mnonstrict{}
   \includegraphics[width=\columnwidth]{img/error-count-nonstrict-row--te-density-diff.pdf}
-  \medskip
+  \smallskip
+
   \mstrict{}
@@
 
-\section{Discussion}
+\section{Answers to Research Questions}
 \label{s:discussion}
 
-To sum up, the data supports the following conclusions:
-
-\subsection{The Good News}
-
-\paragraph{Good: No Internal Error}
-
-FILL hope to never see \code{InternalError}.
-Have zero, yay!
-
-um ... very concerning if this is the top takeaway
-
-\paragraph{Good: No Apparent Turmoil With Upgrades / Downgrades}
-
-But, mode switches are very rare!
-Really too soon to draw any conclusion, this is only a lack of bad.
-
-\paragraph{Do Types Improve Quality of Life?}
-
-\FILL{} compare \mnonstrict{} and \mstrict{}.
-Use edit range~(\cref{t:type-error-survival} for NS vs S)
-and error density~(\cref{f:error-density})
-
-Uh oh, no accumulated FS errors in nocheck.
-Edits vs FS counts are the same in all three modes.
-
-\paragraph{... Something About Add/Drop in Edit Range}
-
-\FILL{}
-
-\subsection{The Bad News}
-
-\paragraph{Bad: Strict too Strict}
-
-\FILL{}
-
-Comparing NS vs S, the syntax error \% is about the same (15 vs 9)
-but the type mismatch \% is far off (2 vs 20).
-S is probably noisy.
-
-Data model is probably part of this.
-
-\paragraph{Big Bad: Low Adoption}
-
-poor advertising,
-unclear value proposition,
-nocheck code is apparently doing fine despite lots of forced strict errors,
-... can we say this outright?!
-
-\subsection{Threats to Validity}
+With the data in hand, we can answer the research questions
+from~\cref{s:introduction}.
+The answers confirm several aspects of Luau's design, but
+at the same time offer several directions for future improvements.
+Luau clearly has a powerful typechecker that can efficiently
+analyze thousands of lines of code.
+But, few creators use type analysis directly and the data presents little
+evidence that doing so would improve their quality of life.
+We conjecture that Luau needs further tailoring to match
+common idioms in Roblox scripts.
+Discovering these idioms is a question for future work~(\cref{s:conclusion}).
+
+\begin{description}
+  \item[How many creators choose to use type analysis?]
+    A mere \pct{10} of creators opt-in to type checking.
+    Most of these use \mnonstrict{} mode, and fewer than \pct{1}
+    use \mstrict{} mode.
+
+  \item[How often do projects contain modules with different analysis modes?]
+    Projects rarely combine analysis modes.
+    Only 512 of \nK{340} sessions ($<\pct{1}$) switched to modules with different
+    analysis modes.
+    This number may be a significant underapproximation because we have data
+    only on modules that the creator chose to open, but the widespread use
+    of \mnocheck{} mode suggests that it is not off by much.
+
+  \item[How often do creators turn analysis off?]
+    Creators rarely disable type analysis after opting in.
+    Only 320 of the \nK{340} sessions ($<\pct{1}$)
+    switch to a weaker mode.
+    \FILL{} are they downgrading from S or NS?
+    %% FILL more insight if we match upgrades + downgrades in session?
+
+  \item[For modules that use type analysis:]\leavevmode
+    \begin{description}
+      \item[Which errors arise?]
+        Syntax errors are extremely common~(over \pct{50},~\cref{t:type-error-count}), 
+        followed by type mismatches~(\pct{20} in \mstrict{}),
+        arity errors~(\pct{2}),
+        and failure to unpack on optional value~(\pct{2}).
+        There are several uncommon errors, such as misusing a table operation,
+        and a few that never arise: internal and too-complex errors,
+        duplicating or swapping a generic parameter, use of a deprecated API,
+        and unsafe dynamic property lookup on a class.
+
+      \item[How do creators respond? Which errors persist through edits?]
+        Creators typically fix analysis errors.
+        With few exceptions, increases and decreases in the number of errors
+        balance out~(\cref{f:error-changes,t:type-error-survival,}).
+        Even in \mnocheck{}, most edits that overlap with syntax errors
+        remove at least one error~(\pct{67}) rather than increase or maintain
+        the total.
+        The errors that most-often survive edits are due to
+        inference at a binary operation,
+        sending \code{self} to a function that does not expect it,
+        and failure to unpack an optional value.
+        But the first two seldom arise, and the third is something creators
+        might reasonably ignore while prototyping.
+
+%% Keep percents
+%% '((CannotInferBinaryOperation "0.50")
+%%   (OptionalValueAccess "0.44")
+%%   (FunctionDoesNotTakeSelf "0.40")
+%%   (CannotExtendTable "0.35")
+%%   (UnknownPropButGotLikeProp "0.28")
+%%   (UnknownProperty "0.26")
+%%   (MissingProperties "0.25")
+%%   (FunctionExitsWithoutReturn "0.21")
+%%   (ModuleHasCyclicDependency "0.20")
+%%   (UnknownRequire "0.18")
+%%   (TypeMismatch "0.18")
+%%   (UnknownSymbol "0.15")
+%%   (NotATable "0.12")
+%%   (GenericError "0.11")
+%%   (CannotCallNonFunction "0.10")
+%%   (CountMismatch "0.09")
+%%   (GenericExtraInformation "0.07")
+%%   (IllegalRequire "0.06")
+%%   (SyntaxError "0.01")
+
+    \end{description}
+
+  \item[What impact does type analysis have on the number of \FS{} errors?]
+    Opting in to type analysis has little impact on \FS{} errors.
+    The percent of \FS{} errors in \mnocheck{} code matches the percent of
+    creators using \mnocheck{} mode~(\cref{f:error-by-mode}); whereas,
+    if type analysis helped reduce \FS{} errors, \mnocheck{} would own a
+    higher percentage of the errors.
+    Furthermore, there is no apparent relation between \FS{} errors and edits,
+    regardless of analysis mode~(\cref{f:error-changes}).
+    A randomly-chosen edit has a 1/3 chance of increasing, decreasing, or
+    maintaining the number of \FS{} errors.
+
+\end{description}
+
+\paragraph{Threats to Validity}
 \label{s:threats}
 
-Error counts per-project and per-module include syntax errors.
-Problematic, because syntax errors are uninteresting (easy to fix),
-but common, which makes it likely that our per-keystroke telemetry (1/200) will record them.
-Edit range has specific counts.
-Next time, we should do that for the other aggregates.
-
-We are sampling sessions. Could miss interesting ones.
-
-Sampling by keystroke means different things for different creators.
-Depends how they type.
-Scripted interactions that generate a large number of keystrokes per second
-can swamp our data and are hard to distinguish from genuine interactions.
-Copy paste might just look like a bot.
-%% \QALAN{} I think you said a paste counts as 1 keystroke; is that right?
-
-We have no knowledge whatsoever about interesting events like save run publish.
-Publish events are private.
-
-Changes to errors in the edit range are hard to follow.
-Could add one error but remove another of the same kind,
-then we can't distinguish from ignoring the error.
-
-Edit ranges are coarse, merely the top and bottom edit.
-No idea what happened in between.
-Long time between keystroke events makes this even worse.
-
-Telemetry cannot replace user studies, as there is no way to measure
-creator sentiment, but provide complementary data at scale.
-No idea about intent is a big problem, we don't know which error the creator wanted to fix.
-
-Type analysis reports many errors at once.
-No way to know what the creator is looking at.
-Not even sure they have the analysis widget open.
-
-Internal validity?
-
-External validity?
-
-Construct validity?
-
-Conclusion validity
-
-Ecological validity is high.
+While this study has high ecological validity due to its focus on
+working creators, there are several threats to keep in mind.
+Some threats stem from our method of collecting data; others
+come from the limited scope of the data.
+
+First of all, sampling is necessarily incomplete.
+Though the data includes many thousands of sessions, it
+may have missed a few critical sessions where a creator
+adopted \mstrict{} mode, encountered \code{CodeTooComplex} errors,
+and gave up.
+Furthemore, our method of sampling by keystroke and by module switch
+skews the data toward creators who, for whatever reason, do more
+of these actions.
+We have no access to other events (e.g., run locally, publish) due to privacy
+concerns; future studies may wish to avoid this limitation by actively
+recruiting subjects.
+
+Second, the edit ranges are a coarse approximation of actual edits.
+They begin at the lowest edited line of code and end at the highest
+edited line, even if nothing between those lines changed.
+
+Third, the total counts of type errors include a large number of syntax errors,
+which have little to do with the Luau type system and are presumably easy for
+creators to fix.
+It would be better for the analysis to ignore these errors.
+Indeed, our comparison between \FS{} errors in \mnocheck{} vs.
+\mstrict{} mode~(\cref{f:error-changes}) might tell a very different story
+without all the noise that syntax errors introduce.
+
+Fourth, type analysis reports several errors at once and we have no idea which
+errors, if any, creators intended to fix with their edits.
+Although it appears that creators choose to fix errors,
+the fixes may be a side effect of other plans.
+
+% Telemetry cannot replace user studies, as there is no way to measure
+% creator sentiment, but provide complementary data at scale.
+% No idea about intent is a big problem, we don't know which error the creator wanted to fix.
+
+Lastly, we have two remarks about external validity.
+First, our information on type errors is limited to error names.
+If another language has an error that is roughly similar to the
+\code{CannotCallNonFunction} error but with slightly different
+semantics, then the relative frequency of that error could be different
+than what we observed.
+Second, we know nothing about the creators in our study except that they
+were selected at random from a large and diverse group.
+A dataset based on a targeted subset of users might show entirely different
+characteristics.
 
@@
 \begin{table}[t]\centering
-  \caption{Comparison of telemetry designs. Two orthogonal concerns are when to collect data and whether to add noise to it.}
+  \caption{Comparison of telemetry designs. Two orthogonal concerns are \emph{when} to collect data and \emph{whether} to add noise to enhance privacy.}
   \label{t:telemetry-design}
@@
 time-to-fix~\cite{ab-sigcse-2015}.
+Related prior work used the GRUMPS telemetry system to collect 4.7 million
+keystroke-level actions from learners programming in
+Ada~\cite{tm-iticse-2004,tkdmceg-ascilite-2003}.
+This data is not available.
 
@@
 %\paragraph{TBD}
-%
-%thomas kennedy .... generic usage monitoring ascilite 2003
 %
@@
 More data is needed to experiment with models of \mstrict{}-mode programmers.
+
+\FILL{} future: interviews, accommodate idioms
 

### 2023-09-27 programming23/paper.tex
conclusion: talk alouds, idioms

@@
 errors, if any, creators intended to fix with their edits.
-Although it appears that creators choose to fix errors,
+Although it appears that creators choose to fix errors with their edits,
 the fixes may be a side effect of other plans.
@@
     developers would have no objections to using it and are simply unaware that
-    this mode exists.
+    other modes exist.
 %  \item
@@
   \item
-    Creators tend to fix them immediately rather than dealing with them in a batch
+    Creators tend to fix errors immediately rather than dealing with them in a batch
     or ignoring the errors altogether~(\cref{f:error-density,f:error-changes}).
@@
     Improving precision is an important direction for future work.
-    Until then, the \mstrict{} mode clearly needs to relax its data-asset
+    Until then, \mstrict{} mode clearly needs to relax its data-asset
     analysis because it raises even more errors than \FS{} checking~(\cref{f:tefs-compass}).
@@
   \item
-    Testing the relationship between two counts can yield insights;
-    in~\cref{f:tefs-compass}, the gap between \mstrict{} vs \FS{} errors
+    Testing the relationship between two counts can yield insights.
+    In~\cref{f:tefs-compass}, the gap between \mstrict{} vs \FS{} errors
     revealed an issue with \mstrict{} checking.
@@
     range lets us assess the outcome of edits without revealing any source code.
-    It was especially useful to know the overlap for old and for current type errors.
+    It was especially useful to have overlaps for two sets of type errors: old and current.
   \item
@@
 much more difficult to build and maintain.
-Tracking old and current errors at the module level would be more maintainable,
-but there is a risk that it is too coarse.
+Tracking old and current errors at the module level would be easier,
+though there is a risk that it is too coarse.
 Third, the massive number of syntax errors begs the question of how to reduce
@@
 sampling rate.
-More data is needed to experiment with models of \mstrict{}-mode programmers.
-
-\FILL{} future: interviews, accommodate idioms
-
-% \paragraph{Bad: Noisy Data}
-% 
-% \begin{verbatim}
-%  require(foobar)
-%   foobar is a module script in the data model
-%  require(foo.bar.b)
-%   - could be edit in progress
-%   - could be renamed module
-%   - 
-% \end{verbatim}
+The errors that we do see in \mstrict{} mode call for an in-depth analysis via
+talk-aloud interviews to discover why type mismatches occur and how how Luau
+can better accommodate untyped idioms.
 
@@
 Greenman was supported by
-grant \href{https://nsf.gov/awardsearch/showAward?AWD_ID=2030859&HistoricalAwards=false}{2030859}
+NSF grant \href{https://nsf.gov/awardsearch/showAward?AWD_ID=2030859&HistoricalAwards=false}{2030859}
 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.

### 2023-09-27 programming23/paper.tex
abstract, intro, conclusion

@@
   The standard way to collect feedback at a large scale from creators working on
-  realistic projects is to deploy client-side telemetry; however, we must do
-  so in a way that respects creators' privacy.
-  Whatever telemetry we use to study the Luau language cannot include
-  personally-identifiable information.
-  For example, we cannot monitor source code or error messages as they
-  may contain personal data.
-  Finding ways to harness telemetry in a private way is the central research challenge.
-  %% TODO awkward
+  real projects is to deploy client-side telemetry; however, we cannot
+  scrape personal data or proprietary information, which means we cannot
+  collect source code fragments, error messages, or even filepaths.
+  The research questions are thus how to design a safe telemetry system
+  and whether it can provide useful insights about type errors.
 
@@
   %% What was done that unveiled new knowledge?
-  We designed and implemented a pseudonymized telemetry system for Luau.
+  We designed and implemented a pseudonymized, randomized telemetry system for Luau.
   Telemetry records include a timestamp, a session id, a reason for sending,
-  and a numeric summary of the most recent type analysis.
-  This information lets us follow type analysis over time within a codebase
-  without revealing private information.
-  Over four months in Spring 2023, we deployed the telemetry system and collected
-  1.5 million telemetry records.
+  and a numeric summary of the most recent type analyses.
+  This information lets us study type errors over time
+  without revealing private data.
+  We deployed the system in Roblox Studio during Spring 2023
+  and collected over 1.5 million telemetry records from over 340,000 sessions.
 
@@
   %% capabilities are enabled by the work?
-  We present several findings about Luau.
-  Some confirm hypotheses by the Luau team while others are rather surprising,
-  but all findings support our claim that pseudonymized telemetry is a
-  useful way to collect feedback.
-  One of the less-surprising findings is that types are unpopular:
+  We present several findings about Luau, all of which suggest that telemetry
+  is an effective way to study type error pragmatics.
+  One of the less-surprising findings is that opt-in gradual types are unpopular:
   there is an 100x gap between the number of untyped Luau sessions and the number
   of typed ones.
-  One surprise is that the strict mode for type analysis is very conservative
+  One surprise is that the strict mode for type analysis is overly conservative
   about interactions with data assets.
-  Instead of using the flexible dynamic type, it uses a top type and requires
-  creators to down-cast assets.
-  Lastly, a reassuring finding is that type analysis rarely hits its internal
-  limits on problem size.
+  % Instead of using the flexible dynamic type, it uses a top type and requires
+  % creators to down-cast assets.
+  A reassuring finding is that type analysis rarely hits its internal limits on
+  problem size.
 
@@
   %% or results and evaluation support this work?
-  Our findings are supported by a publicly-available dataset
-  of over 1.5 million telemetry records
-  and by free software for analyzing the data.
+  Our findings are supported by a dataset
+  of over 1.5 million telemetry records.
+  The data and scripts for analyzing it will be available in an artifact.
 
@@
   %% Why does this work matter?
-  Luau is the first large-scale combination of gradual types and semantic
-  subtyping.
   Beyond the immediate benefits to Luau,
   our findings about types and type errors have implications
-  for adoption and ergonomics in other gradual languages.
-  %% mention Elixir, Verse?
-  FILL specifics.
-  %% TODO remove semantic subtyping
-  Our telemetry design is of broad interest, as it gathers useful
-  data without exposing personal information.
+  for adoption and ergonomics in other gradual languages
+  such as TypeScript, Elixir, and Typed Racket.
+  Our telemetry design is of broad interest, as it reports on
+  type errors without revealing sensitive information.
 
@@
 65~million Daily Active Users, and 14~billion hours of engagement in
-April--June 2023.
+April--June 2023~\cite{corp.roblox.com}.
 There are over 5~million experiences available on the platform thanks to a
-worldwide community of 3~million creators~\cite{corp.roblox.com}.
+worldwide community of 3~million creators.
+Many of the creators build experiences for fun or learning and may
+not consider themselves software developers; others are professional
+developers.
 
@@
 an extension of {Lua~5.1~\cite{lua}}.
-The main addition in Luau is a static type system, which infers
+The main addition in Luau is a static type system that infers
 types for all Luau programs on the fly, as creators modify the code.
-These types are used primarily in type-driven tooling such as autocomplete and
-API documentation~\cite{luau-autocomplete}; the typical creator is unaware
-that a type system powers these tools.
-However, creators may write their own types, and they can opt in
-to receiving type error reports.
-
-As discussed in~\cite{bfj-hatra-2021},
-the goals of the {Luau} type system are rather different from
-a traditional type system, which focuses on compilation and memory safety.
-{Luau} has a diverse user community, ranging from
-students in code camps to professional development studios. These
-creators have quite different needs, with different emphases on
-enabling rapid creation and ensuring software quality.
-{Luau} therefore supports three typing modes for user code:
-the hands-off default \mnocheck{},
-a relaxed \mnonstrict{} mode,
-and a rigorous \mstrict{} mode.
-Creators can switch between modes
-gradually~\cite{st-sfp-2006,tfffgksst-snapl-2017}, one module at a time.
-
-%% TODO Luau is migratory, clearer about this
+These types are used primarily in IDE tooling such as autocomplete and
+in API documentation~\cite{luau-autocomplete}, and indeed, creators
+may be unaware that a typechecker is analyzing their code.
+However, creators can opt in to receiving type error reports and they can write
+their own types to guide designs and document their intentions.
+
+Due to the broad community of creators, the goals of the Luau type
+system are rather unique~\cite{bfj-hatra-2021}.
+Whereas a traditional type system focuses on compilation and memory
+safety, Luau takes a lenient approach by default and lets creators
+gradually~\cite{st-sfp-2006,tf-popl-2008} migrate to rigorous checks
+one module at a time.
+In this way, for example,
+students in code camps may begin in the \mnocheck{} mode that reports only
+syntax errors; hobbyists can use \mnonstrict{} mode to identify
+would-be runtime errors; and developers who need to ensure software
+quality can rely on \mstrict{} mode.
+(No matter the mode, a \emph{\FS{}} type analysis still drives the IDE tools.)
+Luau's types aim to support untyped designs, in the spirit of migratory
+typing~\cite{tfffgksst-snapl-2017}, so that creators can switch modes without
+needing to restructure their code.
 
 In this paper, we investigate methods for measuring the effectiveness
-of the {Luau} type system in the development of {Roblox} scripts.
-We want to collect measurements at a large scale, with thousands
-of participants, and we must do so in a way that protects creators' privacy.
+of the {Luau} type system.
+The goal is to collect feedback at a large scale, with thousands
+of participants maintaining real codebases.
+Consequently, the measurements cannot reveal \emph{any} information about
+source code, as it may contain personal data, proprietary algorithms, novel
+game designs, and so on.
 In comparison to prior work~(\cref{s:related}), which
-with few exceptions~\cite{zhlbr-cc-2020,zhlbr-oopsla-2020,hlzbr-ecoop-2021} is either small in scale
-or collects personally identifiable information~(PII),
-%% TODO careful, proprietary matters too!!!
-we performed a large-scale study using \panon{} \emph{telemetry}.
-
-{Roblox Studio} has a telemetry system, which is used to gauge
-the effectiveness of creation features. This system randomly
-determines which sessions should report telemetry, and for those
-sessions, reports telemetry records back with a summary of the
-session. In the case of this study, the telemetry includes data on the
-number of errors at various levels of granularity: in the current edit
-range, in the current file, and in every file which was type
-checked.
-
-The telemetry data we analyzed does not contain any PII.
-It has no source code;
- no source locations;
-no error messages (which may contain source code);
-no error messages;
-no record of the creator's identity, locale, or IP address;
-and no information about what creation the data came from.
-Telemetry records are correlated by session, using a \panon{}
-session identifier.
-
-Most users of {Roblox Studio} do not opt in to type error
-reporting (\pct{90},~\cref{s:data}), and so they do not see the ``squiggly underlining'' that
-indicates a type error site. Nonetheless, the type inference system
-still runs in the background (since it drives autocomplete and other
-type-based tools), letting us see which type errors would have
-been reported and whether users fix these errors over time.
+with few exceptions~(e.g.,\cite{hlzbr-ecoop-2021}) is small in scale
+or collects source code, we performed a large-scale study using \panon{}
+{telemetry}.
+
+Our starting point is a telemetry framework that is built in
+to {Roblox Studio} and currently measures the effectiveness of
+creation features.
+This system randomly determines which sessions should report telemetry, and,
+for those sessions, reports telemetry records back with a summary of the
+session.
+
+In this work, we design telemetry that collects data on type errors without
+exposing source code, source locations, or even error messages.
+The telemetry counts the number and kind of type errors at various
+levels of granularity.
+Furthermore it maintains a client-side approximation of the latest source-code
+edits and uses that to identify type errors that overlap with this edit range.
+Telemetry records are correlated by session, using a \panon{} session
+identifier and a timestamp.
 
@@
 
-Beyond their immediate applications to {Luau},
-answers to these questions have implications for the design
-of gradual types~\cite{st-sfp-2006}, success types~\cite{lindahl2006practical},
+Beyond their immediate revelance to {Luau},
+answers to these questions have broad implications for the design
+of gradually-typed languages.
+Luau represents a large-scale combination of ideas from
+gradual typing~\cite{st-sfp-2006,tfffgksst-snapl-2017,bat-ecoop-2014},
+success typing~\cite{lindahl2006practical},
 and semantic subtyping~\cite{CF05:GentleIntroduction,Jef22:SemanticSubtyping}.
-Luau represents the first large-scale combination of these features,
-and lessons from this experience can inform future applications.
+Lessons from this experience can inform future applications.
 
 At a higher level, this paper is the first to use telemetry 
-to study a typechecker.
+to study a type checker.
 It thus represents a step toward data-driven language design,
@@
 that occured between February and April 2023
-and includes 72 million type analysis errors.
+and covers thousands of type analysis errors
+and millions of \FS{} errors.
 By contrast to typical qualitative methods such as surveys and interviews, it
@@
   \item
-    Design of a low-overhead, (black-box / PII-safe / impersonal)
-    telemetry method, that other
-    researchers can build on.
-
-  \item
-    Lessons from millions of type errors about
-    the adoption of strict type analysis,
-    the usefulness of type errors,
-    and \FILL{}.
-    These findings are especially important for the
-    gradual typing, success typing, and semantic subtyping communities.
-
-  \item
-    \FILL{} any surprises when doing the analysis?
-    Note, no ML for analysis, we do not have a corpus of text.
+    Design of a low-overhead telemetry system
+    that reports on type errors without revealing source code.
+
+  \item
+    Lessons from many thousands of type errors about
+    adoption, persistent errors, and creators' responses.
+
+  \item
+    An available dataset of over 1.5 million telemetry records
+    and scripts to analyze them.
+    \emph{We intend to submit this in an artifact.}
 
 \end{itemize}
-
-The data and the analysis pipeline are freely available.
-\FILL{} link.
 
@@
 
-\begin{figure}[t]\centering
-  \includegraphics[width=.45\textwidth]{img/roblox-studio.png}
-  \includegraphics[width=.45\textwidth]{img/roblox-studio-ide.png}
-
-  \caption{{Roblox Studio 3D creation} tools (left) and IDE (right).}
-  \label{fig:roblox-studio}
-\end{figure}
-
-%% \FILL{} explain the name "creators"?
+%% Most users of {Roblox Studio} do not opt in to type error
+%% reporting (\pct{90},~\cref{s:data}), and so they do not see the ``squiggly underlining'' that
+%% indicates a type error site. Nonetheless, the type inference system
+%% still runs in the background (since it drives autocomplete and other
+%% type-based tools), letting us see which type errors would have
+%% been reported and whether users fix these errors over time.
 
@@
 the type system, it is possible for these edits to introduce type errors.
+
+\begin{figure}[t]\centering
+  \includegraphics[width=.45\textwidth]{img/roblox-studio.png}
+  \includegraphics[width=.45\textwidth]{img/roblox-studio-ide.png}
+  \caption{{Roblox Studio 3D creation} tools (left) and IDE (right).}
+  \label{fig:roblox-studio}
+\end{figure}
 
@@
   \item
+    \FILL{} \mnonstrict{} should have no false positives,
+    \mstrict{} doesn't care.
+    Why so many keeps in NS??!
+
+  \item
     One striking aspect of~\cref{t:type-error-survival} is the large number
@@
 an industrial strength typechecker.
-With +1.5 million telemetry records from over three hundred thousand sessions
+With over 1.5 million telemetry records from over 300,000 sessions
 that took place during a three-month timespan, the data has plenty to say about
@@
     Testing the relationship between two counts can yield insights.
-    In~\cref{f:tefs-compass}, the gap between \mstrict{} vs \FS{} errors
+    The gap between \mstrict{} and \FS{} errors In~\cref{f:tefs-compass}
     revealed an issue with \mstrict{} checking.
@@
     Reporting the overlap between type errors and the current (approximate) edit
-    range lets us assess the outcome of edits without revealing any source code.
+    range lets us assess the outcome of edits without revealing source code.
     It was especially useful to have overlaps for two sets of type errors: old and current.
@@
     The event ordering that timestamps provide was invaluable; e.g., for learning
-    that error deltas bounce up and down~(\cref{f:error-density}).
+    that error deltas bounce up and down rather than rising shaply, then falling~(\cref{f:error-density}).
 \end{itemize}
@@
 Luau recently adopted semantic subtyping to reduce false-positive type
-errors~\cite{luau-subt}, but the current telemetry has no way to tell if this
+errors~\cite{CF05:GentleIntroduction,Jef22:SemanticSubtyping}, but the current
+telemetry has no way to tell if this
 internal extension is enabled or not.
@@
 much more difficult to build and maintain.
-Tracking old and current errors at the module level would be easier,
+Tracking old and current errors at the module level would be far easier,
 though there is a risk that it is too coarse.
-Third, the massive number of syntax errors begs the question of how to reduce
+Third, the massive number of syntax errors begs the question of how to skip
 them.

### 2023-09-28 programming23/paper.tex
draft labels
sec 2 edits

@@
 
+%\titlerunning{DRAFT do not distribute}
+
 % Alphabetical order for authors?
@@
 %\renewcommand{\shortauthors}{...}
+
+%\authorrunning{DRAFT do not distribute}
 
@@
 quality can rely on \mstrict{} mode.
-(No matter the mode, a \emph{\FS{}} type analysis still drives the IDE tools.)
+(No matter the mode, a {\FS{}} type analysis still drives the IDE tools.)
 Luau's types aim to support untyped designs, in the spirit of migratory
 typing~\cite{tfffgksst-snapl-2017}, so that creators can switch modes without
-needing to restructure their code.
+needing to restructure their code in major, potentially-breaking, ways.
 
@@
 combines {3D creation} tools and an Integrated
-Developer Environment (IDE), as seen in \Cref{fig:roblox-studio}.
+Developer Environment (IDE), as seen in \cref{fig:roblox-studio}.
 The IDE includes an optional \emph{Script Analysis} widget that
 reports syntax errors, type errors, and problems identified by
-lint tools. The script editor also (optionally) highlights
-the location in code where reported errors occur.
-Creators have three \emph{analysis modes} (modes) to choose from
-for every script they write:
+lint tools. The main editor widget can also highlight
+source locations in code where reported errors occur.
+
+\paragraph{Type Analysis Modes, \FS{} Analysis}
+
+Creators can opt in to detailed error reports and highlights by
+choosing a \emph{type analysis mode} for each script they write:
 \begin{itemize}
-  \item \mnocheck{}: report only syntax errors,
-  \item \mnonstrict{}: report syntax errors and a subset of ``high confidence'' type errors, or
+  \item \mnocheck{}: report only syntax errors (the default),
+  \item \mnonstrict{}: report syntax errors and a subset of high confidence type errors, or
   \item \mstrict{}: report syntax errors and all type errors.
@@
 
-As an example of nonstrict mode, the following program only reports one error:
+In addition to the main type analysis, Roblox Studio takes a second pass over
+every codebase with a \emph{\FS{}} analysis to infer autocomplete suggestions
+and drive other IDE tools.
+The \FS{} analysis uses rigorous checks along the lines of \mstrict{} mode
+no matter what mode the creator declared for the script in the hope of inferring
+precise types.
+(Internally to Roblox Studio, this analysis is called \emph{forced strict}
+because it ignores the creator's choice and applies strong checks. However,
+the checks are not the same as \mstrict{} mode;
+see~\cref{s:strict-vs-forcedstrict}.)
+When it fails to infer a type, \FS{} defaults to the unknown gradual
+type~\cite{st-sfp-2006}.
+Creators cannot see the errors reported by \FS{} analysis.
+
+As an example of \mnonstrict{} mode, the following program only reports one error:
 \begin{verbatim}
@@
   if condition then x.q = 7 end
-  local y = x.p + x.q --> no type error
-  local z = x.r       --> "Key 'r' not found in table 'x'"
+  local y = x.p + x.q --> OK
+  local z = x.r           --> UnknownProperty: Key 'r' not found in table 'x'
 \end{verbatim}
-but in strict mode it reports two:
+In \mstrict{} mode, it reports two errors:
 \begin{verbatim}
@@
   if condition then x.q = 7 end
-  local y = x.p + x.q --> "Type 'nil' could not be converted into 'number'"
-  local z = x.r       --> "Key 'r' not found in table 'x'"
+  local y = x.p + x.q --> TypeMismatch: Type 'nil' could not be converted into 'number'
+  local z = x.r           --> UnknownProperty: Key 'r' not found in table 'x'
 \end{verbatim}
 In cases like this, where it is undecidable whether there will be a run-time error,
-strict mode errs on the side of reporting an error, and nonstrict mode errs on
+\mstrict{} mode errs on the side of reporting an error, and \mnonstrict{} mode errs on
 the side of suppressing the error.
 
-Both modes report the \verb|Key 'r' not found in table 'x'| error --
+Both modes report the \code{UnknownProperty} error because
 misspellings of property names are common enough to report in both
-strict and nonstrict mode. See~\cite{bfj-hatra-2021}
-for a more detailed discussion of the rationale for strict and nonstrict mode.
-
-Both modes are opt-in. The default for a new script is \mnocheck{} by
-default, but creators can change this default to \mnonstrict{}.
-
-Even in \mnocheck{} mode, {Roblox Studio} performs type inference, since
-the results are needed by type-directed tooling such as autocomplete and
-API documentation. This behind-the-scenes typechecking is always performed
-in strict mode, since it is important that the inferred types be as precise
-as possible. The type errors produced by this pass are always discarded,
-so the verbosity of strict mode is not an issue.
-
-Since this pass is always performed in strict mode, we refer to it as
-\emph{forced strict} mode. Its main use is in autocomplete, so forced
-strict mode is triggered on every keystroke
-
-Scripts come in two favors: \emph{module scripts} and \emph{non-module
-scripts}.  Module scripts provide reusable libraries, which may be
-\emph{required} by other scripts. Since module scripts can require
-other module script, modules form a graph (though it is
-an error in strict mode if the graph is cyclic, and edges are removed to
-make the graph acyclic).
-
-When typechecking is performed for script analysis, any script that
-has been modified is marked as dirty, then any script that is dirty,
-or which transitively requires a dirty module, is typechecked. More
-commonly, when typechecking is performed for autocomplete, only
-the current script is typechecked, since it is the only dirty
-script, and nothing it requires can transitively require it, since
-the module graph is acyclic.
-
-The state of the world in a {Roblox} experience is captured by
-the \emph{data model}, which is a tree of \emph{instances}, such as
-parts, models, meshes, effects, lighting, audio assets, and physics
-constraints such as forces, springs and joints.
-%% The data model can be seen in the Explorer view of~\Cref{fig:roblox-studio}.
-%% ben: where is it? very hard to see
-While an experience is under development, it is typical for the data
-model to be edited (for example instances to be added, deleted, moved
-or renamed). Since the initial shape of the data model tree is reflected in
-the type system, it is possible for these edits to introduce type errors.
+\mstrict{} and \mnonstrict{} mode. See~\cite{bfj-hatra-2021}
+for a more detailed discussion of the rationale for \mstrict{} and
+\mnonstrict{} mode.
 
@@
 
-%% FILL table = object
+\Cref{t:type-error-labels} lists a few of the type errors that
+script analysis can report (10 out of 35 total)
+to give a sense of Luau.
+A \code{SyntaxError} is the only error that can appear in
+\mnocheck{} code.
+Several errors in addition to \code{UnknownProperty} are about
+tables, which are Luau's (and Lua's) primary data structure.
+Tables support arrays, dictionaries, and objects; consequently,
+tables can have methods and can usually be extended.
+The error \code{TypesAreUnrelated} is a special kind of type mismatch
+that arises only during cast, unification, or subtyping check.
+A \code{CountMismatch} occurs when the arguments to a function
+do not match the function's arity.
+Similarly, an \code{IncorrectGenericParamCount} is an arity mismatch
+for a generic type.
+Lastly, \code{CodeTooComplex} occurs when the typechecker hits an
+internal limit on problem size and \code{GenericError} is a catch-all
+label for miscellaneous errors.
+When Luau can provide more context regarding a generic error, it uses
+a second error label called \code{ExtraInformation} to attach
+a brief description.
 
 \begin{table}
-  \caption{Sample of type errors.}
+  \caption{Sample type errors.}
   % https://github.com/Roblox/luau/blob/master/Analysis/include/Luau/Error.h
@@
 
-  %% TODO review selection after analyzing the full data
-
-  %% FILL examples for each error?
   \begin{tabular}{ll}
     Label & Interpretation \\\midrule
-    \code{CodeTooComplex} & Type analysis failed, cannot understand the code\!\!\! \\
-    %% \code{UnificationTooComplex} & Type analysis failed, unification solver hit a limit\!\!\! \\
-    \code{SyntaxError} & Basic parse error, e.g., \code{for if end} \\
-
-    \code{CountMismatch} & Arity mismatch for a function \\
-    \code{IncorrectGenericParamCount} & Arity mismatch for a generic type \\
-    \code{UnknownProperty} & Referenced an invalid field or method  \\
-    \code{OnlyTablesCanHaveMethods} & Tried to attach a method to a non-table \\
-    \code{CannotCallNonFunction} & Called a value that is not a function \\
-    \code{TypesAreUnrelated} & Failed to cast, unify, or check subtyping \\
-    \code{TypeMismatch} & Generic label for other type errors \\
+    \code{TypeMismatch} & Basic type error. \\
+    \code{SyntaxError} & Basic parse error, e.g., \code{for if end}. \\
+    \code{UnknownProperty} & Referenced an invalid field or method.  \\
+    \code{OnlyTablesCanHaveMethods} & Tried to attach a method to a non-table. \\
+    \code{CannotExtendTable} & Tried to extend a sealed table. \\
+    \code{TypesAreUnrelated} & Failed cast, unify, or subtype. \\
+    \code{CountMismatch} & Arity mismatch for a function. \\
+    %% \code{CannotCallNonFunction} & Called a value that is not a function. \\
+    \code{IncorrectGenericParamCount} & Arity mismatch for a generic type. \\
+    \code{CodeTooComplex} & Type analysis failed, cannot understand the code. \\
+    %% \code{UnificationTooComplex} & Type analysis failed, unification solver hit a limit \\
     \code{GenericError}
     & Generic label for other non-type errors, such as\\
-    & \hbox{}~~looping over an unordered table
+    & \hbox{}~~looping over an unordered table.
     %% attempting to extend a type that does not describe a class
@@
 
-%    \code{ExtraInformation} & Follow-on error that provides additional information for an error at the same source location. \\
-%    UnknownSymbol &  \\
-%    NotATable &  \\
-%    CannotExtendTable &  \\
-%    DuplicateTypeDefinition &  \\
-%    FunctionDoesNotTakeSelf &  \\
-%    FunctionRequiresSelf &  \\
-%    OccursCheckFailed &  \\
-%    UnknownRequire &  \\
-%    UnknownPropButGotLikeProp &  \\
-%    InternalError &  \\
-%    DeprecatedApiUsed &  \\
-%    ModuleHasCyclicDependency &  \\
-%    IllegalRequire &  \\
-%    FunctionExitsWithoutReturn &  \\
-%    DuplicateGenericParameter &  \\
-%    CannotInferBinaryOperation &  \\
-%    MissingProperties &  \\
-%    SwappedGenericTypeParameter &  \\
-%    OptionalValueAccess &  \\
-%    MissingUnionProperty &  \\
-%    NormalizationTooComplex &  \\
-%    TypePackMismatch &  \\
-%    DynamicPropertyLookupOnClassesUnsafe &  \\)
   \end{tabular}
 \end{table}
+
+\paragraph{On-the-Fly Typechecking}
+
+Scripts come in two flavors: \emph{module scripts} and \emph{non-module
+scripts}.
+Module scripts provide reusable libraries, which may be
+{imported} by other scripts.
+Module scripts therefore form a graph in general,
+though it is an error in \mstrict{} mode to create an import cycle
+and edges are removed to make the graph acyclic.
+%% \QALAN{}: do all modes remove the edges?
+
+Typechecking analyzes the entire module graph, but it also keeps track of which
+modules have been edited to avoid repeating work.
+Any script that the creator modifies gets marked as dirty,
+and any script that is dirty or that transitively import a dirty
+module gets analyzed by the typechecker.
+In the common case when typechecking is performed for autocomplete,
+only the current script gets checked because it is the only dirty
+script, and because nothing it requires can transitively require it, since
+the module graph is acyclic.
+With this strategy, the typechecker can run after every keystroke without
+slowing down Roblox Studio.
+
+\paragraph{Data Model}
+
+%% TODO data assets ==> data model in rest of paper
+
+The state of the world in a {Roblox} experience is captured by
+the \emph{data model}, which is a tree of {instances}, such as
+parts, models, meshes, effects, lighting, audio assets, and physics
+constraints such as forces, springs and joints.
+Each asset lives in a separate file.
+There may be thousands of assets.
+%% The data model can be seen in the Explorer view of~\cref{fig:roblox-studio}.
+%% \QALAN{} where is it? very hard to see, needs a clear label
+
+While an experience is under development, it is typical for the data
+model to be edited.
+For example, instances may be added or renamed.
+Since the shape of the data model tree is reflected in
+the type system, it is possible for these edits to introduce a mountain
+of type errors across the project.
 
@@
 \subsection{Two Activities: Asset Creation and Scripting}
+\label{s:strict-vs-forcedstrict}
 

### 2023-09-28 programming23/paper.tex
count down/upgrades, fix bug, but need to double-check on server

@@
       \begin{tabular}[t]{l@{~~}ll}
+        %% TODO numbers may be off by +- 100
+        %%    & 341 & contain a mode upgrade \\
+        %%    & 320 & contain a mode downgrade \\
+        %%    & 512 & contain modules with different modes
+        %%  they don't match out/error-density*hasdown.rkt
+        %%  ??? counting downs instead of sessions with downs ???
+        %% TODO fix throughout paper!
         \zerowidth{Among the mixed-mode sessions:} \\
-        & 341 & contain a mode upgrade \\
-        & 320 & contain a mode downgrade \\
+        & 176 & contain a mode upgrade \\
+        & 233 & contain a mode downgrade \\
         & 512 & contain modules with different modes
@@
 in the data are split across type analysis modes.
-\FILL{} add averages.
 
@@
     Creators rarely disable type analysis after opting in.
-    Only 320 of the \nK{340} sessions ($<\pct{1}$)
+    Only 233 of the \nK{340} sessions ($<\pct{1}$)
     switch to a weaker mode.
-    \FILL{} are they downgrading from S or NS?
+    Most of these (\pct{75}) are downgrading from \mstrict{}
+    rather than \mnonstrict{} mode.
     %% FILL more insight if we match upgrades + downgrades in session?
+
+    %% upgrade
+    %% 519 total sessions
+    %% 176 with up
+    %% #hash((nocheck . 0) (nonstrict . 44) (strict . 132))
+    %%  nocheck : 0.00\%
+    %%  strict : 75.00\%
+    %%  nonstrict : 25.00\%
+    %%
+    %% downgrade
+    %% 515 total sessions
+    %% 233 with down
+    %% #hash((nocheck . 0) (nonstrict . 64) (strict . 169))
+    %%  nocheck : 0.00\%
+    %%  strict : 72.53\%
+    %%  nonstrict : 27.47\%
 

### 2023-09-28 programming23/paper.tex
transpose size / dist tables to save space

@@
 % \usepackage[backend=biber]{biblatex}
+\usepackage[export]{adjustbox}
 \usepackage{amssymb}
@@
 
-\section{{Roblox} Context}
+\section{Context: {Roblox} and {Luau}}
 \label{s:context}
@@
 
-\subsection{Limits of telemetry}
+\subsection{Limits of Telemetry}
 
@@
 
-  \begin{tabular}{l@{}r@{~}l@{}r@{~}l}
-                 &  Lines &             &    Edit Range & \\
-    Mean [std]   &  6,497 & \stddev{22} &         3,680 & \stddev{31} \\
-    Median       &  3,115 &             &           926 & \\
-    \pct{99}     & 50,547 &             &        34,725 & \\
-    %% TODO x-axis starts at 1, not zero ... there are no zeros right??!
-      & \includegraphics[width=0.2\columnwidth]{img/lines-distribution.pdf}
-    & & \includegraphics[width=0.2\columnwidth]{img/editrange-distribution.pdf}
+%  \begin{tabular}{l@{}r@{~}l@{}r@{~}l}
+%                 &  Lines &             &    Edit Range & \\
+%    Mean [std]   &  6,497 & \stddev{22} &         3,680 & \stddev{31} \\
+%    Median       &  3,115 &             &           926 & \\
+%    \pct{99}     & 50,547 &             &        34,725 & \\
+%    %% TODO x-axis starts at 1, not zero ... there are no zeros right??!
+%      & \includegraphics[width=0.2\columnwidth]{img/lines-distribution.pdf}
+%    & & \includegraphics[width=0.2\columnwidth]{img/editrange-distribution.pdf}
+%
+%  \end{tabular}
+
+  \begin{tabular}[t]{lrrrl}
+    ~          & Mean [std] & Median & \pct{99} Pct. & Distribution \\
+    Lines      & 6,497 \stddev{22} & 3,115 & 50,547 & \includegraphics[width=0.14\columnwidth,valign=M]{img/lines-distribution.pdf} \\
+    Edit Range & 3,680 \stddev{31} & 926   & 34,725 & \includegraphics[width=0.14\columnwidth,valign=M]{img/editrange-distribution.pdf}
+    %% TODO x-axis starts at 1, not zero
+    %% TODO much bigger font or much smaller plot!!!
 
@@
 
-  \begin{tabular}{l@{}r@{~}l@{}r@{~}l} \\
-                 & Time Span (sec)  &       & Record Count  \\
-    Mean [std]   &     3,184 & \stddev{16}  &     286 & [583] \\
-    Median       &       845 &              &     138        \\
-    \pct{99}     &    35,450 &              &   3,302        \\
-    %% TODO x-axis starts at 1, not zero ... there are no zeros right??!
-    & \includegraphics[width=0.2\columnwidth]{img/timespan-distribution.pdf}
-    & & \includegraphics[width=0.2\columnwidth]{img/event-count-distribution.pdf}
+%  \begin{tabular}{l@{}r@{~}l@{}r@{~}l} \\
+%                 & Time Span (sec)  &       & Record Count  \\
+%    Mean [std]   &     3,184 & \stddev{16}  &     286 & [583] \\
+%    Median       &       845 &              &     138        \\
+%    \pct{99}     &    35,450 &              &   3,302        \\
+%    %% TODO x-axis starts at 1, not zero ... there are no zeros right??!
+%    & \includegraphics[width=0.2\columnwidth]{img/timespan-distribution.pdf}
+%    & & \includegraphics[width=0.2\columnwidth]{img/event-count-distribution.pdf}
+%
+%  \end{tabular}
+
+  \begin{tabular}{lrrrl} \\
+    ~               & Mean [std] & Median & \pct{99} Pct. & Distribution \\
+    Time Span (sec) & 3,184 \stddev{16} & 845 & 35,450 & \includegraphics[width=0.14\columnwidth,valign=M]{img/timespan-distribution.pdf} \\
+    Record Count    & 286 [583]         & 138 & 3,302  & \includegraphics[width=0.14\columnwidth,valign=M]{img/event-count-distribution.pdf}
+    %% TODO x-axis starts at 1
+    %% TODO huge font, or smaller plot
 
@@
     \begin{description}
-      \item[Which errors arise?]
+      \item[- Which errors arise?]
         Syntax errors are extremely common~(over \pct{50},~\cref{t:type-error-count}), 
@@
 
-      \item[How do creators respond? Which errors persist through edits?]
+      \item[- How do creators respond? Which errors persist through edits?]
         Creators typically fix analysis errors.

### 2023-09-28 programming23/paper.tex
telemetry record fig
data asset -> data model, but both terms are useful
DM = roblox-specific thing
data asset = casual language, easier to make sense of in abstract + conclusion

@@
 
-%% TODO data assets ==> data model in rest of paper
-
 The state of the world in a {Roblox} experience is captured by
@@
     \\\toprule
+    \multicolumn{6}{c}{\upbracefill} & \multicolumn{1}{c}{\upbracefill} \\[-0.5ex]
+    \multicolumn{6}{c}{fixed-size} & \multicolumn{1}{c}{variable-size}
   \end{tabular}
@@
     Improving precision is an important direction for future work.
-    Until then, \mstrict{} mode clearly needs to relax its data-asset
+    Until then, \mstrict{} mode clearly needs to relax its data model
     analysis because it raises even more errors than \FS{} checking~(\cref{f:tefs-compass}).

### 2023-09-28 programming23/paper.tex
quick error-by-mode bars

@@
   \item
-    Overall Counts: summarizes type errors and forced strict errors during the
+    Overall Counts: summarizes type errors and \FS{} errors during the
     last two invocations of type analysis, and additionally reports the number
@@
     \begin{tabular}[t]{r@{~~}r@{~}l}
-      72,235,735 & {forced strict errors} \\
+      72,235,735 & {\FS{} errors} \\
       37,027,281 & in current module & [\pct{51.26}] \\
@@
   \end{tabular}
-  \caption{Type errors and forced strict errors across all telemetry records.}
+  \caption{Type errors and \FS{} errors across all telemetry records.}
   \label{f:count-analysis-errors}
@@
 and breaks them down by location in the codebase.
-In total, there are nearly 600K type errors.
-These appeared in modules that opted in to typechecking with either the
-\mnonstrict{} or \mstrict{} mode.
-The number of forced strict errors is much larger, 72M, because this analysis
+In total, there are over 590,000 type errors.
+%% delete? upcoming figure has the same info
+% Most of these (\pct{65}) appeared in modules that opted in to typechecking with either the
+% \mnonstrict{} or \mstrict{} mode; the remainder are syntax errors reported
+% by \mnocheck{} mode.
+The number of \FS{} errors is much larger, 72M, because this analysis
 ran on every module using strict type.
-(Furthermore, forced strict runs silently, so creators have no
+(Furthermore, \FS{} runs silently, so creators have no
 awareness of or incentive to
@@
 and a small fraction overlapped with the current edit range~(\pct{5} for
-type errors, \pct{1} for forced strict).
+type errors, \pct{1} for \FS{}).
 
@@
 
-\begin{figure}[t]
+\begin{figure}[t]\centering
+
+  %%%% from nsa aggregate-tefs
   %% strict
@@
 
-  \begin{tabular}{l}
-    600K type errors         = \pct{35} \mnocheck{}, \pct{61} \mnonstrict{}, \pct{4} \mstrict{}
-    \\
-    72M forced scrict errors = \pct{89} \mnocheck{}, \pct{10} \mnonstrict{}, \pct{1} \mstrict{}
-    %% really 0.16 strict, very low!!!
+  \newcommand{\labelbars}[1]{\begin{tabular}[t]{l@{}l} \begin{tabular}[t]{r} \mnocheck{} \\ \mnonstrict{} \\ \mstrict{} \\ ~ \end{tabular} & #1 \end{tabular}}
+
+    \begin{tabular}[t]{cc}
+    %% TODO why text blurry??!
+    Type errors & Background errors \\
+    \labelbars{\includegraphics[valign=M]{img/error-by-mode-te.pdf}}
+    & \labelbars{\includegraphics[valign=M]{img/error-by-mode-fs.pdf}}
   \end{tabular}
@@
 
-\Cref{f:error-by-mode} shows how the type errors and forced strict errors
+\Cref{f:error-by-mode} shows how the type errors and \FS{} errors
 in the data are split across type analysis modes.
@@
 
-Forced strict analysis applies the same typechecker to every module, regardless
+Background analysis applies the same typechecker to every module, regardless
 of mode, in the hope of enabling IDE tools.
-Consequently, the breakdown of forced strict errors matches the overall usage
+Consequently, the breakdown of \FS{} errors matches the overall usage
 data~(\cref{f:dataset-overview}): almost \pct{90} for \mnocheck{},
@@
   \end{tabular}
-  \caption{Comparing the change in type errors to the change in forced strict errors.}
+  \caption{Comparing the change in type errors to the change in \FS{} errors.}
   \label{f:tefs-compass}
@@
 Of special interest is the \mstrict{} mode case, because the number of
-type errors and forced strict errors should be nearly the same.
+type errors and \FS{} errors should be nearly the same.
 But, there are two differences between \mstrict{} type checking
-and forced strict checking:
+and \FS{} checking:
 \begin{itemize}
   \item
-    Forced strict checks all dependencies in \mstrict{} mode, no matter
+    Background checks all dependencies in \mstrict{} mode, no matter
     what mode they declare.
-    So, the number of forced strict errors may be higher.
-  \item
-    Forced strict has looser rules for the \emph{data model};
+    So, the number of \FS{} errors may be higher.
+  \item
+    Background checking has looser rules for the \emph{data model};
     that is, for the static assets that a program manipulates.
-    So, the number of forced strict errors may be lower.
+    So, the number of \FS{} errors may be lower.
     %% initial DM, all properties go into a table type; DM -> folder -> model -> part
@@
 \end{itemize}
-Our data contains a surprising number of records where forced strict
+Our data contains a surprising number of records where \FS{}
 reports fewer errors.
@@
 
-\Cref{f:tefs-compass} compares changes in type errors (\tekey{}) to changes in forced
-strict errors (\fskey{}) for the three analysis modes.
+\Cref{f:tefs-compass} compares changes in type errors (\tekey{}) to changes in
+\FS{} errors (\fskey{}) for the three analysis modes.
 To a first approximation, the changes either agree with one another
 or disagree.
-For example, if type errors increase but forced strict errors decrease,
+For example, if type errors increase but \FS{} errors decrease,
 we have a disagreement.
@@
 over time.
-\Cref{f:indy-session} shows the type errors, forced strict errors, and module switches
+\Cref{f:indy-session} shows the type errors, \FS{} errors, and module switches
 in two \mstrict{} mode sessions.
@@
 Type errors appear as orange plus marks.
-Forced strict errors appear as green x marks.
+Background errors appear as green x marks.
 The vertical lines show module switches.
@@
 
-A general observation is that the number of type errors is always less than the number of forced strict
+A general observation is that the number of type errors is always less than the number of \FS{}
 errors, though the numbers are close.

### 2023-09-28 programming23/paper.tex
intro cleanup

@@
 
-{Roblox} is a platform for {shared virtual experiences}, with
+{Roblox} is a platform for {shared virtual experiences} (typically 3D games), with
 65~million Daily Active Users, and 14~billion hours of engagement in
 April--June 2023~\cite{corp.roblox.com}.
-There are over 5~million experiences available on the platform thanks to a
+There are over 5~million distinct user-created programs available on the platform thanks to a
 worldwide community of 3~million creators.
-Many of the creators build experiences for fun or learning and may
+Many of the creators program for fun or learning and may
 not consider themselves software developers; others are professional
@@
 
-Creators build experiences using the
+Creators program using the
 {Luau} programming language~\cite{luau-lang.org},
@@
 In this work, we design telemetry that collects data on type errors without
-exposing source code, source locations, or even error messages.
-The telemetry counts the number and kind of type errors at various
+exposing source code, source locations, or even error message text (which
+may contain revealing information).
+The telemetry counts the number and \emph{kind} of type errors at various
 levels of granularity.
@@
 
-The state of the world in a {Roblox} experience is captured by
+The state of the world in a {Roblox} program is captured by
 the \emph{data model}, which is a tree of {instances}, such as
@@
 
-While an experience is under development, it is typical for the data
+While a program is under development, it is typical for the data
 model to be edited.

### 2023-09-28 programming23/paper.tex
abstract cleanup

@@
   The creators form a broad group, ranging from novices writing
-  their first script to professional developers, thus Luau
+  their first script to professional developers; thus, Luau
   must support a wide audience.
@@
   collect source code fragments, error messages, or even filepaths.
-  The research questions are thus how to design a safe telemetry system
-  and whether it can provide useful insights about type errors.
+  The research questions are thus about how to conduct telemetry that
+  is not invasive and obtain insights from it about type errors.
 

### 2023-09-28 programming23/paper.tex
sec 2

@@
 and drive other IDE tools.
-The \FS{} analysis uses rigorous checks along the lines of \mstrict{} mode
-no matter what mode the creator declared for the script in the hope of inferring
-precise types.
+To infer precise types,
+the \FS{} analysis uses rigorous checks very similar to those of
+\mstrict{} mode (but not identical: see \cref{s:strict-vs-forcedstrict})
+no matter what mode the creator declared for the script.
 (Internally to Roblox Studio, this analysis is called \emph{forced strict}
-because it ignores the creator's choice and applies strong checks. However,
-the checks are not the same as \mstrict{} mode;
-see~\cref{s:strict-vs-forcedstrict}.)
-When it fails to infer a type, \FS{} defaults to the unknown gradual
+because it ignores the creator's choice and applies strong checks.)
+When it fails to infer a type, \FS{} analysis defaults to the unknown gradual
 type~\cite{st-sfp-2006}.
@@
 Any script that the creator modifies gets marked as dirty,
-and any script that is dirty or that transitively import a dirty
+and any script that is dirty or that transitively imports a dirty
 module gets analyzed by the typechecker.
@@
 model to be edited.
-For example, instances may be added or renamed.
+% For example, instances may be added or renamed.
+% --> Commented out to get a better page break.
 Since the shape of the data model tree is reflected in

### 2023-09-28 programming23/paper.tex
sec 3

@@
 be deprecated: without telemetry it may be difficult to know how
-popular a feature is with users, but with telemetry it is
+popular it is, but with telemetry it is
 straightforward.
 
-IDEs use telemetry in a similar fashion to most user-facing
+IDEs use telemetry in a similar fashion\todo{I really don't know what
+  this means. Are you saying VSCode is constantly calling home? Is it
+  telling Racket about my API usage? Do we need anything in this para?} to most user-facing
 applications, for example VSCode~\cite{vsc-telemetry} and
@@
 
-Telemetry for programming languages can be controversial, for example
+Telemetry for programming languages can be controversial:e.g., see
 the lively discussion around telemetry in the Go
@@
 
-Much of the concern about telemetry is because of its potential impact on
-privacy, by revealing Personally Identifiable Information (PII).
-For this reason, it is important that any experiments using telemetry
-not reveal PII:
-no source code;
-no source code locations;
-no error messages (which may contain source code);
-no record of the creator's identity, locale, or IP address;
-and no information about what creation the data came from.
-A good introduction to the privacy implications and tradeoffs
-involved with telemetry is~\cite{transparent-telemetry}.
-
-For performance reasons, there are limits to telemetry data:
-telemetry records have size limits, and the performance impact of
-recording telemetry should be minimal.
-
-In summary, the limitations of telemetry are:
+Telemetry therefore faces at least two major constraints:
 \begin{itemize}
-  \item Data which reveals no PII.
-  \item Fixed-size telemetry records.
-  \item Minimal overhead to record telemtry.
+
+\item It should not reveal private information. This can include
+  Personally Identifying
+  Information (PII) about creators: their identity, location, etc. It also
+  includes trade secrets. Even an error message that contains the name
+  of a function could reveal something the creator would not want to
+  share.
+
+\item For performance reasons, it should have minimal run-time
+  overhead and also transmit small and bounded amounts of data.
 \end{itemize}
+Cox discusses in detail the privacy implications and tradeoffs
+in telemetry~\cite{transparent-telemetry}.
+
+% BEN: I believe the above subsumes all of the below:
+% Much of the concern about telemetry is because of its potential impact on
+% privacy. It can not only reveal Personally Identifiable Information
+% (PII) but also trade secrets, etc. Thus, telemetry should not reveal
+% PII, source code, error messages (which can refer to source), the
+% creator's identity, and so on.
+% % and no information about what creation the data came from.
+
+% For performance reasons, there are limits to telemetry data:
+% telemetry records have size limits, and the performance impact of
+% recording telemetry should be minimal.
+
+% In summary, the limitations of telemetry are:
+% \begin{itemize}
+%   \item Data that don't reveal private information.
+%   \item Fixed-size telemetry records.
+%   \item Minimal overhead to record telemtry.
+% \end{itemize}
+
 In addition, due to the architecture of Roblox Studio
@@
 \begin{figure}[t]\centering
+  % BEN: The left brace is confusing. Is ID not fixed?
   \begin{tabular}{|l|l|l|l|l|l|l|}\hline
@@
 The telemetry for a Roblox Studio session is gathered as a series of
-telemtry records, all with the same \panon{} session
+records, all with the same \panon{} session
 identifier. This allows us to correlate telemetry across a single
@@
 
-In order to avoid swamping our telemetry servers, we randomly throttle
+In order to avoid swamping our telemetry servers, we randomly throttled
 the generation of telemetry records:
@@
 \end{itemize}
-In order to find how file-switching correlated with type errors, we additionally have:
-\begin{itemize}
-  \item
-    every run of the typechecker in an enrolled session which switched file generated a telemetry record.
-\end{itemize}
-
-To track the edit range, we record a start and end position, which we
-update appropriately on every edit. This can result in very large edit
-ranges, for example, if the user edits at the beginning and end of the
-file.
-The main benefit of this strategy is that it ensures constant-size telemetry records.
-
-Each telemetry record contains seven kinds of data.
-Refer to~\cref{f:tdata} for an illustration:
+In order to find how file-switching correlated with type errors, we
+additionally recorded
+every run of the typechecker in an enrolled session which switched
+file generated a telemetry record.\todo{I cannot parse this!}
+
+\Cref{f:tdata} shows the structure of our telemetry records:
 \begin{enumerate}
   \item
-    ID: a \panon{} identifier for the current session
+    ID: a \panon{} identifier for the current session.
   \item
@@
   \item
-    Mode: type analysis mode of the current file, which is either \mnocheck{},
+    Mode: type analysis mode of the current file: \mnocheck{},
     \mnonstrict{}, or \mstrict{}.
@@
 \end{enumerate}
+
+To track the edit range, we record a start and end position, which we
+update appropriately on every edit. This can result in very large edit
+ranges, for example, if the user edits at the beginning and end of the
+file.
+The main benefit of this strategy is that it ensures constant-size telemetry records.
 

### 2023-09-28 programming23/paper.tex
sec 5

@@
 With this telemetry data, we investigate research questions about
-the adoption and benefits of type analysis:
+the adoption and benefits of type analysis: \todo{Some of these are flawed. We don't know anything about creators. Adjust
+  to match section 5.}
 \begin{description}
@@
 
+\todo{No identity $\Rightarrow$ no ``creators'', only sessions.}
 With the data in hand, we can answer the research questions
@@
 at the same time offer several directions for future improvements.
-Luau clearly has a powerful typechecker that can efficiently
+
+Luau clearly\todo{How can we say?} has a powerful typechecker that can efficiently
 analyze thousands of lines of code.
@@
     Projects rarely combine analysis modes.
-    Only 512 of \nK{340} sessions ($<\pct{1}$) switched to modules with different
+    Only 512 of \nK{340} sessions ($<\pct{0.15}$) switched to modules with different
     analysis modes.
@@
     Creators rarely disable type analysis after opting in.
-    Only 233 of the \nK{340} sessions ($<\pct{1}$)
+    Only 233 of the \nK{340} sessions ($<\pct{0.07}$)
     switch to a weaker mode.
-    Most of these (\pct{75}) are downgrading from \mstrict{}
-    rather than \mnonstrict{} mode.
+    Most of these (\pct{75}) downgrade from \mstrict{} to
+    \mnonstrict{} % BEN: confirm!
+    rather than from \mnonstrict{} mode.
     %% FILL more insight if we match upgrades + downgrades in session?
@@
 
-  \item[For modules that use type analysis:]\leavevmode
+  \item[For modules that use type analysis (\mstrict{} and \mnonstrict{}):]\leavevmode
     \begin{description}
@@
         the total.
-        The errors that most-often survive edits are due to
+        The errors that most often survive edits are due to
         inference at a binary operation,
@@
 \mstrict{} mode~(\cref{f:error-changes}) might tell a very different story
-without all the noise that syntax errors introduce.
+without all the noise that syntax errors introduce.\todo{Why didn't
+  YOU just filter them, then?!?}
 
@@
 errors, if any, creators intended to fix with their edits.
-Although it appears that creators choose to fix errors with their edits,
-the fixes may be a side effect of other plans.
+Thus, fixes may be a side effect of other plans, as seen in prior work [CITE].
+% Cite the Marceau SIGCSE 2011 paper. This is the [DIFF] code from the rubric.
 
@@
 First, our information on type errors is limited to error names.
-If another language has an error that is roughly similar to the
+If another language has an error that is roughly similar to
+the\todo{Huh? It's similar yet different?}
 \code{CannotCallNonFunction} error but with slightly different
@@
 characteristics.
+
+Ultimately, we view our work as primarily \emph{formative}. It gives
+us a first look at usage, and generates several questions that can
+lead to hypotheses. Answering those would require much more intrusive
+techniques, or controlled studies, or other means.
 

### 2023-09-29 programming23/paper.tex
sec 3 edits

@@
 scripts}.
+(Though, throughout this paper we use the word ``module''
+to refer to arbitrary scripts.)
 Module scripts provide reusable libraries, which may be
@@
 
-\subsection{Limits of Telemetry}
-
 Telemetry allows an application to ``phone home'' with data
@@
 and feature uptake. A typical usage is in deciding whether an API can
-be deprecated: without telemetry it may be difficult to know how
-popular it is, but with telemetry it is
-straightforward.
-
-IDEs use telemetry in a similar fashion\todo{I really don't know what
-  this means. Are you saying VSCode is constantly calling home? Is it
-  telling Racket about my API usage? Do we need anything in this para?} to most user-facing
-applications, for example VSCode~\cite{vsc-telemetry} and
-IntelliJ~\cite{intellij-telemetry} report telemetry.
-See~\cref{s:related} for futher comparisons among telemetry designs.
-
-Telemetry for programming languages can be controversial:e.g., see
-the lively discussion around telemetry in the Go
+be deprecated; without telemetry, it may be difficult to know how
+popular the API is, but with telemetry it is straightforward.
+Telemetry for programming languages can, however, be controversial.
+See, for example, the lively discussion around telemetry in the Go
 toolchain~\cite{golang-telemetry}.
+
+Roblox Studio has a framework for reporting telemetry.
+Various features in the IDE use telemetry to measure their effectiveness.
 The Luau open source toolchain does \emph{not} report telemetry,
@@
 
-Telemetry therefore faces at least two major constraints:
+We added a telemetry subsystem to Roblox Studio in order to study type errors
+in Luau code.
+This section explains the constraints our telemetry had to meet~(\cref{s:tel-limitations})
+and the final design~(\cref{s:telemetry-records}).
+
+\subsection{Limitations for Luau Telemetry}
+\label{s:tel-limitations}
+
+Telemetry in Roblox Studio faces three major constraints:
 \begin{itemize}
 
-\item It should not reveal private information. This can include
-  Personally Identifying
-  Information (PII) about creators: their identity, location, etc. It also
-  includes trade secrets. Even an error message that contains the name
-  of a function could reveal something the creator would not want to
-  share.
-
-\item For performance reasons, it should have minimal run-time
-  overhead and also transmit small and bounded amounts of data.
+  \item
+    It \emph{must not reveal private information}.
+    This can include Personally Identifying Information (PII) about creators: their identity, location, etc.
+    It also includes trade secrets. Even an error message that contains the
+    name of a function can reveal something the creator does not want to share.
+
+  \item
+    On the client side, telemetry computations \emph{must run quickly enough} to avoid
+    slowing down the editing experience.
+
+  \item
+    Telemetry \emph{must transmit a small amount of data}---not too many records
+    with not too much information inside---to avoid overloading the servers.
+
 \end{itemize}
-Cox discusses in detail the privacy implications and tradeoffs
-in telemetry~\cite{transparent-telemetry}.
-
-% BEN: I believe the above subsumes all of the below:
-% Much of the concern about telemetry is because of its potential impact on
-% privacy. It can not only reveal Personally Identifiable Information
-% (PII) but also trade secrets, etc. Thus, telemetry should not reveal
-% PII, source code, error messages (which can refer to source), the
-% creator's identity, and so on.
-% % and no information about what creation the data came from.
-
-% For performance reasons, there are limits to telemetry data:
-% telemetry records have size limits, and the performance impact of
-% recording telemetry should be minimal.
-
-% In summary, the limitations of telemetry are:
-% \begin{itemize}
-%   \item Data that don't reveal private information.
-%   \item Fixed-size telemetry records.
-%   \item Minimal overhead to record telemtry.
-% \end{itemize}
+Cox~\cite{transparent-telemetry} discusses in detail the privacy implications
+and tradeoffs in telemetry.
 
@@
 \begin{itemize}
-  \item The typechecker does not have access to some lifecycle events,
-    such as save, quit and publish.
-  \item The typechecker does not have access to GUI state, for example
-    whether the Script Analysis widget is visible.
+  \item
+    Lifecycle events including as save, run, quit, and publish.\footnote{Publish events
+    cannot be reported at any rate because they are public, and could be used
+    to match a telemetry session to a creator.}
+  \item
+    GUI state, in particular, whether the Script Analysis widget is visible.
 \end{itemize}
+Our type error telemetry has to meet all of the above constraints.
 
@@
 \begin{figure}[t]\centering
-  % BEN: The left brace is confusing. Is ID not fixed?
   \begin{tabular}{|l|l|l|l|l|l|l|}\hline
@@
     \\\toprule
-    \multicolumn{6}{c}{\upbracefill} & \multicolumn{1}{c}{\upbracefill} \\[-0.5ex]
+    \multicolumn{6}{c}{\raisebox{.5\normalbaselineskip}[0pt][0pt]{$\underbrace{\hspace{9cm}}$}} &
+    \multicolumn{1}{c}{\raisebox{.5\normalbaselineskip}[0pt][0pt]{$\underbrace{\hspace{3cm}}$}} \\[-0.5ex]
     \multicolumn{6}{c}{fixed-size} & \multicolumn{1}{c}{variable-size}
@@
 
-The telemetry for a Roblox Studio session is gathered as a series of
-records, all with the same \panon{} session
-identifier. This allows us to correlate telemetry across a single
+The type error telemetry for a Roblox Studio session is gathered as a series of
+records, all with the same \panon{} session identifier.
+This allows us to correlate telemetry across a single
 session, but not between sessions.
-
-In order to avoid swamping our telemetry servers, we randomly throttled
-the generation of telemetry records:
+In order to avoid swamping the telemetry servers, the client reports on a subset
+of telemetry events via two levels of uniform random sampling:
 \begin{itemize}
   \item
-    1\% of Roblox Studio sessions were enrolled in generating telemetry, and
-  \item
-    0.5\% of runs of the typechecker (approximately 1 in 200 keystrokes)
-      in an enrolled session generated a telemetry record.
+    1\% of Roblox Studio sessions generate \emph{any} type error telemetry, and
+  \item
+    0.5\% of all type analyses (approximately 1 in 200 keystrokes)
+    in an enrolled session generate a telemetry record.
 \end{itemize}
-In order to find how file-switching correlated with type errors, we
-additionally recorded
-every run of the typechecker in an enrolled session which switched
-file generated a telemetry record.\todo{I cannot parse this!}
+In addition to sending telemetry for randomly-selected keystrokes,
+our system sends a telemetry record every time the session changes
+focus from one module to another.
+These records provide important context for mode switches:
+if one record uses \mstrict{} mode but the next uses \mnocheck{},
+then it is critical to know whether the creator downgraded modes
+or simply switched to another module.
 
@@
 To track the edit range, we record a start and end position, which we
-update appropriately on every edit. This can result in very large edit
-ranges, for example, if the user edits at the beginning and end of the
-file.
-The main benefit of this strategy is that it ensures constant-size telemetry records.
-
-\paragraph{NOTES} %% FILL
-This kind of telemetry data is coarse-grained.
-For instance, it does not let us distinguish between edits that ignore a type
+update appropriately on every edit. This can result in very large and
+imprecise edit ranges, for example, if the user edits at the beginning and end
+of the file.
+The upshot of this strategy is that it reduces the size and complexity of
+telemetry records because there is only one inerval for errors to overlap with.
+
+This telemetry design is admittedly coarse-grained.
+For instance, it does not distinguish between edits that ignore a type
 error from edits that remove one error while introducing another.
 We acknowledge this and other threats in~\cref{s:threats}.
-The main advantages of the telemetry design are the small size of each record
-and complete the lack of PII.
-Furthermore, despite the limitations, Studio's telemetry supports a variety of
+The main advantages of this telemetry are the small size of each record
+and complete the lack of private information.
+Furthermore, despite the limitations, this telemetry supports a variety of
 interesting conclusions that we showcase in the next section.
@@
     \FILL{} which survive the longest, most keeps???
+
+  \item
+    \FILL{} beware counts include module switches. Why???
+    Is that defensible at all?
 
@@
 
+\FILL{} cannot match sessions to codebases
+
 While this study has high ecological validity due to its focus on
@@
 
-In other professional-grade IDEs, such as VSCode~\cite{vscode-telemetry}
-and \code{.NET}~\cite{dotnet-telemetry}, telemetry may include all sorts of
+In other professional-grade IDEs, such as VSCode~\cite{vscode-telemetry},
+IntelliJ~\cite{intellij-telemetry}, and \code{.NET}~\cite{dotnet-telemetry}, telemetry may include all sorts of
 data: error reports, source code fragments, timestamps, and filepaths.

### 2023-09-29 programming23/paper.tex
sec 4 getting started

@@
 
-Roblox Studio collected type error telemetry in Spring 2023,
-between February and April.
+Roblox Studio collected type error telemetry between February
+and April 2023.
 Every Studio session had a small random chance of generating
@@
 randomly on each keystroke.
-In total, we collected 1.5 million telemetry records.
-Roughly two-thirds of records are due to keystrokes.
+In total, we collected over 1.5 million telemetry records.
+Roughly two thirds of records are due to keystrokes.
 
@@
 Each vertical bar counts the number of records generated per hour according to
-a client-side timestamp.\todo{Do we need all this detail?}
+a client-side timestamp.
+%% \todo{Do we need all this detail?}
+%% the vertical bars suggest we have lots of day-job developers, will point this out below
 The bars are labeled with a California time zone because that is the locale of
 the Roblox servers, which parsed the timestamps.
-Most buckets contain roughly one thousand records, but some have very few
-and a handful of others exceed three thousand.
+Most buckets contain roughly one thousand records, but the number
+varies between a few dozen at minimum at over three thousand.
 %% \FILL{} investigate one tower, what's happening?? all one session?
 There is a sharp drop in early April when the experiment was finished
-and the telemetry probe was removed (though there is a long tail as
-developers uopdated Roblox Studio).
+and the telemetry probe was removed, though there is a long tail as
+developers updated {Roblox Studio}.
 We do not know which timezone the records originated in, but since the counts
@@
 of Roblox developers are following a Western Hemisphere schedule.
+The regular pattern of the peaks furthermore suggests that many creators
+follow a regular schedule; this may be due to a large community of
+professional developers.
 The counts for weekends (shaded regions) are often the tallest,
@@
 
-A close inspection of the data revealed a few anomalies.
-For the most part, we omit the strange data from futher
-analysis.\todo{I'm inclined to rewrite this to be much more concise.}\todo{agreed}
-
-First, some records have odd timestamps according to the server's records.
-Every telemetry record comes with two timestamps: a client-side timestamp
-from when Roblox Studio created it and a server-side timestamp from when
-the server posted it to its database.
-The server timestamps are little use for our analysis because they bunch
-together; it is common for several records to have identical server
-timestamps.
-However, server timestamps are useful for auditing because they should
-be relatively close in value to the client timestamps.
-For a small number of records (<100? \FILL{}), though, the client is off by a large
-margin: up to one week behind and up to one day ahead of the server.
-We attribute the large delays to issues when sending telemetry (lost internet
-connection, etc.) and smaller offsets to incorrect time settings on the client
-machine.
-In any event, we do not filter these records from further
-analysis.\todo{To be clear, you do NOT filter? Because the section is
-  about what you DO filter.}
-
-Second, some records have identical (client) timestamps.
-This can happen legitimately if type analysis runs several times
-in the same millisecond, but it is unlikely that a human would
-enter keystrokes or switch modules quickly enough.
-We keep the first record with a distinct timestamp and discard the others.
-%% TODO wait, better deduplicate by session and not globally!!!
-
-%% out/negative-edit-range.txt
-%% actual value is NOT negative, but large positive over 4 billion
-%% example: 4294967293
-Third, the edit ranges are negative in a few hundred records~(1,533).
-This is likely due to the user deleting a large block of code.
-Since the issue affects so few records, we omit the negative ranges
-from analysis (but, we do use other parts of the records; for example
-the timestamps influence~\cref{f:records-per-hour}).
+A close inspection of the data revealed a two anomalies
+that we removed from further analysis:
+\begin{enumerate}
+  \item
+    Some records have identical timestamps and session IDs.
+    We keep the first such record and discard the rest.
+    (A human would have to switch modules twice or enter several keystrokes
+     within one millisecond to generate two such records legitimately.)
+
+   \item
+     A few hundred records~(1,533) have negative edit ranges.
+     These are likely due to large deletions.
+     Since the issue affects so few records, we simply ignore
+     their edit ranges. We do, however, use their overall error counts and other uncorrupted fields.
+
+\end{enumerate}
+
+%% not important for submission
+% One related observation, for which \emph{do not} sanitize the data, is
+% that some records have odd timestamps according to the server's records.
+% Every telemetry record comes with two timestamps: a client-side timestamp
+% from when Roblox Studio created it and a server-side timestamp from when
+% the server posted it to its database.
+% The server timestamps are little use for our analysis because they bunch
+% together; it is common for several records to have identical server
+% timestamps.
+% However, server timestamps are useful for auditing because they should
+% be relatively close in value to the client timestamps.
+% For a small number of records (<100? \FILL{}), though, the client is off by a large
+% margin: up to one week behind and up to one day ahead of the server.
+% We attribute the large delays to issues when sending telemetry (lost internet
+% connection, etc.) and smaller offsets to incorrect time settings on the client
+% machine.
 
@@
 We discuss these in turn.
-Codebase size and session length follow power-law
-distributions,\todo{Please confirm w/ a statistical test. See Clauset
-  et al.: arXiv:0706.1062.} with
-many small items and a few very large items.
+Codebase size and session length have long-tailed distributions
+with many small items and few very large ones.
+%% TODO check %% follow power-law distributions~\cite{csn-siamr-2009} 
+%% https://aaronclauset.github.io/powerlaws/
+%% bin version, is this relevant? sounds like no from abtract https://arxiv.org/pdf/1208.3524.pdf
 The number of type analysis errors is somewhat low~(explained in the next
@@
 
-\begin{figure}[t]\centering
+\begin{table}[t]\centering
+  \caption{Size of analyzed code: number of files, number of lines, and lines in edit range}
+  \label{t:codebase-size}
   % ;; (list* (max* vv) (median < vv) mm (stddev/mean mm vv) (percentile* vv)))
@@
 
-  \caption{Size of analyzed code: number of files, number of lines, and lines in edit range}
-  \label{f:codebase-size}
-\end{figure}
-
-\Cref{f:codebase-size} summarizes the size of projects in the dataset.
+\end{table}
+
+%% TODO figure --> table
+
+\Cref{t:codebase-size} summarizes the size of projects in the dataset.
 The top half of the figure reports the mean, standard deviation, median,
@@
 }
-The bottom half shows zoomed-in distributions of the line and edit range
+The distributions zoom in zoomed-in distributions of the line and edit range
 sizes.
-For example, the x-axis of the first plot ranges from 0 to 1,000 lines and
+For example, the x-axis of the first plot counts up to 1,000 lines and
 the y-axis counts up to 1,000 telemetry records.
@@
 Unsurprisingly, the numbers come with large standard
-deviations~([std] in~\cref{f:codebase-size}).
+deviations~([std] in~\cref{t:codebase-size}).
 The median values are more reasonable, with roughly 3,000 lines of code
@@
 
-\Cref{f:session-size} outlines the size of the Roblox Studio sessions in the
-data.
-There are two useful notions of size: real time in seconds (between the first
-and last record) and the total number of records.
+\Cref{f:session-size} outlines the size of sessions using two metrics:
+real time in seconds between the first and last record, and the total number of
+records.
 The top of~\cref{f:session-size} reports the mean, standard deviation,

### 2023-09-29 programming23/paper.tex
sec 4,5,7: creators -> sessions

@@
 records in three sessions.
-It is entirely possible that all these errors came from one creator who
+It is entirely possible that all these errors came from one session that
 deliberately pushed the limits of type analysis.
@@
 regardless of mode.
-No matter whether a creator is using \mnocheck{}, \mnonstrict{},
+No matter whether a session is using \mnocheck{}, \mnonstrict{},
 or \mstrict{} mode, edits seem to balance increases and decreases in the number
@@
     This confirms that changes to the number of errors are balanced.
-    Creators tend to fix errors highlighted by type analysis
+    Sessions tend to fix errors highlighted by type analysis
     even in the stricter analysis modes.
@@
 
-\todo{No identity $\Rightarrow$ no ``creators'', only sessions.}
 With the data in hand, we can answer the research questions
@@
 analyze thousands of lines of code.
-But, few creators use type analysis directly and the data presents little
-evidence that doing so would improve their quality of life.
+But, few sessions use type analysis directly and the data presents little
+evidence that doing so would improve creators' quality of life.
 We conjecture that Luau needs further tailoring to match
@@
   \item[How many creators choose to use type analysis?]
-    A mere \pct{10} of creators opt-in to type checking.
+    A mere \pct{10} of sessions (and of telemetry records overall) opt-in to
+    type checking~(\cref{f:dataset-overview}).
     Most of these use \mnonstrict{} mode, and fewer than \pct{1}
@@
   \item[How often do projects contain modules with different analysis modes?]
-    Projects rarely combine analysis modes.
+    Sessions rarely combine analysis modes.
     Only 512 of \nK{340} sessions ($<\pct{0.15}$) switched to modules with different
@@
   \item[How often do creators turn analysis off?]
-    Creators rarely disable type analysis after opting in.
+    Sessions rarely disable type analysis after opting in.
     Only 233 of the \nK{340} sessions ($<\pct{0.07}$)
@@
       \item[- How do creators respond? Which errors persist through edits?]
-        Creators typically fix analysis errors.
+        Sessions typically fix analysis errors.
         With few exceptions, increases and decreases in the number of errors
@@
     The percent of \FS{} errors in \mnocheck{} code matches the percent of
-    creators using \mnocheck{} mode~(\cref{f:error-by-mode}); whereas,
+    sessions using \mnocheck{} mode~(\cref{f:error-by-mode}); whereas,
     if type analysis helped reduce \FS{} errors, \mnocheck{} would own a
@@
 errors, if any, creators intended to fix with their edits.
-Thus, fixes may be a side effect of other plans, as seen in prior work [CITE].
-% Cite the Marceau SIGCSE 2011 paper. This is the [DIFF] code from the rubric.
+Thus, fixes may be a side effect of other plans, as seen in prior work~\cite{mfk-sigcse-2011}.
+% This is the [DIFF] code from Marceau SIGCSE 2011.
 
@@
 than what we observed.
+%% ben: not sure what to say ... may remove the error names comment
 Second, we know nothing about the creators in our study except that they
@@
 an industrial strength typechecker.
-With over 1.5 million telemetry records from over 300,000 sessions
+With over 1.5 million telemetry records from over 340,000 sessions
 that took place during a three-month timespan, the data has plenty to say about
 type errors and their overlap with program edits.
-At the same time, the data is carefully restricted to avoid leaking personal
+At the same time, the data is limited to avoid leaking personal
 and proprietary information: we know the number and kind of type errors
@@
   \item
-    Nearly all creators (\pct{90}) use the default ``no check''
+    Nearly all sessions (\pct{90}) use the default ``no check''
     typechecking mode~(\cref{f:dataset-overview}).
@@
   \item
-    Creators tend to fix errors immediately rather than dealing with them in a batch
+    Sessions tend to fix errors immediately rather than dealing with them in a batch
     or ignoring the errors altogether~(\cref{f:error-density,f:error-changes}).

### 2023-09-29 programming23/paper.tex
RQs: creator -> session

@@
 With this telemetry data, we investigate research questions about
-the adoption and benefits of type analysis: \todo{Some of these are flawed. We don't know anything about creators. Adjust
-  to match section 5.}
+the adoption and benefits of type analysis:
 \begin{description}
   \item[RQ1.]
-    How many creators use type analysis?
+    How many sessions use type analysis?
     How often do projects contain modules with different
     analysis modes?
-    How often do creators turn analysis off?
+    How often do sessions turn analysis off?
   \item[RQ2.]
@@
     which errors arise,
-    how do creators respond,
+    how do sessions respond,
     and which errors tend to persist despite subsequent edits?
@@
     and scripts to analyze them.
-    \emph{We intend to submit this in an artifact.}
+    \emph{We intend to submit these in an artifact.}
 
@@
 \begin{description}
-  \item[How many creators choose to use type analysis?]
+  \item[How many sessions use type analysis?]
     A mere \pct{10} of sessions (and of telemetry records overall) opt-in to
@@
 
-  \item[How often do creators turn analysis off?]
+  \item[How often do sessions turn analysis off?]
     Sessions rarely disable type analysis after opting in.
@@
 
-      \item[- How do creators respond? Which errors persist through edits?]
+      \item[- How do sessions respond? Which errors persist through edits?]
         Sessions typically fix analysis errors.

### 2023-09-29 programming23/paper.tex
threat: sessions vs people

@@
 
-\FILL{} cannot match sessions to codebases
-
 While this study has high ecological validity due to its focus on
@@
 of these actions.
-We have no access to other events (e.g., run locally, publish) due to privacy
-concerns; future studies may wish to avoid this limitation by actively
+We have no access to other events, e.g., run locally~(\cref{s:tel-limitations});
+future studies may wish to avoid this limitation by actively
 recruiting subjects.
@@
 
-Third, the total counts of type errors include a large number of syntax errors,
+Third, session-based data has limitations.
+A creator who closes and reopens Roblox Studio every day has
+a much higher chance of generating telemetry than one who
+leaves Studio open all week.
+We have no way to reliably check whether multiple sessions are
+reporting on the same codebase.
+Sessions may end abruptly, for instance because of machine reboot,
+instead of ending when the creator was finished editing code.
+
+Fourth, the total counts of type errors include a large number of syntax errors,
 which have little to do with the Luau type system and are presumably easy for
@@
 
-Fourth, type analysis reports several errors at once and we have no idea which
+Fifth, type analysis reports several errors at once and we have no idea which
 errors, if any, creators intended to fix with their edits.

### 2023-09-29 programming23/paper.tex
sec 4: 4.0 to 4.3 arghh

@@
 \newcommand{\nK}[1]{#1\textsc{k}}
-\newcommand{\stddev}[1]{[\nK{#1}]}
+\newcommand{\stddev}[1]{\nK{#1}}
 \newcommand{\gcell}[1]{\cellcolor{green!20}#1}
@@
 
-A close inspection of the data revealed a two anomalies
-that we removed from further analysis:
+A close inspection of the data revealed two anomalies:
 \begin{enumerate}
@@
   %% (2nd biggest has 41K files "31,650,473,913,529")
-
 %% Files has HUGE numbers thanks to data assets. Forget it.
@@
 % 51,761 &              &
-%   & \includegraphics[width=0.2\columnwidth]{img/files-distribution.pdf}
-
-%  \begin{tabular}{l@{}r@{~}l@{}r@{~}l}
-%                 &  Lines &             &    Edit Range & \\
-%    Mean [std]   &  6,497 & \stddev{22} &         3,680 & \stddev{31} \\
-%    Median       &  3,115 &             &           926 & \\
-%    \pct{99}     & 50,547 &             &        34,725 & \\
-%    %% TODO x-axis starts at 1, not zero ... there are no zeros right??!
-%      & \includegraphics[width=0.2\columnwidth]{img/lines-distribution.pdf}
-%    & & \includegraphics[width=0.2\columnwidth]{img/editrange-distribution.pdf}
-%
-%  \end{tabular}
-
-  \begin{tabular}[t]{lrrrl}
-    ~          & Mean [std] & Median & \pct{99} Pct. & Distribution \\
-    Lines      & 6,497 \stddev{22} & 3,115 & 50,547 & \includegraphics[width=0.14\columnwidth,valign=M]{img/lines-distribution.pdf} \\
-    Edit Range & 3,680 \stddev{31} & 926   & 34,725 & \includegraphics[width=0.14\columnwidth,valign=M]{img/editrange-distribution.pdf}
+  \begin{tabular}[t]{lrrrrl}
+    ~          & Mean  & Stddev & Median & \pct{99} Pct. & Distribution \\
+    Lines      & 6,497 & \stddev{22} & 3,115 & 50,547 & \includegraphics[width=0.14\columnwidth,valign=M]{img/lines-distribution.pdf} \\
+    Edit Range & 3,680 & \stddev{31} & 926   & 34,725 & \includegraphics[width=0.14\columnwidth,valign=M]{img/editrange-distribution.pdf}
     %% TODO x-axis starts at 1, not zero
     %% TODO much bigger font or much smaller plot!!!
-
   \end{tabular}
-
 \end{table}
 
-%% TODO figure --> table
-
 \Cref{t:codebase-size} summarizes the size of projects in the dataset.
-The top half of the figure reports the mean, standard deviation, median,
+The middle four columns report the mean, standard deviation, median,
 and \percentile{99} for the number of lines
@@
 }
-The distributions zoom in zoomed-in distributions of the line and edit range
+The plots on the right show zoomed-in distributions of the line and edit range
 sizes.
-For example, the x-axis of the first plot counts up to 1,000 lines and
-the y-axis counts up to 1,000 telemetry records.
+For example, the x-axis of the first plot counts up to 1,000 lines (not showing
+a long tail to the right) and the y-axis counts up to 1,000 telemetry records
+(not showing a tall spike on the left).
 
@@
 while the smallest have merely 1 line of code.
-Unsurprisingly, the numbers come with large standard
-deviations~([std] in~\cref{t:codebase-size}).
+Unsurprisingly, these wide-ranging numbers come with large standard
+deviations.
 The median values are more reasonable, with roughly 3,000 lines of code
@@
 
-\begin{figure}[t]\centering
+\begin{table}[t]\centering
+  \caption{Session size in seconds and in number of records.}
+  \label{t:session-size}
   %% NOTE max time = 1 million seconds = 15 days ...  +100 are over 2 days
-
-%  \begin{tabular}{l@{}r@{~}l@{}r@{~}l} \\
-%                 & Time Span (sec)  &       & Record Count  \\
-%    Mean [std]   &     3,184 & \stddev{16}  &     286 & [583] \\
-%    Median       &       845 &              &     138        \\
-%    \pct{99}     &    35,450 &              &   3,302        \\
-%    %% TODO x-axis starts at 1, not zero ... there are no zeros right??!
-%    & \includegraphics[width=0.2\columnwidth]{img/timespan-distribution.pdf}
-%    & & \includegraphics[width=0.2\columnwidth]{img/event-count-distribution.pdf}
-%
-%  \end{tabular}
-
-  \begin{tabular}{lrrrl} \\
-    ~               & Mean [std] & Median & \pct{99} Pct. & Distribution \\
-    Time Span (sec) & 3,184 \stddev{16} & 845 & 35,450 & \includegraphics[width=0.14\columnwidth,valign=M]{img/timespan-distribution.pdf} \\
-    Record Count    & 286 [583]         & 138 & 3,302  & \includegraphics[width=0.14\columnwidth,valign=M]{img/event-count-distribution.pdf}
+  \begin{tabular}{lrrrrl} \\
+    ~               & Mean & Stddev & Median & \pct{99} Pct. & Distribution \\
+    Time Span (sec) & 3,184 & \stddev{16} & 845 & 35,450 & \includegraphics[width=0.14\columnwidth,valign=M]{img/timespan-distribution.pdf} \\
+    Record Count    & 286 & 583         & 138 & 3,302  & \includegraphics[width=0.14\columnwidth,valign=M]{img/event-count-distribution.pdf}
     %% TODO x-axis starts at 1
     %% TODO huge font, or smaller plot
-
   \end{tabular}
-
-  \caption{Session size in seconds and in number of records.}
-  \label{f:session-size}
-\end{figure}
-
-\Cref{f:session-size} outlines the size of sessions using two metrics:
+\end{table}
+
+\Cref{t:session-size} outlines the size of sessions using two metrics:
 real time in seconds between the first and last record, and the total number of
 records.
-The top of~\cref{f:session-size} reports the mean, standard deviation,
-median, and \percentile{99}; the bottom shows size distributions, zoomed in
-to focus on at most 1,000 units (x-axis) and 1,000 sessions
-(y-axis).\todo{Again, confused.}
-
-Similar to the data on project size, there is huge variation in session
-length.
+The middle columns report the mean, standard deviation,
+median, and \percentile{99}.
+Similar to the previous table~(\cref{t:codebase-size}), the distributions focus
+on the lower-left fragment of the data, which again has a very long tail.
+%% comment on long-tail record counts?
 The longest sessions last several days and/or contain thousands of records,
 while the shortest last a few milliseconds or include just one record.
-Likewise, both plots in the figure follow a power-law distribution
-(especially for record count).
-A typical session runs on the order of minutes and consists of a few dozen events.
-The median session time is approximately 14 minutes.
-The median number of
-records is 138.
+% Likewise, both plots in the figure follow a power-law distribution
+% (especially for record count).
+A median session runs on the order of minutes and consists of a few dozen events.
+%% The median time is 14 minutes and the median 
+%% The median number of records is 138.
 % These numbers are reasonable.
@@
 
-\begin{figure}[t]
+\begin{table}[t]\centering
+  \caption{Current type errors and \FS{} errors across all telemetry records.}
+  \label{t:count-analysis-errors}
   \begin{tabular}[t]{ll} \\
-    \begin{tabular}[t]{r@{~~}r@{~}l}
+    \begin{tabular}[t]{r@{~~}r@{~}r}
       595,137 & type errors \\
-      289,698 & in current module & [\pct{48.68}] \\
-       30,924 & in edit range & [\pct{5.20}]
+      289,698 & \hbox{}~in module & [\pct{48.68}] \\
+       30,924 & \hbox{}~in edit range & [\pct{5.20}]
     \end{tabular}
-    \begin{tabular}[t]{r@{~~}r@{~}l}
+    \begin{tabular}[t]{r@{~}r@{~}r}
       72,235,735 & {\FS{} errors} \\
-      37,027,281 & in current module & [\pct{51.26}] \\
-       1,111,178 & in edit range & [\pct{1.54}]
+      37,027,281 & \hbox{}~in module & [\pct{51.26}] \\
+       1,111,178 & \hbox{}~in edit range & [\pct{1.54}]
     \end{tabular}
   \end{tabular}
-  \caption{Type errors and \FS{} errors across all telemetry records.}
-  \label{f:count-analysis-errors}
-\end{figure}
-
-\FILL{} need old and current in edit range
-
-\Cref{f:count-analysis-errors} lists the number of type analysis errors
+\end{table}
+
+\Cref{t:count-analysis-errors} lists the number of type analysis errors
 and breaks them down by location in the codebase.
 In total, there are over 590,000 type errors.
-%% delete? upcoming figure has the same info
-% Most of these (\pct{65}) appeared in modules that opted in to typechecking with either the
-% \mnonstrict{} or \mstrict{} mode; the remainder are syntax errors reported
-% by \mnocheck{} mode.
-The number of \FS{} errors is much larger, 72M, because this analysis
-ran on every module using strict type.
-(Furthermore, \FS{} runs silently, so creators have no
-awareness of or incentive to
-fix these errors.)
+The number of \FS{} errors is much larger, at 72 million, because this analysis
+uses rigorous checks and runs on every module no matter what type analysis
+mode the module declares.
+(Furthermore, \FS{} checks run silently, so creators have no
+awareness of or incentive to fix these errors.)
 Approximately half the analysis errors occurred in the current module,
@@
 
-The large fraction of errors in the current module is encouraging.
-It suggests that errors appear locally, as the result of edits to nearby
-code (as opposed to edits in one module raising errors in another), and
-that creators fix the errors before switching focus to another
+The large fraction of errors in the current module is encouraging
+for two reasons.
+First, it suggests that errors appear locally, as the result of edits to nearby
+code, rather than as the result of edits to far-away code that lives in another
 module.
-The low fraction of errors in the edit range suggests that errors rarely
-persist through nearby edits.
+Second, it suggests that creators typically fix errors before switching
+to another module.
+
+The low fraction of errors in the edit range sends a mixed message:
+\begin{itemize}
+  \item
+    On one hand, it suggests that edits successfully remove errors.
+    This is true, however, only for type errors and not for \FS{} errors.
+    There are 11,479 more type errors from the previous analysis
+    that overlap with the current edit range~(\pct{27} difference),
+    but 172,678 fewer previous \FS{} errors~(\pct{15} difference).
+    Since creators can see the type errors but not the \FS{}
+    errors before making edits, we conclude that the visibility
+    makes a difference.
+    \Cref{s:type-error-survival} explores edits and type errors in more detail.
+
+  \item
+    On the other hand, the low overlap between errors (whether current or previous)
+    and edits says that edits rarely target highlighted code.
+    There are several possible explanations:
+    analysis might report several errors and the edits only target one or two of them;
+    creators might ignore errors entirely and edit other code; or
+    creators might choose to edit a non-highlighted location to fix an error.
+    Our data cannot narrow down the answer.
+
+\end{itemize}
 
@@
 
-Creators have three analysis modes to choose from.
-Furthermore, they can switch between modes at any time.
-According to~\cref{s:type-analysis-modes}, however, usage is
-extremely skewed toward the type-free \mnocheck{} mode.
-Nearly \pct{90} of all records use the \mnocheck{} mode
-and \pct{90} of all sessions use \mnocheck{} exclusively.
-Among the rest, roughly \pct{10} of all records use \mnonstrict{} and only a
-tiny fraction (\pct{0.4}) use \mstrict{}.
-The per-session data is similar: \pct{9} use \mnonstrict{} exclusively
-and \pct{0.2} use \mstrict{}.
-
-These adoption numbers indicate that there is a communication gap;
-as long as type checking is opt-in, most developers stay opted out.
-Once Luau type checking is opt-out rather than opt-in, it will be interesting
-to revisit these numbers.
+Creators have three analysis modes to choose from
+and can switch between modes at any time.
+\Cref{f:dataset-overview} shows, however, that
+usage is extremely skewed toward the default \mnocheck{} mode.
+Nearly \pct{90} of all records use \mnocheck{}, while \pct{10} use
+\mnonstrict{} and only a tiny fraction~(\pct{0.4}) use \mstrict{} mode.
+Grouping these records into sessions shows that
+\pct{90} of all sessions use \mnocheck{} exclusively,
+\pct{9} use \mnonstrict{} exclusively,
+and \pct{0.2} use \mstrict{} exclusively.
+
+These adoption numbers indicate that as long as type checking is opt-in, most
+creators stay opted out.
+In the future, Luau plans to make \mnonstrict{} the default mode.
+It will be interesting to revisit mode usage at that point.
 
@@
 Most sessions (\pct{99.82}) stick to a single analysis mode.
-These sessions never change the mode in the current module and never switch
-focus to a module with a different analysis mode.
-Among the mixed-mode sessions, most (\pct{80}) switch to modules with
-a different mode, about half contain at least one edit that upgrades
-the typing strictness in the
-current mode, and about half contain a downgrade.
-There are 619 upgrades in total and 583 downgrades.
-\FILL{} Ben check how often the downgrades follow the upgrades.
+They never change the mode in the current module and never
+switch focus to a module with a different analysis mode.
+Among the other, mixed-mode sessions, most of those (\pct{80}) switch to a
+module with a different mode, about half contain at least one edit that
+upgrades to a more strict mode, and about half contain a downgrade to a
+less-strict mode.
+There are 263 total upgrades (across all sessions) and 320 total downgrades.
 
@@
 number of type errors.
-Creators might get discouraged or overwhelmed by a high error count and revert to
-\mnocheck{} mode.
+Creators might get discouraged or overwhelmed by a high error count.
 
 The data does not support this explanation.
-On average, the mode-upgrades in the data resulted in 3 additional
+On average, mode-upgrades resulted in 3 additional
 type errors (stddev 7, median 0).
-The worst-case increase was 57 type errors, which is very high
-but exceptional.
-In another exception, upgrading modes removed three type errors
-(possibly due to other edits being rolled in with the
-mode change).
+The worst-case increase was quite high at 57 type errors, but exceptional.
+In another exception, upgrading modes removed three type errors,
+possibly due to other edits being rolled in with the
+mode change.
 
@@
 But, creators evidently switch modes only when the code is already
-in well-typed shape.
+in a well-typed state.
 
@@
 %% FILL need \%s out of all module switches.
+%% count-modswitch in downgrade-count.rkt
 %% How many module switches exit a module with errors?
@@
     switch to a weaker mode.
-    Most of these (\pct{75}) downgrade from \mstrict{} to
-    \mnonstrict{} % BEN: confirm!
-    rather than from \mnonstrict{} mode.
-    %% FILL more insight if we match upgrades + downgrades in session?
+    One-fourth of the downgrades switch from \mstrict{}
+    to \mnonstrict{}, and therefore keep some type analysis.
+    The remainder switch from \mstrict{} to \mnocheck{} (\pct{50})
+    or from \mnonstrict{} to \mnocheck{} (\pct{25}).
+    %% FILL more insight if we match ups and downs together??
 
@@
     %% 176 with up
+    %% 263 total up
     %% #hash((nocheck . 0) (nonstrict . 44) (strict . 132))
@@
     %%  nonstrict : 25.00\%
+    %%%% #hash(((strict nocheck) . 88)
+    %%%%       (nonstrict . 0)
+    %%%%       (strict . 0)
+    %%%%       ((nonstrict nocheck) . 44)
+    %%%%       (nocheck . 0)
+    %%%%       ((strict nonstrict) . 44))
+    %%%%  (strict nonstrict) : 25.00\%
+    %%%%  nocheck : 0.00\%
+    %%%%  (nonstrict nocheck) : 25.00\%
+    %%%%  strict : 0.00\%
+    %%%%  nonstrict : 0.00\%
+    %%%%  (strict nocheck) : 50.00\%
     %%
@@
     %% 233 with down
+    %% 320 total down
     %% #hash((nocheck . 0) (nonstrict . 64) (strict . 169))
@@
     %%  nonstrict : 27.47\%
+    %%%% #hash(((strict nocheck) . 116)
+    %%%%       (nonstrict . 0)
+    %%%%       (strict . 0)
+    %%%%       ((nonstrict nocheck) . 64)
+    %%%%       (nocheck . 0)
+    %%%%       ((strict nonstrict) . 53))
+    %%%%  (strict nonstrict) : 22.75\%
+    %%%%  nocheck : 0.00\%
+    %%%%  (nonstrict nocheck) : 27.47\%
+    %%%%  strict : 0.00\%
+    %%%%  nonstrict : 0.00\%
+    %%%%  (strict nocheck) : 49.79\%
 

### 2023-09-30 programming23/paper.tex
type error in range + popularity

@@
 \newcommand{\zerowidth}[1]{\makebox[0pt][l]{#1}}
+\newcommand{\zeroheight}[1]{\raisebox{0pt}[0pt][0pt]{\(#1\)}}
 \newcommand{\nK}[1]{#1\textsc{k}}
@@
 \newcommand{\gbox}[1]{\colorbox{green!20}{#1}}
-\newcommand{\ybox}[1]{\colorbox{yellow!18}{#1\vphantom{Ap}}}
-\newcommand{\obox}[1]{\colorbox{orange!29}{#1\vphantom{Ap}}}
-\newcommand{\rbox}[1]{\colorbox{red!30}{#1\vphantom{Ap}}}
+\newcommand{\ybox}[1]{\colorbox{yellow!18}{#1}}
+\newcommand{\obox}[1]{\colorbox{orange!29}{#1}}
+\newcommand{\rbox}[1]{\colorbox{red!30}{#1}}
 \newcommand{\tefsfont}[1]{\textsc{#1}}
@@
 \newcommand{\panon}{pseudonymized}
+\newcommand{\symbox}[1]{{\makebox[0.8em][c]{{#1}}}}
+\newcommand{\addsym}{\symbox{\(\uparrow\)}}
+\newcommand{\keepsym}{\symbox{\(=\)}}
+\newcommand{\dropsym}{\symbox{\(\downarrow\)}}
+% \newcommand{\dsymbox}[1]{\makebox[0.8em][c]{#1}}
+% \newcommand{\dropsym}{\dsymbox{\raisebox{0ex}[0.9em]{\raisebox{0.5ex}{\vrule width 0.6em height 0.6pt}}}}
 
@@
 \paragraph{Number of Type Analysis Errors}
-%\label{s:count-analysis-errors}
+\label{s:count-analysis-errors}
 
@@
 
-  \newcommand{\labelbars}[1]{\begin{tabular}[t]{l@{}l} \begin{tabular}[t]{r} \mnocheck{} \\ \mnonstrict{} \\ \mstrict{} \\ ~ \end{tabular} & #1 \end{tabular}}
+  \newcommand{\labelbars}[1]{\begin{tabular}[t]{l@{}l} \raisebox{2ex}{\begin{tabular}[t]{r}\mnocheck{}\\\mnonstrict{}\\\mstrict{}\end{tabular}} & #1 \end{tabular}}
 
@@
   \end{tabular}
-  \caption{FILL}
+  \caption{Type and \FS{} errors grouped by mode.}
   \label{f:error-by-mode}
@@
 
-\Cref{f:error-by-mode} shows how the type errors and \FS{} errors
-in the data are split across type analysis modes.
-
-Most type errors come from \mnonstrict{} mode, despite it having an order
-of magnitude fewer users that \mnocheck{}.
-This is unsurprising because \mnocheck{} reports only syntax errors,
-whereas \mnonstrict{} reports several others kinds~(detailed in the next section).
-Finally, \mstrict{} mode accounts for a small fraction of type errors, despite
-being more picky, because so few creators use it.
-
-Background analysis applies the same typechecker to every module, regardless
-of mode, in the hope of enabling IDE tools.
-Consequently, the breakdown of \FS{} errors matches the overall usage
-data~(\cref{f:dataset-overview}): almost \pct{90} for \mnocheck{},
-\pct{10} for \mnonstrict{}, and \pct{1} for \mstrict{}.
-Type analysis finds a large number of errors in \mnocheck{} code;
-unfortunately, we have no insight into whether those errors are true
-positives or whether the analysis led to useful IDE suggestions.
-
-\FILL{} average number of FS errors is
-worst in \mnonstrict{}, 49, vs 48 nocheck and 7 strict.
-median is 3 NS, 1 nocheck, 1 strict.
-We might optimistically think that nocheck accumulates
-lots of errors, but no that does not seem to be the case,
-NS and nocheck both apparently fine with latent FS errors.
+Having seen the total number of errors in the dataset and the popularity of the
+three analysis modes, a next question is how many errors each mode
+accounts for.
+There should be a few type errors in the typical \mnocheck{} record (all of
+which would be syntax errors) and many more in the \mnonstrict{} and \mstrict{}
+records.
+For \FS{} errors, we would expect very few in \mstrict{} mode and very many in
+\mnocheck{} mode.
+Whereas \mnocheck{} mode performs little analysis and therefore gives fews hints
+as to how to fix \FS{} errors, \mstrict{} mode checks should closely match
+\FS{} checks.
+
+\Cref{f:error-by-mode} divides errors across modes.
+By the left plot, \mnonstrict{} accounts for two thirds of all type errors.
+This confirms our expectation.
+%% because \mnocheck{} reports only syntax errors and
+%% \mstrict{} covers only a small number of telemetry records.
+The right plot, by contrast, does not support our hypothesis that \mnocheck{}
+records have more \FS{} errors than the other modes.
+Though \mnocheck{} does have the most \FS{} errors, its proportion matches
+the proportion of \mnocheck{} records in the
+dataset~(\cref{f:total-records})---if every record had exactly one \FS{} error,
+the plot would be the same.
+
+Digging further into \FS{} errors, the median number of such errors by mode is
+1 in \mnocheck{}, 3 in \mnonstrict{}, and 1 in \mstrict{}.
+Records in \mnocheck{} mode evidently do not have significantly more \FS{}
+errors.
+Furthermore, type analysis does not seem to reduce \FS{} errors.
 
@@
 \begin{table}[t]
-  \caption{Number of telemetry records that add, keep, or drop (remove) specific type errors from their edit range.}
+  \caption{Number of telemetry records that increase (\addsym{}), preserve
+  (\keepsym{}), or decrease (\dropsym{}) the amount of a specific type
+  error from the edit range.}
   \label{t:type-error-survival}
-  \begin{tabular}{l@{}r@{}r@{}rr@{}r@{}rr@{}r@{}r}
+
+  \begin{tabular}{lr@{~}r@{~}rr@{~}r@{~}rr@{~}r@{~}r}
     & \zerowidth{\mnocheck{}} & & & \zerowidth{\mnonstrict{}} & & & \zerowidth{\mstrict{}} & & \\
-    & \rbox{Add} & \ybox{Keep} & \gbox{Drop} & \rbox{Add} & \ybox{Keep} & \gbox{Drop} & \rbox{Add} & \ybox{Keep} & \gbox{Drop} \\\midrule
-    \code{CannotCallNonFunction} & {0} & {0} & {0} & {17} & {4} & {20} & {1} & {0} & {1} \\
-    \code{CannotExtendTable} & {0} & {0} & {0} & {7} & {9} & {10} & {0} & {0} & {1} \\
-    \code{CannotInferBinaryOperation} & {0} & {0} & {0} & {1} & {2} & {1} & {7} & {6} & {9} \\
-    \code{CountMismatch} & {0} & {0} & {0} & {223} & {48} & {269} & {12} & {2} & {17} \\
-    \code{DuplicateTypeDefinition} & {0} & {0} & {0} & {1} & {0} & {0} & {0} & {0} & {0} \\
-    \code{FunctionDoesNotTakeSelf} & {0} & {0} & {0} & {4} & {6} & {5} & {0} & {0} & {0} \\
-    \code{FunctionExitsWithoutReturn} & {0} & {0} & {0} & {9} & {3} & {6} & {11} & {5} & {8} \\
-    \code{GenericError} & {0} & {0} & {0} & {206} & {52} & {203} & {8} & {1} & {9} \\
-    \code{GenericExtraInformation} & {0} & {0} & {0} & {48} & {7} & {39} & {2} & {0} & {1} \\
-    \code{IllegalRequire} & {0} & {0} & {0} & {16} & {3} & {33} & {0} & {0} & {2} \\
-    \code{IncorrectGenericParamCount} & {0} & {0} & {0} & {0} & {0} & {0} & {1} & {0} & {1} \\
-    \code{MissingProperties} & {0} & {0} & {0} & {11} & {3} & {11} & {6} & {4} & {6} \\
-    \code{MissingUnionProperty} & {0} & {0} & {0} & {1} & {0} & {2} & {0} & {0} & {0} \\
-    \code{ModuleHasCyclicDependency} & {0} & {0} & {0} & {20} & {9} & {15} & {0} & {0} & {1} \\
-    \code{NotATable} & {0} & {0} & {0} & {7} & {2} & {8} & {2} & {0} & {1} \\
-    \code{OccursCheckFailed} & {0} & {0} & {0} & {0} & {0} & {0} & {0} & {0} & {1} \\
-    \code{OnlyTablesCanHaveMethods} & {0} & {0} & {0} & {1} & {0} & {2} & {0} & {0} & {0} \\
-    \code{OptionalValueAccess} & {0} & {0} & {0} & {40} & {52} & {27} & {6} & {2} & {7} \\
-    \code{SyntaxError} & {8349} & {314} & {19220} & {983} & {46} & {2658} & {16} & {0} & {71} \\
-    \code{TypeMismatch} & {0} & {0} & {0} & {159} & {62} & {129} & {21} & {10} & {28} \\
-    \code{TypesAreUnrelated} & {0} & {0} & {0} & {1} & {0} & {0} & {0} & {0} & {1} \\
-    \code{UnknownPropButGotLikeProp} & {0} & {0} & {0} & {26} & {18} & {21} & {1} & {0} & {1} \\
-    \code{UnknownProperty} & {0} & {0} & {0} & {347} & {180} & {385} & {19} & {19} & {35} \\
-    \code{UnknownRequire} & {0} & {0} & {0} & {91} & {39} & {85} & {12} & {3} & {4} \\
-    \code{UnknownSymbol} & {0} & {0} & {0} & {2348} & {496} & {2973} & {49} & {19} & {55}
+    & \rbox{\addsym{}} & \ybox{\keepsym{}} & \gbox{\dropsym{}}
+    & \rbox{\addsym{}} & \ybox{\keepsym{}} & \gbox{\dropsym{}}
+    & \rbox{\addsym{}} & \ybox{\keepsym{}} & \gbox{\dropsym{}} \\\midrule
+    \code{CannotCallNonFunction} & {-} & {-} & {-} & {12} & {2} & {13} & {1} & {-} & {1} \\
+    \code{CannotExtendTable} & {-} & {-} & {-} & {6} & {8} & {1} & {-} & {-} & {1} \\
+    \code{CannotInferBinaryOperation} & {-} & {-} & {-} & {-} & {1} & {-} & {5} & {6} & {6} \\
+    \code{CountMismatch} & {-} & {-} & {-} & {184} & {42} & {157} & {9} & {2} & {9} \\
+    \code{DuplicateTypeDefinition} & {-} & {-} & {-} & {1} & {-} & {-} & {-} & {-} & {-} \\
+    \code{ExtraInformation} & {-} & {-} & {-} & {46} & {6} & {33} & {1} & {-} & {1} \\
+    \code{FunctionDoesNotTakeSelf} & {-} & {-} & {-} & {3} & {4} & {1} & {-} & {-} & {-} \\
+    \code{FunctionExitsWithoutReturning} & {-} & {-} & {-} & {6} & {1} & {5} & {7} & {3} & {5} \\
+    \code{GenericError} & {-} & {-} & {-} & {177} & {47} & {149} & {5} & {-} & {8} \\
+    \code{IllegalRequire} & {-} & {-} & {-} & {7} & {1} & {11} & {-} & {-} & {-} \\
+    \code{IncorrectGenericParameterCount} & {-} & {-} & {-} & {-} & {-} & {-} & {1} & {-} & {1} \\
+    \code{MissingProperties} & {-} & {-} & {-} & {8} & {3} & {6} & {5} & {3} & {3} \\
+    \code{ModuleHasCyclicDependency} & {-} & {-} & {-} & {9} & {5} & {8} & {-} & {-} & {-} \\
+    \code{NotATable} & {-} & {-} & {-} & {4} & {1} & {5} & {2} & {-} & {1} \\
+    \code{OccursCheckFailed} & {-} & {-} & {-} & {-} & {-} & {-} & {-} & {-} & {1} \\
+    \code{OnlyTablesCanHaveMethods} & {-} & {-} & {-} & {-} & {-} & {2} & {-} & {-} & {-} \\
+    \code{OptionalValueAccess} & {-} & {-} & {-} & {21} & {45} & {15} & {4} & {2} & {4} \\
+    \code{SyntaxError} & {-} & {-} & {8290} & {-} & {-} & {1149} & {-} & {-} & {29} \\
+    \code{TypeMismatch} & {-} & {-} & {-} & {103} & {51} & {80} & {13} & {6} & {18} \\
+    \code{UnknownPropButFoundLikeProp} & {-} & {-} & {-} & {20} & {17} & {13} & {-} & {-} & {-} \\
+    \code{UnknownProperty} & {-} & {-} & {-} & {256} & {156} & {208} & {16} & {18} & {22} \\
+    \code{UnknownRequire} & {-} & {-} & {-} & {43} & {30} & {37} & {5} & {3} & {3} \\
+    \code{UnknownSymbol} & {-} & {-} & {-} & {1992} & {438} & {1797} & {38} & {18} & {35} \\
   \end{tabular}
@@
 
-Equipped with an overall picture of the dataset and the skewed usage
-of \mnocheck{}, \mnonstrict{}, and \mstrict{} modes, we are ready to explore
-the type error data.
-Our main focus is on the intersection between type errors
-and edits to the programs.\todo{The next two sentences are in some
-  sense our main RQs, right? I would call them out more explicitly and
-  w/ more justification.}
-When a type error highlights part of the code and the next edit range includes
-that code, we want to know whether the error survives the edit.
-Similarly, we want to know when an edit introduces a type error.
-
-\Cref{t:type-error-survival} counts the number of times that types
-errors overlap with the current edit range.
-There are three sets of columns, corresponding to the \mnocheck{},
-\mnonstrict{}, and \mstrict{} type analysis modes.
-Within each set, the Add columns counts the number of telemetry records
-that introduce a certain type error.
-For example, seventeen of the \mnonstrict{} records add
-at least one \code{CannotCallNonFunction} error to the current edit range.
-The Keep columns counts records that have the same, non-zero number of type errors
-before and after.
-Lastly, the Remove columns count records that decrease the number of type errors.
-
-All three counts are imprecise because our telemetry does not track the exact
-identity of a type error.
-Suppose a creator removes one \code{CannotCallNonFunction} error but then
-immediately introduces another on a different line.
-\FILL{} beware, don't say this same thing Nx throughout the paper.
-\Cref{t:type-error-survival} reports this pair of edits as a Keep rather than
-an Add and a Remove because we cannot distinguish two errors of the same kind.
+The main focus of our telemetry design is to learn how type errors intersect
+with edit ranges.
+Indeed, most of the fields in a telemetry record are dedicated to this
+topic~(\cref{f:tdata}), and it helps to answer RQ2 from the introduction:
+\begin{description}
+  \item[RQ2.]
+    For modules that use type analysis:
+    which errors arise,
+    how do sessions respond,
+    and which errors tend to persist despite subsequent edits?
+\end{description}
+
+Using the results of the previous and current type analysis from
+each telemetry record,~\cref{t:type-error-survival}
+categorizes errors that overlap with the edit range.
+The three events of interest are: when edits introduce a type
+error, when edits preserve a type error, and when edits remove a
+type error.
+Hence, the table has three columns for each analysis mode.
+In the \mnonstrict{} columns, for example, seventeen records
+introduce (\addsym{}) at least one \code{CannotCallNonFunction} error.
+Four records keep the number for that error at the same, nonzero level,
+and twenty reduce the number of \code{CannotCallNonFunction} errors.
+
+All three counts are imprecise because our telemetry system cannot distinguish
+particular type errors.
+If a creator removes one \code{NotATable} error but then
+immediately introduces another on a different line,
+then the centor (\keepsym{}) column gets incremented
+instead of the increase (\addsym{}) and decrease (\dropsym{}) columns.
+
+This table is based on the 996,164 records generated from keystroke events.
+It ignores records based on module switches because those do not have
+meaningful edit ranges.
+One implication of the randomized strategy is that we may not notice
+when a creator fixes an error; the fix must be selected by the
+telemetry system.
+%% \QALAN{}: what does an edit range mean for a module-switch record?
+%%  there are 2 modules involved but only one range!
 
@@
 
-%% \paragraph{Most Type Errors are Typos}
 \begin{itemize}
   \item
-    \FILL{} which survive the longest, most keeps???
-
-  \item
-    \FILL{} beware counts include module switches. Why???
-    Is that defensible at all?
-
-  \item
-    \FILL{} \mnonstrict{} should have no false positives,
-    \mstrict{} doesn't care.
-    Why so many keeps in NS??!
-
-  \item
-    One striking aspect of~\cref{t:type-error-survival} is the large number
-    of \code{SyntaxError}s and the related errors \code{UnknownSymbol} and
-    \code{UnknownProperty}.
-    Most type errors in the data are likely due to typos.
-    \Cref{s:type-error-count} discusses this point in detail.
-    For now, we remark that \code{SyntaxError} is the only possible error
-    in \mnocheck{} mode (by design) and observe that these errors are
-    usually in the Drop column and seldom in the Keep column.\todo{So?}
-
-  \item
-    The numbers are relatively low in general, which means that
-    few errors in the dataset overlap with the edit range
-    The highest value in the \mstrict{} columns is 71, meaning 71 records
-    out of 6,505 total~(\cref{f:dataset-overview} drop a syntax error.
-    The \mnonstrict{} and \mnocheck{} columns are similar.
-    %% FILL show percentages for all three modes?
-    \FILL{} what does this mean? Not sure if bad, or inconclusive.
-
-  \item
-    %% Drop >= Add >= Rem ~= 27 triples
-    %% other nonzero      ~= 13 triples
-    When type errors overlap with the edit range,
-    we would hope the number of Adds and Drops is roughly
-    equal and the number of Keeps is low.
-    This is indeed the case.
-    In most Add/Keep/Drop triples, the Drop column has the highest
-    value and the Keep column has the lowest value.
-    The optimistic conclusion is that creators are fixing the code
-    highlighted by errors.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%  STOP!  ROUGH TEXT BELOW     %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-  \item
-    After the syntax errors, \code{CountMismatch},
-    \code{TypeMismatch}, and \code{GenericError} are the most common
-    \mnonstrict{} and \mstrict{} errors.
-    \FILL{} so what.
-
-  \item
-    \FILL{} which errors break the Drop > Add > Keep ``preferred order''?
-
-  \item
-    (NOTE may want to save relative popularity for next section.)
-    Some errors occur more often in \mnonstrict{} than \mstrict{}:
-    \code{CannotCallNonFunction},
-    \code{CannotExtendTable},
-    \code{GenericExtraInformation},
-    \code{FunctionDoesNotTakeSelf},
-    \code{IllegalRequire},
-    \code{ModuleHasCyclicDependency},
-    \code{OptionalValueAccess},
-    \code{UnknownPropButGotLikeProp}.
-    \QALAN{} are any of these surprising, or expected from the bigger number of
-    \mnonstrict{} records?
-
-  \item
-    A few errors are more common in \mstrict{} than \mnonstrict{}:
-    \code{CannotInferBinaryOperation},
-    \code{FunctionExitsWithoutReturn},
-    \code{IncorrectGenericParamCount}.
-    \FILL{} so what.
+    The numbers in the table are low overall.
+    For instance, the highest \mstrict{} count is 38
+    records out of the 6,505 total~(\cref{f:dataset-overview}).
+    \Cref{s:count-analysis-errors} discussions the implications
+    of this low overlap rate for future work.
+
+  \item
+    \code{SyntaxError} and the related errors \code{UnknownSymbol} and
+    \code{UnknownProperty} are the most popular.
+    If the edit range errors are indicative of type errors at large,
+    then most errors in the data are likely due to typos.
+    These errors also survive (\keepsym{}) the most often.
+
+    \subitem The random-keystroke strategy of our telemetry may be inflating
+    the number of typos because it has no way to avoid reporting edits that
+    happen in the middle of things.
+    For example, a creator who writes a method call
+    \code{bucket.countFish()} letter-by-letter will see errors
+    including \code{UnknownSymbol(buck)} and \code{UnknownProperty(bucket.c)}.
+    Telemetry may pick these up;~\cref{s:conclusion} suggests ways
+    to avoid doing so in the future.
+
+  \item
+    After the syntax errors, \code{CountMismatch}, \code{TypeMismatch}, and
+    \code{GenericError} are the next most common \mnonstrict{} and \mstrict{}
+    errors.
+    A \code{CountMismatch} sends the wrong number of arguments to a function.
+    It may simply be a common error, but it may also be due to our random-keystroke
+    telemetry.
+    The high incidence of \code{TypeMismatch} and \code{GenericError} calls for
+    further study in future work to determine precisely which error arose.
+    (Some of the \code{GenericError}s come with context strings in the form of
+    \code{ExtraInformation} errors, but telemetry does not report these strings.)
+    %% GenericError = only a few things: extend non-class, iterate non-table, ...?
+
+  \item
+    \code{OptionalValueAccess} frequently persists through edits (\keepsym{}).
+    Creators may be ignoring this error because they can deduce that the
+    optional value will be present at runtime.
+
+  \item
+    %% Drop >= Add >= Rem ~= 18
+    %% other nonzero      ~= 21 triples
+    Ideally, the highest number in each column group should
+    be the number of decreases (\dropsym{}).
+    A decrease should happen when the creator sees an error,
+    edits around the highlighted code, and fixes it.
+    However, decreases are highest in only 18 of the 39 nonzero groups.
+
+%%  \item
+%%    \FILL{} \mnonstrict{} should have no false positives,
+%%    \mstrict{} doesn't care.
+%%    Why so many keeps in NS??!
+
+%%  \item
+%%    A few errors are more common in \mstrict{} than \mnonstrict{}:
+%%    \code{CannotInferBinaryOperation},
+%%    \code{FunctionExitsWithoutReturn},
+%%    \code{IncorrectGenericParamCount}.
+%%    \FILL{} so what.
+
+%%   \item
+%%     (NOTE may want to save relative popularity for next section.)
+%%     Some errors occur more often in \mnonstrict{} than \mstrict{}:
+%%     \code{CannotCallNonFunction},
+%%     \code{CannotExtendTable},
+%%     \code{GenericExtraInformation},
+%%     \code{FunctionDoesNotTakeSelf},
+%%     \code{IllegalRequire},
+%%     \code{ModuleHasCyclicDependency},
+%%     \code{OptionalValueAccess},
+%%     \code{UnknownPropButGotLikeProp}.
+%%     \QALAN{} are any of these surprising, or expected from the bigger number of
+%%     \mnonstrict{} records?
 
 \end{itemize}
-
-\paragraph{Do Error Highlights Help?}
-
-\FILL{} data seems to say yes.
 
@@
   %% NOTE: popularity looks similar when [over all records] vs. [restricted to module switch records]
-  \caption{Type error popularity for \mnonstrict{} and \mstrict{} modes (all of the type errors in \mnocheck{} mode are \code{SyntaxError}s).}
+  \caption{Type error popularity for \mnonstrict{} and \mstrict{} modes.}
+  %% {(all of the type errors in \mnocheck{} mode are \code{SyntaxError}s).}
   \label{t:type-error-count}
@@
 
-\Cref{t:type-error-count} reports the percent of
-type errors that overlap with the edit range.
-This table is based on the same data as the previous table~(\cref{t:type-error-survival}),
-but focuses on the overall popularity of type errors
-rather than their relation to edits.
-The left column in each row names one kind of type error
-that overlaps with an edit range.
-The right column shows what percentage of all overlapping
-errors the current error took up.
-
-There are two subtables, for \mnonstrict{} and \mstrict{} mode.
+\Cref{t:type-error-count} reports the percent of type errors that overlap with
+the edit range.
+The purpose of this table is to discover all specific errors in the dataset,
+thus it includes both keystroke-based and module-switch-based records---even
+though edit ranges are not well-defined for the latter.
+
+There are two subtables for \mnonstrict{} and \mstrict{} mode.
 The table does not report \mnocheck{} because all its type errors are
@@
   \item
-    As we noted above, the most popular errors are syntax errors
-    (\FILL{} \pct{0} \mnonstrict{}, \pct{0} \mstrict{})..
-    This is undoubtedly due to the stochiastic nature
-    of our telemetry, which has no way to avoid
-    reporting edits that happen in the middle of things.
-    For example, a creator who writes a method call
-    \code{bucket.countFish()} letter-by-letter will
-    get type analysis to find errors like \code{UnknownSymbol(buck)}
-    and \code{UnknownProperty(bucket.c)}.
-    These are transient issues, but telemetry may catch them.
-
-  \item
-    \FILL{} huge syntax error gap between S nonstrict.
+    Syntax errors, in particular \code{UnknownSymbol}, are much more
+    common in \mnonstrict{} than \mstrict{}.
+    This may be because \mnonstrict{} reports fewer errors in general.
+
+  \item
+    Despite being less strict, \mnonstrict{} reported four errors that
+    never appeared in the \mstrict{} records:
+    \code{DuplicateTypeDefinition}, \code{OnlyTablesCanHaveMethods},
+    \code{MissingUnionProperty}, and \code{FunctionDoesNotTakeSelf}.
+    We attribute this to the much larger number of \mnonstrict{} records in the
+    dataset.
 
@@
     Several other errors have higher \mstrict{} percentages for the
-    same reason, for example, \code{CannotInferBinaryOperation}.
-
-  \item
-    Two errors that appear in the \mstrict{} records
-    never appear in \mnonstrict{}: \code{OccursCheckFailed} and \code{IncorrectGenericParamCount}.
-    We attribute these to more advanced uses of types in \mstrict{} mode.
-
-  \item
-    Four errors from the \mnonstrict{} records never appeared in \mstrict{} mode:
-    \code{DuplicateTypeDefinition}, \code{OnlyTablesCanHaveMethods},
-    \code{MissingUnionProperty}, and \code{FunctionDoesNotTakeSelf}.
-    (\QALAN{} is this interesting, or just a result of having less strict data?)
-
-  \item
-    \FILL{} which errors that NEVER showed up?
-    Nothing for \code{CodeTooComplex} or \code{UnificationTooComplex} is great news.
-    But what about the others?
-    \code{DynamicPropertyLookupOnClassesUnsafe}?
-    \code{SwappedGenericTypeParameter}?
+    same reason, such as \code{CannotInferBinaryOperation}, and
+    two errors appeared only in \mstrict{}:
+    \code{OccursCheckFailed} and \code{IncorrectGenericParamCount}.
+
+  \item
+    Several Luau errors never appeared in edit ranges.
+    Four of these deal with internal typechecker limits, which we discuss
+    below.
+    The others are for: dynamic property lookups,
+    deprecated APIs, swapped generic parameters,
+    functions that require a self, and type pack mismatches.
+    We would not expect these to appear often.
+
+    % \code{DynamicPropertyLookupOnClassesUnsafe},
+    % \code{SwappedGenericTypeParameter},
+    % \code{DeprecatedApiUsed},
+    % \code{DuplicateGenericParameter},
+    % \code{FunctionRequiresSelf}, and
+    % \code{TypePackMismatch}.
 
@@
 For the specific case of \code{CodeTooComplex}, our telemetry tracked project-wide
-counts and found only 26 of these errors, which were spread across eleven
+counts and found only 26 errors, which were spread across eleven
 records in three sessions.
-It is entirely possible that all these errors came from one session that
+It is entirely possible that they all came from one session that
 deliberately pushed the limits of type analysis.
-Whatever the reason, the overall number is low enough to suggest that the
-internal limits are not posing barriers to typechecker adoption.
-
-\subsection{Type Error Density}
+%% Whatever the reason, the overall number is low enough to suggest that the
+%% internal limits are not posing barriers to typechecker adoption.
+
+\subsection{Density of Type Errors}
 
@@
   %% - nocheck indeed the highest > NS > S
+  %% TODO draw units, bigger font
 
@@
   \includegraphics[width=\columnwidth]{img/error-count-strict-row--te-density-diff.pdf}
-  \caption{Changes in type error density (errors in module / lines in module) over time (seconds).}
+  \caption{Changes in type error density (errors-in-module / lines-in-module) over time.}
   \label{f:error-density}

### 2023-09-30 programming23/paper.tex
mitesh comments

@@
 \newcommand{\percentile}[1]{{#1}th~percentile}
+\newcommand{\ptile}[1]{\textsc{p}{#1}}
 \newcommand{\modefont}[1]{\texttt{#1}}
@@
 misspellings of property names are common enough to report in both
-\mstrict{} and \mnonstrict{} mode. See~\cite{bfj-hatra-2021}
-for a more detailed discussion of the rationale for \mstrict{} and
-\mnonstrict{} mode.
+\mstrict{} and \mnonstrict{} mode.
+\mnocheck{} reports no error because the program is syntactically valid.
+See~\cite{bfj-hatra-2021} for a more detailed discussion of the rationale
+behind the type systems.
 
@@
 % 51,761 &              &
+
+  %% FILL P99 or P01
   \begin{tabular}[t]{lrrrrl}
-    ~          & Mean  & Stddev & Median & \pct{99} Pct. & Distribution \\
+    ~          & Mean  & Stddev & Median & \ptile{99} & Distribution \\
     Lines      & 6,497 & \stddev{22} & 3,115 & 50,547 & \includegraphics[width=0.14\columnwidth,valign=M]{img/lines-distribution.pdf} \\
@@
 projects may contain hundreds of files that define data assets.  For the
-record, the median file count is 7,678 and the 99th percentile is 51,761 files.
+record, the median file count is 7,678 and the \percentile{99} is 51,761 files.
 }
@@
   \begin{tabular}{lrrrrl} \\
-    ~               & Mean & Stddev & Median & \pct{99} Pct. & Distribution \\
+    ~               & Mean & Stddev & Median & \ptile{99} & Distribution \\
     Time Span (sec) & 3,184 & \stddev{16} & 845 & 35,450 & \includegraphics[width=0.14\columnwidth,valign=M]{img/timespan-distribution.pdf} \\
@@
     or ignoring the errors altogether~(\cref{f:error-density,f:error-changes}).
+    This should give Roblox confidence to make \mnonstrict{} the default in the future.
     % Changes to the number of errors balance one another~(\cref{f:error-density});

### 2023-09-30 programming23/paper.tex
shrinking 6

@@
 % vim: set fileencoding=utf-8 :
+
+% TODO BEFORE SUBMISSION
+% be consistent between "IDs" and "ids" [SK prefers "IDs" to avoid it being read as a word]
+
+% TODO BEFORE FINAL
+% - uncomment acks
 
@@
 
-Type error telemetry in Roblox Studio has two distinguishing characteristics:
-it includes no personal information (PII), and it sends data on
+Our telemetry has two distinguishing characteristics:
+it includes no private information (both PII and source code details),
+and it sends data on
 randomly-selected runs of the type checker rather than specific events of interest.
 To the best of our knowledge, this design occupies a unique position relative to prior
-work~(\cref{t:telemetry-design}).
+work~(\cref{t:telemetry-design}).\todo{Where do these labels come
+  from?!? There are no ``prior work'' citations either!}
 
@@
 IntelliJ~\cite{intellij-telemetry}, and \code{.NET}~\cite{dotnet-telemetry}, telemetry may include all sorts of
-data: error reports, source code fragments, timestamps, and filepaths.
+data: error reports, source code fragments, time\-stamps, and filepaths.
 Users may be able to opt out of some telemetry, but the details depend on the
@@
 have enabled dozens of contributions to the area of CS
-education~\cite{bask-icer-2018}, ranging from categorization of novice
-errors~\cite{mk-fie-2014,m-masters-2016} to analyses of error frequency and
-time-to-fix~\cite{ab-sigcse-2015}.
-Related prior work used the GRUMPS telemetry system to collect 4.7 million
-keystroke-level actions from learners programming in
-Ada~\cite{tm-iticse-2004,tkdmceg-ascilite-2003}.
-This data is not available.
+education~\cite{bask-icer-2018}. However, these operate at the
+opposite end of privacy.
+
+% BEN: saving space.
+% , ranging from categorization of novice
+% errors~\cite{mk-fie-2014,m-masters-2016} to analyses of error frequency and
+% time-to-fix~\cite{ab-sigcse-2015}.
+% Related prior work used the GRUMPS telemetry system to collect 4.7 million
+% keystroke-level actions from learners programming in
+% Ada~\cite{tm-iticse-2004,tkdmceg-ascilite-2003}.
+% This data is not available.
+% BEN: Not sure what the above sentence means and whether we should
+% keep it. Is it that you can't freely get Blackbox, you have to ask
+% Neil Brown for permission or something?
 
@@
 session ids.
-The data is useful for learning about the frequency of specific events and the
-distribution of events across clients (e.g., how many \code{CodeTooComplex}
-errors arise? what is the average number of type errors?).
-However, the lack of timestamps and ids means that it cannot track edits
-over time (a feature that is quite useful to us, for example, in~\cref{f:error-density}).
+While useful for learning about the frequency and distribution of
+specific events, the lack of timestamps and ids means it cannot track
+edits over time (which we use in \cref{f:error-density}).
+% BEN: commented for space.
+% it cannot track 
+% The data is useful for learning about the frequency of specific events and the
+% distribution of events across clients (e.g., how many \code{CodeTooComplex}
+% errors arise? what is the average number of type errors?).
+% However, the lack of timestamps and ids means that it cannot track edits
+% over time (a feature that is quite useful to us, for example, in~\cref{f:error-density}).
 % In~\cref{f:error-density}, for example, we see that increases to the number

### 2023-09-30 programming23/paper.tex
more sec 6 cleanup and misc

@@
 % be consistent between "IDs" and "ids" [SK prefers "IDs" to avoid it being read as a word]
+% even more so between "ID" and "id" [NB: "id" is a word in English]
+% \usepackage[sort]{natbib} or otherwise manually fix cites to be in ascending order
 
@@
 Our approach to data analysis is inspired by Marceau
-et~al.~\cite{mfk-onward-2011,mfk-sigcse-2011}, who used keystroke-level data
-collected over 6 weeks to analyze students' responses to runtime errors
-in DrRacket~\cite{fcffksf-jfp-2002}.
-Similarly, Macedo et~al.~\cite{mcpcsprs-abz-2020,mcpcsprs-scp-2021} capture
-edit sessions from the online Alloy4Fun editor and present several automated
-analyses.
-A major challenge in our setting, however, is that edit sequences are
-incomplete---there may be minutes-long gaps between two telemetry records.
-Several more-recent works have studied edit sequences
-to draw conclusions about errors and intent, e.g.,~\cite{wk-koli-2020,lgk-pj-2022,rsgl-cpp-2020}.
-
-The BitFit project collects data about beginning students interactions with a
-Java IDE~\cite{ekc-wccce-2016,anna-russo-kennedy-ms-2006}; however, unlike Blackbox, it does
-not collect source code.
-BitFit data consists of a timestamp, a session id, and counter values showing
-the number of attempts at a question, the number of hints requested, the number
-of compiles and runs, and the number of submission attempts.
-On one hand, this data is more informative than Roblox Studio's because it
-targets specific events rather than random keystrokes.
-On the other hand, there are no details about errors to show what students
-struggled with.
+et~al.~\cite{mfk-onward-2011,mfk-sigcse-2011} and
+Macedo et~al.~\cite{mcpcsprs-abz-2020,mcpcsprs-scp-2021}
+for Racket and Alloy4Fun, respectively. Other more recent work has
+also studied edit sequences to infer intent~\cite{wk-koli-2020,lgk-pj-2022,rsgl-cpp-2020}.
+A major difference is that we have only samples, not full sequences,
+which significantly impacts our study methods.\todo{Reasonable rewrite?}
+
+The BitFit project collects data about beginning students interactions in
+Java~\cite{ekc-wccce-2016,anna-russo-kennedy-ms-2006}; it does
+not collect source code.\todo{Does it sample? If not, say so!}
+However, its data do not contain details about errors.\todo{Really?!?}
 Other logging-based studies of Java programs use the full error message to
-draw conclusions~\cite{bgimgm-cse-2016,dlc-iticse-2014}.
-Roblox Studio cannot follow this approach because strings in errors may leak
-personal information.
-
-Much work on the benefits of static types focuses on a small number
-of case studies~\cite{w-popl-1986,hw-scp-2004,td-tosem-2001},
+draw conclusions~\cite{bgimgm-cse-2016,dlc-iticse-2014}, which we
+cannot use to preserve privacy.
+
+Much work~\cite{empirical-types,heeren-thesis} on the benefits of static types focuses on
+small case studies~\cite{w-popl-1986,hw-scp-2004,td-tosem-2001},
 inverviews~\cite{cdhhjklwya-hatra-2020,gstf-hatra-2021,cams-oopsla-2020},
-or static corpuses~\cite{rppd-fse-2014,bhmvv-toplas-2019,bmvv-arxiv-2019}
-(for literature reviews, see~\cite{empirical-types,heeren-thesis}).
-Controlled experiments, for example comparing statically-typed Java and dynamically-typed
-Groovy~\cite{khrts-icpc-2012}, are a notable exception.
-Lerner et~al.~\cite{lfgc-pldi-2007} apply type error repair tool on over two thousand
-of programs written during five homework exercises.
+or static
+corpuses~\cite{rppd-fse-2014,bhmvv-toplas-2019,bmvv-arxiv-2019}.\todo{Why
+  are we citing arXiv?}
+Controlled experiments are rare;
+Kleinschmager, et~al.\cite{khrts-icpc-2012} is a notable exception.
+Lerner et~al.~\cite{lfgc-pldi-2007} apply a type error repair tool on
+over two thousand programs.
 Seidel et~al.~\cite{sscwj-oopsla-2017,sjw-jfp-2018} use five thousand programs
 to train data-driven methods for localizing errors.
-These large studies with several thousands of programs are
-nevertheless on a much smaller scale than Roblox Studio.
+These are all on a vastly smaller scale than our study and also study
+different issues.
 
@@
 %% -----------------------------------------------------------------------------
-\acks
-
-Thanks to Benjamin Chung for several helpful discussions about data analysis
-and effective plotting.
-Greenman was supported by
-NSF grant \href{https://nsf.gov/awardsearch/showAward?AWD_ID=2030859&HistoricalAwards=false}{2030859}
-to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
+% \acks
+
+% Thanks to Benjamin Chung for several helpful discussions about data analysis
+% and effective plotting.
+% Greenman was supported by
+% NSF grant \href{https://nsf.gov/awardsearch/showAward?AWD_ID=2030859&HistoricalAwards=false}{2030859}
+% to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
 

### 2023-09-30 programming23/paper.tex
4.7 4.8 prose

@@
 % 51,761 &              &
-
-  %% FILL P99 or P01
   \begin{tabular}[t]{lrrrrl}
@@
   \includegraphics[width=\columnwidth]{img/error-count-strict-row--te-density-diff.pdf}
-  \caption{Changes in type error density (errors-in-module / lines-in-module) over time.}
+  \caption{Changes in type error density (errors / lines) over time (seconds).}
   \label{f:error-density}
@@
 
-Moving on to type errors across the projects, and not just in the edit range,
-a first interesting question is to study how the overall number of errors
-changes over time.
-But, the sheer number of errors can be skewed by codebase size.
-To give a better basis for project-to-project comparisons, we focus
-on the density of errors rather than the count, where \emph{density} is
-the total number of type errors divided by the lines of code in the project.
-
-\Cref{f:error-density} presents a zoomed-in view of type error density as it
-changes over time in every session of the dataset.
-The x-axis counts time in seconds from the start of the session.
-The y-axis shows the change in density from the previous telemetry record
-in the session to the next.
-Changes can be positive or negative depending on whether the edits added
-or removed errors.
-The plots cut off at$\pm{}50$ density and 150 seconds.to focus on the
-bulk of the data.
-%% Sadly even with the zoom, it's too hard to follow one session ... the plot only gives a big-picture impression.
+Since errors and edits may not overlap, either because the highlights
+are misleading or because the creator ignores the errors, it is important
+to see how the total errors change over time.
+We measure \emph{error density} rather than error count, however,
+to normalize for the wide range of codebase sizes~(\cref{t:codebase-size}).
+Density is the number of type errors divided by the number of lines.
+
+\Cref{f:error-density} plots changes to error density for every session in the dataset.
+The x axis counts time in seconds from the start of the session.
+Most sessions last a few dozen seconds~(\cref{t:session-size}), so the axis
+ends at 150 seconds.
+The y axis shows changes to errors over time; a positive number means errors
+increased since the last record, and a negative number represents a decrease.
+While there are a few errors that go outside the plot bounds
+(max \mnocheck{}: 101, max \mnonstrict{}: 3577, max \mstrict{}: 87), we focus
+on $\pm{}50$ errors because most of the mass is in that region.
+The points are color-coded to give each session the same color, but this is not
+very useful because there are many more sessions than colors.
 
@@
 
-Due to the relative adoption of type analysis, there is much more
-data for the weak \mnocheck{} mode than for \mnonstrict{},
-and in turn for \mnonstrict{} relative to \mstrict{}.
-Still, the plots reveal interesting points:
+% Due to the relative adoption of type analysis, there is much more
+% data for the weak \mnocheck{} mode than for \mnonstrict{},
+% and in turn for \mnonstrict{} relative to \mstrict{}.
+% Still, the plots reveal interesting points:
 
@@
     smaller.
-    Why?
-    There are also fewer sessions (\mstrict{} makes that very clear).
-
-  \item
-    \mnonstrict{} has the biggest number of up/down deltas in the 10+ range.
+    These long-running sessions are not introducing errors.
+    %% TODO check #mod switch in long sessions
+
+  \item
+    \mnonstrict{} has the biggest fluctuations in errors, whereas \mstrict{}
+    has the smallest.
+    One explanation is that some \mnonstrict{} sessions ignore type analysis
+    (in particular the one with over 3,000 errors!) whereas most \mstrict{}
+    sessions pay close attention to it.
+
+  %% \item
+  %%   TODO plot FS with a much bigger range ... tells a different story about buildup?
+
 \end{itemize}
@@
   % FS mod: strict & 488 & [\pct{36.53}] & 344 & [\pct{25.75}] & 504 & [\pct{37.72}] \\
-  \begin{tabular}{l@{~}lr@{}rr@{}rr@{}r}
-    & & \zerowidth{Add} & & \zerowidth{Keep} & & \zerowidth{Drop} \\\midrule
-    type err. & \mnocheck{}   & 20085 & [\pct{49.83}] & 906 & [\pct{2.25}] & 19317 & [\pct{47.92}] \\
-              & \mnonstrict{} & 13030 & [\pct{40.22}] & 5513 & [\pct{17.02}] & 13852 & [\pct{42.76}] \\
+  \begin{tabular}{l@{~~}lr@{}rr@{}rr@{}r}
+    & & \zerowidth{\rbox{\addsym{}}} & & \zerowidth{\ybox{\keepsym{}}} & & \zerowidth{\gbox{\dropsym{}}} \\\midrule
+    type      & \mnocheck{}   & 20085 & [\pct{49.83}] & 906 & [\pct{2.25}] & 19317 & [\pct{47.92}] \\
+    error     & \mnonstrict{} & 13030 & [\pct{40.22}] & 5513 & [\pct{17.02}] & 13852 & [\pct{42.76}] \\
               & \mstrict{}    & 419 & [\pct{39.31}] & 230 & [\pct{21.58}] & 417 & [\pct{39.12}] \\
@@
     \FS       & \mnocheck{}   & 259865 & [\pct{39.09}] & 175643 & [\pct{26.42}] & 229271 & [\pct{34.49}] \\
-              & \mnonstrict{} & 33988 & [\pct{39.74}] & 20738 & [\pct{24.25}] & 30798 & [\pct{36.01}] \\
+    error     & \mnonstrict{} & 33988 & [\pct{39.74}] & 20738 & [\pct{24.25}] & 30798 & [\pct{36.01}] \\
               & \mstrict{}    & 488 & [\pct{36.53}] & 344 & [\pct{25.75}] & 504 & [\pct{37.72}] \\
@@
 
-The apparent visual symmetry in the density plots~(\cref{f:error-density})
+The visual symmetry in the density plots~(\cref{f:error-density})
 suggests that edits add and remove errors with rougly equal frequency,
 regardless of mode.
-No matter whether a session is using \mnocheck{}, \mnonstrict{},
-or \mstrict{} mode, edits seem to balance increases and decreases in the number
-of type errors.
 \Cref{f:error-changes} explores this question of balance in detail.
-For each of the three modes, it reports the percent of edits that increase (Add),
-maintain at a nonzero level (Keep), or decrease (Drop) the number of type
-errors in the project.
+For each of the three modes, it reports the percent of edits that increase (\addsym{}),
+maintain (\keepsym{}), or decrease (\dropsym{}) the number of type
+errors in the module.
+It also reports counts for \FS{} checks to see whether fixing type errors fixes
+\FS{} errors as well.
+The ``maintain'' rows include cases where there are zero errors before and after.
 
@@
   \item
-    The Adds and Drops are remarkably balanced: across all three modes, they are
-    at most \pct{2} apart from one another.
-    This confirms that changes to the number of errors are balanced.
-    Sessions tend to fix errors highlighted by type analysis
-    even in the stricter analysis modes.
-
-  \item
-    The percentage of Keeps is somewhat high: nearly \pct{10} in \mnocheck{}
-    and \pct{20} in \mnonstrict{} and \mstrict{}.
-    Evidently, the mode plays a role here.
-    Outside of \mnocheck{}, type errors are more likely to persist through edits.
+    The type error increases and decreases are remarkably balanced: across all
+    three modes, they are at most \pct{2} apart.
+    While some of this balance is due to module-switch records
+    (if the two modules have different error counts, there will be a change),
+    most of the data is from keystrokes; thus it appears that sessions tend to
+    fix errors highlighted by type analysis.
+
+  \item
+    \mstrict{} mode has a higher percent of  (\keepsym{}) records.
+    It may be that errors are more likely to persist in this mode.
+    %% TODO check if actually zero
 

### 2023-09-30 programming23/paper.tex
shrank sec 7

--- 2023-09-30 programming23/paper.tex
@@
 This paper presents the first large-scale analysis of developers' interactions with
-an industrial strength typechecker.
-With over 1.5 million telemetry records from over 340,000 sessions
-that took place during a three-month timespan, the data has plenty to say about
-type errors and their overlap with program edits.
-At the same time, the data is limited to avoid leaking personal
-and proprietary information: we know the number and kind of type errors
-that occurred, but nothing about the line(s) of code that the typechecker
-complained about.
-
-On the object level, the analysis offers several lessons for the Luau
+an industrial strength typechecker. It is nevertheless intentionally
+limited by the nature and frequency of sampling. Despite these restrictions,
+the analysis offers several lessons for the Luau
 developers and other maintainers of gradually-typed languages:
@@
     Testing the relationship between two counts can yield insights.
-    The gap between \mstrict{} and \FS{} errors In~\cref{f:tefs-compass}
+    The gap between \mstrict{} and \FS{} errors in~\cref{f:tefs-compass}
     revealed an issue with \mstrict{} checking.
@@
 While one could start modeling with the \mnocheck{} and \mnonstrict{}
-data, but there are too few \mstrict{} records at the moment due to our uniform
+data, there are too few \mstrict{} records at the moment due to our uniform
 sampling rate.

### 2023-09-30 programming23/paper.tex
sec 4 pass done, removed timelines

@@
 \newcommand{\tekey}{\tefsfont{te}}
-\newcommand{\fskey}{\tefsfont{fs}}
+\newcommand{\fskey}{\tefsfont{bg}}
 \newcommand{\QALAN}{\FILL{} Alan:}
@@
 creation features.
+%% \QALAN: can we be more specific? what's an example ... API usage?
 This system randomly determines which sessions should report telemetry, and,
@@
 \label{s:context}
+
 % https://create.roblox.com/docs/scripting/luau
-
-% TODO quick reference table
-% \begin{table}
-%   \caption{Modes and analyses}
-%   \label{t:mode-analyze}
-% 
-%   Mode: nocheck ...
-%   Analysis: nocheck ... forced strict
-% 
-% \end{table}
 
@@
               & \mstrict{}    & 419 & [\pct{39.31}] & 230 & [\pct{21.58}] & 417 & [\pct{39.12}] \\
-    \\
+    \\[-2ex]
     \FS       & \mnocheck{}   & 259865 & [\pct{39.09}] & 175643 & [\pct{26.42}] & 229271 & [\pct{34.49}] \\
@@
   \item
-    \mstrict{} mode has a higher percent of  (\keepsym{}) records.
-    It may be that errors are more likely to persist in this mode.
+    \mstrict{} mode has a higher percent of ``maintain'' (\keepsym{}) records.
+    Errors may be more likely to persist in this mode.
     %% TODO check if actually zero
+
+  \item
+    In \FS{} mode, the percent of ``maintain'' records is high across
+    the board and the percents of increases and decreases are lower.
+    No matter the mode, reducing type errors is not guaranteed to reduce
+    \FS{} errors as well.
+    The next section further compares type and \FS{} errors.
 
@@
     \mnocheck{} & \mnonstrict{} & \mstrict{} \\
-    \includegraphics[width=0.3\columnwidth]{img/compass-nocheck.pdf}
+    \includegraphics[width=0.26\columnwidth]{img/compass-nocheck.pdf}
     &
-    \includegraphics[width=0.3\columnwidth]{img/compass-nonstrict.pdf}
+    \includegraphics[width=0.26\columnwidth]{img/compass-nonstrict.pdf}
     &
-    \includegraphics[width=0.3\columnwidth]{img/compass-strict.pdf}
+    \includegraphics[width=0.26\columnwidth]{img/compass-strict.pdf}
   \end{tabular}
@@
 
-Since Roblox Studio runs each codebase through two passes of type analysis,
-it is interesting to compare their results.
-Of special interest is the \mstrict{} mode case, because the number of
-type errors and \FS{} errors should be nearly the same.
-But, there are two differences between \mstrict{} type checking
-and \FS{} checking:
+\Cref{f:tefs-compass} takes a closer look at type errors (\tekey{}) and \FS{} errors (\fskey{}).
+For every pair of events in each session, we compare the change in
+type errors to the change in \FS{} errors and choose one of five categories.
+At a high level, the question at hand is whether the changes \emph{agree} or \emph{disagree}.
+The categories go one step further and separate zero and non-zero changes:
+
+\noindent{}\begin{tabular}{lll}
+  1. & \code{zero} &  neither \tekey{} nor \fskey{} changed \\
+  2. & \code{te only} & \tekey{} changed but \fskey{} remained the same \\
+  3. & \code{bg only} & \tekey{} remained the same but \fskey{} changed \\
+  4. & \code{dis} & \tekey{} and \fskey{} changed in opposite directions \\
+  5. & \code{agree} & \tekey{} and \fskey{} changed in the same direction
+\end{tabular}
+\smallskip{}
+
+Of particular interest is how \mstrict{} and \FS{} checks compares.
+The two analyses are similar, but \FS{} analysis uses strong checks
+on all dependencies and has looser rules for the data model~(\cref{s:context}):
+whereas \mstrict{} requires casts to confirm the types of data assets,
+\FS{} gives data the dynamic gradual type.
+
+\paragraph{Observations}
+
 \begin{itemize}
   \item
-    Background checks all dependencies in \mstrict{} mode, no matter
-    what mode they declare.
-    So, the number of \FS{} errors may be higher.
-  \item
-    Background checking has looser rules for the \emph{data model};
-    that is, for the static assets that a program manipulates.
-    So, the number of \FS{} errors may be lower.
-    %% initial DM, all properties go into a table type; DM -> folder -> model -> part
-    %% get dot-driven autocomplete
+    All three plots have a nonzero percentage of \code{te only} items.
+    In \mstrict{} mode, the percentage jumps up to nearly \pct{25}.
+    Since \FS{} analysis is more lenient only for data model changes, these must
+    be edits that affect the model in a way that raises a type error.
+    The high percentage in \mstrict{} may be a source of frustration.
+    %% Naturally, \mstrict{} mode has the largest number of such edits because
+    %% its type checker is the most careful.
+
+  \item
+    In the first plot, for \mnocheck{},
+    half the records are \code{zero}, nearly half are \code{bg only}, and only a few
+    land in the other categories.
+    While the number of zeros is a bit high, the high \code{bg only} and low
+    \code{agree} counts make sense because \mnocheck{} reports few errors.
+
+  \item
+    In \mnonstrict{}, the number of agreements jumps up to nearly \pct{20}.
+    This is reassuring because type and \FS{} analysis should agree to some extent,
+    but the level of agreement is still rather low.
+
 \end{itemize}
-Our data contains a surprising number of records where \FS{}
-reports fewer errors.
-Data model edits are apparently a regular occurrence in Luau programs.
-
-\Cref{f:tefs-compass} compares changes in type errors (\tekey{}) to changes in
-\FS{} errors (\fskey{}) for the three analysis modes.
-To a first approximation, the changes either agree with one another
-or disagree.
-For example, if type errors increase but \FS{} errors decrease,
-we have a disagreement.
-The plots futher categorize agreements and disagreements into
-the following five cases:
-\begin{enumerate}
-  \item
-    \emph{zero}: neither \tekey{} nor \fskey{} changed
-  \item
-    \emph{te only}: \tekey{} changed but \fskey{} remained the same
-  \item 
-    \emph{fs only}: \tekey{} remained the same but \fskey{} changed
-  \item
-    \emph{dis}: \tekey{} and \fskey{} changed in opposite directions
-  \item 
-    \emph{agree}: \tekey{} and \fskey{} changed in the same direction
-\end{enumerate}
-Both \emph{zero} and \emph{agree} are agreements.
-The others are disagreements.
-The plots show what percent of all telemetry records fell into these five categories.
-
-\paragraph{Observations}
-
-\begin{itemize}
-  \item
-    All three plots have a nonzero percentage of \emph{te only} items.
-    In \mstrict{} mode, the percentage jumps up to nearly \pct{25}.
-    Since \fskey{} is more lenient only for data model changes, these must
-    be edits that affect the model in a way that raises a type error.
-    Naturally, \mstrict{} mode has the largest number of such edits because
-    its type checker is the most careful.
-
-  \item
-    In the first plot, for \mnocheck{},
-    half the records are \emph{zero\/}s, nearly half are \emph{fs only}, and only a few
-    land in the other categories.
-    While the number of zeros is a bit high,
-    the \emph{fs. only} spike makes sense because type checking is supposed to complain
-    sparingly in \mnocheck{} mode.
-
-  \item
-    In the \mnonstrict{} plot, the number of agreements jumps up.
-    This is reassuring.
-    \FILL{}.
-
-\end{itemize}
-
-Users of \mstrict{} mode may find it frustrating that the type checker complains
-about data model edits.
-Though we cannot see the exact reason for the complaints, the checker errs on
-the conservative side.
-Thus we expect the number of useful reports is low.
-This may be a barrier to adaption.
-
-\subsection{Session Timelines}
-
-\begin{figure}[t]\centering
-  %% TODO any way to cluster timelines to avoid the "creepiness" of spying on one user?
-%  \mnonstrict{}\\
-%  \includegraphics[width=\columnwidth]{img/timeline-nonstrict.pdf}
-
-  \medskip{}
-  \mstrict{}\\
-  \includegraphics[width=\columnwidth]{img/timeline-strict.pdf}
-
-  \caption{Two example sessions.}
-  \label{f:indy-session}
-\end{figure}
-
-Another insightful visualization that the telemetry supports is plots of individual sessions
-over time.
-\Cref{f:indy-session} shows the type errors, \FS{} errors, and module switches
-in two \mstrict{} mode sessions.
-The x-axis shows time in seconds from the start of the session.
-The y-axis counts errors.
-Type errors appear as orange plus marks.
-Background errors appear as green x marks.
-The vertical lines show module switches.
-
-In the first plot, the error counts are low until about the 3,000s mark,
-at which point an edit (not a module switch) leads to 15 errors.
-These 15 errors disappear two records later.
-Then, after two module switches and some edits, the number of errors spikes again.
-In the second plot, the number of errors spikes in the beginning due to edits,
-then remains low for the next few thousand seconds.
-
-A general observation is that the number of type errors is always less than the number of \FS{}
-errors, though the numbers are close.
-This should happen in \mstrict{} mode.
 
@@
     Timestamps give useful, low-level insights about the frequency of keystrokes,
-    the length of sessions, and actions over time~(\cref{f:indy-session}).
+    the length of sessions, and actions over time.
     The event ordering that timestamps provide was invaluable; e.g., for learning

### 2023-09-30 programming23/paper.tex
resize figures

@@
 \begin{figure}[t]\centering
-  \includegraphics[width=.45\textwidth]{img/roblox-studio.png}
-  \includegraphics[width=.45\textwidth]{img/roblox-studio-ide.png}
+  \includegraphics[width=.35\textwidth]{img/roblox-studio.png}
+  \hspace{1cm}
+  \includegraphics[width=.35\textwidth]{img/roblox-studio-ide.png}
   \caption{{Roblox Studio 3D creation} tools (left) and IDE (right).}
@@
   %% NOTE max time = 1 million seconds = 15 days ...  +100 are over 2 days
-  \begin{tabular}{lrrrrl} \\
+  \begin{tabular}{lrrrrl}
     ~               & Mean & Stddev & Median & \ptile{99} & Distribution \\
@@
   \label{t:count-analysis-errors}
-  \begin{tabular}[t]{ll} \\
+  \begin{tabular}[t]{ll}
     \begin{tabular}[t]{r@{~~}r@{~}r}
@@
 \begin{figure}[t]\centering
-  \begin{subfigure}[t]{\columnwidth}
-    \begin{tabular}[t]{ll}
-      %% FILL bar chart? rectangle?
+  \begin{subfigure}[b]{0.48\columnwidth}
       \begin{tabular}[t]{r@{~~}l@{~}r}
@@
       \end{tabular}
-      \begin{tabular}[t]{r@{~~}l@{~~}r}
-      \end{tabular}
-    \end{tabular}
-    \caption{Analysis modes across the 1,504,736 records.}
+    \caption{Analysis modes in 1,504,736 records.}
     \label{f:total-records}
   \end{subfigure}
-
-  \begin{subfigure}[t]{\columnwidth}
-    \begin{tabular}[t]{ll} \\
+  \begin{subfigure}[b]{0.48\columnwidth}
       \begin{tabular}[t]{r@{~~}r@{~}r}
@@
       \end{tabular}
-      \begin{tabular}[t]{l@{~~}ll}
-        %% TODO numbers may be off by +- 100
-        %%    & 341 & contain a mode upgrade \\
-        %%    & 320 & contain a mode downgrade \\
-        %%    & 512 & contain modules with different modes
-        %%  they don't match out/error-density*hasdown.rkt
-        %%  ??? counting downs instead of sessions with downs ???
-        %% TODO fix throughout paper!
-        \zerowidth{Among the mixed-mode sessions:} \\
-        & 176 & contain a mode upgrade \\
-        & 233 & contain a mode downgrade \\
-        & 512 & contain modules with different modes
-      \end{tabular}
-    \end{tabular}
-    \caption{Analysis mode(s) for the 347,598 sessions.}
+    \caption{Analysis mode(s) in 347,598 sessions.}
     \label{f:total-sessions}
   \end{subfigure}
+
+  {Among the mixed-mode sessions:
+        176 contain a mode upgrade,
+        233 contain a mode downgrade, and
+        512 contain modules with different modes.}
 
@@
     Type errors & Background errors \\
-    \labelbars{\includegraphics[valign=M]{img/error-by-mode-te.pdf}}
-    & \labelbars{\includegraphics[valign=M]{img/error-by-mode-fs.pdf}}
+    \labelbars{\includegraphics[width=0.20\columnwidth,valign=M]{img/error-by-mode-te.pdf}}
+    & \labelbars{\includegraphics[width=0.20\columnwidth,valign=M]{img/error-by-mode-fs.pdf}}
   \end{tabular}
@@
   \includegraphics[width=\columnwidth]{img/error-count-nocheck-row--te-density-diff.pdf}
-  \smallskip
 
@@
   \includegraphics[width=\columnwidth]{img/error-count-nonstrict-row--te-density-diff.pdf}
-  \smallskip
 
@@
     \mnocheck{} & \mnonstrict{} & \mstrict{} \\
-    \includegraphics[width=0.26\columnwidth]{img/compass-nocheck.pdf}
+    \includegraphics[width=0.24\columnwidth]{img/compass-nocheck.pdf}
     &
-    \includegraphics[width=0.26\columnwidth]{img/compass-nonstrict.pdf}
+    \includegraphics[width=0.24\columnwidth]{img/compass-nonstrict.pdf}
     &
-    \includegraphics[width=0.26\columnwidth]{img/compass-strict.pdf}
+    \includegraphics[width=0.24\columnwidth]{img/compass-strict.pdf}
   \end{tabular}

### 2023-09-30 programming23/paper.tex
sec 4

@@
 % \usepackage[sort]{natbib} or otherwise manually fix cites to be in ascending order
+% redo fig 5 to have better resolution
 
@@
 error from edits that remove one error while introducing another.
-We acknowledge this and other threats in~\cref{s:threats}.
+We acknowledge this and other threats in~\cref{s:threats}.\todo{Do we?
+Double check.}
 The main advantages of this telemetry are the small size of each record
@@
 
-Roblox Studio collected type error telemetry between February
+We collected type error telemetry in Roblox Studio between February
 and April 2023.
@@
 In total, we collected over 1.5 million telemetry records.
-Roughly two thirds of records are due to keystrokes.
+Roughly two thirds are due to keystrokes.
 
@@
 Each vertical bar counts the number of records generated per hour according to
-a client-side timestamp.
-%% \todo{Do we need all this detail?}
-%% the vertical bars suggest we have lots of day-job developers, will point this out below
-The bars are labeled with a California time zone because that is the locale of
-the Roblox servers, which parsed the timestamps.
-Most buckets contain roughly one thousand records, but the number
-varies between a few dozen at minimum at over three thousand.
+a client-side timestamp.\todo{Not just per day?!?}
+The bars are labeled with a California time zone.
+Buckets\todo{What is a BUCKET as opposed to a BAR?}
+contain a few dozen to over 3,000 records, typically about 1,000.
 %% \FILL{} investigate one tower, what's happening?? all one session?
-There is a sharp drop in early April when the experiment was finished
-and the telemetry probe was removed, though there is a long tail as
-developers updated {Roblox Studio}.
-We do not know which timezone the records originated in, but since the counts
-tend to peak midday California time it seems that the majority
-of Roblox developers are following a Western Hemisphere schedule.
-The regular pattern of the peaks furthermore suggests that many creators
-follow a regular schedule; this may be due to a large community of
+Telemetry ended in early April; the long tail is due to developers
+updating Roblox Studio later.
+% BEN: I see no basis for this; nor is this at all interesting.
+% We do not know which timezone the records originated in, but since the counts
+% tend to peak midday California time it seems that the majority
+% of Roblox developers are following a Western Hemisphere schedule.
+The regular pattern of the peaks suggests that many creators
+follow a regular schedule; these may be
 professional developers.
-The counts for weekends (shaded regions) are often the tallest,
-as Roblox has a significant school-aged creator community.
+The tall weekend (shaded) regions could be because
+Roblox has a significant school-aged creator community.
 
@@
     (A human would have to switch modules twice or enter several keystrokes
-     within one millisecond to generate two such records legitimately.)
+     within 1ms to generate two such records legitimately.)
 
@@
 We discuss these in turn.
-Codebase size and session length have long-tailed distributions
-with many small items and few very large ones.
-%% TODO check %% follow power-law distributions~\cite{csn-siamr-2009} 
-%% https://aaronclauset.github.io/powerlaws/
-%% bin version, is this relevant? sounds like no from abtract https://arxiv.org/pdf/1208.3524.pdf
-The number of type analysis errors is somewhat low~(explained in the next
-section), but sufficient to proceed with futher analysis.
+% Commenting these out because that's literally the rest of the
+% subsection. Say anything you want from here over there!
+% Codebase size and session length have long-tailed
+% distributions\todo{DID YOU DO A STATISTICAL TEST.}
+% with many small items and few very large ones.
+% %% TODO check %% follow power-law distributions~\cite{csn-siamr-2009} 
+% %% https://aaronclauset.github.io/powerlaws/
+% %% bin version, is this relevant? sounds like no from abtract https://arxiv.org/pdf/1208.3524.pdf
+% The number of type analysis errors is somewhat low~(explained in the next
+% section), but sufficient to proceed with futher analysis.
 
@@
 \Cref{t:codebase-size} summarizes the size of projects in the dataset.
-The middle four columns report the mean, standard deviation, median,
+The four numeric columns report the mean, standard deviation, median,
 and \percentile{99} for the number of lines
@@
 
-\Cref{t:session-size} outlines the size of sessions using two metrics:
+\Cref{t:session-size} similarly outlines the size of sessions using two metrics:
 real time in seconds between the first and last record, and the total number of
 records.
-The middle columns report the mean, standard deviation,
-median, and \percentile{99}.
-Similar to the previous table~(\cref{t:codebase-size}), the distributions focus
+The plots focus
 on the lower-left fragment of the data, which again has a very long tail.
@@
 First, it suggests that errors appear locally, as the result of edits to nearby
-code, rather than as the result of edits to far-away code that lives in another
+code, rather than as the result of edits to code that lives in another
 module.
@@
     creators might choose to edit a non-highlighted location to fix an error.
-    Our data cannot narrow down the answer.
+    While non-privacy-protecting analysis have answered such
+    questions~\cite{mfk-sigcse-2011}, our data cannot narrow down the answer.
 
@@
 records have more \FS{} errors than the other modes.
-Though \mnocheck{} does have the most \FS{} errors, its proportion matches
+Though \mnocheck{} does have the most \FS{} errors, its \emph{proportion} matches
 the proportion of \mnocheck{} records in the
@@
 errors.
-Furthermore, type analysis does not seem to reduce \FS{} errors.
+Furthermore, type analysis does not seem to reduce \FS{}
+errors.\todo{Isn't this a bit surprising?}
 
@@
 Indeed, most of the fields in a telemetry record are dedicated to this
-topic~(\cref{f:tdata}), and it helps to answer RQ2 from the introduction:
-\begin{description}
-  \item[RQ2.]
-    For modules that use type analysis:
-    which errors arise,
-    how do sessions respond,
-    and which errors tend to persist despite subsequent edits?
-\end{description}
+topic~(\cref{f:tdata}), and are directed to RQ2 on type errors
+(\cref{s:introduction}).
 
@@
 
-All three counts are imprecise because our telemetry system cannot distinguish
-particular type errors.
-If a creator removes one \code{NotATable} error but then
-immediately introduces another on a different line,
-then the centor (\keepsym{}) column gets incremented
-instead of the increase (\addsym{}) and decrease (\dropsym{}) columns.
+% BEN: double-check, I think this says all you were trying to say.
+% "typically" because we COULD have gotten lucky.
+The counts are imprecise because of telemetry granularity.
+If a creator removes one error but then
+immediately introduces another on a different line, this typically manifests as
+a \keepsym{} rather than one \addsym{} and one \dropsym{}.
 
@@
   \item
-    \code{SyntaxError} and the related errors \code{UnknownSymbol} and
-    \code{UnknownProperty} are the most popular.
+    \code{SyntaxError} and the related \code{UnknownSymbol} and
+    \code{UnknownProperty} are most popular.
     If the edit range errors are indicative of type errors at large,
@@
 
-    \subitem The random-keystroke strategy of our telemetry may be inflating
-    the number of typos because it has no way to avoid reporting edits that
-    happen in the middle of things.
+    Our telemetry strategy likely inflates
+    the number of typos because it cannot determine when a creator is
+    mid-edit.
     For example, a creator who writes a method call
-    \code{bucket.countFish()} letter-by-letter will see errors
-    including \code{UnknownSymbol(buck)} and \code{UnknownProperty(bucket.c)}.
-    Telemetry may pick these up;~\cref{s:conclusion} suggests ways
-    to avoid doing so in the future.
+    \code{bucket.countFish()} letter-by-letter will generate an
+    \code{UnknownSymbol} error for \code{buck} and
+    \code{UnknownProperty} for \code{bucket.c}, which may end up sampled.
+    \Cref{s:conclusion} suggests ways
+    to avoid this in the future.
 
@@
     A \code{CountMismatch} sends the wrong number of arguments to a function.
-    It may simply be a common error, but it may also be due to our random-keystroke
-    telemetry.
+    It may simply be a common error, but it may also be a
+    telemetry artifact.
     The high incidence of \code{TypeMismatch} and \code{GenericError} calls for
@@
     Ideally, the highest number in each column group should
-    be the number of decreases (\dropsym{}).
-    A decrease should happen when the creator sees an error,
-    edits around the highlighted code, and fixes it.
+    be decreases (\dropsym{}).
+    A decrease should happen when the creator sees an error and
+    fixes it.
     However, decreases are highest in only 18 of the 39 nonzero groups.
@@
 \Cref{t:type-error-count} reports the percent of type errors that overlap with
-the edit range.
+the edit range.\todo{I DO NOT UNDERSTAND ANYTHING HERE.}
 The purpose of this table is to discover all specific errors in the dataset,
@@
     Syntax errors, in particular \code{UnknownSymbol}, are much more
-    common in \mnonstrict{} than \mstrict{}.
+    common in \mnonstrict{} than \mstrict{}.\todo{This is NOT a
+      ``syntax'' error? It's a typo!}
     This may be because \mnonstrict{} reports fewer errors in general.
@@
 records in three sessions.
-It is entirely possible that they all came from one session that
-deliberately pushed the limits of type analysis.
+Note that all three sessions \emph{could} have been from the same codebase.
+% BEN: I see no reason for this speculation. The only important thing
+% is the fact that this could have been all the same codebase.
+%It is entirely possible that they all came from one session that
+%deliberately pushed the limits of type analysis.
 %% Whatever the reason, the overall number is low enough to suggest that the
@@
     (if the two modules have different error counts, there will be a change),
-    most of the data is from keystrokes; thus it appears that sessions tend to
+    most of the data is from keystrokes; thus, it appears that sessions tend to
     fix errors highlighted by type analysis.
@@
 \Cref{f:tefs-compass} takes a closer look at type errors (\tekey{}) and \FS{} errors (\fskey{}).
+\todo{I'm kinda' baffled.}
 For every pair of events in each session, we compare the change in
@@
 
+\newpage
+
 \bibliography{bib}

### 2023-09-30 programming23/paper.tex
merge 4.9 with 4.8

@@
 \subsection{Do Edits Tend to Add Type Errors?}
+\label{s:tefs-compass}
 
@@
     In \FS{} mode, the percent of ``maintain'' records is high across
-    the board and the percents of increases and decreases are lower.
+    the board and the increases and decreases are lower.
     No matter the mode, reducing type errors is not guaranteed to reduce
     \FS{} errors as well.
-    The next section further compares type and \FS{} errors.
+    This is surprising for \mstrict{} mode, and likely relates to
+    the data model; see below for details.
 
@@
 
-\subsection{Two Activities: Asset Creation and Scripting}
-\label{s:strict-vs-forcedstrict}
-
-\begin{figure}[t]\centering
-  %% TODO
-  %% - thin bars
-  %% - colors for good , bad (disagree) , neutral
-  \begin{tabular}{lll}
-    \mnocheck{} & \mnonstrict{} & \mstrict{} \\
-    \includegraphics[width=0.24\columnwidth]{img/compass-nocheck.pdf}
-    &
-    \includegraphics[width=0.24\columnwidth]{img/compass-nonstrict.pdf}
-    &
-    \includegraphics[width=0.24\columnwidth]{img/compass-strict.pdf}
-  \end{tabular}
-  \caption{Comparing the change in type errors to the change in \FS{} errors.}
-  \label{f:tefs-compass}
-\end{figure}
-
-\Cref{f:tefs-compass} takes a closer look at type errors (\tekey{}) and \FS{} errors (\fskey{}).
-\todo{I'm kinda' baffled.}
-For every pair of events in each session, we compare the change in
-type errors to the change in \FS{} errors and choose one of five categories.
-At a high level, the question at hand is whether the changes \emph{agree} or \emph{disagree}.
-The categories go one step further and separate zero and non-zero changes:
-
-\noindent{}\begin{tabular}{lll}
-  1. & \code{zero} &  neither \tekey{} nor \fskey{} changed \\
-  2. & \code{te only} & \tekey{} changed but \fskey{} remained the same \\
-  3. & \code{bg only} & \tekey{} remained the same but \fskey{} changed \\
-  4. & \code{dis} & \tekey{} and \fskey{} changed in opposite directions \\
-  5. & \code{agree} & \tekey{} and \fskey{} changed in the same direction
-\end{tabular}
-\smallskip{}
-
-Of particular interest is how \mstrict{} and \FS{} checks compares.
-The two analyses are similar, but \FS{} analysis uses strong checks
-on all dependencies and has looser rules for the data model~(\cref{s:context}):
-whereas \mstrict{} requires casts to confirm the types of data assets,
-\FS{} gives data the dynamic gradual type.
-
-\paragraph{Observations}
-
+\paragraph{Data Model Types: Strict vs. Background Analysis}
+
+Based on the description of \FS{} analysis from~\cref{s:context},
+it should raise \emph{at least as many errors} as \mstrict{} mode
+because it uses \mstrict{} checks for all dependencies.
+For example, if a \mstrict{} module imports from a \mnocheck{} one,
+then \mstrict{} analysis does not analyze the import but
+\FS{} analysis does.
+
+In \cref{f:error-changes}, however, nearly \pct{3} of \mstrict{}-mode records
+increase the number of \mstrict{} errors without increasing \FS{} errors.
+Further inspection reveals that the situation is a bit worse, as
+some records increase \mstrict{} errors without changing the \FS{} errors:
 \begin{itemize}
   \item
-    All three plots have a nonzero percentage of \code{te only} items.
-    In \mstrict{} mode, the percentage jumps up to nearly \pct{25}.
-    Since \FS{} analysis is more lenient only for data model changes, these must
-    be edits that affect the model in a way that raises a type error.
-    The high percentage in \mstrict{} may be a source of frustration.
-    %% Naturally, \mstrict{} mode has the largest number of such edits because
-    %% its type checker is the most careful.
-
-  \item
-    In the first plot, for \mnocheck{},
-    half the records are \code{zero}, nearly half are \code{bg only}, and only a few
-    land in the other categories.
-    While the number of zeros is a bit high, the high \code{bg only} and low
-    \code{agree} counts make sense because \mnocheck{} reports few errors.
-
-  \item
-    In \mnonstrict{}, the number of agreements jumps up to nearly \pct{20}.
-    This is reassuring because type and \FS{} analysis should agree to some extent,
-    but the level of agreement is still rather low.
-
+    \pct{22} of \mstrict{} records increase (\addsym{}) the number of type errors
+    but maintain (\keepsym{}) the number of \FS{} errors.
 \end{itemize}
+The reason for the discrepancy is that \mstrict{} mode and \FS{} analysis
+assign different types to the data model: \mstrict{} assigns a top type
+and requires downcasts at use sites, whereas \FS{} analysis uses the
+dynamic gradual type.
+Dealing with these data model errors may be a source of frustration
+for creators using \mstrict{} mode.
 
@@
     Until then, \mstrict{} mode clearly needs to relax its data model
-    analysis because it raises even more errors than \FS{} checking~(\cref{f:tefs-compass}).
+    analysis because it raises even more errors than \FS{} checking~(\cref{s:tefs-compass}).
 \end{itemize}
@@
     Testing the relationship between two counts can yield insights.
-    The gap between \mstrict{} and \FS{} errors in~\cref{f:tefs-compass}
+    The gap between \mstrict{} and \FS{} errors in~\cref{s:tefs-compass}
     revealed an issue with \mstrict{} checking.

### 2023-09-30 programming23/paper.tex
cleanup

@@
 some records increase \mstrict{} errors without changing the \FS{} errors:
-\begin{itemize}
-  \item
     \pct{22} of \mstrict{} records increase (\addsym{}) the number of type errors
     but maintain (\keepsym{}) the number of \FS{} errors.
-\end{itemize}
+
 The reason for the discrepancy is that \mstrict{} mode and \FS{} analysis

### 2023-09-30 programming23/paper.tex
checkpoint, todo cleanup

@@
 error from edits that remove one error while introducing another.
-We acknowledge this and other threats in~\cref{s:threats}.\todo{Do we?
-Double check.}
+We acknowledge this and other threats in~\cref{s:threats}.
 The main advantages of this telemetry are the small size of each record
@@
 \Cref{f:records-per-hour} provides a time-ordered distribution of the data.
-Each vertical bar counts the number of records generated per hour according to
-a client-side timestamp.\todo{Not just per day?!?}
-The bars are labeled with a California time zone.
-Buckets\todo{What is a BUCKET as opposed to a BAR?}
-contain a few dozen to over 3,000 records, typically about 1,000.
+The x-axis is labeled with a California time zone; the thin vertical lines
+mark the start of a new day.
+Each of the blue bars in the plot bar is a bucket that counts the number of
+records generated per hour according to a client-side timestamp.
+Buckets contain a few dozen to over 3,000 records, typically about 1,000.
 %% \FILL{} investigate one tower, what's happening?? all one session?
@@
 % BEN: I see no basis for this; nor is this at all interesting.
+%% ben: basis = it's where the peaks are
 % We do not know which timezone the records originated in, but since the counts
@@
 Furthermore, type analysis does not seem to reduce \FS{}
-errors.\todo{Isn't this a bit surprising?}
+errors; this is surprising and~\cref{s:tefs-compass} has more to say about it.
 
@@
 % "typically" because we COULD have gotten lucky.
+% ben: I'd written "immediatly introduces" to rule out getting lucky. Deleted "immediately" and went with your version.
 The counts are imprecise because of telemetry granularity.
 If a creator removes one error but then
-immediately introduces another on a different line, this typically manifests as
+introduces another on a different line, this typically manifests as
 a \keepsym{} rather than one \addsym{} and one \dropsym{}.
@@
 % This is the [DIFF] code from Marceau SIGCSE 2011.
+Edits that fix one error but introduce another are also indistinguishable
+from edits that maintain the number of errors.
 

### 2023-09-30 programming23/paper.tex
fix undef ref, clear 4.6 todos

@@
 
-\Cref{t:type-error-count} reports the percent of type errors that overlap with
-the edit range.\todo{I DO NOT UNDERSTAND ANYTHING HERE.}
-The purpose of this table is to discover all specific errors in the dataset,
-thus it includes both keystroke-based and module-switch-based records---even
-though edit ranges are not well-defined for the latter.
-
-There are two subtables for \mnonstrict{} and \mstrict{} mode.
-The table does not report \mnocheck{} because all its type errors are
-by definition \code{SyntaxError}s.
+\Cref{t:type-error-count} sorts type errors by their popularity
+in \mnonstrict{} and \mstrict{} mode.
+The purpose of this table is to discover and rank all errors that
+appear in the dataset, thus it is based on both keystroke and
+module-switch telemetry records even though edit ranges are not well-defined
+for the latter.
+The table does not include \mnocheck{} because all its type errors are
+\code{SyntaxError}s by definition.
 
@@
   \item
-    Syntax errors, in particular \code{UnknownSymbol}, are much more
-    common in \mnonstrict{} than \mstrict{}.\todo{This is NOT a
-      ``syntax'' error? It's a typo!}
-    This may be because \mnonstrict{} reports fewer errors in general.
+    Syntax errors, in particular \code{UnknownSymbol} (for an unbound variable),
+    are much more common in \mnonstrict{} than \mstrict{}.
+    This may be because \mnonstrict{} reports fewer errors
+    to dilute the proportion of unknown symbols.
 
@@
 \label{s:tefs-compass}
+\label{s:strict-vs-forcedstrict}
 

### 2023-09-30 programming23/paper.tex
sec 5 todo cleanup

@@
     & & \zerowidth{\rbox{\addsym{}}} & & \zerowidth{\ybox{\keepsym{}}} & & \zerowidth{\gbox{\dropsym{}}} \\\midrule
-    type      & \mnocheck{}   & 20085 & [\pct{49.83}] & 906 & [\pct{2.25}] & 19317 & [\pct{47.92}] \\
-    error     & \mnonstrict{} & 13030 & [\pct{40.22}] & 5513 & [\pct{17.02}] & 13852 & [\pct{42.76}] \\
+    type error  & \mnocheck{}   & 20085 & [\pct{49.83}] & 906 & [\pct{2.25}] & 19317 & [\pct{47.92}] \\
+    (in module) & \mnonstrict{} & 13030 & [\pct{40.22}] & 5513 & [\pct{17.02}] & 13852 & [\pct{42.76}] \\
                 & \mstrict{}    & 419 & [\pct{39.31}] & 230 & [\pct{21.58}] & 417 & [\pct{39.12}] \\
@@
     error       & \mnonstrict{} & 33988 & [\pct{39.74}] & 20738 & [\pct{24.25}] & 30798 & [\pct{36.01}] \\
-              & \mstrict{}    & 488 & [\pct{36.53}] & 344 & [\pct{25.75}] & 504 & [\pct{37.72}] \\
+    (in module) & \mstrict{}    & 488 & [\pct{36.53}] & 344 & [\pct{25.75}] & 504 & [\pct{37.72}] \\
   \end{tabular}
@@
 
-Luau clearly\todo{How can we say?} has a powerful typechecker that can efficiently
-analyze thousands of lines of code.
+Luau has an efficient type checker that can
+analyze thousands of lines of code~(\cref{t:codebase-size}).
 But, few sessions use type analysis directly and the data presents little
@@
 creators to fix.
-It would be better for the analysis to ignore these errors.
-Indeed, our comparison between \FS{} errors in \mnocheck{} vs.
-\mstrict{} mode~(\cref{f:error-changes}) might tell a very different story
-without all the noise that syntax errors introduce.\todo{Why didn't
-  YOU just filter them, then?!?}
+Our comparison between \FS{} errors in \mnocheck{} vs.
+\mstrict{} mode~(\cref{f:error-changes}) might tell a different story
+without the noise that syntax errors introduce, but we cannot easily
+filter them because we have only the total number (not the kind)
+for errors outside the edit range.
 
@@
 
-Lastly, we have two remarks about external validity.
-First, our information on type errors is limited to error names.
-If another language has an error that is roughly similar to
-the\todo{Huh? It's similar yet different?}
-\code{CannotCallNonFunction} error but with slightly different
-semantics, then the relative frequency of that error could be different
-than what we observed.
-%% ben: not sure what to say ... may remove the error names comment
-Second, we know nothing about the creators in our study except that they
+Lastly, regarding external validity,
+we know nothing about the creators in our study except that they
 were selected at random from a large and diverse group.

### 2023-09-30 programming23/paper.tex
macros for ID (\uid) and IDE (\IDE)

@@
 % TODO BEFORE SUBMISSION
-% be consistent between "IDs" and "ids" [SK prefers "IDs" to avoid it being read as a word]
-% even more so between "ID" and "id" [NB: "id" is a word in English]
 % \usepackage[sort]{natbib} or otherwise manually fix cites to be in ascending order
@@
 % - uncomment acks
+% - update all refs to artifacts to no longer be in future tense
 
@@
 
-\title{Type Error Telemetry at Scale}
+\title{Privacy-Respecting Type Error Telemetry at Scale}
 % lean, lite, modest, non-intrusive, private, ...
@@
 
+\newcommand{\uid}{\textsc{id}}
+\newcommand{\IDE}{\textsc{ide}}
 \newcommand{\code}[1]{\texttt{#1}}
@@
   %% addressed by others (if at all)?
-  Since Luau is currently being used by many creators, we want to collect their feedback
+  Since Luau is currently being used by many creators, we want to collect data
   to improve the language and, in particular, the type system.
-  The standard way to collect feedback at a large scale from creators working on
-  real projects is to deploy client-side telemetry; however, we cannot
+  The standard way to collect data is to deploy client-side telemetry; however, we cannot
   scrape personal data or proprietary information, which means we cannot
@@
   %% What was done that unveiled new knowledge?
-  We designed and implemented a pseudonymized, randomized telemetry system for Luau.
+  We designed and implemented a pseudonymized, randomly-sampling telemetry system for Luau.
   Telemetry records include a timestamp, a session id, a reason for sending,
@@
 types for all Luau programs on the fly, as creators modify the code.
-These types are used primarily in IDE tooling such as autocomplete and
+These types are used primarily in \IDE{} tooling such as autocomplete and
 in API documentation~\cite{luau-autocomplete}, and indeed, creators
@@
 quality can rely on \mstrict{} mode.
-(No matter the mode, a {\FS{}} type analysis still drives the IDE tools.)
+(No matter the mode, a {\FS{}} type analysis still drives the \IDE{} tools.)
 Luau's types aim to support untyped designs, in the spirit of migratory
@@
 In comparison to prior work~(\cref{s:related}), which
-with few exceptions~(e.g.,\cite{hlzbr-ecoop-2021}) is small in scale
+with few exceptions is small in scale
 or collects source code, we performed a large-scale study using \panon{}
@@
     Design of a low-overhead telemetry system
-    that reports on type errors without revealing source code.
+    that reports on type errors without revealing potentially private information.
 
@@
 combines {3D creation} tools and an Integrated
-Developer Environment (IDE), as seen in \cref{fig:roblox-studio}.
-The IDE includes an optional \emph{Script Analysis} widget that
+Developer Environment (\IDE{}), as seen in \cref{fig:roblox-studio}.
+The \IDE{} includes an optional \emph{Script Analysis} widget that
 reports syntax errors, type errors, and problems identified by
@@
 
-\paragraph{Type Analysis Modes, \FS{} Analysis}
+\paragraph{Type Analysis Modes, Background Analysis}
 
@@
 
-In addition to the main type analysis, Roblox Studio takes a second pass over
-every codebase with a \emph{\FS{}} analysis to infer autocomplete suggestions
-and drive other IDE tools.
-To infer precise types,
-the \FS{} analysis uses rigorous checks very similar to those of
-\mstrict{} mode (but not identical: see \cref{s:strict-vs-forcedstrict})
-no matter what mode the creator declared for the script.
-(Internally to Roblox Studio, this analysis is called \emph{forced strict}
-because it ignores the creator's choice and applies strong checks.)
-When it fails to infer a type, \FS{} analysis defaults to the unknown gradual
-type~\cite{st-sfp-2006}.
-Creators cannot see the errors reported by \FS{} analysis.
+\begin{figure}[t]\centering
+  \includegraphics[width=.35\textwidth]{img/roblox-studio.png}
+  \hspace{1cm}
+  \includegraphics[width=.35\textwidth]{img/roblox-studio-ide.png}
+  \caption{{Roblox Studio 3D creation} tools (left) and \IDE{} (right).}
+  \label{fig:roblox-studio}
+\end{figure}
 
@@
 
-\begin{figure}[t]\centering
-  \includegraphics[width=.35\textwidth]{img/roblox-studio.png}
-  \hspace{1cm}
-  \includegraphics[width=.35\textwidth]{img/roblox-studio-ide.png}
-  \caption{{Roblox Studio 3D creation} tools (left) and IDE (right).}
-  \label{fig:roblox-studio}
-\end{figure}
+In addition to the main type analysis, Roblox Studio takes a second pass over
+every codebase with a \emph{\FS{}} analysis to infer autocomplete suggestions
+and drive other \IDE{} tools.
+To infer precise types,
+the \FS{} analysis uses rigorous checks very similar to those of
+\mstrict{} mode (but not identical: see \cref{s:strict-vs-forcedstrict})
+no matter what mode the creator declared for the script.
+(Internally to Roblox Studio, this analysis is called \emph{forced strict}
+because it ignores the creator's choice and applies strong checks.)
+When it fails to infer a type, \FS{} analysis defaults to the unknown gradual
+type~\cite{st-sfp-2006}.
+Creators cannot see the errors reported by \FS{} analysis.
 
@@
 
-Scripts come in two flavors: \emph{module scripts} and \emph{non-module
-scripts}.
-(Though, throughout this paper we use the word ``module''
-to refer to arbitrary scripts.)
-Module scripts provide reusable libraries, which may be
+Luau modules provide reusable libraries, which may be
 {imported} by other scripts.
-Module scripts therefore form a graph in general,
+Imports therefore form a graph in general,
 though it is an error in \mstrict{} mode to create an import cycle
@@
 Roblox Studio has a framework for reporting telemetry.
-Various features in the IDE use telemetry to measure their effectiveness.
+Various features in the \IDE{} use telemetry to measure their effectiveness.
 The Luau open source toolchain does \emph{not} report telemetry,
@@
 
-We added a telemetry subsystem to Roblox Studio in order to study type errors
+We added a subsystem to Roblox Studio's telemetry in order to study type errors
 in Luau code.
@@
 Telemetry in Roblox Studio faces three major constraints:
-\begin{itemize}
+\begin{enumerate}
 
@@
 
-\end{itemize}
+\end{enumerate}
 Cox~\cite{transparent-telemetry} discusses in detail the privacy implications
@@
 
-In addition, due to the architecture of Roblox Studio
+In addition, due to the architecture of Roblox Studio,
 there is some information that is not available to the
@@
   \item
-    GUI state, in particular, whether the Script Analysis widget is visible.
+    GUI state: in particular, whether the Script Analysis widget is visible.
 \end{itemize}
@@
   \begin{tabular}{|l|l|l|l|l|l|l|}\hline
-    ID & Time & Mode & Reason & Size Info & Overall Counts & Edit Range Counts
+    \uid{} & Time & Mode & Reason & Size Info & Overall Counts & Edit Range Counts
     \\\toprule
@@
 The type error telemetry for a Roblox Studio session is gathered as a series of
-records, all with the same \panon{} session identifier.
+records, all with the same per-session \panon{} session identifier.
 This allows us to correlate telemetry across a single
@@
   \item
-    ID: a \panon{} identifier for the current session.
+    \uid{}: a \panon{} identifier for the current session.
   \item
@@
 Furthermore, despite the limitations, this telemetry supports a variety of
-interesting conclusions that we showcase in the next section.
+inferences that we showcase in the next section.
 
@@
 mark the start of a new day.
-Each of the blue bars in the plot bar is a bucket that counts the number of
-records generated per hour according to a client-side timestamp.
-Buckets contain a few dozen to over 3,000 records, typically about 1,000.
+Each day has 24 records, one per hour, represented by a blue bar
+indicating the number of records generated in that hour.
+Each hour has a few dozen to over 3,000 records, typically about
+1,000.\footnote{We conjecture the very low bars are due to outage on
+  the logging server, but need to investigate further.}
 %% \FILL{} investigate one tower, what's happening?? all one session?
@@
   \item
-    Some records have identical timestamps and session IDs.
+    Some records have identical timestamps and session \uid{}s.
     We keep the first such record and discard the rest.
@@
 
-  {Among the mixed-mode sessions:
+  {Among the other sessions:
         176 contain a mode upgrade,
@@
 switch focus to a module with a different analysis mode.
-Among the other, mixed-mode sessions, most of those (\pct{80}) switch to a
+Among the other sessions, most of those (\pct{80}) switch to a
 module with a different mode, about half contain at least one edit that
@@
 
-\subsection{Type Error Popularity}
+\subsection{Type Error Frequency}
 \label{s:type-error-count}
@@
 
-\Cref{t:type-error-count} sorts type errors by their popularity
+\Cref{t:type-error-count} sorts type errors by
+the frequency with which they appear
 in \mnonstrict{} and \mstrict{} mode.
 The purpose of this table is to discover and rank all errors that
-appear in the dataset, thus it is based on both keystroke and
-module-switch telemetry records even though edit ranges are not well-defined
+appear in the dataset; thus, it is based on both keystroke and
+module-switch telemetry records, even though edit ranges are not well-defined
 for the latter.
@@
 
-In other professional-grade IDEs, such as VSCode~\cite{vscode-telemetry},
+In other professional-grade \IDE{}s, such as VSCode~\cite{vscode-telemetry},
 IntelliJ~\cite{intellij-telemetry}, and \code{.NET}~\cite{dotnet-telemetry}, telemetry may include all sorts of
@@
 license agreement.
-Furthermore, at least in VSCode, IDE extensions can report their own telemetry.
+Furthermore, at least in VSCode, \IDE{} extensions can report their own telemetry.
 While this data can be invaluable for discovering bugs in production, it
 must be handled with extreme care.
-
 The Blackbox project similarly collects telemetry in broad strokes.
@@
 any comments above the main class) along with the compiler's response~\cite{bkmu-sigcse-2014}.
-%% Though somewhat intrusive,
 The millions of programs in the Blackbox dataset
-have enabled dozens of contributions to the area of CS
-education~\cite{bask-icer-2018}. However, these operate at the
-opposite end of privacy.
+have, however, enabled dozens of contributions to the area of CS
+education~\cite{bask-icer-2018}.
 
@@
 Unlike Roblox Studio, transparent telemetry includes no timestamps and no 
-session ids.
+session \uid{}s.
 While useful for learning about the frequency and distribution of
-specific events, the lack of timestamps and ids means it cannot track
+specific events, the lack of timestamps and \uid{}s means it cannot track
 edits over time (which we use in \cref{f:error-density}).
@@
 % errors arise? what is the average number of type errors?).
-% However, the lack of timestamps and ids means that it cannot track edits
-% over time (a feature that is quite useful to us, for example, in~\cref{f:error-density}).
 % In~\cref{f:error-density}, for example, we see that increases to the number

### 2023-09-30 programming23/paper.tex
sec 2

@@
 
-In addition to the main type analysis, Roblox Studio takes a second pass over
-every codebase with a \emph{\FS{}} analysis to infer autocomplete suggestions
-and drive other IDE tools.
-To infer precise types,
-the \FS{} analysis uses rigorous checks very similar to those of
-\mstrict{} mode (but not identical: see \cref{s:strict-vs-forcedstrict})
-no matter what mode the creator declared for the script.
-(Internally to Roblox Studio, this analysis is called \emph{forced strict}
-because it ignores the creator's choice and applies strong checks.)
-When it fails to infer a type, \FS{} analysis defaults to the unknown gradual
-type~\cite{st-sfp-2006}.
-Creators cannot see the errors reported by \FS{} analysis.
+\begin{figure}[t]\centering
+  \includegraphics[width=.35\textwidth]{img/roblox-studio.png}
+  \hspace{1cm}
+  \includegraphics[width=.35\textwidth]{img/roblox-studio-ide.png}
+  \caption{{Roblox Studio 3D creation} tools (left) and IDE (right).}
+  \label{fig:roblox-studio}
+\end{figure}
 
@@
 
-\begin{figure}[t]\centering
-  \includegraphics[width=.35\textwidth]{img/roblox-studio.png}
-  \hspace{1cm}
-  \includegraphics[width=.35\textwidth]{img/roblox-studio-ide.png}
-  \caption{{Roblox Studio 3D creation} tools (left) and IDE (right).}
-  \label{fig:roblox-studio}
-\end{figure}
+In addition to the main type analysis, Roblox Studio takes a second pass over
+every codebase with a \emph{\FS{}} analysis to infer autocomplete suggestions
+and drive other IDE tools.
+To infer precise types,
+the \FS{} analysis uses rigorous checks very similar to those of
+\mstrict{} mode (but not identical: see \cref{s:strict-vs-forcedstrict})
+no matter what mode the creator declared for the script.
+(Internally to Roblox Studio, this analysis is called \emph{forced strict}
+because it ignores the creator's choice and applies strong checks.)
+When it fails to infer a type, \FS{} analysis defaults to the unknown gradual
+type~\cite{st-sfp-2006}.
+Creators cannot see the errors reported by \FS{} analysis.
 

### 2023-09-30 programming23/paper.tex
related work todos

@@
   \includegraphics[width=.35\textwidth]{img/roblox-studio-ide.png}
-  \caption{{Roblox Studio 3D creation} tools (left) and IDE (right).}
+  \caption{{Roblox Studio 3D creation} tools (left) and \IDE{} (right).}
   \label{fig:roblox-studio}
@@
 every codebase with a \emph{\FS{}} analysis to infer autocomplete suggestions
-and drive other IDE tools.
+and drive other \IDE{} tools.
 To infer precise types,
@@
 We conjecture that Luau needs further tailoring to match
-common idioms in Roblox scripts.
+common idioms in untyped code.
 Discovering these idioms is a question for future work~(\cref{s:conclusion}).
@@
 Sessions may end abruptly, for instance because of machine reboot,
-instead of ending when the creator was finished editing code.
+instead of ending when the creator was finished editing.
 
@@
 randomly-selected runs of the type checker rather than specific events of interest.
-To the best of our knowledge, this design occupies a unique position relative to prior
-work~(\cref{t:telemetry-design}).\todo{Where do these labels come
-  from?!? There are no ``prior work'' citations either!}
-
-In other professional-grade IDEs, such as VSCode~\cite{vscode-telemetry},
-IntelliJ~\cite{intellij-telemetry}, and \code{.NET}~\cite{dotnet-telemetry}, telemetry may include all sorts of
+To the best of our knowledge, this design occupies a unique position relative
+to prior work, the two main strands of which we describe below and summarize
+in~\cref{t:telemetry-design}.
+
+In a \emph{classic} telemetry system, such as the ones in
+VSCode~\cite{vscode-telemetry}, IntelliJ~\cite{intellij-telemetry}, and
+\code{.NET}~\cite{dotnet-telemetry},
+telemetry may include all sorts of
 data: error reports, source code fragments, time\-stamps, and filepaths.
@@
 any comments above the main class) along with the compiler's response~\cite{bkmu-sigcse-2014}.
-%% Though somewhat intrusive,
 The millions of programs in the Blackbox dataset
 have enabled dozens of contributions to the area of CS
-education~\cite{bask-icer-2018}. However, these operate at the
-opposite end of privacy.
+education~\cite{bask-icer-2018}, but its telemetry does little to respect privacy.
 
@@
 
-At the opposite end of the privacy spectrum, Go's design for \emph{transparent
-telemetry} reports only counter values~\cite{transparent-telemetry}.
+At the opposite end of the privacy spectrum, Go's design for \emph{transparent}
+telemetry reports only counter values~\cite{transparent-telemetry}.
 Unlike Roblox Studio, transparent telemetry includes no timestamps and no 
-session ids.
+session \uid{}s.
 While useful for learning about the frequency and distribution of
-specific events, the lack of timestamps and ids means it cannot track
+specific events, the lack of timestamps and \uid{}s means it cannot track
 edits over time (which we use in \cref{f:error-density}).
@@
 A major difference is that we have only samples, not full sequences,
-which significantly impacts our study methods.\todo{Reasonable rewrite?}
+which significantly impacts our study methods.
 
 The BitFit project collects data about beginning students interactions in
-Java~\cite{ekc-wccce-2016,anna-russo-kennedy-ms-2006}; it does
-not collect source code.\todo{Does it sample? If not, say so!}
-However, its data do not contain details about errors.\todo{Really?!?}
+Java~\cite{ekc-wccce-2016,anna-russo-kennedy-ms-2006}.
+It does not collect source code, nor does it collect error messages;
+the telemetry simply counts all (i.e., without sampling) compile events, hint
+requests, etc. in a transparent~\cite{transparent-telemetry} manner.
 Other logging-based studies of Java programs use the full error message to
@@
 or static
-corpuses~\cite{rppd-fse-2014,bhmvv-toplas-2019,bmvv-arxiv-2019}.\todo{Why
-  are we citing arXiv?}
+corpuses~\cite{rppd-fse-2014,bhmvv-toplas-2019}
 Controlled experiments are rare;

### 2023-09-30 programming23/paper.tex
sec 5

@@
 Luau has an efficient type checker that can
-analyze thousands of lines of code~(\cref{t:codebase-size}).
+analyze thousands of lines of code~(\cref{t:codebase-size}) quickly
+enough to be used on-the-fly in an IDE.
 But, few sessions use type analysis directly and the data presents little
@@
 reporting on the same codebase.
-Sessions may end abruptly, for instance because of machine reboot,
-instead of ending when the creator was finished editing code.
+Sessions may end abruptly, e.g., because of machine reboot,
+instead of ending when the creator was done editing code.
 

### 2023-09-30 programming23/paper.tex

--- 2023-09-30 programming23/paper.tex
@@
 reporting on the same codebase.
-Sessions may end abruptly, for instance because of machine reboot,
-instead of ending when the creator was finished editing.
+Sessions may end abruptly, e.g., because of machine reboot,
+instead of ending when the creator was done editing code.
 

### 2023-09-30 programming23/paper.tex
fix resolution, fig 5
sec 6

@@
 privacy~\cite{zhlbr-cc-2020,epk-ccs-2014,wblj-usenix-2017}.
-Recent work that is especially relevant to the session-based data collected
-by Roblox Studio shows how to add noise to traces~\cite{zhlbr-cc-2020},
+Recent work about session-based data that could be added
+to Roblox Studio shows how to add noise to traces~\cite{zhlbr-cc-2020},
 how to account for known relationships between events~\cite{zhlbr-oopsla-2020},
-and how to choose the noise amount on the client side~\cite{hlzbr-ecoop-2021}.
+and how to choose the noise on the client side~\cite{hlzbr-ecoop-2021}.
 Applying differential privacy is subtle, however.
-While we can likely fuzz the total number of type errors, if we were to
+While we could fuzz the total number of type errors, if we were to
 fuzz the number of \code{CodeTooComplex} errors, it might drastically
-affect our conclusions.
+affect our inferences.
 
@@
 A major difference is that we have only samples, not full sequences,
-which significantly impacts our study methods.
+which significantly impacts our study methods and inferences.
 
@@
 the telemetry simply counts all (i.e., without sampling) compile events, hint
-requests, etc. in a transparent~\cite{transparent-telemetry} manner.
+requests, etc. While it does operate in a
+transparent~\cite{transparent-telemetry} manner, it only provides
+coarse-grained information that would be insufficient in our study.
 Other logging-based studies of Java programs use the full error message to
 draw conclusions~\cite{bgimgm-cse-2016,dlc-iticse-2014}, which we
-cannot use to preserve privacy.
+cannot use.
 

### 2023-09-30 programming23/paper.tex
sec 7

@@
 % \usepackage[sort]{natbib} or otherwise manually fix cites to be in ascending order
-% redo fig 5 to have better resolution
 
@@
 internal extension is enabled or not.
-Three other aspects are worth rethinking.
-First, it would be useful to have fine-grained counts for \FS{} analysis
+
+Three other aspects are also worth rethinking:
+\begin{enumerate}
+
+  \item It would be useful to have fine-grained counts for \FS{} analysis
 errors and to know the specific reason behind errors such as \code{TypeMismatch}.
 But, this additional data could double the size of telemetry records.
-Second, although tracking edit ranges was useful, it made the telemetry system
+
+\item Although tracking edit ranges was useful, it made the telemetry system
 much more difficult to build and maintain.
@@
 though there is a risk that it is too coarse.
-Third, the massive number of syntax errors begs the question of how to skip
+
+\item The massive number of syntax errors begs the question of how to skip
 them.
@@
 Another option is to ignore records that have a syntax error in the edit range.
-
-In future work, it would be interesting to build statistical models of
+\end{enumerate}
+
+It would also be interesting to build statistical models of
 the ``average'' programmer using each typing mode and compare their error
@@
 The errors that we do see in \mstrict{} mode call for an in-depth analysis via
-talk-aloud interviews to discover why type mismatches occur and how how Luau
+talk-aloud interviews to discover why type mismatches occur and how Luau
 can better accommodate untyped idioms.

### 2023-10-01 programming23/paper.tex
reorder cites

@@
 % TODO BEFORE SUBMISSION
-% \usepackage[sort]{natbib} or otherwise manually fix cites to be in ascending order
 
@@
 Luau represents a large-scale combination of ideas from
-gradual typing~\cite{st-sfp-2006,tfffgksst-snapl-2017,bat-ecoop-2014},
+gradual typing~\cite{bat-ecoop-2014,st-sfp-2006,tfffgksst-snapl-2017},
 success typing~\cite{lindahl2006practical},
@@
 To deal with pathologies such as the worst-case time for ML-style type
-inference~\cite{m-popl-1990,ktu-caap-1990}, the typechecker
+inference~\cite{ktu-caap-1990,m-popl-1990}, the typechecker
 has internal limits that restrict the problems it will attempt to solve.
@@
 techniques from differential
-privacy~\cite{zhlbr-cc-2020,epk-ccs-2014,wblj-usenix-2017}.
+privacy~\cite{epk-ccs-2014,wblj-usenix-2017,zhlbr-cc-2020}.
 Recent work about session-based data that could be added
@@
 Our approach to data analysis is inspired by Marceau
-et~al.~\cite{mfk-onward-2011,mfk-sigcse-2011} and
+et~al.~\cite{mfk-sigcse-2011,mfk-onward-2011} and
 Macedo et~al.~\cite{mcpcsprs-abz-2020,mcpcsprs-scp-2021}
 for Racket and Alloy4Fun, respectively. Other more recent work has
-also studied edit sequences to infer intent~\cite{wk-koli-2020,lgk-pj-2022,rsgl-cpp-2020}.
+also studied edit sequences to infer intent~\cite{lgk-pj-2022,rsgl-cpp-2020,wk-koli-2020}.
 A major difference is that we have only samples, not full sequences,
@@
 
-Much work~\cite{empirical-types,heeren-thesis} on the benefits of static types focuses on
-small case studies~\cite{w-popl-1986,hw-scp-2004,td-tosem-2001},
-inverviews~\cite{cdhhjklwya-hatra-2020,gstf-hatra-2021,cams-oopsla-2020},
+Much work~\cite{heeren-thesis,empirical-types} on the benefits of static types focuses on
+small case studies~\cite{hw-scp-2004,td-tosem-2001,w-popl-1986},
+inverviews~\cite{cdhhjklwya-hatra-2020,cams-oopsla-2020,gstf-hatra-2021},
 or static
-corpuses~\cite{rppd-fse-2014,bhmvv-toplas-2019}
+corpuses~\cite{bhmvv-toplas-2019,rppd-fse-2014}
 Controlled experiments are rare;
@@
 over two thousand programs.
-Seidel et~al.~\cite{sscwj-oopsla-2017,sjw-jfp-2018} use five thousand programs
+Seidel et~al.~\cite{sjw-jfp-2018,sscwj-oopsla-2017} use five thousand programs
 to train data-driven methods for localizing errors.

### 2023-10-02 programming23/paper.tex
error density FS, +-200

@@
 % TODO BEFORE FINAL
+% - explain edit range computations
 % - uncomment acks
@@
 \authorinfo{(\email{benjamin.l.greenman@gmail.com}) was a CIFellows 2020
-postdoc at Brown Univesity and is now an assistant professor
-at the recently-named Kahlert School of Computing at the University of Utah.}
+postdoc at Brown University and is now an assistant professor in the
+Kahlert School of Computing at the University of Utah.}
 \affiliation[a]{Brown University, Providence, RI, USA}
@@
     Edit Range & 3,680 & \stddev{31} & 926   & 34,725 & \includegraphics[width=0.14\columnwidth,valign=M]{img/editrange-distribution.pdf}
-    %% TODO x-axis starts at 1, not zero
-    %% TODO much bigger font or much smaller plot!!!
+    %% TODO x-axis starts at 1, not zero ... do we have zeros?! may need to explain
   \end{tabular}
@@
     %% TODO x-axis starts at 1
-    %% TODO huge font, or smaller plot
   \end{tabular}
@@
     \begin{tabular}[t]{cc}
-    %% TODO why text blurry??!
     Type errors & Background errors \\
@@
   \label{t:type-error-count}
-  \begin{tabular}{ll}
+  \begin{tabular}{l@{~~}l}
     \mnonstrict{} & \mstrict{} \\
@@
 
-%\newpage
+\newpage
+
+% \appendix
 % 
-%\appendix
+% \section{Data Details}
 % 
-%\section{Data Details}
+% Tips for interpreting the data:
 % 
-%Tips for interpreting the data:
-%
-%\begin{itemize}
+% \begin{itemize}
+%   \item
+%     A \code{"-"} entry denotes a zero.
+%     This data was not in the telemetry record.
+%     Kibana filled in the dash.
+%   \item
+%     Timestamps are California local time.
+%     %% (Daylight savings?)
+%   \item
 %   \item
@@
 %   \item
-%  \item
-%\end{itemize}
+% \end{itemize}
 % 
-%\newpage
+% \section{More Plots}
+% 
+% \begin{figure}[t]\centering
+%   %% +-200 background errors, same as ever! balanced
+%   \mnocheck{}
+%   \includegraphics[width=\columnwidth]{img/error-count-nocheck-row--fs-density-diff.pdf}
+% 
+%   \mnonstrict{}
+%   \includegraphics[width=\columnwidth]{img/error-count-nonstrict-row--fs-density-diff.pdf}
+% 
+%   \mstrict{}
+%   \includegraphics[width=\columnwidth]{img/error-count-strict-row--fs-density-diff.pdf}
+%   \caption{Changes in \FS{} error density (errors / lines) over time (seconds).}
+%   \label{f:error-density}
+% \end{figure}
 

### 2023-10-12 programming23/paper.tex
fig 6 caption
delete comment
power law check
except for lines, they all have p>0.2 which is what Clauset et al said was good

lines has p=0.1

did not add prose to the paper, this doesn't seem important after all
but maybe we'll have space to include in the final version

--- 2023-10-02 programming23/paper.tex
@@
 \begin{figure}[t]\centering
-  %% TODO bigger text
   %% TODO weird artifact (|) in x-min label
@@
 We discuss these in turn.
-% Commenting these out because that's literally the rest of the
-% subsection. Say anything you want from here over there!
-% Codebase size and session length have long-tailed
-% distributions\todo{DID YOU DO A STATISTICAL TEST.}
-% with many small items and few very large ones.
-% %% TODO check %% follow power-law distributions~\cite{csn-siamr-2009} 
-% %% https://aaronclauset.github.io/powerlaws/
-% %% bin version, is this relevant? sounds like no from abtract https://arxiv.org/pdf/1208.3524.pdf
-% The number of type analysis errors is somewhat low~(explained in the next
-% section), but sufficient to proceed with futher analysis.
 
@@
 The largest ones have over 50,000 lines of code
-while the smallest have merely 1 line of code.
+while the smallest have zero lines of code.
+%% TODO filter the zero lines? what happens in that data?? run a query
 Unsurprisingly, these wide-ranging numbers come with large standard
@@
 and 1,000 lines in edit ranges.
+
+%% power law test~\cite{csn-siamr-2009,g-jss-2015}
+%%  lines     xmin = 2 alpha = 1.46 p = 0.1 (low!)
+%%  editrange xmin = 1 alpha = 1.48 p = 0.73
 
@@
     Record Count    & 286 & 583         & 138 & 3,302  & \includegraphics[width=0.14\columnwidth,valign=M]{img/event-count-distribution.pdf}
-    %% TODO x-axis starts at 1
   \end{tabular}
@@
 while the shortest last a few milliseconds or include just one record.
-% Likewise, both plots in the figure follow a power-law distribution
-% (especially for record count).
+%% 0 ms = 1 event
 A median session runs on the order of minutes and consists of a few dozen events.
-%% The median time is 14 minutes and the median 
-%% The median number of records is 138.
-% These numbers are reasonable.
+
+%% power law test~\cite{csn-siamr-2009,g-jss-2015}
+%%  timespan    xmin = 1 alpha = 2.88 p = 0.23
+%%  event-count xmin = 1 alpha = 1.33 p = 0.86
 
@@
 
-% BEN: double-check, I think this says all you were trying to say.
-% "typically" because we COULD have gotten lucky.
-% ben: I'd written "immediatly introduces" to rule out getting lucky. Deleted "immediately" and went with your version.
 The counts are imprecise because of telemetry granularity.
@@
 Note that all three sessions \emph{could} have been from the same codebase.
-% BEN: I see no reason for this speculation. The only important thing
-% is the fact that this could have been all the same codebase.
-%It is entirely possible that they all came from one session that
-%deliberately pushed the limits of type analysis.
-%% Whatever the reason, the overall number is low enough to suggest that the
-%% internal limits are not posing barriers to typechecker adoption.
 
@@
   %% - nocheck indeed the highest > NS > S
-  %% TODO draw units, bigger font
 
@@
   \includegraphics[width=\columnwidth]{img/error-count-strict-row--te-density-diff.pdf}
-  \caption{Changes in type error density (errors / lines) over time (seconds).}
+  \caption{Changes in type error density ($\frac{\mbox{\#errors}}{\mbox{\#lines}}$) over time (seconds).}
   \label{f:error-density}

### 2023-10-14 programming23/paper.tex
prio + poplar

@@
 
+Prio and Poplar demonstrate a telemetry method to compute public aggregates
+over private data using a middle layer of
+servers~\cite{cb-nsdi-2017,bbcgi-sp-2021}.
+Clients can depend on strong privacy guarantees even if other clients and some
+servers (but not all) are malicious.
+The design has been deployed in Mozilla Firefox to monitor blocklist
+usage~\cite{mozilla-prio}.
+Luau telemetry might benefit from a redesign to use middle-layer servers,
+though it is unclear what data from type analysis would support meaningful
+aggregates.
+%% In particular, the data would still need to be relatively small to trasmit.
+
 Cooperative bug isolation is a method for designing

### 2023-10-14 programming23/paper.tex
rewording prio: cut 'strong' add 'private'

@@
 servers~\cite{cb-nsdi-2017,bbcgi-sp-2021}.
-Clients can depend on strong privacy guarantees even if other clients and some
+Clients can depend on privacy guarantees even if other clients and some
 servers (but not all) are malicious.
@@
 Luau telemetry might benefit from a redesign to use middle-layer servers,
-though it is unclear what data from type analysis would support meaningful
+though it is unclear what private data from type analysis would support meaningful
 aggregates.

### 2023-12-05 programming23/paper.tex
sec5, label RQs

@@
 \begin{description}
-  \item[How many sessions use type analysis?]
+  \item[RQ1.]\leavevmode
+    \begin{description}
+      \item[- How many sessions use type analysis?]
         A mere \pct{10} of sessions (and of telemetry records overall) opt-in to
@@
 
-  \item[How often do projects contain modules with different analysis modes?]
+      \item[- How often do projects contain modules with different modes?]
         Sessions rarely combine analysis modes.
@@
 
-  \item[How often do sessions turn analysis off?]
+      \item[- How often do sessions turn analysis off?]
         Sessions rarely disable type analysis after opting in.
@@
         %%%%  (strict nocheck) : 49.79\%
-
-  \item[For modules that use type analysis (\mstrict{} and \mnonstrict{}):]\leavevmode
+  \end{description}
+
+  \item[RQ2. For modules that use type analysis (\mstrict{} and \mnonstrict{}):]\leavevmode
     \begin{description}
@@
 
-  \item[What impact does type analysis have on the number of \FS{} errors?]
+  \item[RQ3. What impact does type analysis have on the number of \FS{} errors?]
     Opting in to type analysis has little impact on \FS{} errors.

### 2023-12-05 programming23/paper.tex
sec 5 clarify 'contradiction'

@@
         analysis modes.
-        This number may be a significant underapproximation because we have data
+        This number may be an underapproximation because we have data
         only on modules that the creator chose to open, but the widespread use
         of \mnocheck{} mode suggests that it is not off by much.
+        If $\pct{90}$ of \emph{all} modules indeed use \mnocheck{}, then multi-mode
+        projects must be rare.
 

### 2023-12-05 programming23/paper.tex
typo fixes

@@
 
+\newcommand{\etal}{et~al.}
 \newcommand{\uid}{\textsc{id}}
@@
   must support a wide audience.
-  As part of its efforts to support all kinds programmers, Luau includes an
+  As part of its efforts to support all kinds of programmers, Luau includes an
   optional, gradual type system and goes to great lengths to minimize false
@@
 The upshot of this strategy is that it reduces the size and complexity of
-telemetry records because there is only one inerval for errors to overlap with.
+telemetry records because there is only one interval for errors to overlap with.
 
 This telemetry design is admittedly coarse-grained.
-For instance, it does not distinguish between edits that ignore a type
+For instance, it does not distinguish edits that ignore a type
 error from edits that remove one error while introducing another.
@@
 The main advantages of this telemetry are the small size of each record
-and complete the lack of private information.
+and the complete lack of private information.
 Furthermore, despite the limitations, this telemetry supports a variety of
@@
 \begin{table}[t]\centering
-  \caption{Size of analyzed code: number of files, number of lines, and lines in edit range}
+  \caption{Size of analyzed code: number of lines and lines in edit range}
   \label{t:codebase-size}
@@
     creators might choose to edit a non-highlighted location to fix an error.
-    While non-privacy-protecting analysis have answered such
+    While non-privacy-protecting analyses have answered such
     questions~\cite{mfk-sigcse-2011}, our data cannot narrow down the answer.
@@
 small case studies~\cite{hw-scp-2004,td-tosem-2001,w-popl-1986},
-inverviews~\cite{cdhhjklwya-hatra-2020,cams-oopsla-2020,gstf-hatra-2021},
+interviews~\cite{cdhhjklwya-hatra-2020,cams-oopsla-2020,gstf-hatra-2021},
 or static
-corpuses~\cite{bhmvv-toplas-2019,rppd-fse-2014}
+corpuses~\cite{bhmvv-toplas-2019,rppd-fse-2014}.
 Controlled experiments are rare;
-Kleinschmager, et~al.\cite{khrts-icpc-2012} is a notable exception.
-Lerner et~al.~\cite{lfgc-pldi-2007} apply a type error repair tool on
+Kleinschmager \etal{}~\cite{khrts-icpc-2012} is a notable exception.
+Lerner \etal{}~\cite{lfgc-pldi-2007} apply a type error repair tool on
 over two thousand programs.
-Seidel et~al.~\cite{sjw-jfp-2018,sscwj-oopsla-2017} use five thousand programs
+Seidel \etal{}~\cite{sjw-jfp-2018,sscwj-oopsla-2017} use five thousand programs
 to train data-driven methods for localizing errors.
@@
 % \cite{}
-% heeren
 % constraint-based type inference;
@@
     The event ordering that timestamps provide was invaluable; e.g., for learning
-    that error deltas bounce up and down rather than rising shaply, then falling~(\cref{f:error-density}).
+    that error deltas bounce up and down rather than rising sharply, then falling~(\cref{f:error-density}).
 \end{itemize}

### 2023-12-06 programming23/paper.tex
artifact todo outline, edit buckets/hours discussion

@@
 mark the start of a new day.
-Each day has 24 records, one per hour, represented by a blue bar
-indicating the number of records generated in that hour.
+For each of the 24 hours in each day, there is one blue bar
+indicating the number of records generated during that hour.
 Each hour has a few dozen to over 3,000 records, typically about
@@
 This paper presents the first large-scale analysis of developers' interactions with
-an industrial strength typechecker. It is nevertheless intentionally
-limited by the nature and frequency of sampling. Despite these restrictions,
+an industrial strength typechecker in a privacy-respecting way.
+Despite the intentionally-restricted data,
 the analysis offers several lessons for the Luau
 developers and other maintainers of gradually-typed languages:
+%% TODO more lessons?
 \begin{itemize}

### 2023-12-07 programming23/paper.tex
import cycle prose

@@
 {imported} by other scripts.
-Imports therefore form a graph in general,
-though it is an error in \mstrict{} mode to create an import cycle
-and edges are removed to make the graph acyclic.
-%% \QALAN{}: do all modes remove the edges?
+Imports therefore form a graph in general, though the analysis
+removes edges to ensure an acyclic graph.
+In \mstrict{} mode, it is an error to create an import cycle.
 
@@
 module gets analyzed by the typechecker.
-In the common case when typechecking is performed for autocomplete,
-only the current script gets checked because it is the only dirty
-script, and because nothing it requires can transitively require it, since
-the module graph is acyclic.
+When typechecking is performed for autocomplete, the common
+case is that only the current script gets checked because it
+is the only dirty script.
+(Since the module graph is acyclic, nothing that the current
+script requires can transitively require it.)
 With this strategy, the typechecker can run after every keystroke without

### 2023-12-08 programming23/artifact/main.tex
checkpoint on artifact, bib, edits

\documentclass{article}

\usepackage{graphicx}
\usepackage[margin=2cm]{geometry}
\usepackage{verbatim}
\usepackage[export]{adjustbox}

\begin{document}

\newcommand{\modefont}[1]{\texttt{#1}}
\newcommand{\mnocheck}{\modefont{nocheck}}
\newcommand{\mnonstrict}{\modefont{nonstrict}}
\newcommand{\mstrict}{\modefont{strict}}

%% ---
\subsection*{Figure 3: Records per Hour}

\includegraphics[width=\columnwidth]{out/row-distribution.pdf}

TODO reason for sending

%% ---
\subsection*{Table 2: Size of Analyzed Code}

\includegraphics[width=0.4\columnwidth]{out/lines-distribution.pdf}
\includegraphics[width=0.4\columnwidth]{out/editrange-distribution.pdf}

TODO mean stddev median P99

%% ---
\subsection*{Table 3: Session Size}

\includegraphics[width=0.4\columnwidth]{out/timespan-distribution.pdf}
\includegraphics[width=0.4\columnwidth]{out/event-count-distribution.pdf}

TODO mean stddev median P99

%% ---
\subsection*{Table 4: Current Type Errors and Background Errors}

TODO

%% ---
\subsection*{Figure 4: Overview of Mode Usage}

TODO

%% ---
\subsection*{Figure 5: Type and Background Errors Grouped by Mode}

{ \newcommand{\labelbars}[1]{\begin{tabular}[t]{l@{}l} \raisebox{2ex}{\begin{tabular}[t]{r}\mnocheck{}\\\mnonstrict{}\\\mstrict{}\end{tabular}} & #1 \end{tabular}}
  \begin{tabular}[t]{cc}
    Type errors & Background errors \\
    \labelbars{\includegraphics[width=0.20\columnwidth,valign=M]{out/error-by-mode-te.pdf}}
    & \labelbars{\includegraphics[width=0.20\columnwidth,valign=M]{out/error-by-mode-fs.pdf}}
  \end{tabular}
}

%% ---
\subsection*{Table 5: Specific Errors in Edit Range}

TODO

%% ---
\subsection*{Table 6: Type Error Popularity}

TODO

{\footnotesize
\verbatiminput{out/ctc-info.txt}
}

%% ---
\subsection*{Figure 6: Type Error Density}

\includegraphics[width=\columnwidth]{out/error-count-nocheck-row--te-density-diff.pdf}
\includegraphics[width=\columnwidth]{out/error-count-nonstrict-row--te-density-diff.pdf}
\includegraphics[width=\columnwidth]{out/error-count-strict-row--te-density-diff.pdf}

\end{document}

### 2023-12-08 programming23/paper.tex
checkpoint on artifact, bib, edits

@@
 
-\documentclass[english,submission,cleveref]{programming}
+\documentclass[english,cleveref]{programming}
 %% use 'submission' for initial submission, remove it for camera-ready (see 5.1)
@@
     and scripts to analyze them.
-    \emph{We intend to submit these in an artifact.}
+    \emph{TODO We intend to submit these in an artifact.}
 
@@
     \code{GenericError}
-    & Generic label for other non-type errors, such as\\
-    & \hbox{}~~looping over an unordered table.
+    & Generic label for other non-type errors,\\
+    & \hbox{}~~such as looping over an unordered table.
     %% attempting to extend a type that does not describe a class
@@
 Each hour has a few dozen to over 3,000 records, typically about
-1,000.\footnote{We conjecture the very low bars are due to outage on
+1,000.
+{We conjecture the very low bars are due to outage on
   the logging server, but need to investigate further.}
+The tall bars are FILL TODO.
 %% \FILL{} investigate one tower, what's happening?? all one session?
@@
   \begin{tabular}[t]{lrrrrl}
-    ~          & Mean  & Stddev & Median & \ptile{99} & Distribution \\
+    ~          & Mean  & Stddev & Median & \ptile{99} & Distribution \\\midrule
     Lines      & 6,497 & \stddev{22} & 3,115 & 50,547 & \includegraphics[width=0.14\columnwidth,valign=M]{img/lines-distribution.pdf} \\
@@
   \begin{tabular}{lrrrrl}
-    ~               & Mean & Stddev & Median & \ptile{99} & Distribution \\
+    ~               & Mean & Stddev & Median & \ptile{99} & Distribution \\\midrule
     Time Span (sec) & 3,184 & \stddev{16} & 845 & 35,450 & \includegraphics[width=0.14\columnwidth,valign=M]{img/timespan-distribution.pdf} \\
@@
            156,883 & \mnonstrict{}        & [\pct{10.43}] \\
-             6,505 & \mstrict{}           & [\pct{ 0.43}]
+             6,505 & \mstrict{}           & [\pct{ 0.43}] \\
+             {}
       \end{tabular}
@@
     records out of the 6,505 total~(\cref{f:dataset-overview}).
-    \Cref{s:count-analysis-errors} discussions the implications
+    \Cref{s:count-analysis-errors} discusses the implications
     of this low overlap rate for future work.
@@
     \code{SyntaxError} and the related \code{UnknownSymbol} and
-    \code{UnknownProperty} are most popular.
+    \code{UnknownProperty} are the most common.
     If the edit range errors are indicative of type errors at large,
@@
   \begin{tabular}{l@{~~}l}
-    \mnonstrict{} & \mstrict{} \\
+    \mnonstrict{} & \mstrict{} \\\midrule
     \begin{tabular}[t]{l@{~}r}
@@
 ends at 150 seconds.
+%% TODO explain why 150 ... shows some tail but not whole thing
 The y axis shows changes to errors over time; a positive number means errors
@@
 \begin{figure}[t]\centering
+  %% code/error-count.rkt
   % TODO use bars instead, percent only, don't care about counts
@@
 In a \emph{classic} telemetry system, such as the ones in
-VSCode~\cite{vscode-telemetry}, IntelliJ~\cite{intellij-telemetry}, and
+VSCode~\cite{vscode-telemetry}, IntelliJ~\cite{intellij-telemetry},
+Eclipse~\cite{eclipse-udc}, and
 \code{.NET}~\cite{dotnet-telemetry},
@@
 must be handled with extreme care.
-The Blackbox project similarly collects telemetry in broad strokes.
-Every Java program submitted for compilation goes into the Blackbox database (minus
-any comments above the main class) along with the compiler's response~\cite{bkmu-sigcse-2014}.
-The millions of programs in the Blackbox dataset
-have enabled dozens of contributions to the area of CS
-education~\cite{bask-icer-2018}, but its telemetry does little to respect privacy.
-
-% BEN: saving space.
-% , ranging from categorization of novice
-% errors~\cite{mk-fie-2014,m-masters-2016} to analyses of error frequency and
-% time-to-fix~\cite{ab-sigcse-2015}.
-% Related prior work used the GRUMPS telemetry system to collect 4.7 million
-% keystroke-level actions from learners programming in
-% Ada~\cite{tm-iticse-2004,tkdmceg-ascilite-2003}.
-% This data is not available.
-% BEN: Not sure what the above sentence means and whether we should
-% keep it. Is it that you can't freely get Blackbox, you have to ask
-% Neil Brown for permission or something?
 
@@
 edits over time (which we use in \cref{f:error-density}).
-% BEN: commented for space.
-% it cannot track 
-% The data is useful for learning about the frequency of specific events and the
-% distribution of events across clients (e.g., how many \code{CodeTooComplex}
-% errors arise? what is the average number of type errors?).
-% In~\cref{f:error-density}, for example, we see that increases to the number
-% of type errors are quickly followed by decreases.
-% Without timestamps, we would be unable to rule out the possibility that users
-% build up a mountain of type errors and then slowly reduce them to zero.
 
@@
 privacy~\cite{epk-ccs-2014,wblj-usenix-2017,zhlbr-cc-2020}.
-Recent work about session-based data that could be added
-to Roblox Studio shows how to add noise to traces~\cite{zhlbr-cc-2020},
+Recent work shows how to add noise to traces~\cite{zhlbr-cc-2020},
 how to account for known relationships between events~\cite{zhlbr-oopsla-2020},
 and how to choose the noise on the client side~\cite{hlzbr-ecoop-2021}.
+Roblox Studio could benefit from these techniques.
 Applying differential privacy is subtle, however.
@@
 telemetry systems~\cite{liblit-thesis}.
-The focus is not on privacy, but rather on how to collect a small amount of
-data from each user and perform statistically-sound analyses.
+The goal is to collect a small amount of data from each user and perform
+statistically-sound analyses; privacy is not a main concern.
 Each feature of interest \emph{within} a user's codebase has a uniform-random
@@
 
-The BitFit project collects data about beginning students interactions in
-Java~\cite{ekc-wccce-2016,anna-russo-kennedy-ms-2006}.
-It does not collect source code, nor does it collect error messages;
-the telemetry simply counts all (i.e., without sampling) compile events, hint
-requests, etc. While it does operate in a
-transparent~\cite{transparent-telemetry} manner, it only provides
-coarse-grained information that would be insufficient in our study.
-Other logging-based studies of Java programs use the full error message to
-draw conclusions~\cite{bgimgm-cse-2016,dlc-iticse-2014}, which we
-cannot use.
+There is a rich body of work that uses IDE instrumentation to study programmers
+(e.g.,~\cite{hl-icse-2010,j-ieeesoft-2013,kje-ase-2010, mmlb-swvis-2014, ncvjd-ecoop-2013, rl-icse-2008, we-xp-2004}).
+WatchDog records the time that programmers spend reading, writing, and testing
+code~\cite{blpgz-icse-2016}.
+TDD-Guide collects similar data to teach testing methods~\cite{mdk-xp-2008}.
+Examplar logs programs to give students feedback on their tests~\cite{pgnwk-koli-2022,wk-icer-2019}.
+The Blackbox project collects Java programs and has enabled
+dozens of contributions to the area of CS education~\cite{bask-icer-2018,bkmu-sigcse-2014}.
+Other logging-based studies of Java programs draw conclusions from compiler error
+messages~\cite{bgimgm-cse-2016,dlc-iticse-2014}.
+The GRUMPS telemetry system collected 4.7 million keystroke-level actions from
+Ada learners~\cite{tkdmceg-ascilite-2003,tm-iticse-2004}.
+None of the above are sufficiently private to work for Roblox Studio
+The BitFit project counts notable events (compile program, ask for hint, etc.)
+for Java~\cite{ekc-wccce-2016,anna-russo-kennedy-ms-2006}.
+While it does operate in a transparent~\cite{transparent-telemetry} manner, it
+only provides coarse-grained information that would be insufficient in our
+study.
 
@@
 % https://link.springer.com/content/pdf/10.1023/a:1008040416962.pdf
-
-% \cite{}
-% constraint-based type inference;
-% general te criteria:
-% - target audience (novice / expert)
-% - output format (text / viz)
-% - interactivity (yes / no)
-% - heuristics (yes / no)
-% - primary-or-external (most type errors are easy, some could really use dedicated help)
 

### 2023-12-08 programming23/artifact/main.tex
artifact checkpoint

@@
 
-TODO reason for sending
+Lines 1-2 of \texttt{out/overview.txt} give the reason for sending:
+
+\begin{verbatim}
+ 1504736 total logs = 1341348 nocheck + 156883 nonstrict + 6505 strict
+  508572 due to module switch
+\end{verbatim}
 
@@
 
-\includegraphics[width=0.4\columnwidth]{out/lines-distribution.pdf}
-\includegraphics[width=0.4\columnwidth]{out/editrange-distribution.pdf}
+From \texttt{out/summary-of-size-distributions.rktd}:
 
-TODO mean stddev median P99
+{\footnotesize
+\verbatiminput{out/summary-of-size-distributions.rktd}
+}
+
+\begin{minipage}{0.5\columnwidth}
+  \textbf{Lines:}\\
+  \includegraphics[width=\columnwidth]{out/lines-distribution.pdf}
+\end{minipage}\begin{minipage}{0.5\columnwidth}
+  \textbf{Edit Range:}\\
+  \includegraphics[width=\columnwidth]{out/editrange-distribution.pdf}
+\end{minipage}
 
@@
 
+See previous section for mean, stddev, median, and P99.
+
+\begin{minipage}{0.5\columnwidth}
+  \textbf{Timespan:}
   \includegraphics[width=0.4\columnwidth]{out/timespan-distribution.pdf}
+\end{minipage}\begin{minipage}{0.5\columnwidth}
+  \textbf{Event Count / Record Count:}
   \includegraphics[width=0.4\columnwidth]{out/event-count-distribution.pdf}
-
-TODO mean stddev median P99
+\end{minipage}
 
@@
 
-TODO
+Lines 3-8 of \texttt{out/overview.txt}:
+
+\begin{verbatim}
+ 72235735 total forced strict type errors, 37027281 in module,
+ curr 1111178 in edit regions,
+ olde 938500 in edit regions
+ 595137 total type errors, 289698 in module,
+ curr 14917 non-stx + 16007 syntax in edit regions,
+ olde 13836 non-stx + 28565 syntax in edit regions
+\end{verbatim}
 
 %% ---
-\subsection*{Figure 4: Overview of Mode Usage}
+\subsection*{Figure 4: Overview of Type Analysis Modes}
 
-TODO
+Line 1 of \texttt{out/overview.txt} partitions logs by mode:
+
+\begin{verbatim}
+ 1504736 total logs = 1341348 nocheck + 156883 nonstrict + 6505 strict
+\end{verbatim}
+
+Lines 9-13 of \texttt{out/overview.txt} partition sessions and report on upgrades and downgrades,
+\emph{but} the final two numbers are incorrect (upgrade and downgrade) because they include
+module switches:
+
+\begin{verbatim}
+ 347598 sessions
+  346956 single mode = 313509 nocheck + 32902 nonstrict + 545 strict
+  512 multi mode projects
+  341 mode upgrades
+  320 mode downgrades
+\end{verbatim}
+
+Correct up/down-grade numbers are in \texttt{out/downgrade-count.txt}:
+
+\includeverbatim{out/downgrade-count.txt}
+
+TODO ^^^
 

### 2023-12-08 programming23/artifact/main.tex
artifact more more more

@@
 
+Plot from \texttt{code/row-distribution.rkt}, which uses the dataset
+\texttt{out/row-distribution.rktd}, which comes from running
+\texttt{prepare.rkt --mode main}.
+
 \includegraphics[width=\columnwidth]{out/row-distribution.pdf}
 
-Lines 1-2 of \texttt{out/overview.txt} give the reason for sending:
+Lines 1-2 of \texttt{out/overview.txt} (output of \texttt{prepare.rkt --mode main})
+give the reason for sending:
 
@@
 
-From \texttt{out/summary-of-size-distributions.rktd}:
+From \texttt{out/summary-of-size-distributions.rktd},\\
+which is the output of \texttt{code/sdupdate.rkt}\\
+on \texttt{out/size-distributions.rktd}:
 
@@
 \begin{minipage}{0.5\columnwidth}
-  \textbf{Timespan:}
+  \textbf{Timespan:}\\
   \includegraphics[width=0.4\columnwidth]{out/timespan-distribution.pdf}
 \end{minipage}\begin{minipage}{0.5\columnwidth}
-  \textbf{Event Count / Record Count:}
+  \textbf{Event Count / Record Count:}\\
   \includegraphics[width=0.4\columnwidth]{out/event-count-distribution.pdf}
@@
 
-Correct up/down-grade numbers are in \texttt{out/downgrade-count.txt}:
+Correct up/down-grade numbers are in \texttt{out/downgrade-count.txt},
+which is the output of \texttt{code/downgrade-count.rkt} after running
+the following commands to generate data:
 
-\includeverbatim{out/downgrade-count.txt}
+\begin{verbatim}
+PLTSTDERR="error info@luau" racket prepare.rkt --mode divide-sessions
+PLTSTDERR="error info@luau" racket prepare.rkt --mode session-query
+PLTSTDERR="error info@luau" racket prepare.rkt --mode aggregate-te
+\end{verbatim}
 
-TODO ^^^
+% \includeverbatim{out/downgrade-count.txt}
+
+% TODO
 
@@
 \subsection*{Figure 5: Type and Background Errors Grouped by Mode}
+
+These plots come from \texttt{code/error-by-mode.rkt}, which uses
+hard-coded data from \texttt{out/overview.txt}:
 
@@
 
+Output from \texttt{code/type-error-survival.rkt}
+run on \texttt{code/type-error-survival-ss-*.rktd}:
+
 TODO
@@
 
+Output from \texttt{code/type-error-count.rkt}
+run on \texttt{code/te-editrange-ss-*.rktd}:
+
 TODO
+
+\subsubsection*{Internal Limits, Code Too Complex}
+
+Data from \texttt{out/ctc-info.txt}, which comes from
+\texttt{prepare.rkt --mode main}:
 
@@
 
-\includegraphics[width=\columnwidth]{out/error-count-nocheck-row--te-density-diff.pdf}
-\includegraphics[width=\columnwidth]{out/error-count-nonstrict-row--te-density-diff.pdf}
-\includegraphics[width=\columnwidth]{out/error-count-strict-row--te-density-diff.pdf}
+Plot output from \texttt{code/error-count.rkt}
+on \texttt{out/error-density-ss-*.rktd}.
+(It also prints text output.):
+
+\includegraphics[width=0.8\columnwidth]{out/error-count-nocheck-row--te-density-diff.pdf}
+\includegraphics[width=0.8\columnwidth]{out/error-count-nonstrict-row--te-density-diff.pdf}
+\includegraphics[width=0.8\columnwidth]{out/error-count-strict-row--te-density-diff.pdf}
+
+\subsection*{Figure 7: Type Error Density}
+
+Shuffled text output from \texttt{code/error-count.rkt}
+on \texttt{out/error-density-ss-*.rktd}:
+
+\begin{verbatim}
+TE
+  nocheck & 48378 & (\pct{46.83}) & 9440 & [\pct{9.14}] & 45479 & [\pct{44.03}] \\
+FS
+  nocheck & 270763 & (\pct{36.99}) & 238696 & [\pct{32.61}] & 222588 & [\pct{30.41}] \\
+TE mod
+  nocheck & 20085 & (\pct{49.83}) & 906 & [\pct{2.25}] & 19317 & [\pct{47.92}] \\
+FS mod
+  nocheck & 259865 & (\pct{39.09}) & 175643 & [\pct{26.42}] & 229271 & [\pct{34.49}] \\
+TE
+  nonstrict & 19491 & (\pct{39.63}) & 9567 & [\pct{19.45}] & 20121 & [\pct{40.91}] \\
+FS
+  nonstrict & 35330 & (\pct{37.28}) & 29483 & [\pct{31.11}] & 29955 & [\pct{31.61}] \\
+TE mod
+  nonstrict & 13030 & (\pct{40.22}) & 5513 & [\pct{17.02}] & 13852 & [\pct{42.76}] \\
+FS mod
+  nonstrict & 33988 & (\pct{39.74}) & 20738 & [\pct{24.25}] & 30798 & [\pct{36.01}] \\
+TE
+  strict & 733 & (\pct{39.18}) & 368 & [\pct{19.67}] & 770 & [\pct{41.15}] \\
+FS
+  strict & 574 & (\pct{33.71}) & 561 & [\pct{32.94}] & 568 & [\pct{33.35}] \\
+TE mod
+  strict & 419 & (\pct{39.31}) & 230 & [\pct{21.58}] & 417 & [\pct{39.12}] \\
+FS mod
+  strict & 488 & (\pct{36.53}) & 344 & [\pct{25.75}] & 504 & [\pct{37.72}] \\
+\end{verbatim}
 

### 2023-12-08 programming23/paper.tex
artifact more more more

@@
     \end{tabular}
-    \begin{tabular}[t]{r@{~}r@{~}r}
+    \begin{tabular}[t]{r@{~~}r@{~}r}
       72,235,735 & {\FS{} errors} \\

### 2023-12-08 programming23/artifact/main.tex
artifact data

@@
 
+\usepackage{xcolor}
+\usepackage{colortbl}
 \usepackage{graphicx}
@@
 
+\newcommand{\code}[1]{\texttt{#1}}
+\newcommand{\pct}[1]{$#1\,\%$}
 \newcommand{\modefont}[1]{\texttt{#1}}
@@
 \newcommand{\mstrict}{\modefont{strict}}
+\newcommand{\gcell}[1]{\cellcolor{green!20}#1}
+\newcommand{\ycell}[1]{\cellcolor{yellow!18}#1}
+\newcommand{\ocell}[1]{\cellcolor{orange!29}#1}
+\newcommand{\rcell}[1]{\cellcolor{red!30}$\!\!$#1$\!\!$}
+\newcommand{\zerowidth}[1]{\makebox[0pt][l]{#1}}
+\let\midrule=\hrule
 
@@
   \textbf{Lines:}\\
-  \includegraphics[width=\columnwidth]{out/lines-distribution.pdf}
+  \includegraphics[width=0.4\columnwidth]{out/lines-distribution.pdf}
 \end{minipage}\begin{minipage}{0.5\columnwidth}
   \textbf{Edit Range:}\\
-  \includegraphics[width=\columnwidth]{out/editrange-distribution.pdf}
+  \includegraphics[width=0.4\columnwidth]{out/editrange-distribution.pdf}
 \end{minipage}
@@
 \emph{but} the final two numbers are incorrect (upgrade and downgrade) because they include
-module switches:
+data from module switches:
 
@@
 
-% \includeverbatim{out/downgrade-count.txt}
+Output:
 
-% TODO
+\verbatiminput{out/downgrade-count.txt}
 
@@
 
-TODO
+\begin{tabular}{lr@{~}r@{~}rr@{~}r@{~}rr@{~}r@{~}r}
+  & \zerowidth{\mnocheck{}} & & & \zerowidth{\mnonstrict{}} & & & \zerowidth{\mstrict{}} & & \\
+  & \rcell{Add} & \ycell{Keep} & \gcell{Drop} & \rcell{Add} & \ycell{Keep} & \gcell{Drop} & \rcell{Add} & \ycell{Keep} & \gcell{Drop} \\\hline
+  TypeMismatch & {0} & {0} & {0} & {103} & {51} & {80} & {13} & {6} & {18} \\
+  UnknownSymbol & {0} & {0} & {0} & {1992} & {438} & {1797} & {38} & {18} & {35} \\
+  UnknownProperty & {0} & {0} & {0} & {256} & {156} & {208} & {16} & {18} & {22} \\
+  NotATable & {0} & {0} & {0} & {4} & {1} & {5} & {2} & {0} & {1} \\
+  CannotExtendTable & {0} & {0} & {0} & {6} & {8} & {1} & {0} & {0} & {1} \\
+  OnlyTablesCanHaveMethods & {0} & {0} & {0} & {0} & {0} & {2} & {0} & {0} & {0} \\
+  DuplicateTypeDefinition & {0} & {0} & {0} & {1} & {0} & {0} & {0} & {0} & {0} \\
+  CountMismatch & {0} & {0} & {0} & {184} & {42} & {157} & {9} & {2} & {9} \\
+  FunctionDoesNotTakeSelf & {0} & {0} & {0} & {3} & {4} & {1} & {0} & {0} & {0} \\
+  OccursCheckFailed & {0} & {0} & {0} & {0} & {0} & {0} & {0} & {0} & {1} \\
+  UnknownRequire & {0} & {0} & {0} & {43} & {30} & {37} & {5} & {3} & {3} \\
+  IncorrectGenericParameterCount & {0} & {0} & {0} & {0} & {0} & {0} & {1} & {0} & {1} \\
+  SyntaxError & {0} & {0} & {8290} & {0} & {0} & {1149} & {0} & {0} & {29} \\
+  UnknownPropButFoundLikeProp & {0} & {0} & {0} & {20} & {17} & {13} & {0} & {0} & {0} \\
+  GenericError & {0} & {0} & {0} & {177} & {47} & {149} & {5} & {0} & {8} \\
+  CannotCallNonFunction & {0} & {0} & {0} & {12} & {2} & {13} & {1} & {0} & {1} \\
+  ExtraInformation & {0} & {0} & {0} & {46} & {6} & {33} & {1} & {0} & {1} \\
+  ModuleHasCyclicDependency & {0} & {0} & {0} & {9} & {5} & {8} & {0} & {0} & {0} \\
+  IllegalRequire & {0} & {0} & {0} & {7} & {1} & {11} & {0} & {0} & {0} \\
+  FunctionExitsWithoutReturning & {0} & {0} & {0} & {6} & {1} & {5} & {7} & {3} & {5} \\
+  CannotInferBinaryOperation & {0} & {0} & {0} & {0} & {1} & {0} & {5} & {6} & {6} \\
+  MissingProperties & {0} & {0} & {0} & {8} & {3} & {6} & {5} & {3} & {3} \\
+  OptionalValueAccess & {0} & {0} & {0} & {21} & {45} & {15} & {4} & {2} & {4} \\
+\end{tabular}
 
@@
 Output from \texttt{code/type-error-count.rkt}
-run on \texttt{code/te-editrange-ss-*.rktd}:
+run on \texttt{code/te-editrange-ss-*.rktd}.
 
-TODO
+This script produces output for all three modes: \mnocheck{}, \mnonstrict{}, and \mstrict{}.
+The paper combines \mnonstrict{} and \mstrict{} into one table.
+
+\subsubsection*{\mnocheck{}}
+
+unused code IllegalRequire
+unused code UnknownSymbol
+unused code TypeMismatch
+unused code NormalizationTooComplex
+unused code DynamicPropertyLookupOnClassesUnsafe
+unused code MissingUnionProperty
+unused code UnknownProperty
+unused code TypePackMismatch
+unused code DuplicateTypeDefinition
+unused code UnknownRequire
+unused code TypesAreUnrelated
+unused code MissingProperties
+unused code FunctionExitsWithoutReturning
+unused code CannotExtendTable
+unused code FunctionDoesNotTakeSelf
+unused code UnificationTooComplex
+unused code GenericError
+unused code NotATable
+unused code OnlyTablesCanHaveMethods
+unused code CodeTooComplex
+unused code InternalError
+unused code FunctionRequiresSelf
+unused code UnknownPropButFoundLikeProp
+unused code CannotInferBinaryOperation
+unused code IncorrectGenericParameterCount
+unused code OptionalValueAccess
+unused code DuplicateGenericParameter
+unused code CountMismatch
+unused code OccursCheckFailed
+unused code DeprecatedApiUsed
+unused code CannotCallNonFunction
+unused code ModuleHasCyclicDependency
+unused code SwappedGenericTypeParameter
+unused code ExtraInformation
+NEVER APPEAR: (IllegalRequire UnknownSymbol TypeMismatch NormalizationTooComplex DynamicPropertyLookupOnClassesUnsafe MissingUnionProperty UnknownProperty TypePackMismatch DuplicateTypeDefinition UnknownRequire TypesAreUnrelated MissingProperties FunctionExitsWithoutReturning CannotExtendTable FunctionDoesNotTakeSelf UnificationTooComplex GenericError NotATable OnlyTablesCanHaveMethods CodeTooComplex InternalError FunctionRequiresSelf UnknownPropButFoundLikeProp CannotInferBinaryOperation IncorrectGenericParameterCount OptionalValueAccess DuplicateGenericParameter CountMismatch OccursCheckFailed DeprecatedApiUsed CannotCallNonFunction ModuleHasCyclicDependency SwappedGenericTypeParameter ExtraInformation)
+
+\begin{tabular}{lr}
+  \code{SyntaxError} & \pct{100.00} \\
+\end{tabular}
+
+\subsubsection*{\mnonstrict{}}
+
+unused code NormalizationTooComplex
+unused code DynamicPropertyLookupOnClassesUnsafe
+unused code TypePackMismatch
+unused code UnificationTooComplex
+unused code CodeTooComplex
+unused code InternalError
+unused code FunctionRequiresSelf
+unused code IncorrectGenericParameterCount
+unused code DuplicateGenericParameter
+unused code OccursCheckFailed
+unused code DeprecatedApiUsed
+unused code SwappedGenericTypeParameter
+NEVER APPEAR: (NormalizationTooComplex DynamicPropertyLookupOnClassesUnsafe TypePackMismatch UnificationTooComplex CodeTooComplex InternalError FunctionRequiresSelf IncorrectGenericParameterCount DuplicateGenericParameter OccursCheckFailed DeprecatedApiUsed SwappedGenericTypeParameter)
+
+\begin{tabular}{lr}
+  \code{UnknownSymbol} & \pct{62.13} \\
+  \code{SyntaxError} & \pct{15.42} \\
+  \code{UnknownProperty} & \pct{8.28} \\
+  \code{UnknownRequire} & \pct{3.13} \\
+  \code{TypeMismatch} & \pct{2.44} \\
+  \code{CountMismatch} & \pct{2.26} \\
+  \code{OptionalValueAccess} & \pct{2.07} \\
+  \code{GenericError} & \pct{2.06} \\
+  \code{UnknownPropButFoundLikeProp} & \pct{0.44} \\
+  \code{CannotExtendTable} & \pct{0.42} \\
+  \code{ExtraInformation} & \pct{0.32} \\
+  \code{ModuleHasCyclicDependency} & \pct{0.23} \\
+  \code{IllegalRequire} & \pct{0.20} \\
+  \code{NotATable} & \pct{0.15} \\
+  \code{CannotCallNonFunction} & \pct{0.15} \\
+  \code{MissingProperties} & \pct{0.09} \\
+  \code{FunctionExitsWithoutReturning} & \pct{0.07} \\
+  \code{FunctionDoesNotTakeSelf} & \pct{0.07} \\
+  \code{MissingUnionProperty} & \pct{0.02} \\
+  \code{CannotInferBinaryOperation} & \pct{0.02} \\
+  \code{OnlyTablesCanHaveMethods} & \pct{0.01} \\
+  \code{DuplicateTypeDefinition} & \pct{0.00} \\
+  \code{TypesAreUnrelated} & \pct{0.00} \\
+\end{tabular}
+
+\subsubsection*{\mstrict{}}
+
+unused code NormalizationTooComplex
+unused code DynamicPropertyLookupOnClassesUnsafe
+unused code MissingUnionProperty
+unused code TypePackMismatch
+unused code DuplicateTypeDefinition
+unused code FunctionDoesNotTakeSelf
+unused code UnificationTooComplex
+unused code OnlyTablesCanHaveMethods
+unused code CodeTooComplex
+unused code InternalError
+unused code FunctionRequiresSelf
+unused code DuplicateGenericParameter
+unused code DeprecatedApiUsed
+unused code SwappedGenericTypeParameter
+NEVER APPEAR: (NormalizationTooComplex DynamicPropertyLookupOnClassesUnsafe MissingUnionProperty TypePackMismatch DuplicateTypeDefinition FunctionDoesNotTakeSelf UnificationTooComplex OnlyTablesCanHaveMethods CodeTooComplex InternalError FunctionRequiresSelf DuplicateGenericParameter DeprecatedApiUsed SwappedGenericTypeParameter)
+
+\begin{tabular}{lr}
+  \code{UnknownSymbol} & \pct{23.97} \\
+  \code{TypeMismatch} & \pct{20.46} \\
+  \code{UnknownProperty} & \pct{18.88} \\
+  \code{SyntaxError} & \pct{9.31} \\
+  \code{CannotInferBinaryOperation} & \pct{6.94} \\
+  \code{MissingProperties} & \pct{4.04} \\
+  \code{FunctionExitsWithoutReturning} & \pct{2.99} \\
+  \code{OptionalValueAccess} & \pct{2.99} \\
+  \code{CountMismatch} & \pct{2.99} \\
+  \code{UnknownRequire} & \pct{2.37} \\
+  \code{GenericError} & \pct{2.28} \\
+  \code{NotATable} & \pct{1.32} \\
+  \code{ExtraInformation} & \pct{0.35} \\
+  \code{UnknownPropButFoundLikeProp} & \pct{0.26} \\
+  \code{IllegalRequire} & \pct{0.18} \\
+  \code{IncorrectGenericParameterCount} & \pct{0.18} \\
+  \code{CannotCallNonFunction} & \pct{0.18} \\
+  \code{TypesAreUnrelated} & \pct{0.09} \\
+  \code{CannotExtendTable} & \pct{0.09} \\
+  \code{OccursCheckFailed} & \pct{0.09} \\
+  \code{ModuleHasCyclicDependency} & \pct{0.09} \\
+\end{tabular}
 
@@
 \includegraphics[width=0.8\columnwidth]{out/error-count-nocheck-row--te-density-diff.pdf}
+
 \includegraphics[width=0.8\columnwidth]{out/error-count-nonstrict-row--te-density-diff.pdf}
+
 \includegraphics[width=0.8\columnwidth]{out/error-count-strict-row--te-density-diff.pdf}
@@
 
-Shuffled text output from \texttt{code/error-count.rkt}
+From reordered text output from \texttt{code/error-count.rkt}
 on \texttt{out/error-density-ss-*.rktd}:

### 2023-12-09 programming23/artifact/main.tex
dockerfile

@@
 
+Current errors in edit range = (14917 + 16007) = 30924
+
 %% ---
@@
 \end{verbatim}
+
+There are 642 = (347598 - 346956) multi-mode sessions.
 
@@
     Type errors & Background errors \\
-    \labelbars{\includegraphics[width=0.20\columnwidth,valign=M]{out/error-by-mode-te.pdf}}
-    & \labelbars{\includegraphics[width=0.20\columnwidth,valign=M]{out/error-by-mode-fs.pdf}}
+    \labelbars{\includegraphics[width=0.20\columnwidth,valign=M]{out/error-by-mode-te.png}}
+    & \labelbars{\includegraphics[width=0.20\columnwidth,valign=M]{out/error-by-mode-fs.png}}
   \end{tabular}
@@
 Output from \texttt{code/type-error-survival.rkt}
-run on \texttt{code/type-error-survival-ss-*.rktd}:
+run on \texttt{code/type-error-survival-ss-*.rktd}
+and then sorted alphabetically:
 
@@
   & \rcell{Add} & \ycell{Keep} & \gcell{Drop} & \rcell{Add} & \ycell{Keep} & \gcell{Drop} & \rcell{Add} & \ycell{Keep} & \gcell{Drop} \\\hline
+  CannotCallNonFunction & {0} & {0} & {0} & {12} & {2} & {13} & {1} & {0} & {1} \\
+  CannotExtendTable & {0} & {0} & {0} & {6} & {8} & {1} & {0} & {0} & {1} \\
+  CannotInferBinaryOperation & {0} & {0} & {0} & {0} & {1} & {0} & {5} & {6} & {6} \\
+  CountMismatch & {0} & {0} & {0} & {184} & {42} & {157} & {9} & {2} & {9} \\
+  DuplicateTypeDefinition & {0} & {0} & {0} & {1} & {0} & {0} & {0} & {0} & {0} \\
+  ExtraInformation & {0} & {0} & {0} & {46} & {6} & {33} & {1} & {0} & {1} \\
+  FunctionDoesNotTakeSelf & {0} & {0} & {0} & {3} & {4} & {1} & {0} & {0} & {0} \\
+  FunctionExitsWithoutReturning & {0} & {0} & {0} & {6} & {1} & {5} & {7} & {3} & {5} \\
+  GenericError & {0} & {0} & {0} & {177} & {47} & {149} & {5} & {0} & {8} \\
+  IllegalRequire & {0} & {0} & {0} & {7} & {1} & {11} & {0} & {0} & {0} \\
+  IncorrectGenericParameterCount & {0} & {0} & {0} & {0} & {0} & {0} & {1} & {0} & {1} \\
+  MissingProperties & {0} & {0} & {0} & {8} & {3} & {6} & {5} & {3} & {3} \\
+  ModuleHasCyclicDependency & {0} & {0} & {0} & {9} & {5} & {8} & {0} & {0} & {0} \\
+  NotATable & {0} & {0} & {0} & {4} & {1} & {5} & {2} & {0} & {1} \\
+  OccursCheckFailed & {0} & {0} & {0} & {0} & {0} & {0} & {0} & {0} & {1} \\
+  OnlyTablesCanHaveMethods & {0} & {0} & {0} & {0} & {0} & {2} & {0} & {0} & {0} \\
+  OptionalValueAccess & {0} & {0} & {0} & {21} & {45} & {15} & {4} & {2} & {4} \\
+  SyntaxError & {0} & {0} & {8290} & {0} & {0} & {1149} & {0} & {0} & {29} \\
   TypeMismatch & {0} & {0} & {0} & {103} & {51} & {80} & {13} & {6} & {18} \\
+  UnknownPropButFoundLikeProp & {0} & {0} & {0} & {20} & {17} & {13} & {0} & {0} & {0} \\
+  UnknownProperty & {0} & {0} & {0} & {256} & {156} & {208} & {16} & {18} & {22} \\
+  UnknownRequire & {0} & {0} & {0} & {43} & {30} & {37} & {5} & {3} & {3} \\
   UnknownSymbol & {0} & {0} & {0} & {1992} & {438} & {1797} & {38} & {18} & {35} \\
-  UnknownProperty & {0} & {0} & {0} & {256} & {156} & {208} & {16} & {18} & {22} \\
-  NotATable & {0} & {0} & {0} & {4} & {1} & {5} & {2} & {0} & {1} \\
-  CannotExtendTable & {0} & {0} & {0} & {6} & {8} & {1} & {0} & {0} & {1} \\
-  OnlyTablesCanHaveMethods & {0} & {0} & {0} & {0} & {0} & {2} & {0} & {0} & {0} \\
-  DuplicateTypeDefinition & {0} & {0} & {0} & {1} & {0} & {0} & {0} & {0} & {0} \\
-  CountMismatch & {0} & {0} & {0} & {184} & {42} & {157} & {9} & {2} & {9} \\
-  FunctionDoesNotTakeSelf & {0} & {0} & {0} & {3} & {4} & {1} & {0} & {0} & {0} \\
-  OccursCheckFailed & {0} & {0} & {0} & {0} & {0} & {0} & {0} & {0} & {1} \\
-  UnknownRequire & {0} & {0} & {0} & {43} & {30} & {37} & {5} & {3} & {3} \\
-  IncorrectGenericParameterCount & {0} & {0} & {0} & {0} & {0} & {0} & {1} & {0} & {1} \\
-  SyntaxError & {0} & {0} & {8290} & {0} & {0} & {1149} & {0} & {0} & {29} \\
-  UnknownPropButFoundLikeProp & {0} & {0} & {0} & {20} & {17} & {13} & {0} & {0} & {0} \\
-  GenericError & {0} & {0} & {0} & {177} & {47} & {149} & {5} & {0} & {8} \\
-  CannotCallNonFunction & {0} & {0} & {0} & {12} & {2} & {13} & {1} & {0} & {1} \\
-  ExtraInformation & {0} & {0} & {0} & {46} & {6} & {33} & {1} & {0} & {1} \\
-  ModuleHasCyclicDependency & {0} & {0} & {0} & {9} & {5} & {8} & {0} & {0} & {0} \\
-  IllegalRequire & {0} & {0} & {0} & {7} & {1} & {11} & {0} & {0} & {0} \\
-  FunctionExitsWithoutReturning & {0} & {0} & {0} & {6} & {1} & {5} & {7} & {3} & {5} \\
-  CannotInferBinaryOperation & {0} & {0} & {0} & {0} & {1} & {0} & {5} & {6} & {6} \\
-  MissingProperties & {0} & {0} & {0} & {8} & {3} & {6} & {5} & {3} & {3} \\
-  OptionalValueAccess & {0} & {0} & {0} & {21} & {45} & {15} & {4} & {2} & {4} \\
 \end{tabular}
@@
 
-\includegraphics[width=0.8\columnwidth]{out/error-count-nocheck-row--te-density-diff.pdf}
+\includegraphics[width=0.8\columnwidth]{out/error-count-nocheck-row--te-density-diff.png}
 
-\includegraphics[width=0.8\columnwidth]{out/error-count-nonstrict-row--te-density-diff.pdf}
+\includegraphics[width=0.8\columnwidth]{out/error-count-nonstrict-row--te-density-diff.png}
 
-\includegraphics[width=0.8\columnwidth]{out/error-count-strict-row--te-density-diff.pdf}
+\includegraphics[width=0.8\columnwidth]{out/error-count-strict-row--te-density-diff.png}
 
@@
 From reordered text output from \texttt{code/error-count.rkt}
-on \texttt{out/error-density-ss-*.rktd}:
+on \texttt{out/error-density-ss-*.rktd}.
+Here is the raw, unsorted output:
 

### 2023-12-09 programming23/artifact/runall.sh
artifact nearly done

@@ -1,0 +1,20 @@
+echo "deleting everything in out/ ..."
+echo "=== before"
+ls out/
+rm out/*.*
+echo "=== after"
+
+PLTSTDERR="error info@luau" racket code/prepare.rkt
+PLTSTDERR="error info@luau" racket code/prepare.rkt --mode modswitch
+PLTSTDERR="error info@luau" racket code/prepare.rkt --mode divide-sessions
+PLTSTDERR="error info@luau" racket code/prepare.rkt --mode session-query
+PLTSTDERR="error info@luau" racket code/prepare.rkt --mode aggregate-te
+
+racket code/row-distribution.rkt
+racket code/sdupdate.rkt
+racket code/downgrade-count.rkt
+racket code/error-by-mode.rkt
+racket code/type-error-survival.rkt
+racket code/type-error-count.rkt
+racket code/error-count.rkt
+

### 2023-12-09 programming23/paper.tex
acks
artifact checkpoint
error count ignore module switch data ... needs work to incorporate

@@
 % TODO BEFORE SUBMISSION
+% - units on all plots
 
@@
 The type error telemetry for a Roblox Studio session is gathered as a series of
-records, all with the same per-session \panon{} session identifier.
+records, all with the same  \panon{} session identifier.
 This allows us to correlate telemetry across a single
@@
   \item
-    \uid{}: a \panon{} identifier for the current session.
+    \uid{}: a \panon{} identifier (16-digit random number) for the current session.
   \item
@@
 Most sessions last a few dozen seconds~(\cref{t:session-size}), so the axis
-ends at 150 seconds.
-%% TODO explain why 150 ... shows some tail but not whole thing
+ends at 150 seconds to show some but not all of the long tail.
 The y axis shows changes to errors over time; a positive number means errors
@@
 on $\pm{}50$ errors because most of the mass is in that region.
-The points are color-coded to give each session the same color, but this is not
-very useful because there are many more sessions than colors.
+The points are color-coded by their session.
+There are many colors througout the plot, which shows that the data comes
+from many different sessions and is not dominated by a few high-traffic sessions.
 
@@
 
-  %% \item
-  %%   TODO plot FS with a much bigger range ... tells a different story about buildup?
+  \item
+    When a color appears above the x axis, the same color usually appears below it as well.
+    For example, \mnocheck{} has a cluster of purple points near the 20 second mark
+    and $\pm 30$ density, which suggests that one session repeatedly added and removed errors.
+    The next subsection conducts a thorough analysis.
 
@@
 
+  %% with ALL data, modswitch + no switch
   % TE: nocheck & 48378 & [\pct{46.83}] & 9440 & [\pct{9.14}] & 45479 & [\pct{44.03}] \\
@@
   % FS mod: strict & 488 & [\pct{36.53}] & 344 & [\pct{25.75}] & 504 & [\pct{37.72}] \\
+
+  %% with NON-moduleswitch data
+% TE
+%   nocheck & 961 & (\pct{38.50}) & 694 & [\pct{27.80}] & 841 & [\pct{33.69}] \\
+% FS
+%   nocheck & 156422 & (\pct{37.19}) & 153965 & [\pct{36.61}] & 110194 & [\pct{26.20}] \\
+% TE mod
+%   nocheck & 0 & (\pct{na}) & 0 & [\pct{na}] & 0 & [\pct{na}] \\
+% FS mod
+%   nocheck & 153913 & (\pct{36.96}) & 152735 & [\pct{36.68}] & 109807 & [\pct{26.37}] \\
+% TE
+%   nonstrict & 8623 & (\pct{37.67}) & 5263 & [\pct{22.99}] & 9004 & [\pct{39.34}] \\
+% FS
+%   nonstrict & 18927 & (\pct{37.41}) & 17791 & [\pct{35.17}] & 13872 & [\pct{27.42}] \\
+% TE mod
+%   nonstrict & 7185 & (\pct{36.61}) & 4768 & [\pct{24.30}] & 7672 & [\pct{39.09}] \\
+% FS mod
+%   nonstrict & 18573 & (\pct{37.09}) & 17604 & [\pct{35.15}] & 13905 & [\pct{27.76}] \\
+% TE
+%   strict & 442 & (\pct{38.27}) & 244 & [\pct{21.13}] & 469 & [\pct{40.61}] \\
+% FS
+%   strict & 296 & (\pct{31.66}) & 360 & [\pct{38.50}] & 279 & [\pct{29.84}] \\
+% TE mod
+%   strict & 251 & (\pct{35.35}) & 204 & [\pct{28.73}] & 255 & [\pct{35.92}] \\
+% FS mod
+%   strict & 272 & (\pct{32.04}) & 317 & [\pct{37.34}] & 260 & [\pct{30.62}] \\
+
   \begin{tabular}{l@{~~}lr@{}rr@{}rr@{}r}
@@
 %% -----------------------------------------------------------------------------
-% \acks
-
-% Thanks to Benjamin Chung for several helpful discussions about data analysis
-% and effective plotting.
-% Greenman was supported by
-% NSF grant \href{https://nsf.gov/awardsearch/showAward?AWD_ID=2030859&HistoricalAwards=false}{2030859}
-% to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
+\acks
+
+Thanks to Benjamin Chung for several helpful discussions about data analysis
+and effective plotting.
+Greenman was supported by
+NSF grant \href{https://nsf.gov/awardsearch/showAward?AWD_ID=2030859&HistoricalAwards=false}{2030859}
+to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
 

### 2023-12-12 programming23/paper.tex
ctc: developer not necessarily to blame
revisions

@@
 \newcommand{\pct}[1]{\SI{#1}{\percent}}
+\newcommand{\nanpct}{n/a}
 \newcommand{\percentile}[1]{{#1}th~percentile}
@@
 quality can rely on \mstrict{} mode.
-(No matter the mode, a {\FS{}} type analysis still drives the \IDE{} tools.)
+Also, no matter the mode, a {\FS{}} type analysis runs to guide \IDE{} tools~(\cref{s:context}).
 Luau's types aim to support untyped designs, in the spirit of migratory
@@
     An available dataset of over 1.5 million telemetry records
-    and scripts to analyze them.
-    \emph{TODO We intend to submit these in an artifact.}
+    and scripts to analyze them: \url{https://zenodo.org/doi/10.5281/zenodo.10275213}
 
@@
     \code{IncorrectGenericParamCount} & Arity mismatch for a generic type. \\
-    \code{CodeTooComplex} & Type analysis failed, cannot understand the code. \\
+    \code{CodeTooComplex} & Type analysis failed~(\cref{s:code-too-complex}). \\
     %% \code{UnificationTooComplex} & Type analysis failed, unification solver hit a limit \\
@@
   \item
-    Telemetry \emph{must transmit a small amount of data}---not too many records
-    with not too much information inside---to avoid overloading the servers.
+    Telemetry \emph{must transmit a small amount of data} to avoid overloading the servers.
+    This constrains both the number of records the system can transmit and the size of each record.
 
 \end{enumerate}
-Cox~\cite{transparent-telemetry} discusses in detail the privacy implications
-and tradeoffs in telemetry.
+See Cox~\cite{transparent-telemetry} for further discussion of
+the privacy implications and tradeoffs in telemetry.
 
@@
   \item
-    Lifecycle events including as save, run, quit, and publish.\footnote{Publish events
+    Lifecycle events including save, run, quit, and publish.
+    {(Publish events
     cannot be reported at any rate because they are public, and could be used
-    to match a telemetry session to a creator.}
+    to match a telemetry session to a creator.)}
   \item
@@
 Our type error telemetry has to meet all of the above constraints.
+This means that it cannot send event-based telemetry on IDE lifecycle events
+(due to Studio's architecture) and it cannot send full results from every type
+analysis because that would overwhelm the data-collection server.
 
@@
   \item
-    \uid{}: a \panon{} identifier (16-digit random number) for the current session.
+    \uid{}: a \panon{} identifier (15-digit random number) for the current session.
   \item
@@
   \item
-    Overall Counts: summarizes type errors and \FS{} errors during the
-    last two invocations of type analysis, and additionally reports the number
+    Overall Counts: summary of type errors and \FS{} errors during the
+    last two invocations of type analysis, and the number
     of times type analysis hit an internal limit on problem
@@
   \item
-    Edit Range Counts: these list specific errors that arose in the last two
+    Edit Range Counts: list of specific errors that arose in the last two
     type analyses and overlap with the current edit range.
@@
 of the file.
+We acknowledge this and other threats in~\cref{s:threats}.
 The upshot of this strategy is that it reduces the size and complexity of
 telemetry records because there is only one interval for errors to overlap with.
-
-This telemetry design is admittedly coarse-grained.
-For instance, it does not distinguish edits that ignore a type
-error from edits that remove one error while introducing another.
-We acknowledge this and other threats in~\cref{s:threats}.
-The main advantages of this telemetry are the small size of each record
-and the complete lack of private information.
-Furthermore, despite the limitations, this telemetry supports a variety of
-inferences that we showcase in the next section.
 
@@
 1,000.
-{We conjecture the very low bars are due to outage on
-  the logging server, but need to investigate further.}
-The tall bars are FILL TODO.
-%% \FILL{} investigate one tower, what's happening?? all one session?
-Telemetry ended in early April; the long tail is due to developers
-updating Roblox Studio later.
-% BEN: I see no basis for this; nor is this at all interesting.
-%% ben: basis = it's where the peaks are
-% We do not know which timezone the records originated in, but since the counts
-% tend to peak midday California time it seems that the majority
-% of Roblox developers are following a Western Hemisphere schedule.
+We conjecture the very low bars are due to outage on
+the logging server and that the very tall bars are due to
+user scripts that generated a huge number of keystrokes.
+%% TODO investigate one tower, what's happening?? all one session?
+Telemetry ended in early April; however, there are short bars in the mid-April
+range from developers who waited a few days to update Roblox Studio.
+
 The regular pattern of the peaks suggests that many creators
@@
 
-There is a huge amount of variation across projects.
+There is a huge amount of variation across sessions.
 The largest ones have over 50,000 lines of code
@@
 mode the module declares.
-(Furthermore, \FS{} checks run silently, so creators have no
-awareness of or incentive to fix these errors.)
+Furthermore, \FS{} checks run silently, so creators have no
+awareness of or incentive to fix these errors.
 Approximately half the analysis errors occurred in the current module,
-and a small fraction overlapped with the current edit range~(\pct{5} for
+and a small fraction of these overlapped with the current edit range~(\pct{5} for
 type errors, \pct{1} for \FS{}).
-
-The large fraction of errors in the current module is encouraging
-for two reasons.
+The other errors point to different modules in the codebase.
+
+The fraction of errors in the current module is relatively high because projects
+contain many modules, all of which might contain latent type errors.
+This high fraction is encouraging for two reasons.
 First, it suggests that errors appear locally, as the result of edits to nearby
@@
 module.
-Second, it suggests that creators typically fix errors before switching
+Second, it suggests that creators fix errors before switching
 to another module.
@@
   \item
-    On one hand, it suggests that edits successfully remove errors.
+    On one hand, it may be that edits successfully remove errors.
     This is true, however, only for type errors and not for \FS{} errors.
@@
 type errors (stddev 7, median 0).
+This data excludes module switch records and considers only keystroke events.
 The worst-case increase was quite high at 57 type errors, but exceptional.
@@
 Furthermore, type analysis does not seem to reduce \FS{}
-errors; this is surprising and~\cref{s:tefs-compass} has more to say about it.
+errors.
+This is surprising;~\cref{s:tefs-compass} explores further.
 
@@
 Indeed, most of the fields in a telemetry record are dedicated to this
-topic~(\cref{f:tdata}), and are directed to RQ2 on type errors
-(\cref{s:introduction}).
+topic~(\cref{f:tdata}), and are directed to RQ2 on type errors .
 
@@
 If a creator removes one error but then
-introduces another on a different line, this typically manifests as
-a \keepsym{} rather than one \addsym{} and one \dropsym{}.
+introduces another on a different line, this can manifest as
+a keep (\keepsym{}) rather than one add (\addsym{)} and one drop (\dropsym{}).
 
@@
   \item
-    \code{SyntaxError} and the related \code{UnknownSymbol} and
+    \code{SyntaxError} and the related errors \code{UnknownSymbol} and
     \code{UnknownProperty} are the most common.
     If the edit range errors are indicative of type errors at large,
-    then most errors in the data are likely due to typos.
-    These errors also survive (\keepsym{}) the most often.
-
-    Our telemetry strategy likely inflates
-    the number of typos because it cannot determine when a creator is
-    mid-edit.
+    then most errors in the data are likely due to typos during edits.
+    These errors also survive (\keepsym{}) the more than any other.
+
+    Curiously, the number of syntax errors never decreases or stays the same.
+    This may be because the easy fix is to press undo, and telemetry cannot
+    see undo events.
+
+    %% TODO awkward
+    Our telemetry strategy likely inflates the number of typos because it cannot
+    determine when a creator is mid-edit.
     For example, a creator who writes a method call
@@
     \code{UnknownSymbol} error for \code{buck} and
-    \code{UnknownProperty} for \code{bucket.c}, which may end up sampled.
+    \code{UnknownProperty} for \code{bucket.c}, which may end up getting sampled.
     \Cref{s:conclusion} suggests ways
@@
   \item
-    Syntax errors, in particular \code{UnknownSymbol} (for an unbound variable),
+    Typo-level errors, in particular \code{UnknownSymbol} (for an unbound variable),
     are much more common in \mnonstrict{} than \mstrict{}.
@@
     Despite being less strict, \mnonstrict{} reported four errors that
-    never appeared in the \mstrict{} records:
-    \code{DuplicateTypeDefinition}, \code{OnlyTablesCanHaveMethods},
-    \code{MissingUnionProperty}, and \code{FunctionDoesNotTakeSelf}.
+    never appeared in \mstrict{} records:
+    \code{DuplicateTypeDefinition}, \code{FunctionDoesNotTakeSelf},
+    \code{MissingUnionProperty}, and \code{OnlyTablesCanHaveMethods}.
     We attribute this to the much larger number of \mnonstrict{} records in the
@@
 has internal limits that restrict the problems it will attempt to solve.
-Hitting a limit triggers one of the following errors:
+Hitting a limit triggers one of the following errors along with a message
+that the analysis failed to understand the code:
 \code{CodeTooComplex},
@@
   \mnocheck{}
-  \includegraphics[width=\columnwidth]{img/error-count-nocheck-row--te-density-diff.pdf}
+  \includegraphics[width=\columnwidth]{img/error-count-nocheck-row--te-density-diff-no-switch.pdf}
 
   \mnonstrict{}
-  \includegraphics[width=\columnwidth]{img/error-count-nonstrict-row--te-density-diff.pdf}
+  \includegraphics[width=\columnwidth]{img/error-count-nonstrict-row--te-density-diff-no-switch.pdf}
 
   \mstrict{}
-  \includegraphics[width=\columnwidth]{img/error-count-strict-row--te-density-diff.pdf}
+  \includegraphics[width=\columnwidth]{img/error-count-strict-row--te-density-diff-no-switch.pdf}
   \caption{Changes in type error density ($\frac{\mbox{\#errors}}{\mbox{\#lines}}$) over time (seconds).}
@@
 \Cref{f:error-density} plots changes to error density for every session in the dataset.
+The data excludes module-switch telemetry records to keep the focus on edits.
 The x axis counts time in seconds from the start of the session.
@@
 While there are a few errors that go outside the plot bounds
-(max \mnocheck{}: 101, max \mnonstrict{}: 3577, max \mstrict{}: 87), we focus
+(max \mnocheck{}: 51, max \mnonstrict{}: 3586, max \mstrict{}: 87), we focus
 on $\pm{}50$ errors because most of the mass is in that region.
@@
 There are many colors througout the plot, which shows that the data comes
-from many different sessions and is not dominated by a few high-traffic sessions.
+from many different sessions rather than from a few high-traffic sessions.
 
@@
   \item
+    Only a small fraction of \mnocheck{} records have nonzero type error density.
+    This is in line with~\cref{t:type-error-survival}.
+    Module switches in \mnocheck{} have nonzero density more often, but still
+    rarely (\pct{3}).
+    %% 34372 / 1337552 = 100 * 0.025664048949124968
+
+  \item
     After the 40-second mark, changes to type error density become much
@@
     These long-running sessions are not introducing errors.
-    %% TODO check #mod switch in long sessions
-
-  \item
-    \mnonstrict{} has the biggest fluctuations in errors, whereas \mstrict{}
-    has the smallest.
+
+  \item
+    \mnonstrict{} has a bigger fluctuations in errors than \mstrict{}.
     One explanation is that some \mnonstrict{} sessions ignore type analysis
@@
     sessions pay close attention to it.
-
-  \item
-    When a color appears above the x axis, the same color usually appears below it as well.
-    For example, \mnocheck{} has a cluster of purple points near the 20 second mark
-    and $\pm 30$ density, which suggests that one session repeatedly added and removed errors.
-    The next subsection conducts a thorough analysis.
 
@@
   %% with NON-moduleswitch data
-% TE
-%   nocheck & 961 & (\pct{38.50}) & 694 & [\pct{27.80}] & 841 & [\pct{33.69}] \\
-% FS
-%   nocheck & 156422 & (\pct{37.19}) & 153965 & [\pct{36.61}] & 110194 & [\pct{26.20}] \\
-% TE mod
-%   nocheck & 0 & (\pct{na}) & 0 & [\pct{na}] & 0 & [\pct{na}] \\
-% FS mod
-%   nocheck & 153913 & (\pct{36.96}) & 152735 & [\pct{36.68}] & 109807 & [\pct{26.37}] \\
-% TE
-%   nonstrict & 8623 & (\pct{37.67}) & 5263 & [\pct{22.99}] & 9004 & [\pct{39.34}] \\
-% FS
-%   nonstrict & 18927 & (\pct{37.41}) & 17791 & [\pct{35.17}] & 13872 & [\pct{27.42}] \\
-% TE mod
-%   nonstrict & 7185 & (\pct{36.61}) & 4768 & [\pct{24.30}] & 7672 & [\pct{39.09}] \\
-% FS mod
-%   nonstrict & 18573 & (\pct{37.09}) & 17604 & [\pct{35.15}] & 13905 & [\pct{27.76}] \\
-% TE
-%   strict & 442 & (\pct{38.27}) & 244 & [\pct{21.13}] & 469 & [\pct{40.61}] \\
-% FS
-%   strict & 296 & (\pct{31.66}) & 360 & [\pct{38.50}] & 279 & [\pct{29.84}] \\
-% TE mod
-%   strict & 251 & (\pct{35.35}) & 204 & [\pct{28.73}] & 255 & [\pct{35.92}] \\
-% FS mod
-%   strict & 272 & (\pct{32.04}) & 317 & [\pct{37.34}] & 260 & [\pct{30.62}] \\
+  % TE nocheck & 961 & (\pct{38.50}) & 694 & [\pct{27.80}] & 841 & [\pct{33.69}] \\
+  % TE nonstrict & 8623 & (\pct{37.67}) & 5263 & [\pct{22.99}] & 9004 & [\pct{39.34}] \\
+  % TE strict & 442 & (\pct{38.27}) & 244 & [\pct{21.13}] & 469 & [\pct{40.61}] \\
+
+  % TE mod nocheck & 0 & (\pct{na}) & 0 & [\pct{na}] & 0 & [\pct{na}] \\
+  % TE mod nonstrict & 7185 & (\pct{36.61}) & 4768 & [\pct{24.30}] & 7672 & [\pct{39.09}] \\
+  % TE mod strict & 251 & (\pct{35.35}) & 204 & [\pct{28.73}] & 255 & [\pct{35.92}] \\
+
+  % FS nocheck & 156422 & (\pct{37.19}) & 153965 & [\pct{36.61}] & 110194 & [\pct{26.20}] \\
+  % FS nonstrict & 18927 & (\pct{37.41}) & 17791 & [\pct{35.17}] & 13872 & [\pct{27.42}] \\
+  % FS strict & 296 & (\pct{31.66}) & 360 & [\pct{38.50}] & 279 & [\pct{29.84}] \\
+
+  % FS mod nocheck & 153913 & (\pct{36.96}) & 152735 & [\pct{36.68}] & 109807 & [\pct{26.37}] \\
+  % FS mod nonstrict & 18573 & (\pct{37.09}) & 17604 & [\pct{35.15}] & 13905 & [\pct{27.76}] \\
+  % FS mod strict & 272 & (\pct{32.04}) & 317 & [\pct{37.34}] & 260 & [\pct{30.62}] \\
 
@@
     & & \zerowidth{\rbox{\addsym{}}} & & \zerowidth{\ybox{\keepsym{}}} & & \zerowidth{\gbox{\dropsym{}}} \\\midrule
-    type error  & \mnocheck{}   & 20085 & [\pct{49.83}] & 906 & [\pct{2.25}] & 19317 & [\pct{47.92}] \\
-    (in module) & \mnonstrict{} & 13030 & [\pct{40.22}] & 5513 & [\pct{17.02}] & 13852 & [\pct{42.76}] \\
-                & \mstrict{}    & 419 & [\pct{39.31}] & 230 & [\pct{21.58}] & 417 & [\pct{39.12}] \\
+    type error  & \mnocheck{}   &      0 & [\nanpct{}  ] &      0 & [\nanpct{}]   &      0 & [\nanpct{}] \\
+    (in module) & \mnonstrict{} &   7185 & [\pct{36.61}] &   4786 & [\pct{24.30}] &   7672 & [\pct{39.09}] \\
+                & \mstrict{}    &    251 & [\pct{35.35}] &    204 & [\pct{28.73}] &    255 & [\pct{35.92}] \\
     \\[-2ex]
-    \FS         & \mnocheck{}   & 259865 & [\pct{39.09}] & 175643 & [\pct{26.42}] & 229271 & [\pct{34.49}] \\
-    error       & \mnonstrict{} & 33988 & [\pct{39.74}] & 20738 & [\pct{24.25}] & 30798 & [\pct{36.01}] \\
-    (in module) & \mstrict{}    & 488 & [\pct{36.53}] & 344 & [\pct{25.75}] & 504 & [\pct{37.72}] \\
+    \FS         & \mnocheck{}   & 153913 & [\pct{36.96}] & 152735 & [\pct{36.68}] & 109807 & [\pct{26.37}] \\
+    error       & \mnonstrict{} &  18573 & [\pct{37.09}] &  17604 & [\pct{35.15}] &  13905 & [\pct{27.76}] \\
+    (in module) & \mstrict{}    &    272 & [\pct{32.04}] &    317 & [\pct{37.34}] &    260 & [\pct{30.62}] \\
   \end{tabular}
@@
   \item
-    The type error increases and decreases are remarkably balanced: across all
-    three modes, they are at most \pct{2} apart.
-    While some of this balance is due to module-switch records
-    (if the two modules have different error counts, there will be a change),
-    most of the data is from keystrokes; thus, it appears that sessions tend to
-    fix errors highlighted by type analysis.
+    None of our \mnocheck{} samples captured a change in syntax errors in the
+    current module.
+    The nonzero changes in density above~(1802 total, shown in \cref{f:error-density}) all affected
+    other modules.
+    %% TODO doublecheck really zero??!
+    %% QALAN what would this mean ... how can one module change stx errors in another??
+    %% 961 + 841 = 1802
+
+  \item
+    The type error increases and decreases are evenly balanced in
+    \mnonstrict{} and \mstrict{} modes ($\pm\pct{3}$).
+    Since this data is from keystrokes rather than module switches,
+    it appears that sessions tend to fix errors highlighted by type analysis.
 
@@
     Errors may be more likely to persist in this mode.
-    %% TODO check if actually zero
-
-  \item
-    In \FS{} mode, the percent of ``maintain'' records is high across
-    the board and the increases and decreases are lower.
+    %% NOTE these "=" are all maintaining a nonzero number of errors
+
+  \item
+    In \FS{} mode, the percent of ``maintain'' records is higher across
+    the board and roughly equal to the percents of increases and decreases
+    (about one-third each).
     No matter the mode, reducing type errors is not guaranteed to reduce
@@
 increase the number of \mstrict{} errors without increasing \FS{} errors.
+%% 3% = 35% - 32%
 Further inspection reveals that the situation is a bit worse, as
 some records increase \mstrict{} errors without changing the \FS{} errors:
-    \pct{22} of \mstrict{} records increase (\addsym{}) the number of type errors
+\pct{16} of all \mstrict{} records increase (\addsym{}) the number of type errors
 but maintain (\keepsym{}) the number of \FS{} errors.
+%% 16% = from code/compass.rkt
 
@@
 enough to be used on-the-fly in an IDE.
-But, few sessions use type analysis directly and the data presents little
-evidence that doing so would improve creators' quality of life.
+But, few sessions use type analysis directly.
 We conjecture that Luau needs further tailoring to match
@@
         or from \mnonstrict{} to \mnocheck{} (\pct{25}).
+        Then again, sessions rarely turn analysis on as well:
+        only 176 ($<\pct{0.05}$) switch to a stronger mode.
         %% FILL more insight if we match ups and downs together??
@@
     regardless of analysis mode~(\cref{f:error-changes}).
-    A randomly-chosen edit has a 1/3 chance of increasing, decreasing, or
+    A randomly-chosen edit has a roughly 1/3 chance of increasing, decreasing, or
     maintaining the number of \FS{} errors.
@@
 of these actions.
-We have no access to other events, e.g., run locally~(\cref{s:tel-limitations});
+We have no access to other events, e.g., local runs~(\cref{s:tel-limitations});
 future studies may wish to avoid this limitation by actively
-recruiting subjects.
+recruiting participants.
 
@@
 VSCode~\cite{vscode-telemetry}, IntelliJ~\cite{intellij-telemetry},
-Eclipse~\cite{eclipse-udc}, and
-\code{.NET}~\cite{dotnet-telemetry},
+\code{.NET}~\cite{dotnet-telemetry}
+and
+Eclipse~\cite{eclipse-udc},
 telemetry may include all sorts of
@@
 There is a rich body of work that uses IDE instrumentation to study programmers
-(e.g.,~\cite{hl-icse-2010,j-ieeesoft-2013,kje-ase-2010, mmlb-swvis-2014, ncvjd-ecoop-2013, rl-icse-2008, we-xp-2004}).
+(e.g.,~\cite{pgnwk-koli-2022,wk-icer-2019,hl-icse-2010,j-ieeesoft-2013,kje-ase-2010, mmlb-swvis-2014, ncvjd-ecoop-2013, rl-icse-2008, we-xp-2004}).
 WatchDog records the time that programmers spend reading, writing, and testing
@@
 TDD-Guide collects similar data to teach testing methods~\cite{mdk-xp-2008}.
-Examplar logs programs to give students feedback on their tests~\cite{pgnwk-koli-2022,wk-icer-2019}.
 The Blackbox project collects Java programs and has enabled
@@
 Despite the intentionally-restricted data,
-the analysis offers several lessons for the Luau
-developers and other maintainers of gradually-typed languages:
-%% TODO more lessons?
+the analysis offers several lessons for Luau in
+particular and for other gradual languages~(e.g.~\cite{bat-ecoop-2014,cdv-elixir-types-2023,tfffgksst-snapl-2017}) in general:
 \begin{itemize}
@@
     other modes exist.
+    Luau, and perhaps other languages, must take care to advertise gradual
+    types and demonstrate how to use them effectively to increase adoption.
 %  \item
@@
     This should give Roblox confidence to make \mnonstrict{} the default in the future.
+    Other languages should consider adding a nonstrict mode that minimizes false positives
+    as a bridge toward strict type checking.
     % Changes to the number of errors balance one another~(\cref{f:error-density});
@@
     analysis because it raises even more errors than \FS{} checking~(\cref{s:tefs-compass}).
+    Other languages should avoid making the same mistake.
 \end{itemize}
@@
 more metadata.
-Luau recently adopted semantic subtyping to reduce false-positive type
+For example, Luau recently adopted semantic subtyping to reduce false-positive type
 errors~\cite{CF05:GentleIntroduction,Jef22:SemanticSubtyping}, but the current
@@
 
-  \item It would be useful to have fine-grained counts for \FS{} analysis
-errors and to know the specific reason behind errors such as \code{TypeMismatch}.
-But, this additional data could double the size of telemetry records.
+  \item It would be useful to track the identity of errors (to know exactly which ones get fixed),
+    to have fine-grained counts for \FS{} analysis errors, and to
+    record the specific reason behind errors such as \code{TypeMismatch}.
+    But, this additional data could easily double the size of telemetry records.
 
@@
 
+% \begin{figure}[t]\centering
+%   YES SWITCH
+% 
+%   \mnocheck{}
+%   \includegraphics[width=\columnwidth]{img/error-count-nocheck-row--te-density-diff-yes-switch.pdf}
+% 
+%   \mnonstrict{}
+%   \includegraphics[width=\columnwidth]{img/error-count-nonstrict-row--te-density-diff-yes-switch.pdf}
+% 
+%   \mstrict{}
+%   \includegraphics[width=\columnwidth]{img/error-count-strict-row--te-density-diff-yes-switch.pdf}
+%   \caption{Changes in type error density ($\frac{\mbox{\#errors}}{\mbox{\#lines}}$) over time (seconds).}
+%   \label{f2:error-density}
+% \end{figure}
+% 
+% \begin{figure}[t]\centering
+%   ALL
+% 
+%   \mnocheck{}
+%   \includegraphics[width=\columnwidth]{img/error-count-nocheck-row--te-density-diff-all.pdf}
+% 
+%   \mnonstrict{}
+%   \includegraphics[width=\columnwidth]{img/error-count-nonstrict-row--te-density-diff-all.pdf}
+% 
+%   \mstrict{}
+%   \includegraphics[width=\columnwidth]{img/error-count-strict-row--te-density-diff-all.pdf}
+%   \caption{Changes in type error density ($\frac{\mbox{\#errors}}{\mbox{\#lines}}$) over time (seconds).}
+%   \label{f3:error-density}
+% \end{figure}
+
 \newpage

### 2023-12-12 programming23/paper.tex
missing period

@@
 Ada learners~\cite{tkdmceg-ascilite-2003,tm-iticse-2004}.
-None of the above are sufficiently private to work for Roblox Studio
+None of the above are sufficiently private to work for Roblox Studio.
 The BitFit project counts notable events (compile program, ask for hint, etc.)

### 2023-12-12 programming23/paper.tex
blackbox, sec 3 end para, redundancies

@@
 of the file.
-We acknowledge this and other threats in~\cref{s:threats}.
 The upshot of this strategy is that it reduces the size and complexity of
 telemetry records because there is only one interval for errors to overlap with.
+
+This telemetry design is admittedly coarse-grained.
+For instance, it does not distinguish between edits that remove one error
+while introducing another.
+We acknowledge this and other threats in~\cref{s:threats}.
+The main advantages of this telemetry are the small size of each record and
+the complete lack of private information.
+Furthermore, despite the limitations, this telemetry supports a variety of
+inferences that we showcase in the next section.
 
@@
 and twenty reduce the number of \code{CannotCallNonFunction} errors.
-
-The counts are imprecise because of telemetry granularity.
-If a creator removes one error but then
-introduces another on a different line, this can manifest as
-a keep (\keepsym{}) rather than one add (\addsym{)} and one drop (\dropsym{}).
 
@@
 
-Luau has an efficient type checker that can
-analyze thousands of lines of code~(\cref{t:codebase-size}) quickly
-enough to be used on-the-fly in an IDE.
-But, few sessions use type analysis directly.
-We conjecture that Luau needs further tailoring to match
-common idioms in untyped code.
-Discovering these idioms is a question for future work~(\cref{s:conclusion}).
-
 \begin{description}
@@
         use \mstrict{} mode.
+
+        Luau has an efficient type checker that can
+        analyze thousands of lines of code~(\cref{t:codebase-size}) quickly
+        enough to be used on-the-fly in an IDE.
+        But, few sessions use type analysis directly.
+        We conjecture that Luau needs further tailoring to match
+        common idioms in untyped code.
+        Discovering these idioms is a question for future work~(\cref{s:conclusion}).
 
@@
 edited line, even if nothing between those lines changed.
+On a related note, the computatation that determines whether an old
+type error overlaps with the current edit range has many edge cases,
+especially to handle deletions, and may mis-count.
+This could explain why~\cref{t:type-error-survival} reports zero syntax errors
+that overlap with the current edit range; if the fix is to delete code,
+then the edit range is empty.
 
@@
 There is a rich body of work that uses IDE instrumentation to study programmers
-(e.g.,~\cite{pgnwk-koli-2022,wk-icer-2019,hl-icse-2010,j-ieeesoft-2013,kje-ase-2010, mmlb-swvis-2014, ncvjd-ecoop-2013, rl-icse-2008, we-xp-2004}).
+(e.g.,~\cite{pgnwk-koli-2022,wk-icer-2019,hl-icse-2010,j-ieeesoft-2013,kje-ase-2010, mmlb-swvis-2014, ncvjd-ecoop-2013, rl-icse-2008, we-xp-2004,bgimgm-cse-2016,dlc-iticse-2014}).
+The Blackbox project has accumulated terabytes of Java code over a ten-year span,
+and has thereby enabled dozens of contributions to CS education~\cite{bask-icer-2018,bkmu-sigcse-2014}.
+The GRUMPS telemetry system collected over four million keystroke-level actions to study
+student programmers~\cite{tkdmceg-ascilite-2003,tm-iticse-2004}.
 WatchDog records the time that programmers spend reading, writing, and testing
@@
 TDD-Guide collects similar data to teach testing methods~\cite{mdk-xp-2008}.
-The Blackbox project collects Java programs and has enabled
-dozens of contributions to the area of CS education~\cite{bask-icer-2018,bkmu-sigcse-2014}.
-Other logging-based studies of Java programs draw conclusions from compiler error
-messages~\cite{bgimgm-cse-2016,dlc-iticse-2014}.
-The GRUMPS telemetry system collected 4.7 million keystroke-level actions from
-Ada learners~\cite{tkdmceg-ascilite-2003,tm-iticse-2004}.
 None of the above are sufficiently private to work for Roblox Studio.
@@
 While it does operate in a transparent~\cite{transparent-telemetry} manner, it
-only provides coarse-grained information that would be insufficient in our
+provides only coarse-grained information that would be insufficient in our
 study.

### 2023-12-12 programming23/paper.tex
units on plots
grant++ diff cmd

@@
 This telemetry design is admittedly coarse-grained.
-For instance, it does not distinguish between edits that remove one error
+For instance, it does not distinguish between edits that
+ignore a type error from edits that remove one error
 while introducing another.
@@
 user scripts that generated a huge number of keystrokes.
-%% TODO investigate one tower, what's happening?? all one session?
+%% two biggest towers (7050 6691) are mostly from 2 sessions:
+%% (("\"178,020,742,609,772\"" . 6079))
+%% (("\"458,983,252,701,986\"" . 5085))
 Telemetry ended in early April; however, there are short bars in the mid-April
@@
   \includegraphics[width=\columnwidth]{img/error-count-strict-row--te-density-diff-no-switch.pdf}
-  \caption{Changes in type error density ($\frac{\mbox{\#errors}}{\mbox{\#lines}}$) over time (seconds).}
+  \caption{Changes in type error density ($\frac{\mbox{\#curr - \#old errors}}{\mbox{\#lines}}$) over time (seconds).}
   \label{f:error-density}
@@
 and effective plotting.
-Greenman was supported by
-NSF grant \href{https://nsf.gov/awardsearch/showAward?AWD_ID=2030859&HistoricalAwards=false}{2030859}
+We gratefully acknowledge support from
+NSF grant \href{https://www.nsf.gov/awardsearch/showAward?AWD_ID=2227863&HistoricalAwards=false}{SHF-2227863}
+and NSF grant \href{https://nsf.gov/awardsearch/showAward?AWD_ID=2030859&HistoricalAwards=false}{2030859}
 to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.

### 2023-12-15 programming23/artifact/runall.sh
run script update files

@@ -1,16 +1,20 @@
 echo "deleting everything in out/ ..."
 echo "=== before"
 ls out/
-rm out/*.*
+rm -f out/*.*
 echo "=== after"
 
 PLTSTDERR="error info@luau" racket code/prepare.rkt
+PLTSTDERR="error info@luau" racket code/prepare.rkt --mode size-distro
 PLTSTDERR="error info@luau" racket code/prepare.rkt --mode modswitch
 PLTSTDERR="error info@luau" racket code/prepare.rkt --mode divide-sessions
 PLTSTDERR="error info@luau" racket code/prepare.rkt --mode session-query
+PLTSTDERR="error info@luau" racket code/prepare.rkt --mode session-fold
+PLTSTDERR="error info@luau" racket code/prepare.rkt --mode count-te-editrange 
 PLTSTDERR="error info@luau" racket code/prepare.rkt --mode aggregate-te
 
 racket code/row-distribution.rkt
+racket code/size-distribution.rkt
 racket code/sdupdate.rkt
 racket code/downgrade-count.rkt
 racket code/error-by-mode.rkt
@@ -18,3 +22,4 @@ racket code/type-error-survival.rkt
 racket code/type-error-count.rkt
 racket code/error-count.rkt
 
+echo "DONE BUILDING, next step = compile pdf outside docker"

### 2023-12-19 programming23/paper.tex
reword 'hobbyist'

@@
 
-% TODO BEFORE SUBMISSION
-% - units on all plots
-
 % TODO BEFORE FINAL
-% - explain edit range computations
-% - uncomment acks
 % - update all refs to artifacts to no longer be in future tense
@@
 one module at a time.
-In this way, for example,
-students in code camps may begin in the \mnocheck{} mode that reports only
-syntax errors; hobbyists can use \mnonstrict{} mode to identify
-would-be runtime errors; and developers who need to ensure software
-quality can rely on \mstrict{} mode.
-Also, no matter the mode, a {\FS{}} type analysis runs to guide \IDE{} tools~(\cref{s:context}).
+The \mnocheck{} mode reports only syntax errors,
+the \mnonstrict{} mode reports would-be runtime errors and skips over
+other issues,
+and the \mstrict{} mode reports a variety of potential errors.
+In addition, no matter the mode, a {\FS{}} type analysis runs to guide \IDE{} tools~(\cref{s:context}).
 Luau's types aim to support untyped designs, in the spirit of migratory

### 2023-12-21 programming23/paper.tex
update daily users + hours for Sep

@@
 {Roblox} is a platform for {shared virtual experiences} (typically 3D games), with
-65~million Daily Active Users, and 14~billion hours of engagement in
-April--June 2023~\cite{corp.roblox.com}.
+70~million Daily Active Users, and 16~billion hours of engagement in
+July--September 2023~\cite{corp.roblox.com}.
 There are over 5~million distinct user-created programs available on the platform thanks to a

### 2024-01-31 programming23/paper.tex
update clsfile, cite artifact

@@
 
-% TODO BEFORE FINAL
-% - update all refs to artifacts to no longer be in future tense
-
 % The paper on using Luau telemetry to measure the effectiveness of type error reporting.
 
-\documentclass[english,cleveref]{programming}
-%% use 'submission' for initial submission, remove it for camera-ready (see 5.1)
+\documentclass[crc,english,cleveref]{programming}
 
@@
 
-%%%%%%%%%%%%%%%%%%
+%%% The following is specific to Programming Journal, Volume 8, Issue 3 and the paper
+%%% 'Privacy-Respecting Type Error Telemetry at Scale'
+%%% by Ben Greenman, Alan Jeffrey, Shriram Krishnamurthi, and Mitesh Shah.
+%%%
 \paperdetails{
-  %% perspective options are: art, sciencetheoretical, scienceempirical, engineering.
-  %% Choose exactly the one that best describes this work. (see 2.1)
-  perspective=scienceempirical,
-  %% State one or more areas, separated by a comma. (see 2.2)
-  %% Please see list of areas in http://programming-journal.org/cfp/
-  %% The list is open-ended, so use other areas if yours is/are not listed.
-  area={Data mining for programming, Telemetry},
-  %% You may choose the license for your paper (see 3.)
-  %% License options include: cc-by (default), cc-by-nc
-  % license=cc-by,
+  submitted=2023-10-01,
+  published=2024-02-15,
+  year=2024,
+  volume=8,
+  issue=3,
+  articlenumber=12,
 }
-%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%
-%% These data are provided by the editors. May be left out on submission.
-%\paperdetails{
-%  submitted=2023-1-31,
-%  published=2016-10-11,
-%  year=2016,
-%  volume=1,
-%  issue=1,
-%  articlenumber=1,
-%}
-%%%%%%%%%%%%%%%%%%
 
@@
 Kahlert School of Computing at the University of Utah.}
+% \orcid{0000-0001-7078-9287}
 \affiliation[a]{Brown University, Providence, RI, USA}
 \affiliation[b]{University of Utah, Salt Lake City, UT, USA}
-% \orcid{0000-0001-7078-9287}
 
@@
 \author[c]{Mitesh Shah}
+% \orcid{0009-0000-6084-123X}
 \authorinfo{(\email{mshah@roblox.com}) is Senior Engineering Director, Programmability, at Roblox.}
@@
   of over 1.5 million telemetry records.
-  The data and scripts for analyzing it will be available in an artifact.
+  The data and scripts for analyzing it are available in an artifact.
 
@@
 \begin{figure}[t]\centering
-  %% TODO weird artifact (|) in x-min label
+  %% TODO weird artefact (|) in x-min label
   %% code/row-distribution.rkt
@@
 
+\paragraph{Data Availability Statement}
+
+The artifact for this paper contains the full telemetry dataset
+and scripts for reproducing the analysis in~\cref{s:data}.
+The source code for the paper itself is available through Software
+Heritage~\cite{luau-swhid}.
+
 %% -----------------------------------------------------------------------------

### 2024-01-31 programming23/paper.tex
proofreading

@@
 not consider themselves software developers; others are professional
-developers.
+developers who work full time with the Roblox platform.
 
@@
 gradually~\cite{st-sfp-2006,tf-popl-2008} migrate to rigorous checks
-one module at a time.
-The \mnocheck{} mode reports only syntax errors,
+one module at a time via three analysis modes.
+The default \mnocheck{} mode reports only syntax errors,
 the \mnonstrict{} mode reports would-be runtime errors and skips over
@@
 levels of granularity.
-Furthermore it maintains a client-side approximation of the latest source-code
+Furthermore, it maintains a client-side approximation of the latest source-code
 edits and uses that to identify type errors that overlap with this edit range.
-Telemetry records are correlated by session, using a \panon{} session
-identifier and a timestamp.
+Each record contains a \panon{} session identifier and a timestamp.
 
@@
     errors?
-    For example, do \FS{} errors pile up in unanalyzed projects?
+    For example, do \FS{} errors pile up in unanalyzed (\mnocheck{}) projects?
 \end{description}
@@
 is not restricted to users' \emph{perceptions} about their work and it is not
-limited to a small number of users.
+limited to a small number of users performing synthetic tasks.
 
@@
   \item
-    An available dataset of over 1.5 million telemetry records
-    and scripts to analyze them: \url{https://zenodo.org/doi/10.5281/zenodo.10275213}
+    A dataset of over 1.5 million telemetry records and scripts to analyze them~\cite{luau-artifact}.
 
@@
 Creators can opt in to detailed error reports and highlights by
-choosing a \emph{type analysis mode} for each script they write:
+choosing a \emph{type analysis mode} for each script.
+The modes are:
 \begin{itemize}
   \item \mnocheck{}: report only syntax errors (the default),
-  \item \mnonstrict{}: report syntax errors and a subset of high confidence type errors, or
+  \item \mnonstrict{}: report syntax errors and a subset of high confidence type errors, and
   \item \mstrict{}: report syntax errors and all type errors.
 \end{itemize}
-Each run of type analysis can report several errors.
+Each run of the type analysis can report several errors.
 There is no guarantee that creators read every error.
 In fact, creators who close the Script Analysis widget
-will see highlights in their code but no further details.
+can see highlights in their code but no further details about the errors.
 
@@
 
- As an example of \mnonstrict{} mode, the following program only reports one error:
+ As an example of \mnonstrict{} mode, the following program reports only one error:
 \begin{verbatim}
@@
 \end{verbatim}
-In cases like this, where it is undecidable whether there will be a run-time error,
-\mstrict{} mode errs on the side of reporting an error, and \mnonstrict{} mode errs on
+In programs like the above, where it is undecidable whether there will be a run-time error,
+\mstrict{} mode errs on the side of reporting an error and \mnonstrict{} mode errs on
 the side of suppressing the error.
@@
 \mstrict{} and \mnonstrict{} mode.
-\mnocheck{} reports no error because the program is syntactically valid.
+The \mnocheck{} mode reports no error because the program is syntactically valid.
 See~\cite{bfj-hatra-2021} for a more detailed discussion of the rationale
@@
 no matter what mode the creator declared for the script.
-(Internally to Roblox Studio, this analysis is called \emph{forced strict}
-because it ignores the creator's choice and applies strong checks.)
 When it fails to infer a type, \FS{} analysis defaults to the unknown gradual
@@
 Creators cannot see the errors reported by \FS{} analysis.
+Internally to Roblox Studio, this analysis is called \emph{forced strict}
+because it ignores the creator's choice and applies strong checks.
 
@@
 \mnocheck{} code.
-Several errors in addition to \code{UnknownProperty} are about
+Several errors including \code{UnknownProperty} are about
 tables, which are Luau's (and Lua's) primary data structure.
-Tables support arrays, dictionaries, and objects; consequently,
+Tables encode arrays, dictionaries, and objects; consequently,
 tables can have methods and can usually be extended.
@@
 for a generic type.
-Lastly, \code{CodeTooComplex} occurs when the typechecker hits an
-internal limit on problem size and \code{GenericError} is a catch-all
+A \code{CodeTooComplex} occurs when the typechecker hits an
+internal limit on problem size. Lastly, \code{GenericError} is a catch-all
 label for miscellaneous errors.
@@
     & Generic label for other non-type errors,\\
-    & \hbox{}~~such as looping over an unordered table.
+    & \hbox{}~~e.g., looping over an unordered table.
     %% attempting to extend a type that does not describe a class
@@
 {imported} by other scripts.
-Imports therefore form a graph in general, though the analysis
+Imports therefore form a graph in general, though type analysis
 removes edges to ensure an acyclic graph.
@@
 Since the shape of the data model tree is reflected in
-the type system, it is possible for these edits to introduce a mountain
+the type system, it is possible for these edits to introduce a huge number
 of type errors across the project.
@@
 
-Telemetry allows an application to ``phone home'' with data
+Telemetry allows an application to phone home with data
 summarizing usage patterns, such performance data, crash reporting,
-and feature uptake. A typical usage is in deciding whether an API can
+and feature uptake. A typical use-case is in deciding whether an API can
 be deprecated; without telemetry, it may be difficult to know how
@@
 See Cox~\cite{transparent-telemetry} for further discussion of
-the privacy implications and tradeoffs in telemetry.
+the privacy implications and tradeoffs of telemetry for programming languages.
 
@@
 
-\Cref{f:tdata} shows the structure of our telemetry records:
+\Cref{f:tdata} shows the structure of telemetry records:
 \begin{enumerate}
@@
 The plots focus
-on the lower-left fragment of the data, which again has a very long tail.
+on the lower-left fragment of the dataset, which again has a very long tail.
 %% comment on long-tail record counts?
@@
     that overlap with the current edit range~(\pct{27} difference),
-    but 172,678 fewer previous \FS{} errors~(\pct{15} difference).
-    Since creators can see the type errors but not the \FS{}
-    errors before making edits, we conclude that the visibility
+    but 172,678 fewer overlapping previous \FS{} errors~(\pct{15} difference).
+    Since creators can see type errors but not \FS{}
+    errors before making edits, we conclude that visibility
     makes a difference.
@@
     While non-privacy-protecting analyses have answered such
-    questions~\cite{mfk-sigcse-2011}, our data cannot narrow down the answer.
+    questions in the past~\cite{mfk-sigcse-2011}, our data cannot narrow down an answer.
 
@@
 less-strict mode.
-There are 263 total upgrades (across all sessions) and 320 total downgrades.
+There are 263 total upgrades across all sessions and 320 total downgrades.
 
@@
 This data excludes module switch records and considers only keystroke events.
-The worst-case increase was quite high at 57 type errors, but exceptional.
+The worst-case increase was quite high, at 57 type errors, but exceptional.
 In another exception, upgrading modes removed three type errors,
@@
 \Cref{f:error-by-mode} divides errors across modes.
-By the left plot, \mnonstrict{} accounts for two thirds of all type errors.
+The left plot shows that \mnonstrict{} accounts for two thirds of all type errors.
 This confirms our expectation.
@@
 Indeed, most of the fields in a telemetry record are dedicated to this
-topic~(\cref{f:tdata}), and are directed to RQ2 on type errors .
+topic~(\cref{f:tdata}), and are directed toward RQ2 on type errors.
 
@@
 
-    Curiously, the number of syntax errors never decreases or stays the same.
-    This may be because the easy fix is to press undo, and telemetry cannot
+    \subitem Curiously, the number of syntax errors never decreases or stays the same.
+    This may be because the easy fix is to press undo; telemetry does not
     see undo events.
@@
     %% TODO awkward
-    Our telemetry strategy likely inflates the number of typos because it cannot
+    \subitem The number of typos is inflated by our telemetry strategy because it cannot
     determine when a creator is mid-edit.
@@
     \code{UnknownSymbol} error for \code{buck} and
-    \code{UnknownProperty} for \code{bucket.c}, which may end up getting sampled.
+    \code{UnknownProperty} for \code{bucket.c}, both of which may end up
+    getting sampled.
     \Cref{s:conclusion} suggests ways
-    to avoid this in the future.
+    to avoid reporting typos in the future.
 
@@
     further study in future work to determine precisely which error arose.
-    (Some of the \code{GenericError}s come with context strings in the form of
-    \code{ExtraInformation} errors, but telemetry does not report these strings.)
+    Some of the \code{GenericError}s come with context strings in the form of
+    \code{ExtraInformation} errors, but telemetry does not report these strings.
     %% GenericError = only a few things: extend non-class, iterate non-table, ...?
@@
   \item
-    \code{TypeMismatch} is far more common in \mstrict{} (\pct{20.46})
-    than \mnonstrict{} (\pct{2.44}) because \mstrict{} mode
+    \code{TypeMismatch} is far more common in \mstrict{} mode (\pct{20.46})
+    than \mnonstrict{} (\pct{2.44}) because \mstrict{}
     has additional type constraints.
@@
     below.
-    The others are for: dynamic property lookups,
+    The others are for
+    dynamic property lookups,
     deprecated APIs, swapped generic parameters,
@@
 To deal with pathologies such as the worst-case time for ML-style type
-inference~\cite{ktu-caap-1990,m-popl-1990}, the typechecker
+inference~\cite{ktu-caap-1990,m-popl-1990}, the Luau typechecker
 has internal limits that restrict the problems it will attempt to solve.
@@
 are misleading or because the creator ignores the errors, it is important
-to see how the total errors change over time.
+to see how the total number errors changes over time.
 We measure \emph{error density} rather than error count, however,
@@
 Most sessions last a few dozen seconds~(\cref{t:session-size}), so the axis
-ends at 150 seconds to show some but not all of the long tail.
+ends at 150 seconds to show a bit of the long tail.
 The y axis shows changes to errors over time; a positive number means errors
@@
 \FS{} errors as well.
-The ``maintain'' rows include cases where there are zero errors before and after.
+The ``maintain'' rows exclude cases where there are zero errors before and after.
 
@@
 With the data in hand, we can answer the research questions
-from~\cref{s:introduction}.
-The answers confirm several aspects of Luau's design, but
-at the same time offer several directions for future improvements.
+from~\cref{s:introduction}:
 
@@
 
-        Luau has an efficient type checker that can
+        \subitem Luau has an efficient type checker that can
         analyze thousands of lines of code~(\cref{t:codebase-size}) quickly
@@
       \item[- How do sessions respond? Which errors persist through edits?]
-        Sessions typically fix analysis errors.
+        Sessions appear to fix analysis errors.
         With few exceptions, increases and decreases in the number of errors
@@
         the total.
-        The errors that most often survive edits are due to
+        The errors that survive edits most often are due to
         inference at a binary operation,
@@
         and failure to unpack an optional value.
-        But the first two seldom arise, and the third is something creators
+        But, the first two seldom arise and the third is something creators
         might reasonably ignore while prototyping.
@@
 
-  \item[RQ3. What impact does type analysis have on the number of \FS{} errors?]
-    Opting in to type analysis has little impact on \FS{} errors.
+  \item[RQ3. What impact does type analysis have on \FS{} errors?]
+    Opting in to analysis has little impact on the number of \FS{} errors.
     The percent of \FS{} errors in \mnocheck{} code matches the percent of
@@
 and gave up.
-Furthemore, our method of sampling by keystroke and by module switch
+Furthermore, our method of sampling by keystroke and by module switch
 skews the data toward creators who, for whatever reason, do more
@@
 edited line, even if nothing between those lines changed.
-On a related note, the computatation that determines whether an old
+On a related note, the computation that determines whether an old
 type error overlaps with the current edit range has many edge cases,
@@
 probability of contributing telemetry.
-The system-builder can analyze this telemetry with predicates to identify
-notable events and logistic regression to narrow down nondeterministic bugs.
+A system-builder can analyze this telemetry with predicates to identify
+notable events and use logistic regression to narrow down nondeterministic bugs.
 There is a high burden on designers to decide what to capture,
@@
   \item
-    Sessions tend to fix errors immediately rather than dealing with them in a batch
+    Sessions appear to fix errors immediately rather than dealing with them in a batch
     or ignoring the errors altogether~(\cref{f:error-density,f:error-changes}).
-    This should give Roblox confidence to make \mnonstrict{} the default in the future.
+    This observation should give Roblox confidence to make \mnonstrict{} the default in the future.
     Other languages should consider adding a nonstrict mode that minimizes false positives

### 2024-02-09 programming23/paper.tex
luau ++
orcid, biber, final touch ups

@@
 % Not sure why <Programming> is recommending non-standard tools
-% \usepackage[backend=biber]{biblatex}
+% uncomment because publisher asked about missing URL access dates
+\usepackage[backend=biber]{biblatex}
 \usepackage[export]{adjustbox}
@@
 
+\addbibresource{bib.bib}
+
 %%% The following is specific to Programming Journal, Volume 8, Issue 3 and the paper
@@
 
-\author[a,b]{Ben Greenman}
+\author[a,b]{Ben Greenman}[0000-0001-7078-9287]
 \authorinfo{(\email{benjamin.l.greenman@gmail.com}) was a CIFellows 2020
@@
 Kahlert School of Computing at the University of Utah.}
-% \orcid{0000-0001-7078-9287}
 \affiliation[a]{Brown University, Providence, RI, USA}
@@
 
-\author[c]{Alan Jeffrey}
+\author[c]{Alan Jeffrey}[0000-0001-6342-0318]
 \authorinfo{(\email{ajeffrey@roblox.com}) is a Principal Software Engineer at Roblox.}
-% \orcid{0000-0001-6342-0318}
 \affiliation[c]{Roblox, San Mateo, CA, USA}
 
-\author[a]{Shriram Krishnamurthi}
-% \orcid{0000-0001-5184-1975}
+\author[a]{Shriram Krishnamurthi}[0000-0001-5184-1975]
 \authorinfo{(\email{shriram@brown.edu}) is the Vice President of Programming Languages (no, not really) at Brown University.}
 
-\author[c]{Mitesh Shah}
-% \orcid{0009-0000-6084-123X}
+\author[c]{Mitesh Shah}[0009-0000-6084-123X]
 \authorinfo{(\email{mshah@roblox.com}) is Senior Engineering Director, Programmability, at Roblox.}
@@
 Creators can opt in to detailed error reports and highlights by
-choosing a \emph{type analysis mode} for each script.
-The modes are:
+selecting a \emph{type analysis mode} for each script.
+There are three modes that creators may choose from:
 \begin{itemize}
@@
   \item
-    1\% of Roblox Studio sessions generate \emph{any} type error telemetry, and
-  \item
-    0.5\% of all type analyses (approximately 1 in 200 keystrokes)
+    \pct{1} of Roblox Studio sessions generate \emph{any} type error telemetry, and
+  \item
+    \pct{0.5} of all type analyses (approximately 1 in 200 keystrokes)
     in an enrolled session generate a telemetry record.
@@
 and scripts for reproducing the analysis in~\cref{s:data}.
-The source code for the paper itself is available through Software
-Heritage~\cite{luau-swhid}.
 
@@
 
-\bibliography{bib}
+{
+
+\printbibliography
+}
 

### 2024-02-21 programming23/paper.tex
more publisher fixes

@@
     Counting specific type errors helps to identify trends and to confirm that
-    undesirable events rarely happen (e.g.,\code{UnificationTooComplex}).
+    undesirable events rarely happen (e.g.,~\code{UnificationTooComplex}).
   \item
